### 2019-02-26 reports/preliminary_results.Rmd
Har sammanfattat vad jag gjort för att skicka till NH.

---
title: "Preliminary results"
author: "Erik Bulow"
date: '2019-02-25'
output: 
  html_document:
    df_print: "kable"
    toc: true
    toc_float: true
    code_folding: "hide"
abstract: |
  This is a background description of some analysis made within the project. 
  It is written in English due to planned collaboration with NJR.
  I have started with prevouis analysis made by Szilard but have used new data and slightly different methods.
  I have only considered death within 90 days so far.
bibliography: P:/library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
load("../cache/df.RData")
library(tidyverse)
```

# Data

I have used data from the "new linkage data base". Primary prosthesis inserted 2008-2015 were included. Note that we should have no problem with censoring here since death dates are recorded up until February 2018. Hence even if patients were operated during 2015, they are not automatically censored at the end of that year. (Censoring would only occour in the case of migration etc.)
Comorbidity indices based on Elixhauser van Walraven (ECI), Charlson's original index (CCI) and a modified Risk Rx IV index (modified by Anne Garland) were used.

The following inclusion/exclusion criteria were applied:

1. Only primaries (no re-operations).
2. Only OA.
3. No hemi.
4. No second bilaterals.
5. No first bilateral if second bilateral inserted within a year.
6. No surface prosthesis
7. Only age 18-100 years
8. only BMI $\leq 50$
9. Only ASA $\leq 3$
10. Exclude if missing education level
11. Exclude if missing marital status level
12. Exclude if missing type of hospital
13. Exclude if fixation data missing

No imputation applied.

Exact numbers of patients/hips from each inclusion/exclusion step will be calculated after everything is set. 
The final data set has `r nrow(df)` rows with `r ncol (df)` columns:

```{r}
names(df)
```

education = 
* low = pre-gymnasium
* middle = gymnasium
* high = post-gymnasium

civil_status = married/single/widow(er). Registered partnership etc are treated as marriage.

## Data profiling report

A data profiling report is find as an appendix.
Some comments:

### Basic statistics

Most of our variables are categorical (very few numeric)

### Missing data

We do not have any missing data (due to our selection criteria)

### Univariate distributions (numeric variables)

BMI and age are approximately normally distributed. CCI and ECI have very little data (most of the patients have no recorded comorbidities). It should be discussed whether diagnoses set during hospitalization should be included. There are some negative ECI values as well (due to the van Walraven index).

### Bar Chart (categorical variables)

Type of hospital, sex, ASA, civil status, education and prosthesis group looks as expected.

We have few patients with reversed hybrids (unbalanced sample).

We see that there are very few patients with each comorbidity (hypertension is an exception, as well as some more commonly used medications). This is problematic since logistic regression works best with outcome proportions of at least 20 %.

No one gets medically treated for dementia.

### QQ PLot

Those plots indicate (again) that age and BMI are approximately normally distributed, while some other variables are not.

We can see as well that individuals who dies within 90 days have higher ECI, CCI and age on average.

### Corelation analysis

We have too many variables to distinguish. Obviously, each variable is strongly correlated to itself (diagonal line). A cluster of correlated variables are also found in the upper right corner since outcome variables appear in different forms in the data set. We can also assume that some comorbidities are associated and obviously so for the same comorbidity measured by different indices.

### Bivariate distribution

Older people are over-represented among the dead.

## Table 1

I include all values of each comorbidity index for the moment but those should obviously be truncated/transformed later.

Proportions for most variables are roughly the same as in earlier versions, although not exact due to the extended time frame of the study period. 
Comorbidity data differ but this is currently under discussion.

```{r}
library(tableone)
load("../cache/t1.RData")
 t1 %>%
  print(
    showAllLevels = TRUE,
    printToggle = FALSE
  ) %>%
  as_tibble(rownames = "what")
```

## Table 2

We can also make a table similar to table 2. So far we only include univariable models.
Most of the result is consistent with the previous version of the table.

```{r}
load("../cache/t2_univariable.RData")
t2_univariable
```

# Modelling

## Risk of overfitting

We have a lot of potential predictors but very few events to predict. This is problematic and we risk overfittnig the model to the data. Hence, if we develop/train our model using all data and then use the same data to evaluate predictive power, the result will look better than it actually is (applying the model to new data will not work as well).

We can use some proportion of our data as a hold-out sample to use for later validation. It is common to use 75 % of the data for model training and 25 % for evaluation. We can choose however to use 90 % for training and 10 % for evaluation. With the rare event rate, a totally random 90/10-split might lead to extremely few "positive outcomes" (deaths) in the evaluation set. The sampling is therefore stratified on the outcome.

## Downsamling

The training set will now have approximately `r round(nrow(df) * .9)` patients. Only about 0.26 % of those will die within 90 days. A secure bet then will be to assume that everyone will survive. Such a model will have perfect specificity. The sensitivity will of course be 0, but most likely it will anyway if the difference between patients who dies and survive is not clear-cut. Hence, if we are to use any sort of step-wise regression modelling, the estimated coefficients for a model based on all this data, will underestimate the true coefficient values (the model will be too conservative) [@King2001].

A common and simple solution to this problem is to use down-sampling. This is done by matching if the goal is to compare treatments or similar. We can not use matching in our case (we are not comparing groups based on variables that we have not yet selected). We can however simply include all cases and then as many controls (survivors).

An alternative to down-sampling is up-sampling or a combination of the two [@Lunardon2014].
An additional benefit of down-sampling however is that we will also reduce the sample size (without losing too much power). This is beneficial since step-wise regression techniques around 100 variables and a large data set will otherwise be very time consuming.

With a balanced data set, we can no longer cheat by predicting everyone as survivors without a severe loss of specificity. We therefore force the model to do a much better job (not underestimating the beta coefficients in the logistic regression model).

# Uncertainty

A draw-back with the down-sampling however is that we might get a sample that is not representative for the larger population. Our estimates might be very sensitive to the individuals who just happened to show up in the sample. If we perform a step-wise regression on two different samples, it would therefore be possible that each data set will provide two different set of models (based on different predictors) [@Lukacs2010; @Sauerbrei1999; @Zellner2004].

A theoretically simple (although computationally intense) method is to repeat this procedure several times with different samples. We will do that.

## Candidate predictors

I took 1,000 bootstrap replicates of the training data and applied down-sampling to each of those. 
For each sample I used the Aikake Information Criteria as a selection criteria within step-wise regression. I saved a list of the selected variables for each "winning model". Note that the size of the coefficients for those variables are not yet important. The same variable can even have a positive effect in one model, but a negative effect in another (depending on which other variables are included, thus which part of the population would constitute the base line).

Note that variables that are highly correlated could not be included in the same selection step, since it will then be arbitrary which one of those to select. I therefore used four independent step-wise regression procedures to each re-sample evaluating:

1. General variables including comorbidity indices
2. Individual ECI diagnosis
3. Individual CCI diagnosis
4. Individual Risk Rx IV diagnosis

Of course in the first group, it could be argued that the different comorbidity indices are correlated with each other. We did not see this in practice however.

The following table list all assessed variables ordered by the proportion of times for which they were included in the final model chosen by the step-wise procedure.

```{r}
load("../cache/prop_selected.RData")
DT::datatable(prop_selected)
```

It is common to use univariable screening before the step-wise procedure, which was here considered redundant since univariable models are already included in the step-wise procedure (see [@Harrell2015]).

I also tried the Bayesian Information Criteria (BIC) instead of AIC, but it promotes simpler models quite heavily in this case, wherefore fewer predictors will end up in the final model. I did not find this to be desirable yet.

We can see from the table above that most potential predictors end up in less than half of all models. We can illustrate the distribution of proportions for a more intuitive overview:

```{r}
ggplot(prop_selected, aes(value)) + 
  geom_density(fill = "lightblue") + 
  theme_minimal()
```

It is usually recommended to only include variables that are in fact chosen in around 90 %
 (or even 95 % of the cases). This would be age, ASA, medication for inflammatory pain and sex in our case. We could use a 75 % limit however since this is not our final model.

Note that heart infarction and renal disease did not show up here. Those were included last time but we can see from the table above that those were in fact choose in less than 75 % of our Bootstrap-samples. It might be argued that those were in fact important predictors in the original sample. The goal of prediction modelling however is not to evaluate the past but to predict the future!

We now have the following candidate predictors:

```{r}
load("../cache/candidates.RData")
candidates
```

It was actually only ASA = 3 that showed up above (level 2 could not be distinguished from level 1) but we include both levels any way (the data is already available so why not use it?).

# Combined model

The candidate variables above where found from four different starting points. We can now argue that those variables should not be heavily associated with each other. It would have been different if for example heart infarction would have been important if classified both by ECI, CCI and Rx. We could then combine those factors into one (which was done for the previous version of the manuscript). This might also be necessary with a different threshold than 75 %. It might nevertheless be relevant to use step-wise regression once more based on the smaller subset of candidate variables.

It would be possible to do this on the whole training data set. We would still have the same uncertainty problem as above however. Instead we can use "bagging" (bootstrap aggregating), hence repeated bootstrap samples (again), with down-sampling performed to each re-sample and a full main effects model estimated for each re-sample. This is similar to above but the result from each sample is not only a set of candidate predictors but also estimates of each coefficient associated with those variables. Also, models of all possible combinations of potential predictors are tried out, not only the few where AIC were not increased. We get the following results:

```{r}
load("../cache/modsums.RData")
round(modsums$modsum[[1]]$coefmat.full, 2)
```

Hence, only CCI, age and ASA = 3 seems to be important.

# Additional models

Given the results above, we might be tempted do drop the Rx-variables. It would be nice if those were not needed since acquiring this data in clinical practice might be cumbersome. Sex on the other hand is almost always available anyway. Hence, we'll better keep it. This will yield an alternative model for comparison:

```{r}
round(modsums$modsum[[2]]$coefmat.full, 2)
```

Figure 2 in the earlier version of the manuscript also included models with only ASA, CCI, ECI and Risk Rx. We will do too for comparison and can add a model  with age and sex only as well.

# Predictive power

We now 7 models that can be compared based on predictive power, thus by area under the (ROC) curve (AUC).
Let's take another 1,000 bootstrap replicates and compare observations to predictions for those. Note that we do not only use the list of predictors for each model, but also the estimated coefficients. We thus evaluate the models from above, rather than fitting new ones and evaluate again. We do this to decrease overfitting. This will give us 1,000 ROC curves for each model. We could plot all of those but the resulting figure would be quite messy (and time consuming to draw). I therefore fitted a locally estimated scatter-plot smoothing (LOESS) line as an average ROC-curve based on individual bootstrap estimates.

Overfitting will nevertheless occur since we are in fact still using the training data. We thus need to re-calculate AUC-values one final time for the evaluation data set, the 10 % of the original data that we have not looked at yet. We could apply Bootstrap replicates to the test data as well in order to estimate uncertainty/confidence intervals. Those would probably we very wide though and are avoided to save (computational) time.

# Results and interpretation

Averaged ROC-curves for each model is illustrated below. Curves below the dotted line would not occur if the models had been estimated and evaluated using the same data set. This is not the case here and we do in fact have models that perform worse then random. This happens for models with only a few possible values of its predictors, since those models are insensitive to nuances not captured by those limited values. This is also a clear sign of over-fitting (the model is too heavily biased to the initial data and the result can not be generalized to a larger population).

The funny looking shape of the ASA model is due to the fact that we have only three distinct value for this predictor. The model must therefore assign all patients with the same score as either dead or alive after 90 days. This is a blunt tool!

There seems to be no relevant difference between the full and simplified model. This is good, since we can then avoid the cumbersome data retrieval of prescription data if using the model in practice. A simpler model with age and sex only, is almost as good, especially applied to a previously unseen data set (which is what we should care about).

It is evident that ROC curves from the training data over-estimate the predictive power of each model. This is clear since ROC curves for the evaluation set deviates less from the horizontal line.

```{r}
knitr::include_graphics("../graphs/rocs.png")
```

We could  summaries the ROC curves by their area under the curve (AUC). We have used bootstrap replicates and can therefore select the 2.5 % smallest and largest estimates as confidence limits based on the training data. We can present the same data either in a table, or graphically.

```{r}
load("../cache/auc_table.RData")
auc_table
```

```{r}
knitr::include_graphics("../graphs/auc_ci.png")
```

We can conclude that the winning model have the same AUC as the chosen model from the previous version of the manuscript (0.8). We could also conclude however that this is an optimistic estimate since we do not reach that far using the test data. We can also conclude that the full model is in fact no better than the slightly simplified model or the model with age and sex only.

It might seem strange that both CCI and ECI actually underestimate the AUC. We can see above however that corresponding curves based in the evaluation data are very close to the diagonal with just a few exceptions. We might suspect that the larger training set included some extreme values for those indices that became influential for the model building, but which were not present in the evaluation set. We said in the beginning that a common split between training data and evaluation data is 3:1. We had 9:1 (which we knew were risky). We did so because the outcome is so rare and we therefore wanted to maximize the available data for estimation.

Lets have a look at the extremes of the Elixhauser index (CCI is less extreme but have a similar tendency). Most patientd have ECI within the range -4 to 17. Those are excluded here:

```{r}
load("../cache/ECI_extremes.RData")
ECI_extremes %>% 
  filter(!ECI_index_walraven %in% -4:17)
```

Hence, there is only one individual with ECI = -14, and she was assigned to the training set. She had obesity, drug abuse and depression, all associated with a long and happy life according to Van Walraven (Sic!).

The patient with ECI = 33 on the other hand had congestive heart failure, cardiac arrhythmia, hypertension, renal failure and a solid cancer tumor with metastases.

Both of those patients actually survived 90 days after their surgery (which is an indication to why ECI does not work as a predictor of death)!

One way to avaoid exrteme observations for ECI/CCI would be to transform those variables by the Yeo-Johnson transformation. I tried this but it did not help. It might therefore be reasonable tom simply truncate the scale (or even ignore/exclude those extreme cases).

## Separation

Another attempt to illustrate each models ability to classify patients as either dead or alive is depicted below. If a model was able to fully predict death, the blue and red areas would be completely separated. If the model can not predict anything at all, those colors would completely blend into each other. More blue to the right means that people who actually died where also predicted to do so. We can see once more that the separation is larger when the models are applied to the training data, compared to the evaluation data set.

```{r}
knitr::include_graphics("../graphs/separation_auc.png")
```

### 2019-02-26 diagnostics/dataExplorer.R

@@ -3,20 +3,22 @@ load.project()
 
 config <- list(
   "introduce" = list(),
-  "plot_str"   = list(
-    "type"     = "diagonal",
-    "fontSize" = 35,
-    "width"    = 1000,
-    "margin"   = list("left" = 350, "right" = 250)
-  ),
-  "plot_missing"     = list(),
+  # "plot_str"   = list(
+  #   "type"     = "diagonal",
+  #   "fontSize" = 35,
+  #   "width"    = 1000,
+  #   "margin"   = list("left" = 350, "right" = 250)
+  # ),
+  #"plot_missing"     = list(),
   "plot_histogram"   = list(),
-  "plot_qq"          = list(sampled_rows = 1000L),
+  "plot_qq"          = list(),
   "plot_bar"         = list(),
   "plot_correlation" = list("cor_args" = list("use" = "pairwise.complete.obs")),
   # "plot_prcomp"    = list(),
-  "plot_boxplot"     = list(),
-  "plot_scatterplot" = list(sampled_rows = 1000L)
+  "plot_boxplot"     = list()#,
+  # "plot_scatterplot" = list()
 )
 
-DataExplorer::create_report(df, output_dir = "diagnostics", config = config)
+DataExplorer::create_report(
+  df, output_dir = "diagnostics", y = "death90f", config = config
+)

### 2019-03-07 reports/preliminary_results.Rmd
new version with correct comorbidity data. See (other project but related): https://www.pivotaltracker.com/story/show/164274282

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
 load("../cache/df.RData")
 library(tidyverse)
+options(digits = 2)
 ```
@@
 Hence even if patients were operated during 2015, they are not automatically censored at the end of that year.
-(Censoring would only occour in the case of migration etc.)
+(Censoring would only occour in the case of emigration etc.)
 Comorbidity indices based on Elixhauser van Walraven (ECI), Charlson's original index (CCI) and a modified Risk Rx IV index (modified by Anne Garland) were used.
@@
 education =
+
 * low = pre-gymnasium
@@
 
-BMI and age are approximately normally distributed.
-CCI and ECI have very little data (most of the patients have no recorded comorbidities).
-It should be discussed whether diagnoses set during hospitalization should be included.
-There are some negative ECI values as well (due to the van Walraven index).
+BMI and age are approximately normally distributed. There are some negative ECI values as well (due to the van Walraven index).
 
@@
 
-Type of hospital, sex, ASA, civil status, education and prosthesis group looks as expected.
-
-We have few patients with reversed hybrids (unbalanced sample).
+Type of hospital, sex, ASA, civil status, education and looks as expected.
 
@@
 Proportions for most variables are roughly the same as in earlier versions, although not exact due to the extended time frame of the study period.
-Comorbidity data differ but this is currently under discussion.
 
@@
 An additional benefit of down-sampling however is that we will also reduce the sample size (without losing too much power).
-This is beneficial since step-wise regression techniques around 100 variables and a large data set will otherwise be very time consuming.
+This is beneficial since step-wise regression techniques with around 100 variables and a large data set will otherwise be very time consuming.
 
@@
 A draw-back with the down-sampling however is that we might get a sample that is not representative for the larger population.
-Our estimates might be very sensitive to the individuals who just happened to show up in the sample.
-If we perform a step-wise regression on two different samples, it would therefore be possible that each data set will provide two different set of models (based on different predictors) [@Lukacs2010; @Sauerbrei1999; @Zellner2004].
+Our estimates might be very sensitive to the individuals who just happened to end up in the sample.
+If we perform a step-wise regression on two different samples, it would therefore be possible that each data set will provide two different sets of models (based on different predictors) [@Lukacs2010; @Sauerbrei1999; @Zellner2004].
 
@@
 Note that variables that are highly correlated could not be included in the same selection step, since it will then be arbitrary which one of those to select.
-I therefore used four independent step-wise regression procedures to each re-sample evaluating:
+I therefore used four independent step-wise regression procedures to each re-sample:
 
-1. General variables including comorbidity indices
+1. General variables
 2. Individual ECI diagnosis
@@
 4. Individual Risk Rx IV diagnosis
-
-Of course in the first group, it could be argued that the different comorbidity indices are correlated with each other. We did not see this in practice however.
 
@@
 It is usually recommended to only include variables that are in fact chosen in around 90 %
- (or even 95 % of the cases). This would be age, ASA, medication for inflammatory pain and sex in our case. We could use a 75 % limit however since this is not our final model.
-
-Note that heart infarction and renal disease did not show up here.
-Those were included last time but we can see from the table above that those were in fact choose in less than 75 % of our Bootstrap-samples.
-It might be argued that those were in fact important predictors in the original sample.
-The goal of prediction modelling however is not to evaluate the past but to predict the future!
+ (or even 95 % of the cases). We could use a 75 % limit however since this is not our final model.
 
@@
 The candidate variables above where found from four different starting points.
-We can now argue that those variables should not be heavily associated with each other.
-It would have been different if for example heart infarction would have been important if classified both by ECI, CCI and Rx.
-We could then combine those factors into one (which was done for the previous version of the manuscript).
-This might also be necessary with a different threshold than 75 %.
-It might nevertheless be relevant to use step-wise regression once more based on the smaller subset of candidate variables.
+It might therefore be relevant to use step-wise regression once more based on the smaller subset of candidate variables to account for possible dependencies.
 
@@
 
-Hence, only CCI, age and ASA = 3 seems to be important.
-
 # Additional models
 
-Given the results above, we might be tempted do drop the Rx-variables.
-It would be nice if those were not needed since acquiring this data in clinical practice might be cumbersome.
-Sex on the other hand is almost always available anyway.
-Hence, we'll better keep it.
-This will yield an alternative model for comparison:
-
-```{r}
-round(modsums$modsum[[2]]$coefmat.full, 2)
-```
-
-Figure 2 in the earlier version of the manuscript also included models with only ASA, CCI, ECI and Risk Rx. We will do too for comparison and can add a model  with age and sex only as well.
+We will compare our model to models with ASA, CCI, ECI and Risk Rx, as in the previous version of the manuscript. We can include a model with only age and sex as well.
 
@@
 
-We now 7 models that can be compared based on predictive power, thus by area under the (ROC) curve (AUC).
+We now 6 models that can be compared based on predictive power, thus by area under the (ROC) curve (AUC).
 Let's take another 1,000 bootstrap replicates and compare observations to predictions for those.
@@
 Those would probably we very wide though and are avoided to save (computational) time.
+We could do this as a later step before summarising the results however.
 
@@
 
-There seems to be no relevant difference between the full and simplified model.
-This is good, since we can then avoid the cumbersome data retrieval of prescription data if using the model in practice.
-A simpler model with age and sex only, is almost as good, especially applied to a previously unseen data set (which is what we should care about).
-
-It is evident that ROC curves from the training data over-estimate the predictive power of each model. This is clear since ROC curves for the evaluation set deviates less from the horizontal line.
+It is evident that ROC curves from the training data over-estimate the predictive power of each model, as well as that the differnece between models get over-estimated.
 
@@
 We could also conclude however that this is an optimistic estimate since we do not reach that far using the test data.
-We can also conclude that the full model is in fact no better than the slightly simplified model or the model with age and sex only.
 
 It might seem strange that both CCI and ECI actually underestimate the AUC.
-We can see above however that corresponding curves based in the evaluation data are very close to the diagonal with just a few exceptions.
+We can see above however that corresponding curves based on the evaluation data are very close to the diagonal with just a few exceptions.
 We might suspect that the larger training set included some extreme values for those indices that became influential for the model building, but which were not present in the evaluation set.
@@
 
-Lets have a look at the extremes of the Elixhauser index (CCI is less extreme but have a similar tendency). Most patientd have ECI within the range -4 to 17. Those are excluded here:
+Lets have a look at the extremes of the Elixhauser index (CCI is less extreme but have a similar tendency). Most patientd have ECI within the range -7 to 21. Those are excluded here:
 
@@
 ECI_extremes %>% 
-  filter(!ECI_index_walraven %in% -4:17)
+  filter(!ECI_index_walraven %in% -7:21)
 ```
 
-Hence, there is only one individual with ECI = -14, and she was assigned to the training set.
+Hence, there is only one individual with ECI = -11, and she was assigned to the training set.
 She had obesity, drug abuse and depression, all associated with a long and happy life according to Van Walraven (Sic!).
@@
 
-One way to avaoid exrteme observations for ECI/CCI would be to transform those variables by the Yeo-Johnson transformation.
+One way to avaoid extreme observations for ECI/CCI would be to transform those variables by the Yeo-Johnson transformation.
 I tried this but it did not help.
-It might therefore be reasonable tom simply truncate the scale (or even ignore/exclude those extreme cases).
+It might therefore be reasonable to simply truncate the scale (or even ignore/exclude those extreme cases).
 
@@
 ```
+
+# References

### 2019-03-07 src/extremes.R

@@ -19,7 +19,7 @@ cache("ECI_extremes")
 # Find the most HEALTHY ECI-person
 healthy <-
   df_train %>%
-  filter(ECI_index_walraven == -14)
+  filter(ECI_index_walraven == min(ECI_extremes$ECI_index_walraven))
 # comorbids
 healthy %>%
   select(starts_with("ECI_"), -ECI_index_walraven) %>%
@@ -31,7 +31,7 @@ select(healthy, death90f)
 # Find the most SICK ECI-person
 sick <-
   df_train %>%
-  filter(ECI_index_walraven == 33)
+  filter(ECI_index_walraven == max(ECI_extremes$ECI_index_walraven))
 # comorbids
 sick %>%
   select(starts_with("ECI_"), -ECI_index_walraven) %>%

### 2019-03-07 src/table1.R

@@ -1,19 +1,21 @@
 library(ProjectTemplate)
 load.project()
 
+fctrs <- c(
+  "P_ASA",
+  "ECI_index_walraven",
+  "CCI_index_quan_original",
+  "Rx_index_index"
+)
+
 t1 <-
   tableone::CreateTableOne(
-    vars = predictors,
-    factorVars = c(
-      "P_ASA",
-      "ECI_index_walraven",
-      "CCI_index_quan_original",
-      "Rx_index_index"
-    ),
+    vars = unique(c(fctrs, predictors)),
+    factorVars = fctrs,
     data = df
   )
 
-cache("table1")
+cache("t1")
 
 table1 <-
   t1 %>%

### 2019-04-01 reports/article/article.Rmd
Some code cleaning and rearrangements of scripts.

---
title: "Prediction of 90-day mortality after major surgery made simpler: An analysis
of different comorbidity measures based on XXXX patients from the Swedish Hip
Arthroplasty Register"
author: Anne Garland^1,2,3^, Erik Bülow^2,4^, Szilard Nemes^2,4^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
date: '2019-03-12'
bibliography: P:/library.bib
output: 
  bookdown::html_document2:
    toc: true
    toc_float: true
    df_print: kable
    theme: journal
abstract: |
  OBJECTIVE: Early mortality after one of the most common surgical procedures in developed countries, total hip arthroplasty (THA), is dependent on medical comorbidities. Complex diagnosis-based tools such as the Charlson Comorbidity Index (CCI) and the Elixhauser score as well as the prescription-based RxRiskV score have been used to estimate comorbidity, but we here investigate whether easily accessible data that are routinely collected in daily clinical practice are better at predicting early mortality after THA.

  DESIGN: Nationwide retrospective cohort study.

  SETTING: Database combining the Swedish Hip Arthroplasty Registry, the Swedish Patient Registry, and data from Statistics Sweden.

  PARTICIPANTS: 38,735 patients who had received a THA due to primary osteoarthritis between 2008 and 2012.

  PRIMARY OUTCOME MEASURE: Death within 90 days after surgery.

  RESLUTS: Unadjusted cumulative 90-day survival was 99.7 (95% confidence interval: 99.68 to 99.78) and the number of deaths was 109. Best predictive performance for 90-day mortality was found for a model combining age, gender, presence of heart infarction, renal disease, and American Society of Anaesthesiologists (ASA) grade(c=0.80). The established, more complex comorbidity measures RxRiskV Score (c= 0.66), CCI (c=0.64), and Elixhauser (c=0.62) all had inferior predictive performance, while ASA grade alone had a c-statistic of 0.70.

  CONCLUSION: A less demanding comorbidity measure based on age, gender, two important comorbidities, and ASA grade, is better at predicting early postoperative mortality after THA than commonly used comorbidity measures that depend on access to extensive information on ICD-codes or medication prescription. This novel tool could simplify research on mortality after surgical interventions and may result in easy-to-use risk estimators for daily clinical practice.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
source("../../lib/helpers.R")
```

# Affiliation and contact

1.	Department of Orthopaedics, Institute of Surgical Sciences, Uppsala University Hospital, Uppsala, Sweden  
2.	Swedish Hip Arthroplasty Register, Gothenburg, Sweden
3.	Department of Orthopaedics, Visby Hospital, Visby, Sweden 
4.	Department of Orthopaedics, Institute of Clinical Sciences, The Sahlgrenska Academy, University of Gothenburg, Gothenburg, Sweden

**Correspondence:** anne.l.garland@gmail.com, Tel: +46-498-26 80 00, Orthopeadic Department, Visby Hospital, St Göransgatan 10, 621 84 Visby, Sweden  
nils.hailer@surgsci.uu.se  
szilard.nemes@registercentrum.se  
erik.bulow@registercentrum.se  
goran.garellick@registercentrum.se

# Introduction

The presence of comorbidities is associated with a poorer outcome after THA surgery, ie an increased risk of early postoperative mortality, an increased risk of revision [@Inacio2015; @Gordon2013] and inferior patient-reported outcomes [@Gordon2013; @Hofstede2016]. In research settings, comorbidity is commonly measured using multi-facetted diagnosis- or prescription-based coding algorithms [@Bozic2013]. @Inacio2016 performed a study where the ability of three of the most commonly used coding algorithms to predict mortality after total hip arthroplasty (THA) and total knee arthroplasty (TKA), ie, Charlson Comorbidity Index (CCI), Elixhauser Score, and RxRiskV, was evaluated. These complex comorbidity measures are based on the availability of extensive databases including in- and outpatient data on ICD-codes, or detailed information on drug prescriptions prior to surgery. Oftentimes, such databases can only be created by combining several population-based registries, raising both ethical and practical concerns. Repeated modifications of the weighting of certain items contained in the comorbidity measures, and revalidations of the scores have been necessary [@Sundararajan2004; @Deyo1992; @Quan2011; @Cleves1997; @Walraven2009]. This has resulted in numerous variations of the same score, and hence uncertainty regarding the interpretation of, and comparison between different studies. Several universal and arthroplasty-specific risk prediction tools of various degrees of complexity have been introduced in an attempt to quantify the individual patients' risk profile, but none has been broadly accepted [@Manning2016].

An easily applicable tool with few dimensions is thus needed, both in research and in clinical practice. In the context of trauma, prediction tools are common, and it has been possible to reduce the number of variables without losing predictive power [@Gerdin2016]. In this study, we therefore aimed to investigate whether the use of easily accessible data, that are routinely collected, is an alternative to complicated coding algorithms in predicting the risk of early postoperative mortality after THA.

# Methods

## Study design and study population

```{r flowchart, fig.cap = "Flowchart."}
knitr::include_graphics("../../graphs/fig1.png")
```

We performed a nationwide retrospective cohort study (Figure \@ref(fig:flowchart)). Primary outcome measure was ninety-day mortality, the secondary outcome measure was one-year mortality. Patients registered in the Swedish Hip Arthroplasty Register (SHAR) who received a THA due to primary osteoarthritis between 2008 and 2012 were included. Only elective primary THA procedures were included, and all other diagnoses than primary osteoarthritis were excluded, in order to minimize the risk of selection bias. 
Follow-up started on the date of surgery and ended on the day of death, emigration, or December 31st 2012, whichever came first. Only the first surgery was accounted for in bilaterally operated patients in order to avoid dependency issues. Patients were excluded if a second THA surgery was performed within 12 months after the first THA surgery, ie, neither ninety-day nor one year mortality were influenced by any subsequent THA surgery. Information on age, gender, body mass index (BMI), diagnosis codes and prescriptions registered within one year prior to surgery, and socioeconomic background was included in order to calculate and compare the comorbidity measures described below.

## Sources of data

**The SHAR** has collected information on all THA procedures performed in Sweden since 1979. The SHAR has a completeness of 96-98% and has been validated repeatedly [@Soderman2000; @shpr2014; @Soderman2001].

**Statistics Sweden** is a state-owned registry collecting demographic and socioeconomic information on the entire Swedish population, eg, marital status, level of education, and personal and family income.

**The Swedish National Patient Register** was initiated in 1964 and contains information on diagnosis codes and dates of admissions and discharge for all individuals in Sweden. The positive predictive value of the Swedish National Patient Register is estimated around $90 \pm 5~\%$ [@Ludvigsson2011].

**The Drug Register** was initiated in 2005. The register contains information on all collected prescriptions in Sweden. 
Based on the ten-digit personal identity number that all Swedish citizens are assigned at birth or immigration, linkage of information from the databases described above is performed.

## Definitions

**The Charlson Comorbidity Index (CCI)** is a diagnose-based coding algorithm used in research [@Charlson1987]. It was developed through defining numerous clinical conditions and assessing their relevance in the prediction of one-year mortality. Each of the 19 comorbidities was then assigned a weighted score based on their relevance in prediction of one-year all-cause mortality. In the present study the original weighting of the Deyo-modification [@Deyo1992] of the index and the revised weighting according to Quan were both investigated [@Quan2011; @Charlson1987].

**The Elixhauser Score** is a diagnose-based coding algorithm used in research [@Elixhauser1998]. The Elixhauser Score is more detailed than the CCI, with 30 categories included in the original algorithm. In the present study the weighting of categories according to van Walraven was used [@Walraven2009].

**The RxRiskV Score** is a prescription-based coding algorithm used in research [@Fishman2003; @Sloan2003]. It has been argued that a prescription-based measure would be more reliable than a diagnosed based measure [@Iezzoni1997; @Johnson2006].
The American Society of Anaesthesiologists (ASA) physical status classification is a six-category physical status evaluation system that has remained unchanged except for the sixth category (brain-dead person) that was subsequently added for use in transplantation surgery [@Saklad1941]. Information on ASA grade was introduced in the SHAR in 2008, which is why our study period begins that year. ASA grade is routinely assessed pre-operatively in most developed countries. However, ASA grade has not been validated as a predictor of 90-day mortality after THA, and ASA grade is known to have a high degree of internal variability [@Haynes1995]. ASA grade has previously been compared to the CCI, but not with respect to mortality after THA [@Whitmore2014; @Kork2015].

## Statistics

We adhered to the guidelines on statistical analyses of register data [@Ranstam2011c; @Ranstam2011b]. Means, medians and ranges were used to describe continuous data. 95% confidence intervals (CI) described estimation uncertainty. Categorical data were investigated by cross-tabulation and the Chi-square test. The Kaplan-Meier method was used to estimate unadjusted cumulative survival. In order to calculate unadjusted and adjusted odds ratios (OR) with CI logistic regression models were fitted, with adjustment for relevant confounders (age, gender, BMI (Body Mass Index), type of hospital, fixation method, comorbidity, and socioeconomic position). Continuous variables were not categorized in order to strengthen the precision of the statistical analyses [@Altman2006]. 
The performance of the logistic regression models was evaluated by its discrimination capacity using c-statistics. The c-statistic represents the area under the curve (AUC in the plotted ROC [receiver operated characteristic] graph) and stands for the degree of predictability. We performed no imputation for missing data. 
Separate dimensions in the comorbidity algorithms (Elixhauser, CCI, and RxRiskV) that were considered relevant based on lower bounds of CI being larger than one were combined in a larger model together with age, gender, BMI, and ASA. One at a time, the different dimensions were then removed until the optimal combination with the highest c-statistic was found.

## Ethical approval

All patients registered in the SHAR have received information about the register and the type of data that are routinely collected prior to surgery, and all patients have been given the choice not to participate in the registry or associated research. In accordance with the Swedish Patient Data Law from 2009 written informed consent was not obtained from individual patients. Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (2013: 360-13).

# Results

## The study population

After the exclusion process described above (Figure \@ref(fig:flowchart)) 38,735 individuals with an age between 18 and 100 years who hade undergone THA due to primary OA were eligible for analysis. Their mean age was 68.6 years (SD 10.11), and there were somewhat more women (56.2%) than men in the study population. 
The highest proportion of patients with multiple comorbidities (ie, three or more) was identified by the RxRiskV measure (69.9%). The Elixhauser Score identified 5.0% of the study population as suffering from three or more comorbidities, whereas the proportion of patients with three or more comorbidities was 3.3% when calculated according to the CCI. The proportion of patients with an ASA grade above 2 was 15.2%.
See Table 1 for characteristics of the study population.

```{r}
load("../../cache/table1.RData")
prestext(table1)
```

## Ninety-day mortality after THA

Unadjusted cumulative 90-day survival was 99.7 (CI 99.68 to 99.78) and the number of deaths during this period was 109. We found that age (OR 1.1 [CI 1.09-1.14]) and male gender were associated with an increased risk of death (OR for female gender: 0.5 [CI 0.37-0.79]). Patients with a higher degree of comorbidity according to the CCI, the Elixhauser score, and the RxRiskV score had an increased risk of 90-day mortality than patients with a lower score. After adjustment for age, gender, BMI, type of hospital, fixation method, comorbidity, and socioeconomic status only the CCI and ASA grade above 2 remained statistically significantly associated with an increased risk of 90-day mortality. Among the investigated socioeconomic background variables only the marital status of widows (-ers) was statistically significantly associated with an increased risk of 90-day mortality (adjusted OR 2.6 [CI 1.67-3.93]). Unadjusted and adjusted ORs for 90-day mortality with 95% confidence intervals are presented in Table \@ref(tab:tab2).
Exploratory analyses of the excluded 151 patients with an ASA grade above 3 showed that 149 were classified as ASA 4, 2 as ASA 5, and none as ASA 6. 27 individuals with ASA grade 4 to 6 (18%) died during the study period.

```{r tab2}
load("../../cache/t2_univariable.RData")
prestext(t2_univariable)
```

## A combination of age, gender, presence of heart infarction or renal disease, and ASA grade is the best predictor of mortality

We found that the combination of age, gender, presence of heart infarction and renal disease, and ASA grade was the best predictor of both 90-day mortality (c=0.80) and one-year mortality (Figure 2 and Supplementary Figure 1). Introduction of socioeconomic variables such as education level and marital status into the multivariable regression model only marginally increased its predictive power (data where?). Adding information on BMI to the model did not strengthen the predictive power (Supplementary Table 1). 
The Elixhauser score was better at predicting 90-day mortality than any of its separate dimensions, with a c-statistic of 0.62 for the model containing all items of the Elixhauser score (see Table 3A and Figure 2).
When examining different dimensions of the CCI we found that the model containing all items of the CCI performed better in predicting 90-day and one-year mortality than the separate dimensions did individually. The original weighting was somewhat better at predicting 90-day mortality than the weighting according to Quan (c=0.64 vs. 0.61; see Table 3B and Figure 2.)
The prescription-based RxRiskV Score was better at predicting 90-day mortality than the diagnosis-based comorbidity measures (CCI and Elixhauser Score), with a c-statistic of 0.66. The model containing all items included in the RxRiskV Score was better at predicting 90-day mortality than the separate items investigated one at a time (See Table 3C and Figure 2).
The ASA classification alone had a better predictive strength for 90-day mortality (c=0.70) than the more complex comorbidity measures CCI, Elixhauser score, and RxRiskV score (See Figure 2).

# Discussion

A set of easily accessible data is a better predictor of early mortality after major surgery than complex coding algorithms 
In this nationwide cohort study we intended to compare the performance of a set of easily accessible data that are routinely collected in daily clinical practice with complex comorbidity coding algorithms (ie CCI, Elixhauser Score and RxRiskV). The best predictive strength was found for a relatively simple model including age, gender, presence of cardiac infarction or renal disease during the last 12 months prior to THA surgery, and ASA grade (c=0.81). Thios simple model was also better than the above mentioned comorbidity measures at predicting  one-year mortality (Supplementary Figure 1 and Supplementary Table 3).
Comorbidities are known to influence the outcome after THA [@Inacio2015; @Gordon2013; @Hofstede2016]. In order to assess the effect of comorbidity on early mortality after THA different coding algorithms have been proposed in research settings. The coding algorithms are complex, and hence they demand a merge of information on ICD-ocdes and medication prescriptions from several data sources. These coding algorithms are not used in clinical settings since the administrative burden associated with identifying some 30 ICD- or ATC-codes for every patient is not realistic. Thus, comorbidity measures based on patient administrative data are only accessible to researchers, but - even then -observational study designs are hampered by the usual limitations such as incompleteness and inaccuracy of coding [@Bozic2013]. 
In this present study we found that the prescription-based RxRiskV Score performed better than the diagnosis-based comorbidity comorbidity measures CCI and Elixhauser Score in predicting 90-day mortality. The original CCI was somewhat better than the Elixhauser Score in predicting 90-day mortality. This differs from earlier findings by Inacio et al. where the RxRiskV did not perform as well as the CCI and the Elixhauser Score, and where the c- statistics were generally higher than in our study [@Inacio2016]. Such dissimilarities could be explained by the facts that our study population was younger, included more women, and that we only included diagnoses and prescriptions registered one year prior to surgery. Overall, the predictive strength of all investigated diagnose- or prescription-based comorbidity measures was better than the included dimensions investigated separately. To put it differently, in terms of predicting mortality, each comorbidity measure was an improvement over the separate items included in each measure.
We also found that the ASA classification was better at predicting both 90-day and one-year mortality than the more complex coding algorithms, with a c- statistic of 0.70. The ASA classification has been repeatedly compared to the CCI, but no consensus as to which one is superior has been reached, and, to our knowledge, such comparisons have not been performed on THA population [@Whitmore2014; @Kork2015]. Individuals with an ASA grade of 4 to 6 were excluded from our study since those categories describe severe disease, moribund and brain-dead individuals, and it can be questioned whether the classification is correct, and - if it is - whether these patients should ever have received a THA. We thus excluded this very small group.
Obesity is generally known to be associated with a higher risk of morbidity and all-cause mortality [@Must2000]. However, previous studies on primary THA cohorts have not indicated a higher risk of mortality in obese patients, a result that is confirmed in our study (Supplementary Table 1) [@Wallace2014]. An explanation could be that obese patients selected for THA are comparably healthy.

Risk prediction may be useful in the process of patient selection prior to surgery, in the preoperative risk management including a review of current medications, and in perioperative anaesthesia management. A number of risk prediction tools of various complexity for adverse outcomes after total joint replacements have been introduced but none has been broadly accepted [@Manning2016]. In the context of trauma surgery outcome prediction tools are common, and it is seems possible to reduce the number of items without losing predictive power [@Gerdin2016]. Our results indicate that the risk of early postoperative mortality after THA could be assessed by a relatively simple prediction model.

## Strengths and limitations

A strength of this study is its nationwide design with a large cohort of primary THA patients with a reasonable number of events, rendering estimation relatively precise. Our sources of data are highly valid, and the proportion of missing data in our cohort was low [@Soderman2000; @shpr2014; @Soderman2001; @Ludvigsson2011]. Limitations to this study are the potential biases at different levels that are commonly associated with observational data, and the risk of coding errors as expected when dealing with patient administrative data. Selection bias is also an issue in this study, since patients who died on the table during attempted THA surgery may not have been reported to the SHAR, but, judging from clinical experience, such events are extremely uncommon in a population of patients scheduled for elective THA surgery for osteoarthritis. The much higher expected frequency of fatal events during THA surgery on hip fracture patients was one of the main reasons to exclude this group from the present analyses.
It is important to distinguish between explanatory observational research and attempts at predicting individual events such as early mortality after surgical interventions. The combination of parameters in the best-performing model described in the present study may serve as a predictor of mortality on an individual level, but the described combination of parameters would need to be validated in a different sample of individuals. Since this has not yet been done we cannot extrapolate our findings to prediction models in a clinical setting, but aim at performing such additional studies
Our results indicate that in research on mortality after a very common surgical intervention a less complex comorbidity measure consisting of easily accessible data that are routinely collected in daily clinical practice is superior to some of the commonly used diagnose- or prescription-based coding algorithms. It would be interesting to evaluate the ability of our novel set of parameters to predict adverse events and revision rates.

## Conclusion

Our results derived from a nationwide cohort study indicate that a less data demanding comorbidity measure, the combination of age, gender, presence of heart infarction or renal disease and ASA grade, serves is better at predicting early postoperative mortality after THA than comorbidity measures based on more complex coding algorithms.

# Contribution of authors

AG, NH: initiated the study and managed the ethical review board application. EB, SN and AG performed the statistical analyses. GG: Assisted in preparing the review board application. AG and EB drafted the manuscript. AG, EB, SN, NH and GG took part in designing the study and editing the manuscript.

# What this paper adds-box

## What is already known on this subject:

A higher comorbidity burden is associated with increased early mortality after major surgery. Established comorbidity measures rely on complex coding algorithms of data on ICD- and/or prescription codes, and thus they are cumbersome instruments in research and impossible to use in clinical practice.

## What this study adds:

The combination of age, gender, presence of heart infarction or renal disease, and ASA grade was a better predictor of mortality after primary THA than the more complex comorbidity measures Charlson Comorbidity Index, Elixhauser Score and RxRiskV. Thus, a five-item instrument that is easily available in a clinical setting could both simplify research on mortality after major surgery and be developed into a simple risk prediction tool.

# Bibliography

### 2019-04-01 reports/preliminary_results.Rmd
Some code cleaning and rearrangements of scripts.

@@
 (Censoring would only occour in the case of emigration etc.)
-Comorbidity indices based on Elixhauser van Walraven (ECI), Charlson's original index (CCI) and a modified Risk Rx IV index (modified by Anne Garland) were used.
+Comorbidity indices based on the sum of comorbidities recognized by Elixhauser et al.
+(ECI), Charlson's original index (CCI) and a modified Risk Rx IV index (modified by Anne Garland) were used.
+Each of those indices were truncated since higher values are rare.
 
@@
 4. No second bilaterals.
-5. No first bilateral if second bilateral inserted within a year.
+5. No first bilateral if second bilateral inserted within 90 days.
 6. No surface prosthesis
@@
 8. only BMI $\leq 50$
-9. Only ASA $\leq 3$
+9. Only ASA $\leq 5$
 10. Exclude if missing education level
@@
 
+We have also used some individual comorbidities recognized by ECI, CCI and Rx and combined those to identify individual comorbidities using an arbitrary source.
+Those variables are included with suffixes "c_" to their variable names.
+
 ## Data profiling report
@@
 
-Most of our variables are categorical (very few numeric)
+Most of our variables are categorical/binary.
 
@@
 
-I include all values of each comorbidity index for the moment but those should obviously be truncated/transformed later.
-
-Proportions for most variables are roughly the same as in earlier versions, although not exact due to the extended time frame of the study period.
+Note that some EPVs are low, thus that we have very few observed detahs among some levels!
+This can make modelleing hard.
+There is no perfect limit of cases, but 10-20 is commonly recommended and  [@Courvoisier2011] suggested to have at least 5!
 
@@
 
-We have a lot of potential predictors but very few events to predict.
+We have a lot of potential predictors but very few events to predict (low EPV).
 This is problematic and we risk overfittnig the model to the data.
@@
 The sampling is therefore stratified on the outcome.
+
+To have such a small estimation set implies however that the observed outcome will be very rare, and the result of validating the model on that data set will be unstable.
+In this case, the split-sample approach might in fact lead to more confusion.
+It has been argued by [@Austin2017a] that Bootsrtap-validation is superiod in this case.
+Another prominent predictive modelling guru, Max Kuhn however, does argue for the need of a separate validation test.
+The approach used for this project should be further discussed!
 
@@
 
-1. General variables
+1. General variables (including our newly combined comorbidities)
 2. Individual ECI diagnosis
@@
 load("../cache/modsums.RData")
-round(modsums$modsum[[1]]$coefmat.full, 2)
+round(modsums$modsum[[1]]$coefmat.full, 2) %>% 
+  as_tibble(rownames = "coef") %>% 
+  mutate(Est_exp = round(exp(Estimate), 2))
 ```
@@
 ECI_extremes %>% 
-  filter(!ECI_index_walraven %in% -7:21)
+  filter(!ECI_index_sum_all %in% -7:21)
 ```
@@
 
+This visualisation was just made up but a similar approach has been suggested by @Greenhill2011.
+
 ```{r}

### 2019-04-01 diagnostics/00. Data_management _AG_mortality.R

@@ -1,179 +1,0 @@
-# Jag vill med detta skript återskapa tidigare flödesschema från Szilard med gamla data
-# för att kolla så vi gör motsvarande urval etc.
-
-library(ProjectTemplate)
-load.project()
-
-anne_pred_orig <- readRDS(
-  "//RCHOST9/Stat_Arkiv/Z_Arkiv/SHPR/Szilard Nemes/Beslutstöd/Output/data_with_adverse_events_2016-08-22.rds")
-ane_pred <- anne_pred_orig
-
-
-# 001. OppAr < 2007
-anne_pred <- dplyr::filter(anne_pred, OppAr > 2007)
-dim(anne_pred)
-
-# 002. Primary arthros
-anne_pred <- dplyr::filter(anne_pred, DiaGrp == 1)
-dim(anne_pred)
-
-# 003. Only pateinst with 1 oprration
-LPN <- dplyr::filter(anne_pred, OpNr == 2)
-anne_pred <- dplyr::filter(anne_pred, !lpnr %in% LPN$lpnr)
-dim(anne_pred)
-
-# 004. Age between 18 and 100
-anne_pred <- dplyr::filter(anne_pred, (17 < Alder & Alder < 101))
-dim(anne_pred)
-
-# 005. No resurfacing
-anne_pred <- dplyr::filter(anne_pred, ProtGrp != 5)
-dim(anne_pred)
-
-#
-## Calculate BMI
-#
-anne_pred <- dplyr::mutate(anne_pred, bmi =  Vikt/(Langd/100)^2)
-sum(is.na(anne_pred$bmi))
-
-# 006. BMI < 51 and existing
-
-anne_pred <- dplyr::filter(anne_pred, !is.na(bmi)) %>%
-             dplyr::filter( bmi  <= 50)
-dim(anne_pred)
-
-# 007. ASA not 4 & 5
-
-anne_pred <- dplyr::filter(anne_pred, ASA < 4) %>%
-             droplevels()
-dim(anne_pred)
-
-
-
-
-
-### -------------- Format the socioeconomic variables -----------------------###
-
-# 001. Education level
-anne_pred <- dplyr::mutate(anne_pred, Sun2000niva_old = factor(gsub(" ", "", Sun2000niva_old))) %>%
-                  dplyr::mutate(Sun2000niva_old = factor(Sun2000niva_old,
-                                         levels = c('*', 1, 2, 3, 4, 5, 6, 7),
-                                         labels = c('NA',
[…]

### 2019-04-01 lib/best_model.R

@@ -1,0 +1,80 @@
+
+# Find best mdoel ---------------------------------------------------------
+
+# Model averaging with model weights based on AIC from bootstrap replicates
+# Note that each BS replicate is taken after the recipi is applied.
+# The downsmapling means that this might be of limited value, wherefore an
+# additional BS is applied below. We use sqrt(B) outer bootstrappinug replicates
+# and sqrt(B) inner replicates
+bsavg <- function(mod_data) {
+  model <- glm(death90f ~ .,
+               binomial, mod_data, x = TRUE, na.action = "na.fail")
+  dr          <- MuMIn::dredge(model)
+  am          <- model.avg(dr, data = mod_data)
+  Weights(am) <- bootWeights(am, R = floor(sqrt(config$N_bots)))
+  am
+}
+
+
+# Fit model based on variable names and average results based on bootstrap
+mod_avg <- function(nms) {
+
+  mod_avg <-
+    df %>%
+    # sqrt(B) outer bootstrap replicates
+    rsample::bootstraps(
+      ceiling(sqrt(config$N_bots)), strata = "death90f", apparent = TRUE) %>%
+    mutate(
+      recipes  = map(splits, prepper, recipe = reci(df), retain = TRUE),
+      # Only keep variables that are matched by the candidate vector names
+      mod_data = map(recipes, juice, death90f, matches(nms)),
+      am       = map(mod_data, bsavg),
+      w        = map(am, ~ enframe(c(Weights(.))))
+    )
+
+  # Average weights over the outer bootstrap replicates
+  w <-
+    mod_avg %>%
+    unnest(w) %>%
+    group_by(name) %>%
+    summarise(w = mean(value))
+
+  # Update weights for the whole training set wit hthe boostatrap averaged weights
+  fit <- mod_avg$am[mod_avg$id == "Apparent"][[1]]
+  Weights(fit) <- w$w
+  fit
+}
+
+
+# Simple glm with result ni list, help function to below
+glml <- function(nms) {
+  lr(juice(prep(reci(df)), death90f, matches(nms)))
+}
+
+
+pred <- function(..., model) {
+  pred_type <- switch(class(model)[1], "_glm" = "prob", averaging = "response")
+  p <- predict(model, ..., type = pred_type)
+  if (inherits(model, "_glm")) p <- 1 - p$.pred_alive
+  p
+}
+
+
+# Evaluate the final model based on ROC-curves and AUC
[…]

### 2019-04-01 lib/figures.R

@@ -1,0 +1,13 @@
+
+# We have more data than we need for a plot. Simplify by loess model
+# if more than 50 uniqe points on either axis
+loesspreds <- function(data) {
+  if (length(unique(data$sensitivity)) < 50 |
+      length(unique(data$specificity)) < 50) {
+    return(data)
+  }
+  spec <- seq(0, 1, 0.01)
+  sens <- loess(sensitivity ~ specificity, data) %>%
+    predict(spec)
+  tibble(specificity = spec, sensitivity = sens)
+}

### 2019-04-01 lib/modelling.R

@@ -1,0 +1,74 @@
+
+# Find candidate predictors ----------------------------------------------
+
+
+# Logistic regressoin using parsnip
+lr <- function(x) {
+  fit <-
+    logistic_reg() %>%
+    set_engine("glm") %>%
+    fit(death90f ~ ., data = x)
+}
+
+# Identify variables used in the best model according to stepwise regression
+bestfit <- function(x) {
+  vars <- setdiff(names(x), "death90f")
+  best <- stats::step(glm(death90f ~ ., binomial, x), trace = 0)
+
+  tibble(
+    var  = vars,
+    # Fins name matches fom original names (no dummies)
+    incl = map_lgl(vars, ~ any(startsWith(names(coef(best)), .)))
+  )
+}
+
+reci <- function(df, outcome = "death90f") {
+  rec <-
+    recipe(as.formula(paste(outcome, "~ .")), df) %>%
+    step_downsample(all_outcomes())
+
+  # Create dummy variables for categorical data
+  if (sum(vapply(df, is.factor, logical(1))) > 1)
+    rec <- rec %>% step_dummy(all_predictors(), all_numeric())
+  rec
+}
+
+
+# Use bootstrap resampling to find number of times each variable is selected in
+# stepwise regression
+find_predictors <- function(
+  df, rec = reci(df), B = config$N_bots, outcome = "death90f") {
+
+  df %>%
+    rsample::bootstraps(B, strata = outcome) %>%
+    mutate(
+      recipes  = map(splits, prepper, recipe = rec, retain = TRUE),
+      mod_data = map(recipes, juice),
+      included = future_map(mod_data, bestfit)
+    ) %>%
+    unnest(included) %>%
+    spread(var, incl)
+}
+
+# List names of predictors selected at least threshold % of the time
+list_predictors <- function(x, df, outcome = "death90f", thr = .9) {
+
+  winners_all <-
+    x %>%
+    summarise_at(vars(-id), mean) %>%
+    mutate_all(~ . >= thr) %>%
+    c(recursive = TRUE) %>%
+    {names(.)[.]}
+
+  preds    <- setdiff(names(df), outcome)
+  winners  <- preds[map_lgl(preds, ~ any(startsWith(winners_all, .)))]
+  loosers  <- setdiff(preds, winners)
+  extra    <- winners_all[!map_lgl(winners_all, ~ any(startsWith(., winners)))]
+
+  list(winners = winners, loosers = loosers, extra = extra)
+}
+
[…]

### 2019-04-01 munge/01-filter.R

@@ -3,16 +3,17 @@ N <- function() c(n_distinct(df_shpr$LopNr), nrow(df_shpr))                     
 
 # Endast 2008-2015 enligt data/linkage.R
 
-# Endast primäroperationer
+# Endast primäroperationer - preselektion
 df_shpr <-
   df_shpr_orig %>%
-  distinct(LopNr, P_Side, .keep_all = TRUE); N()                                  #
-
-# Endast OA
-df_shpr <- filter(df_shpr, P_DiaGrp == "Primär artros"); N()                      #
+  distinct(LopNr, P_Side, .keep_all = TRUE)
 
 # Endats THA
-df_shpr <- filter(df_shpr, P_ProstType == "Totalprotes"); N()                     #
+df_shpr <- filter(df_shpr, P_ProstType == "Totalprotes"); N()                     # 112818 127671
+
+# Endast OA
+df_shpr <- filter(df_shpr, P_DiaGrp == "Primär artros"); N()                      # 90137 102698
+
 
 # Indikera opnr och tid mellan operationer för de som har två
 df_shpr <-
@@ -27,21 +28,24 @@ df_shpr <-
   ungroup()
 
 # Endast första höft
-df_shpr <- filter(df_shpr, opnr == 1); N()                                       #
+df_shpr <- filter(df_shpr, opnr == 1); N()                                       # 80805
 
-# Ignorera om ytterligare THA inom ett år efter första
-df_shpr <- filter(df_shpr, is.na(time_between) | time_between > 365); N()         #
+# Ignorera om ytterligare THA inom 90 dagar efter första
+df_shpr <- filter(df_shpr, is.na(time_between) | time_between > 90); N()         # 79888
 
-df_shpr <- filter(df_shpr, is.na(P_KVA1) | P_KVA1 != "NFB62 - Primär total ytersättningspr"); N() #
-df_shpr <- filter(df_shpr, between(P_Age, 18, 100)); N()                          #
+# Endast cementfria
 df_shpr <- filter(df_shpr,
                   P_AcetCupCemMix != "Cementfritt",
-                  P_FemStemCemMix != "Cementfritt"); N()                          #
-df_shpr <- filter(df_shpr, P_BMI <= 50); N()                                      #
-df_shpr <- filter(df_shpr, P_ASA <= 3); N()                                       #
-df_shpr <- filter(df_shpr, !is.na(education)); N()                                #
[…]

### 2019-04-01 munge/02-survdata.R

@@ -2,39 +2,40 @@
 
 df <-
   df_shpr %>%
-  mutate_if(
-    is.logical,
-    coalesce, FALSE
-  ) %>%
-  mutate_at(
-    vars(contains("index")),
-    coalesce, 0
-  ) %>%
+  mutate_if(is.logical, coalesce, FALSE) %>%
+  mutate_at(vars(contains("index")), coalesce, 0) %>%
   mutate(
-    P_SurgYear = as.numeric(as.factor(substr(P_SurgDate, 1, 4))),
-    P_ASA = as.factor(P_ASA),
+    P_SurgYear = as.numeric(substr(P_SurgDate, 1, 4)),
+    P_ASA = pmin(P_ASA, 4),
 
     # Civilstånd
     civil_status = as.character(civil_status),
     civil_status =
-      ifelse(civil_status %in% c("divorced", "unmarried"), "single", civil_status) %>%
+      ifelse(
+        civil_status %in% c("divorced", "unmarried"),
+        "single", civil_status) %>%
       as.factor() %>% relevel("married"),
 
-    education = relevel(education, "low"),
-    P_TypeOfHospital = relevel(P_TypeOfHospital, "Universitets- eller regionssjukhus"),
-    P_Gender = relevel(P_Gender, "Man"),
+    education =
+      factor(education, levels(education), c("low", "middle", "high")),
+    P_TypeOfHospital =
+      factor(P_TypeOfHospital, levels(P_TypeOfHospital),
+        c("University", "County", "Rural", "Private"),
+      ),
+    P_Gender = factor(P_Gender, c("Man", "Kvinna"), c("Male", "Female")),
 
     # Survival
-    stime   = as.numeric(coalesce(DateOfDeath, as.Date("2018-02-01")) - P_SurgDate),
+    stime   =
+      as.numeric(coalesce(DateOfDeath, as.Date("2018-02-01")) - P_SurgDate),
     status  = !is.na(DateOfDeath),
 
     # Outcome
-    death90    = status & stime < 90,
-    death365   = status & stime < 365,
+    death90   = status & stime < 90,
+    death365  = status & stime < 365,
     death90f  = factor(death90,  c(FALSE, TRUE), c("alive", "dead")),
     death365f = factor(death365, c(FALSE, TRUE), c("alive", "dead")),
 
-    # rtuncate comorbidity indices
+    # truncate comorbidity indices
[…]

### 2019-04-01 src/extremes.R

@@ -11,7 +11,7 @@ ECI_extremes <-
     eval = df_test,
     .id = "dataset"
   ) %>%
-  count(dataset, ECI_index_walraven) %>%
+  count(dataset, ECI_index_sum_all) %>%
   spread(dataset, n)
 cache("ECI_extremes")
 
@@ -19,10 +19,10 @@ cache("ECI_extremes")
 # Find the most HEALTHY ECI-person
 healthy <-
   df_train %>%
-  filter(ECI_index_walraven == min(ECI_extremes$ECI_index_walraven))
+  filter(ECI_index_sum_all == min(ECI_extremes$ECI_index_sum_all))
 # comorbids
 healthy %>%
-  select(starts_with("ECI_"), -ECI_index_walraven) %>%
+  select(starts_with("ECI_"), -ECI_index_sum_all) %>%
   {names(.)[c(., recursive = TRUE)]}
 # Outcome = alive
 select(healthy, death90f)
@@ -31,10 +31,10 @@ select(healthy, death90f)
 # Find the most SICK ECI-person
 sick <-
   df_train %>%
-  filter(ECI_index_walraven == max(ECI_extremes$ECI_index_walraven))
+  filter(ECI_index_sum_all == max(ECI_extremes$ECI_index_sum_all))
 # comorbids
 sick %>%
-  select(starts_with("ECI_"), -ECI_index_walraven) %>%
+  select(starts_with("ECI_"), -ECI_index_sum_all) %>%
   {names(.)[c(., recursive = TRUE)]}
 # Outcome = alive
 select(sick, death90f)

### 2019-04-01 src/table1.R

@@ -3,27 +3,51 @@ load.project()
 
 fctrs <- c(
   "P_ASA",
-  "ECI_index_walraven",
+  "ECI_index_sum_all",
   "CCI_index_quan_original",
-  "Rx_index_index"
+  "Rx_index_index",
+  "P_SurgYear"
 )
+
+dft1 <-
+  df %>%
+  mutate(
+    Female = P_Gender == "Kvinna",
+    P_ASA = factor(P_ASA, 1:4, c(1:3, "4-5"))
+  ) %>%
+  select(
+    death90f,
+    Female,
+    one_of(predictors),
+    contains("_index_"),
+    starts_with("c_"),
+    starts_with("ECI_"),
+    starts_with("Rx_"),
+    -P_Gender
+  )
 
 t1 <-
   tableone::CreateTableOne(
-    vars = unique(c(fctrs, predictors)),
+    strata = "death90f",
+    vars = setdiff(names(dft1), "death90f"),
     factorVars = fctrs,
-    data = df
+    data = dft1,
+    test = FALSE
   )
-
-cache("t1")
 
 table1 <-
   t1 %>%
   print(
-    showAllLevels = TRUE,
     printToggle = FALSE
   ) %>%
-  as_tibble(rownames = "what")
+  as_tibble(rownames = "what") %>%
+  mutate(
+    what = gsub("= TRUE|P_|c_|TypeOf|P_Surg|ECI_|Rx_", "", what),
+    what = gsub("_", " ", what)
+  ) %>%
+  add_row(what = "Comorbidity indices",      .after = 31) %>%
+  add_row(what = "Combined comorbidities",   .after = 52) %>%
+  add_row(what = "Elixhauser comorbidities", .after = 69) %>%
+  add_row(what = "Risk Rx V comorbidities", .after = 75)
 
-writexl::write_xlsx(table1, "reports/table1.xlsx")
-
+cache_all(c("t1", "table1"))

### 2019-04-01 src/table2.R

@@ -10,12 +10,8 @@ lrm <- function(rhs) {
 }
 
 t2_univariable <-
-  tibble(
-    name = predictors
-  ) %>%
-  mutate(
-    coefs = map(name, lrm)
-  ) %>%
+  tibble(name = predictors) %>%
+  mutate(coefs = map(name, lrm)) %>%
   unnest(coefs) %>%
   mutate(
     var  = map2_chr(name, var, ~gsub(.x, "", .y, fixed = TRUE)),

### 2019-06-19 reports/article/article.Rmd
commit before removing RxRiskV

@@
 ---
-title: "Prediction of 90-day mortality after major surgery made simpler: An analysis
-of different comorbidity measures based on XXXX patients from the Swedish Hip
-Arthroplasty Register"
-author: Anne Garland^1,2,3^, Erik Bülow^2,4^, Szilard Nemes^2,4^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
+title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and extrenaly validated model based on Swedish and English register data"
+author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Adrian Sayers^5,6^, Szilard Nemes^2,4^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
 date: '2019-03-12'
@@
 output: 
-  bookdown::html_document2:
-    toc: true
-    toc_float: true
+  bookdown::word_document2:
+    toc: false
     df_print: kable
-    theme: journal
-abstract: |
-  OBJECTIVE: Early mortality after one of the most common surgical procedures in developed countries, total hip arthroplasty (THA), is dependent on medical comorbidities.
-Complex diagnosis-based tools such as the Charlson Comorbidity Index (CCI) and the Elixhauser score as well as the prescription-based RxRiskV score have been used to estimate comorbidity, but we here investigate whether easily accessible data that are routinely collected in daily clinical practice are better at predicting early mortality after THA.
-
-  DESIGN: Nationwide retrospective cohort study.
-
-  SETTING: Database combining the Swedish Hip Arthroplasty Registry, the Swedish Patient Registry, and data from Statistics Sweden.
-
-  PARTICIPANTS: 38,735 patients who had received a THA due to primary osteoarthritis between 2008 and 2012.
-
-  PRIMARY OUTCOME MEASURE: Death within 90 days after surgery.
-
-  RESLUTS: Unadjusted cumulative 90-day survival was 99.7 (95% confidence interval: 99.68 to 99.78) and the number of deaths was 109.
-Best predictive performance for 90-day mortality was found for a model combining age, gender, presence of heart infarction, renal disease, and American Society of Anaesthesiologists (ASA) grade(c=0.80).
-The established, more complex comorbidity measures RxRiskV Score (c= 0.66), CCI (c=0.64), and Elixhauser (c=0.62) all had inferior predictive performance, while ASA grade alone had a c-statistic of 0.70.
-
-  CONCLUSION: A less demanding comorbidity measure based on age, gender, two important comorbidities, and ASA grade, is better at predicting early postoperative mortality after THA than commonly used comorbidity measures that depend on access to extensive information on ICD-codes or medication prescription.
-This novel tool could simplify research on mortality after surgical interventions and may result in easy-to-use risk estimators for daily clinical practice.
 ---
@@
 ```{r setup, include = FALSE}
-knitr::opts_chunk$set(echo = FALSE)
-library(tidyverse)
-source("../../lib/helpers.R")
+options(
+  digits            = 1,
+  knitr.kable.NA    = '',
+  repos = list(CRAN = "https://cran.rstudio.com/")
+)
+knitr::opts_chunk$set(echo = FALSE, root.dir = '..')
+library(ProjectTemplate)
+load.project()
 ```
@@
 4.	Department of Orthopaedics, Institute of Clinical Sciences, The Sahlgrenska Academy, University of Gothenburg, Gothenburg, Sweden
+5. ...
+6. ...
 
-**Correspondence:** anne.l.garland@gmail.com, Tel: +46-498-26 80 00, Orthopeadic Department, Visby Hospital, St Göransgatan 10, 621 84 Visby, Sweden  
-nils.hailer@surgsci.uu.se  
-szilard.nemes@registercentrum.se  
-erik.bulow@registercentrum.se  
-goran.garellick@registercentrum.se
+**Correspondence:** 
+AG: anne.l.garland@gmail.com, Phone: +46 498 26 80 00, Postal: Orthopeadic Department, Visby Hospital, St Göransgatan 10, 621 84 Visby, Sweden  
+EB: erik.bulow@registercentrum.se  
+SN: szilard.nemes@registercentrum.se  
+NH: nils.hailer@surgsci.uu.se
+
+# Abstract
+
+**OBJECTIVE:** Early mortality after total hip arthroplasty (THA) is associated with comorbidity.
+Diagnosis-based tools such as the Charlson or Elixhauser Comorbidity indices, or the prescription-based RxRisk V score, have been used to quantify this.
+Those instruments are however complex and requires the use of extensive data sets not routinely available in most doctor-patient settings.
+We investigated if a simplified model, could substitute those complex indices to predict early mortality after THA.
+
+**PATIENTS AND METHODS:** We studied `r nrow(df)` patients with THA due to primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` from the Swedish Hip Arthroplasty Register, linked to national the population register, the National Patient Register and the longitudinal integration database for health insurance and labour market studies from Statistics Sweden.
+We used stepwise forward and backward logistic regression in combination with bootstrap averaging (bagging), to develop a prediction model for patient deaths within 90 days after surgery.
+Predictive power was assessed by the area under the reciever operating characteristic curve (AUC).
+The final model was applied to British data for external validation.
+
+**RESLUTS:** The unadjusted cumulative 90-day survival was `r surv90d`.
+Best predictive performance for 90-day mortality was found for a model combining `r coefs_print` (AUC = `r with(aucs, AUCtrain_est[name == "full"])`).
+This mdoel was superior to the established but complex comorbidity measures RxRiskV Score (`r with(aucs, AUCtrain_est[name == "RxRiskV"])`), the Charlson comorbidity index (AUC = `r with(aucs, AUCtrain_est[name == "CCI"])`), and the Elixhauser comorbidity score (AUC = `r with(aucs, AUCtrain_est[name == "ECI"])`).
+
+**CONCLUSION:** We found a relatively simple prediction model of 90 day mortality after total hip arthroplasty.
+This model requires less data and is easier to calculate compared to previously well-known comorbidity indices.
 
@@
 
-The presence of comorbidities is associated with a poorer outcome after THA surgery, ie an increased risk of early postoperative mortality, an increased risk of revision [@Inacio2015; @Gordon2013] and inferior patient-reported outcomes [@Gordon2013; @Hofstede2016].
+The presence of pre-surgery comorbidity is associated with poorer outcome after the insertion of total hip arthroplasty (THA).
+Patients with increased comorbidity an increased risk of early postoperative mortality and revision [@Inacio2015; @Gordon2013], as well as inferior patient-reported outcomes [@Gordon2013; @Hofstede2016].
 In research settings, comorbidity is commonly measured using multi-facetted diagnosis- or prescription-based coding algorithms [@Bozic2013].
-@Inacio2016 performed a study where the ability of three of the most commonly used coding algorithms to predict mortality after total hip arthroplasty (THA) and total knee arthroplasty (TKA), ie, Charlson Comorbidity Index (CCI), Elixhauser Score, and RxRiskV, was evaluated.
-These complex comorbidity measures are based on the availability of extensive databases including in- and outpatient data on ICD-codes, or detailed information on drug prescriptions prior to surgery.
-Oftentimes, such databases can only be created by combining several population-based registries, raising both ethical and practical concerns.
-Repeated modifications of the weighting of certain items contained in the comorbidity measures, and revalidations of the scores have been necessary [@Sundararajan2004; @Deyo1992; @Quan2011; @Cleves1997; @Walraven2009].
-This has resulted in numerous variations of the same score, and hence uncertainty regarding the interpretation of, and comparison between different studies.
-Several universal and arthroplasty-specific risk prediction tools of various degrees of complexity have been introduced in an attempt to quantify the individual patients' risk profile, but none has been broadly accepted [@Manning2016].
+@Inacio2016 studied the ability of three commonly used coding algorithms to predict mortality after THA and total knee arthroplasty: the Charlson (CCI) and Elixhauser (ECI) Comorbidity Indices, as well as RxRiskV.
+These comorbidity measures are quite complex to estimate and are based on the availability of extensive datasets including in- and outpatient data on ICD-codes, or detailed information on drug prescriptions prior to surgery.
+Oftentimes, such datasets can only be created by linking several population-based registries, raising both ethical and practical concerns.
+Also, each of the comorbidity index exist in numerous versions [@Sundararajan2004; @Deyo1992; @Quan2011; @Cleves1997; @Walraven2009].
+Interpretation and comparison between different studies is therefore difficult.
+Comorbidity data have also been used in several universal and arthroplasty-specific risk prediction tools to make risk profiles for individual patients.
+In the context of trauma, prediction tools are common, and it has been possible to reduce the number of variables without losing predictive power [@Gerdin2016].
+No model has so far been broadly accepted for elective THA however [@Manning2016].
+An easily applicable tool with few dimensions is thus needed, both in research and in clinical practice.
+We aimed to find such a model to predict the risk of 90-day postoperative mortality after THA.
 
-An easily applicable tool with few dimensions is thus needed, both in research and in clinical practice.
-In the context of trauma, prediction tools are common, and it has been possible to reduce the number of variables without losing predictive power [@Gerdin2016].
-In this study, we therefore aimed to investigate whether the use of easily accessible data, that are routinely collected, is an alternative to complicated coding algorithms in predicting the risk of early postoperative mortality after THA.
+# Patients and Methods
 
-# Methods
-
-## Study design and study population
-
-```{r flowchart, fig.cap = "Flowchart."}
+```{r flowchart, fig.cap = "Flowchart depicting inclusion criteria and number of patients included in the development phase of the model."}
 knitr::include_graphics("../../graphs/fig1.png")
@@
 
-We performed a nationwide retrospective cohort study (Figure \@ref(fig:flowchart)).
-Primary outcome measure was ninety-day mortality, the secondary outcome measure was one-year mortality.
-Patients registered in the Swedish Hip Arthroplasty Register (SHAR) who received a THA due to primary osteoarthritis between 2008 and 2012 were included.
-Only elective primary THA procedures were included, and all other diagnoses than primary osteoarthritis were excluded, in order to minimize the risk of selection bias.
-Follow-up started on the date of surgery and ended on the day of death, emigration, or December 31st 2012, whichever came first.
-Only the first surgery was accounted for in bilaterally operated patients in order to avoid dependency issues.
-Patients were excluded if a second THA surgery was performed within 12 months after the first THA surgery, ie, neither ninety-day nor one year mortality were influenced by any subsequent THA surgery.
-Information on age, gender, body mass index (BMI), diagnosis codes and prescriptions registered within one year prior to surgery, and socioeconomic background was included in order to calculate and compare the comorbidity measures described below.
+Patients recorded in the Swedish Hip Arthroplasty Register (SHAR) with cemented primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` were included in the development phase of the study (Figure \@ref(fig:flowchart)).
+Only the first THA was accounted for in bilaterally operated patients and patients were excluded if a second THA was inserted within 90 days after the first.
+Hence, the ninety-day mortality were not influenced by any subsequent THA surgery.
+Follow-up started on the day of surgery and ended at death, emigration, or December 31st `r max(df$P_SurgYear)`, whichever came first.
 
-## Sources of data
+Data linkage, based on the unique identity numbers assigned to all Swedish residents [@Ludvigsson2009], were used to collect data from a variaty of sources, as previsouly described by @Cnudde2016.
 
-**The SHAR** has collected information on all THA procedures performed in Sweden since 1979.
-The SHAR has a completeness of 96-98% and has been validated repeatedly [@Soderman2000; @shpr2014; @Soderman2001].
+Age, sex, body mass index (BMI), ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from SHAR, with a completeness of 96-98% [@shpr2017].
+Data on education level (low/middle/high) and civil status, were collected from the longitudinal integration database for health insurance and labour market studies from Statistics Sweden [@Ludvigsson2019].
+The Swedish National Patient Register was used for  comorbidity data during one year before surgery.
+The register contains all relevant diagnoses coded by ICD-10, as well as admissions and discharge dates for in- and outpatient visits in all private and public hospitals [@Ludvigsson2011].
+We used medical data from the national prescription register where all prescriptions are classified by the Anatomic Therapeutic Chemical classification system (ATC).
+Death dates were linked from the national population register.
 
-**Statistics Sweden** is a state-owned registry collecting demographic and socioeconomic information on the entire Swedish population, eg, marital status, level of education, and personal and family income.
+Comorbidity was recognized by individual ICD-10 and ATC codes grouped into 17 categories according to CCI [@Charlson1987; @Deyo1992; @Quan2005], 31 categories according to ECI and 42 categories according to the RxRiskV [@Fishman2003; @Sloan2003] classification.
+RxRiskV medications prescribed for inflammatory pain was excluded since those are also sold over-the-counter in grocery stores, yielding questionable completeness in the prescription register.
+Patients with no hospital visits or no prescriptions were assumed to have no comorbidity.
 
-**The Swedish National Patient Register** was initiated in 1964 and contains information on diagnosis codes and dates of admissions and discharge for all individuals in Sweden.
-The positive predictive value of the Swedish National Patient Register is estimated around $90 \pm 5~\%$ [@Ludvigsson2011].
-
-**The Drug Register** was initiated in 2005. The register contains information on all collected prescriptions in Sweden. 
-Based on the ten-digit personal identity number that all Swedish citizens are assigned at birth or immigration, linkage of information from the databases described above is performed.
-
-## Definitions
-
-**The Charlson Comorbidity Index (CCI)** is a diagnose-based coding algorithm used in research [@Charlson1987].
-It was developed through defining numerous clinical conditions and assessing their relevance in the prediction of one-year mortality.
-Each of the 19 comorbidities was then assigned a weighted score based on their relevance in prediction of one-year all-cause mortality.
-In the present study the original weighting of the Deyo-modification [@Deyo1992] of the index and the revised weighting according to Quan were both investigated [@Quan2011; @Charlson1987].
-
-**The Elixhauser Score** is a diagnose-based coding algorithm used in research [@Elixhauser1998].
-The Elixhauser Score is more detailed than the CCI, with 30 categories included in the original algorithm.
-In the present study the weighting of categories according to van Walraven was used [@Walraven2009].
-
-**The RxRiskV Score** is a prescription-based coding algorithm used in research [@Fishman2003; @Sloan2003].
-It has been argued that a prescription-based measure would be more reliable than a diagnosed based measure [@Iezzoni1997; @Johnson2006].
-The American Society of Anaesthesiologists (ASA) physical status classification is a six-category physical status evaluation system that has remained unchanged except for the sixth category (brain-dead person) that was subsequently added for use in transplantation surgery [@Saklad1941].
-Information on ASA grade was introduced in the SHAR in 2008, which is why our study period begins that year.
-ASA grade is routinely assessed pre-operatively in most developed countries.
-However, ASA grade has not been validated as a predictor of 90-day mortality after THA, and ASA grade is known to have a high degree of internal variability [@Haynes1995].
-ASA grade has previously been compared to the CCI, but not with respect to mortality after THA [@Whitmore2014; @Kork2015].
+Some comorbidities were identified by more than one classification and some distinct comorbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
+We identified 16 broader categories in addition to 5 standalone ECI classes and 14 RxRiskV classes that were kept unchanged.
+Groups were merged according to clinical relevance as to be recognized in a patient-doctor meeting without access to external register data (see supplementary data).
 
@@
 
-We adhered to the guidelines on statistical analyses of register data [@Ranstam2011c; @Ranstam2011b].
-Means, medians and ranges were used to describe continuous data.
-95% confidence intervals (CI) described estimation uncertainty.
-Categorical data were investigated by cross-tabulation and the Chi-square test.
-The Kaplan-Meier method was used to estimate unadjusted cumulative survival.
-In order to calculate unadjusted and adjusted odds ratios (OR) with CI logistic regression models were fitted, with adjustment for relevant confounders (age, gender, BMI (Body Mass Index), type of hospital, fixation method, comorbidity, and socioeconomic position).
-Continuous variables were not categorized in order to strengthen the precision of the statistical analyses [@Altman2006].
-The performance of the logistic regression models was evaluated by its discrimination capacity using c-statistics.
-The c-statistic represents the area under the curve (AUC in the plotted ROC [receiver operated characteristic] graph) and stands for the degree of predictability.
-We performed no imputation for missing data.
-Separate dimensions in the comorbidity algorithms (Elixhauser, CCI, and RxRiskV) that were considered relevant based on lower bounds of CI being larger than one were combined in a larger model together with age, gender, BMI, and ASA.
-One at a time, the different dimensions were then removed until the optimal combination with the highest c-statistic was found.
+We used the Kaplan-Meier estimator to assess unadjusted cumulative survival.
+
+Further analysis was based on logistic regression since no censoring occoured within the 90 day study period.
+
+Univariable models were applied to all predictors individually to estimate odds ratios for descriptive purposes.
+
+All potential predictors were then partioned into four groups where there were assumed to be no colinearity among variables: 1) the 17 combined comorbidities, 2) the 5 ECI comorbidities, 3) the 14 RxRiskV comorbidities, and 4) age, sex, BMI, ASA, type of hospital, year of surgery, education and civil status.
+
+1,000 bootstrap samples were drawn from the initial data set.
+Early mortality is very rare for patients with elective THA however.
+Downsampling stratified on the outcome was therefore applied to each sample to include only as many survivals as deaths.
+This gave us balanced samples and managable computations.
+[@King2001; @Courvoisier2011; @Austin2017a].
+
+A step-wise linear regression procedure based on the Aikake Information Criteria (AIC) was applied to each bootstrap sample to identify important predictors from each of the four groups of variables [@Sauerbrei1999; @Zellner2004; @Lukacs2010].
+Variables that were kept in the final model for at least 75 % of all bootstrap samples were selected as candidates for further consideration.
+
+A full main effects model were made to include all previously selected candidate predictors.
+Two modified versions were considered as well, where age was modeled by resttricted cubic spline with either 2 or 3 knots.
+Three simplified models with age and sex only, were used for comparison, as well as four univariable regression models with ASA, CCI, ECI and RxRiskV alone.
+
+We then took another 1,000 downsampled bootsrtap samples to estimate coefficients for each model.
+Model averaging based on the information criteria was used for the full main effects model and its variants.
+Hence, an automated model selection procedure evaluating all possible subsets of the included main effects, yielding averaged coeffients estimates based on variable importance.
+
+Each model, now with their coefficient estimates fixed, were then evaluated by their predictive power based on their area under the reciever operation characteristics (ROC) curve (AUC).
+Point estimates as well as 95 % confidence intervals were estimated by emperical bootstrap, once again with 1,000 resamples but this time without downsampling.
+Sensitivity and specificity from each bootstrap sample were subjected to a local polynomial regression fit (loess) to illustrate an averaged reciever operation characteristics (ROC) curve for each model.
+
+We used `r substr(R.version.string, 1, 15)` (R Foundation for Statistical   Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC, MuMIn and rFSA [@Lambert2018].
 
@@
 
-All patients registered in the SHAR have received information about the register and the type of data that are routinely collected prior to surgery, and all patients have been given the choice not to participate in the registry or associated research.
-In accordance with the Swedish Patient Data Law from 2009 written informed consent was not obtained from individual patients.
-Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (2013: 360-13).
+Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (271-14 and 360-13).
 
@@
 
-## The study population
+There were `r nrow(df)` patients (Figure \@ref(fig:flowchart)), `r paste(range(df$P_Age), collapse = " - ")` years old, whereof `r round(mean(df$P_Sex == "Female") * 100)` % were female.
+`r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` patients died within 90 days and no one was censored before that.
+The unadjusted cummulative 90-day survival was `r surv90d`.
 
-After the exclusion process described above (Figure \@ref(fig:flowchart)) 38,735 individuals with an age between 18 and 100 years who hade undergone THA due to primary OA were eligible for analysis.
-Their mean age was 68.6 years (SD 10.11), and there were somewhat more women (56.2%) than men in the study population.
-The highest proportion of patients with multiple comorbidities (ie, three or more) was identified by the RxRiskV measure (69.9%).
-The Elixhauser Score identified 5.0% of the study population as suffering from three or more comorbidities, whereas the proportion of patients with three or more comorbidities was 3.3% when calculated according to the CCI.
-The proportion of patients with an ASA grade above 2 was 15.2%.
-See Table 1 for characteristics of the study population.
+Some characteristics of the study population are presented in Table \@ref(tab:tab1).
+`r mean(df$Rx_index_index != 0) * 100` % of all patients had at least one pre-surgery comorbidity according to RxRiskV, but only `r mean(df$CCI_index_quan_original != 0) * 100` % and `r mean(df$ECI_index_sum_all  != 0) * 100` % according to ECI and CCI.
+The proportion of patients with ASA class 3 was `r mean(df$P_ASA == 3) * 100` %.
+Most individual comorbidities were more common among patients who died.
 
-```{r}
-load("../../cache/table1.RData")
-prestext(table1)
+```{r tab1}
+prestext(table1) %>% 
+  knitr::kable(caption = "Baseline demographic information on the study population 2008-2012.
+CCI/ECI = Charlson/Elixhauser Comorbidity Index.
+Individual comorbidities without prefixes were based on combinations of different comorbidity classifications.
+Prefix ECI and Rx(RiskV) indicate classification specific comorbidities.")
 ```
 
-## Ninety-day mortality after THA
-
-Unadjusted cumulative 90-day survival was 99.7 (CI 99.68 to 99.78) and the number of deaths during this period was 109.
-We found that age (OR 1.1 [CI 1.09-1.14]) and male gender were associated with an increased risk of death (OR for female gender: 0.5 [CI 0.37-0.79]).
-Patients with a higher degree of comorbidity according to the CCI, the Elixhauser score, and the RxRiskV score had an increased risk of 90-day mortality than patients with a lower score.
-After adjustment for age, gender, BMI, type of hospital, fixation method, comorbidity, and socioeconomic status only the CCI and ASA grade above 2 remained statistically significantly associated with an increased risk of 90-day mortality.
-Among the investigated socioeconomic background variables only the marital status of widows (-ers) was statistically significantly associated with an increased risk of 90-day mortality (adjusted OR 2.6 [CI 1.67-3.93]).
-Unadjusted and adjusted ORs for 90-day mortality with 95% confidence intervals are presented in Table \@ref(tab:tab2).
-Exploratory analyses of the excluded 151 patients with an ASA grade above 3 showed that 149 were classified as ASA 4, 2 as ASA 5, and none as ASA 6.
-27 individuals with ASA grade 4 to 6 (18%) died during the study period.
+Odds ratios from univariable logistic regresison models are presented in Table \@ref(tab:tab2).
+Higher age, ASA, CCI, ECI and RxRiskV, as well as being male or widow(er), or having cancer, gout, kidney disease or heart infarct, increase the risk of death within 90 days after surgery.
+Private hospitals tend to operate healthier patients and the odds of death decrease slightly by year of surgery.
 
 ```{r tab2}
-load("../../cache/t2_univariable.RData")
-prestext(t2_univariable)
+prestext(t2_univariable) %>% 
+  knitr::kable(caption = "Unadjusted odds ratios (OR) for 90-day mortality with 95 % confidence intervals (CI).
+Individual comorbiditities are only presented if selected in at least 75 % of all bootstrap samples (selected).
+Variables included in the final model are marked with 'FM' in model column.")
 ```
 
-## A combination of age, gender, presence of heart infarction or renal disease, and ASA grade is the best predictor of mortality
+ASA class was the only variable included in the final models by all stage-wise regression procedures from each of the 1,000 bootstrap samples (100 % in column "selected" in Table \@ref(tab:tab2)).
+Most potential predictors were rarely selected and only `r coefs_print`, where selected at least 75 % of the time (Table \@ref(tab:tab2)).
 
-We found that the combination of age, gender, presence of heart infarction and renal disease, and ASA grade was the best predictor of both 90-day mortality (c=0.80) and one-year mortality (Figure 2 and Supplementary Figure 1).
-Introduction of socioeconomic variables such as education level and marital status into the multivariable regression model only marginally increased its predictive power (data where?).
-Adding information on BMI to the model did not strengthen the predictive power (Supplementary Table 1).
-The Elixhauser score was better at predicting 90-day mortality than any of its separate dimensions, with a c-statistic of 0.62 for the model containing all items of the Elixhauser score (see Table 3A and Figure 2).
-When examining different dimensions of the CCI we found that the model containing all items of the CCI performed better in predicting 90-day and one-year mortality than the separate dimensions did individually.
-The original weighting was somewhat better at predicting 90-day mortality than the weighting according to Quan (c=0.64 vs.
-0.61; see Table 3B and Figure 2.)
-The prescription-based RxRiskV Score was better at predicting 90-day mortality than the diagnosis-based comorbidity measures (CCI and Elixhauser Score), with a c-statistic of 0.66.
-The model containing all items included in the RxRiskV Score was better at predicting 90-day mortality than the separate items investigated one at a time (See Table 3C and Figure 2).
-The ASA classification alone had a better predictive strength for 90-day mortality (c=0.70) than the more complex comorbidity measures CCI, Elixhauser score, and RxRiskV score (See Figure 2).
+```{r auc, warning = FALSE}
+knitr::kable(table_auc, caption = "Area Under the Curve (AUC) as a measure of predictive power for the full model compared to simple models with standard comorbidity measures, as well as a reference model with age and sex.
+Age was included as either a main effect, or in the form of restricted cubic splines (rcs) with either 2 or 3 knots.")
+
+```
+
+A model with those covariates, the "full" model, had an out-of-sample AUC of `r aucs$AUCtrain_est[[1]]`.
+The predictive power did not improve with age modeled by restricted cubic splines (Table \@ref(tab:auc) and Figure \@ref(fig:aucci)).
+The full model led to better predictions compared to a simplified model with only age and sex, or any of the conventional comorbidity indices (CCI, ECI and RxRiskV).
+The full model is further described in Table \@ref(tab:fullmodel).
+
+```{r aucci, fig.cap = "Area Under the Curve (AUC) as a measure of predictive power for the full model compared to simple models with standard comorbidity measures, as well as a reference model with age and sex.
+Age was included as either a main effect, or in the form of restricted cubic splines (rcs) with 2 or 3 knots.
+Confidence intervals based on 1,000 non-paramteric bootstrap replicates."}
+
+knitr::include_graphics("../../graphs/auc_ci.png")
+
+```
+
+```{r fullmodel}
+knitr::kable(table_full_model_coefs, caption = "Estimated odds ratios for variables in the full model based on full averages from 1,000 bootstrap samples.")
+```
+
+ROC curves for all models compared to the full model are displayed in Figure \@ref(fig:rocs). Each models ability separate survivors from death are further illustrated in Figure \@ref(fig:sep).
+
+```{r rocs, fig.cap = "ROC curves for each model compared to the full main effects model."}
+knitr::include_graphics("../../graphs/rocs.png")
+```
+
+```{r sep, fig.cap = "Separation"}
+knitr::include_graphics("../../graphs/separation_auc.png")
+```
 
 # Discussion
+
+We found that a univariable main effects logistic regression model with `r coefs_print` was superior to predict death within 90 days after insertion of elective THA.
+Some covariates are not statistically significant themselves (\@ref(tab:fullmodel)) but are still relevant due to unobserved heteregeneity  [@Mood2010].
+
+It is well known that male sex is associated with earlier deaths.
+It is also well known that the remaining life span will decrease with incresed age.
+It is however not obvious that this relation must be striclty linear.
+We therefore allowed for a more flexible relation based on restricted cubic splines, but found that a linear relationship was indeed most sufficient.
+
+ASA class was the only covariate chosen by all stepwise regression procedures.
+This might not be surprising due to its definition.
+It is interesting to note however that a univariable model with ASA grade alone, yielded the lowest predictive power of all models compared.
+An out-of-sample point estimate of its AUC was < 0.5 (\@ref(tab:fullmodel) and \@ref(fig:rocs), meaning that predictions based solely on ASA would be worse than random.
+This shows the important of validation, due to the risk of overfitting, as well as the importance of multivariable adjustment.
+ASA class is routinely assessed pre-operatively in most developed countries.
+It is however known to have a high degree of internal variability [@Haynes1995].
+It has previously been compared to the CCI, but not with respect to mortality after THA [@Whitmore2014; @Kork2015].
+It has been argued that a prescription-based measure would be more reliable than a diagnosed based measure [@Iezzoni1997; @Johnson2006].
+We did confirm that estimation of AUC were more stable for the RxRiskV index compared to CCI and ECI (as seen by the width of the confidence intervals in figure \@ref(fig:aucci)), although point estimates of the latter was marginaly higher.
+This reflects that both coverage and completeness from the prescription register is better compared to the national patient regiseter.
+The first hold records of all medications prescribed, the second only contains information that the doctor deemed  relevant at the patient encounter.
+
+# Further discussion from Garland not yet adjusted to new models
 
@@
 
-## Strengths and limitations
-
 A strength of this study is its nationwide design with a large cohort of primary THA patients with a reasonable number of events, rendering estimation relatively precise.
@@
 
-## Conclusion
-
 Our results derived from a nationwide cohort study indicate that a less data demanding comorbidity measure, the combination of age, gender, presence of heart infarction or renal disease and ASA grade, serves is better at predicting early postoperative mortality after THA than comorbidity measures based on more complex coding algorithms.
@@
 
-AG, NH: initiated the study and managed the ethical review board application.
-EB, SN and AG performed the statistical analyses.
-GG: Assisted in preparing the review board application.
+AG and NH initiated the study and managed the ethical review board application.
+EB, AS, SN, and AG performed the statistical analyses.
 AG and EB drafted the manuscript.
-AG, EB, SN, NH and GG took part in designing the study and editing the manuscript.
-
-# What this paper adds-box
-
-## What is already known on this subject:
-
-A higher comorbidity burden is associated with increased early mortality after major surgery.
-Established comorbidity measures rely on complex coding algorithms of data on ICD- and/or prescription codes, and thus they are cumbersome instruments in research and impossible to use in clinical practice.
-
-## What this study adds:
-
-The combination of age, gender, presence of heart infarction or renal disease, and ASA grade was a better predictor of mortality after primary THA than the more complex comorbidity measures Charlson Comorbidity Index, Elixhauser Score and RxRiskV.
-Thus, a five-item instrument that is easily available in a clinical setting could both simplify research on mortality after major surgery and be developed into a simple risk prediction tool.
+All athours edited and finalized the manuscript.
 

### 2019-06-19 lib/BRLasso.R

@@ -1,0 +1,66 @@
+# Bootstrap ranking Lasso copied (and modified) from
+
+BRLasso <- function(x, y, B = 5, Boots = 100, kfold = 10) {
+
+  varx <- varx0 <- colnames(x)
+  nvar          <- length(varx)
+  n             <- length(y)
+  var_list_all  <- vector("list", B)
+
+  RecordM2 = matrix(0, nrow = B, ncol = nvar)
+  rownames(RecordM2) <- paste0("B", seq_len(B))
+  colnames(RecordM2) <- varx
+
+  for (ij in seq_len(B)) {
+    Nboot <- n
+    RecordM = matrix(0, nrow = Boots, ncol = nvar)
+    rownames(RecordM) = paste0("BSample", seq_len(Boots))
+    colnames(RecordM) = varx
+    for (i in seq_len(Boots)) {
+      repeat {
+        s <- sample(Nboot, replace = TRUE)
+        if (length(table(y[s])) >= 2 & length(table(y[-s])) >= 2)
+          break
+      }
+
+      cvfit <- glmnet::cv.glmnet(x[s,], y[s], nfolds = kfold, family = "binomial")
+      model.final <- cvfit$glmnet.fit
+      nzero       <- as.matrix(coef(model.final, s = cvfit$lambda.min))
+
+      for (i0 in seq_len(dim(nzero)[1])) {
+        for (j0 in seq_len(nvar)) {
+          if (dimnames(nzero)[[1]][i0] == colnames(RecordM)[j0]) {
+            RecordM[i, j0] = nzero[i0,]
+          }
+        }
+      }
+      message("Step 1-Current iteration: ", i)
+    }
+
+    Impor     <- abs(apply(RecordM, 2, mean))
+    score     <- sort(Impor)
+    index     <- seq_along(score)
+    model.pwl <- SiZer::piecewise.linear(x = index, y = score)
+    break_p   <- round(model.pwl$change.point)
+    var_list_all[[ij]] <- names(score[break_p:length(index)])
+    if (length(var_list_all[[ij]]) != 0) {
+      varx0 <- intersect(var_list_all[[ij]], varx0)
+    }
+    Impor0    <- apply(RecordM, 2, mean)[var_list_all[[ij]]]
+    for (gp in seq_along(Impor0)) {
+      RecordM2[ij, colnames(RecordM2) == names(Impor0)[gp]] <- Impor0[gp]
+    }
[…]

### 2019-06-19 lib/globals.R

@@ -2,7 +2,10 @@
 memory.limit(1e10)
 set.seed(132456798)
 future::plan("multiprocess")
-options('na.action' = "na.fail")
+options(
+  digits = 2,
+  na.action = "na.fail"
+)
 
 # Add any project specific configuration here.
 add.config(
@@ -11,6 +14,6 @@ add.config(
 
 # Add project specific configuration that can be overridden from load.project()
 add.config(
-  N_bots         = 10e3,
+  N_bots         = 1e3,
   apply.override = TRUE
 )

### 2019-06-19 lib/helpers.R

@@ -1,12 +1,12 @@
 con <- shar_linkage_con()
 
 predictors <- c(
-  "P_Age", "P_Gender", "P_BMI", "P_ASA", "P_TypeOfHospital", "P_SurgYear",
+  "P_Age", "P_Sex", "P_BMI", "P_ASA", "P_TypeOfHospital", "P_SurgYear",
   "education", "civil_status"
 )
 
 # Cahe list of pobjects at once
-cache_all <- function(x, ...) Vectorize(cache, "variable")
+cache_all <- Vectorize(cache, "variable")
 
 ### Summary function for GLM
 glmCI <- function(model, exponent = FALSE, alpha = 0.05) {
@@ -38,7 +38,7 @@ survEst <- function(fit, times) {
 
 # Change table text to labels that can be presented
 prestext <- function(df) {
-  gs <- function(x, pattern, ...) gsub(pattern, x, ...)
+  gs <- function(x, pattern, replacement) gsub(pattern, replacement, x)
   df %>%
   rename(
     variable = 1,
@@ -48,9 +48,9 @@ prestext <- function(df) {
     level    = tolower(level),
     variable =
       variable %>%
-      gs("P_|Surg|TypeOf", "") %>%
-      gs("_", " ") %>%
-      gs("Gender", "Sex"),
+      gs("Rx_index_index", "RxRiskV") %>%
+      gs("P_|c_|Surg|TypeOf|_index_sum_all|_index_quan_original", "") %>%
+      gs("_", " "),
 
     # Initial capital letter
     variable = paste0(toupper(substr(variable, 1, 1)), substring(variable, 2))

### 2019-06-19 munge/02-survdata.R

@@ -6,7 +6,7 @@ df <-
   mutate_at(vars(contains("index")), coalesce, 0) %>%
   mutate(
     P_SurgYear = as.numeric(substr(P_SurgDate, 1, 4)),
-    P_ASA = pmin(P_ASA, 4),
+    P_ASA = factor(P_ASA),
 
     # Civilstånd
     civil_status = as.character(civil_status),
@@ -22,7 +22,7 @@ df <-
       factor(P_TypeOfHospital, levels(P_TypeOfHospital),
         c("University", "County", "Rural", "Private"),
       ),
-    P_Gender = factor(P_Gender, c("Man", "Kvinna"), c("Male", "Female")),
+    P_Sex = factor(P_Gender, c("Man", "Kvinna"), c("Male", "Female")),
 
     # Survival
     stime   =

### 2019-06-19 munge/03-compositvariabler.R

@@ -20,7 +20,7 @@ df <-
       ECI_hypertension_uncomplicated            |
       ECI_hypertension_complicated              |
       Rx_hypertension,
-    c_kidney =
+    c_kidney_disease =
       ECI_renal_failure                         |
       CCI_renal_disease                         |
       Rx_end_stage_renal_disease,
@@ -30,7 +30,7 @@ df <-
       CCI_diabetes_without_complication         |
       CCI_diabetes_complication                 |
       Rx_diabetes,
-    c_vascular =
+    c_vascular_disease =
       ECI_peripheral_vascular_disorder          |
       CCI_peripheral_vascular_disease           |
       CCI_cerebrovascular_disease               |
@@ -53,7 +53,7 @@ df <-
       CCI_moderate_or_severe_liver_disease      |
       Rx_liver_failure                          |
       Rx_hepatitis_c,
-    c_lung_airways =
+    c_lung_airways_disease =
       ECI_chronic_pulmonary_disease             |
       ECI_pulmonary_circulation_disorder        |
       CCI_chronic_pulmonary_disease             |

### 2019-06-19 munge/04-candidates.R

@@ -1,0 +1,35 @@
+
+# Find candidate predictors -----------------------------------------------
+
+# Screen all comorbidity measures as well as other factors for
+# predictors to include in combined model
+important_factors <-
+  tibble(preds = c("ECI", "CCI", "Rx", "c_")) %>%
+  mutate(data = map(preds, getdata)) %>%
+  add_row(
+    preds = "general",
+    data = list(select(df, death90f, predictors))
+  ) %>%
+  mutate(
+    facts   = map(data, find_predictors, B = config$N_bots),
+    impfact = map2(facts, data, list_predictors, outcome = "death90f", thr = .75)
+  )
+
+# Character vector of names of candidate predictors
+candidates <- c(map(important_factors$impfact, "winners"), recursive = TRUE)
+drop       <- c(map(important_factors$impfact, "loosers"), recursive = TRUE)
+
+cache_all(c("important_factors", "candidates", "drop"))
+
+
+# Proportions of all variables --------------------------------------------
+
+prop_selected <-
+  important_factors$facts %>%
+  map(summarise_at, vars(-id), mean) %>%
+  c(recursive = TRUE) %>%
+  enframe() %>%
+  mutate(value = value * 100) %>%
+  arrange(desc(value))
+
+cache("prop_selected")

### 2019-06-19 munge/05-best_model.R

@@ -1,0 +1,66 @@
+
+# Cancidate names prepared for reular expression matching
+nms <- paste(setdiff(candidates, "P_Age"), collapse = "|")
+
+# Object with models to compare and evaluate
+models <-
+  tribble(
+    ~name,                ~preds,                                                ~ modavg,
+    "full",                paste0(nms, "|P_Age"),                                TRUE,
+    "full (rcs 2)",        paste0(nms, "|P_Age_ns_[12]"),                        TRUE,
+    "full (rcs 3)",        paste0(nms, "|P_Age_ns_[123]"),                       TRUE,
+    "full-cancer",       paste(setdiff(candidates, "c_cancer"), collapse = "|"), TRUE,
+    "simpl",               paste(setdiff(candidates, "P_ASA"), collapse = "|"),  TRUE,
+    "ASA",                 "P_ASA",                                              FALSE,
+    "CCI",                 "CCI_index_quan_original",                            FALSE,
+    "ECI",                 "ECI_index_sum_all",                                  FALSE,
+    "RxRiskV",             "Rx_index_index",                                     FALSE,
+    "age + sex (main)",    "P_Sex|P_Age",                                        FALSE,
+    "age + sex (rcs 2)",   "P_Sex|P_Age_ns_[12]",                                FALSE,
+    "age + sex (rcs 3)",   "P_Sex|P_Age_ns_[123]",                               FALSE
+  ) %>%
+  mutate(
+
+    # Functions and recipes for model fitting (model averaging or GLM)
+    fun          = map(modavg, ~ {if (.) mod_avg else glml}),
+    rec          = map(name, ~ reci(df, rcs = case_when(
+      grepl("2", .) ~ 2,
+      grepl("3", .) ~ 3,
+      TRUE ~ NA_real_)
+    )
+    ),
+    # Model
+    model        = future_map2(fun, preds, ~ .x(.y)),
+    modsum       = map2(modavg, model,
+                        ~ {if (.x) summary(.y) else summary(.y$fit)}),
+    # Evaluate
+    evaluate     = pmap(tibble(model, preds, rec), ~ evaluate(..1, ..2, ..3)),
+    # AUC values from evaluation sets
+    AUCtrain_lo  = map_dbl(evaluate, ~ quantile(.$auc, .025)),
+    AUCtrain_est = map_dbl(evaluate, ~ quantile(.$auc, .5)),
[…]

### 2019-06-19 src/coefs_print.R

@@ -1,0 +1,27 @@
+library(ProjectTemplate)
+load.project()
+
+# Nice text string of coefficients in full model
+cfs       <- names(bestmodelcoefs)[-1]
+
+cfs_dis   <-
+  cfs[grepl("c_|Rx", cfs)] %>%
+  {gsub("c_|Rx_|P_|TRUE|_Female|_X[123]", "", .)} %>%
+  unique() %>%
+  {gsub("_", " ", .)}
+
+cfs_other <-
+  cfs[!grepl("c_|Rx", cfs)] %>%
+  {gsub("c_|Rx_|P_|TRUE|_Female|_X[123]", "", .)} %>%
+  unique()
+
+
+coefs_print <-
+  paste0(
+    paste(cfs_other, collapse = ", "),
+    ", or the precense of ",
+    glue::glue_collapse(cfs_dis, sep = ", ", last = " or "),
+    ", during the year before surgery"
+  )
+
+cache("coefs_print")

### 2019-06-19 src/table2.R

@@ -1,23 +1,53 @@
 library(ProjectTemplate)
 load.project()
 
-
 # Table 2 unadjusted ------------------------------------------------------
 
 lrm <- function(rhs) {
-  glm(as.formula(paste("death90 ~", rhs)), binomial, df) %>%
-    glmCI(TRUE)
+  glm(
+    as.formula(paste("death90 ~", rhs)),
+    binomial, df
+  ) %>%
+  glmCI(TRUE)
 }
 
+# Obs här använder vi objekt som skapas i modelling.R! Inte helt OK egentligen!
+full_model_coefs <- gsub("TRUE|_Female", "", names(bestmodelcoefs)[-1])
+
+vars <-
+  unique(c(
+    predictors,
+    names(df)[grepl("index", names(df))],
+    gsub("_X[0-9]", "", full_model_coefs)
+  ))
+
+# Proportions selected
+props <-
+  prop_selected %>%
+  mutate(
+    clean_name = gsub("[^[:alnum:]]|X", "", name),
+  ) %>%
+  select(-name)
+
 t2_univariable <-
-  tibble(name = predictors) %>%
-  mutate(coefs = map(name, lrm)) %>%
+  tibble(name = vars) %>%
+  mutate(
+    model = ifelse(name %in% gsub("_X[23]", "", full_model_coefs), "FM", ""),
+    coefs = map(name, lrm)
+  ) %>%
   unnest(coefs) %>%
+  mutate(clean_name = gsub("[^[:alnum:]]|TRUE", "", var)) %>%
+  left_join(props, "clean_name") %>%
   mutate(
     var  = map2_chr(name, var, ~gsub(.x, "", .y, fixed = TRUE)),
+    var = gsub("TRUE", "", var),
     name = ifelse(duplicated(name), "", name),
+    name = gsub("P_|c_|index_sum_all|index_quan_original|index_index|TypeOf|Surg", "", name),
+    name = gsub("Rx$", "RxRiskV", name),
+    name = gsub("_", " ", name),
     CI   = paste0("(", round(low, 2), "-", round(high, 2), ")")
   ) %>%
-  select(-low, -high)
+  select(name, var, OR, CI, `selected (%)` = value, model)
+
 
 cache("t2_univariable")

### 2019-08-22 diagnostics/prelresults_brlasso.Rmd
Use BRLasso with some preliminary result to present to SN and OR.

---
title: "Preliminära resultat av BRLasso"
author: "Erik Bulow"
date: '2019-08-22'
output: 
  pdf_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
load("../cache/table1.RData")
table1
```

```{r}
knitr::include_graphics("../graphs/brlasso_roc.png")
```

```{r}
knitr::include_graphics("../graphs/brlasso_auc_ci.png")
```

```{r}
knitr::include_graphics("../graphs/brlasso_separation_auc.png")
```

```{r}
load("../cache/brlasso_tbl_auc.RData")
brlasso_tbl_auc
```

```{r}
load("../cache/brlasso_tbl_coefs.RData")
brlasso_tbl_coefs
```

### 2019-08-22 lib/BRLasso.R

@@ -1,66 +1,0 @@
-# Bootstrap ranking Lasso copied (and modified) from
-
-BRLasso <- function(x, y, B = 5, Boots = 100, kfold = 10) {
-
-  varx <- varx0 <- colnames(x)
-  nvar          <- length(varx)
-  n             <- length(y)
-  var_list_all  <- vector("list", B)
-
-  RecordM2 = matrix(0, nrow = B, ncol = nvar)
-  rownames(RecordM2) <- paste0("B", seq_len(B))
-  colnames(RecordM2) <- varx
-
-  for (ij in seq_len(B)) {
-    Nboot <- n
-    RecordM = matrix(0, nrow = Boots, ncol = nvar)
-    rownames(RecordM) = paste0("BSample", seq_len(Boots))
-    colnames(RecordM) = varx
-    for (i in seq_len(Boots)) {
-      repeat {
-        s <- sample(Nboot, replace = TRUE)
-        if (length(table(y[s])) >= 2 & length(table(y[-s])) >= 2)
-          break
-      }
-
-      cvfit <- glmnet::cv.glmnet(x[s,], y[s], nfolds = kfold, family = "binomial")
-      model.final <- cvfit$glmnet.fit
-      nzero       <- as.matrix(coef(model.final, s = cvfit$lambda.min))
-
-      for (i0 in seq_len(dim(nzero)[1])) {
-        for (j0 in seq_len(nvar)) {
-          if (dimnames(nzero)[[1]][i0] == colnames(RecordM)[j0]) {
-            RecordM[i, j0] = nzero[i0,]
-          }
-        }
-      }
-      message("Step 1-Current iteration: ", i)
-    }
-
-    Impor     <- abs(apply(RecordM, 2, mean))
-    score     <- sort(Impor)
-    index     <- seq_along(score)
-    model.pwl <- SiZer::piecewise.linear(x = index, y = score)
-    break_p   <- round(model.pwl$change.point)
-    var_list_all[[ij]] <- names(score[break_p:length(index)])
-    if (length(var_list_all[[ij]]) != 0) {
-      varx0 <- intersect(var_list_all[[ij]], varx0)
-    }
-    Impor0    <- apply(RecordM, 2, mean)[var_list_all[[ij]]]
-    for (gp in seq_along(Impor0)) {
-      RecordM2[ij, colnames(RecordM2) == names(Impor0)[gp]] <- Impor0[gp]
-    }
[…]

### 2019-08-22 munge/01-filter.R

@@ -28,24 +28,26 @@ df_shpr <-
   ungroup()
 
 # Endast första höft
-df_shpr <- filter(df_shpr, opnr == 1); N()                                       # 80805
+df_shpr <- filter(df_shpr, opnr == 1); N()                                       # 77075
 
 # Ignorera om ytterligare THA inom 90 dagar efter första
-df_shpr <- filter(df_shpr, is.na(time_between) | time_between > 90); N()         # 79888
+df_shpr <- filter(df_shpr, is.na(time_between) | time_between > 90); N()         # 76158
 
 # Endast cementfria
 df_shpr <- filter(df_shpr,
                   P_AcetCupCemMix != "Cementfritt",
-                  P_FemStemCemMix != "Cementfritt"); N()                          # 53718
+                  P_FemStemCemMix != "Cementfritt"); N()                          # 50569
 
-df_shpr <- filter(df_shpr, is.na(P_KVA1) | P_KVA1 != "NFB62 - Primär total ytersättningspr"); N() # 53718
-df_shpr <- filter(df_shpr, between(P_Age, 18, 100)); N()                          # 53718
+df_shpr <- filter(df_shpr, is.na(P_KVA1) | P_KVA1 != "NFB62 - Primär total ytersättningspr"); N() # 50569
+df_shpr <- filter(df_shpr, between(P_Age, 18, 100)); N()                          # 50569
 
-df_shpr <- filter(df_shpr, P_BMI <= 50); N()                                      # 50761
-df_shpr <- filter(df_shpr, P_ASA <= 3); N()                                       # 49956
-df_shpr <- filter(df_shpr, !is.na(education)); N()                                # 49611
-df_shpr <- filter(df_shpr, !is.na(civil_status)); N()                             # 49611
-df_shpr <- filter(df_shpr, !is.na(P_TypeOfHospital)); N()                         # 49235
+df_shpr <- filter(df_shpr, P_BMI <= 50); N()                                      # 47709
+df_shpr <- filter(df_shpr, P_ASA <= 3); N()                                       # 46960
+
+# Dessa skulle vi ju ev kunna imputera ist!
+df_shpr <- filter(df_shpr, !is.na(education)); N()                                # 46621
+df_shpr <- filter(df_shpr, !is.na(civil_status)); N()                             # 46621
+df_shpr <- filter(df_shpr, !is.na(P_TypeOfHospital)); N()                         # 46272
 
[…]

### 2019-08-22 munge/03-compositvariabler.R

@@ -7,78 +7,53 @@ df <-
     c_heart_condition =
       ECI_congestive_heart_failure              |
       CCI_congestive_heart_failure              |
-      Rx_chronic_heart_failure                  |
       ECI_valvular_disease,
     c_heart_infarct =
-      CCI_myocardial_infarction                 |
-      Rx_angina                                 |
-      Rx_ischaemic_heart_disease_hypertension,
+      CCI_myocardial_infarction,
     c_arrythmia =
-      ECI_cardiac_arrhythmias                   |
-      Rx_arrhythmia,
+      ECI_cardiac_arrhythmias,
     c_hypertoni =
       ECI_hypertension_uncomplicated            |
-      ECI_hypertension_complicated              |
-      Rx_hypertension,
+      ECI_hypertension_complicated,
     c_kidney_disease =
       ECI_renal_failure                         |
-      CCI_renal_disease                         |
-      Rx_end_stage_renal_disease,
+      CCI_renal_disease,
     c_diabetes =
       ECI_diabetes_uncomplicated                |
       ECI_diabetes_complicated                  |
       CCI_diabetes_without_complication         |
-      CCI_diabetes_complication                 |
-      Rx_diabetes,
+      CCI_diabetes_complication,
     c_vascular_disease =
       ECI_peripheral_vascular_disorder          |
       CCI_peripheral_vascular_disease           |
-      CCI_cerebrovascular_disease               |
-      Rx_anti_coagulation_therapy               |
-      Rx_anti_platelet_therapy,
+      CCI_cerebrovascular_disease,
     c_cancer =
       ECI_lymphoma                              |
       ECI_metastatic_cancer                     |
       ECI_solid_tumor                           |
       CCI_malingnancy                           |
-      CCI_metastatic_solid_tumor                |
-      Rx_malignancies,
+      CCI_metastatic_solid_tumor,
     c_aids_hiv =
       ECI_aids_hiv                              |
-      CCI_aids_hiv                              |
-      Rx_hiv,
+      CCI_aids_hiv,
     c_liver_disease =
       ECI_liver_disease                         |
       CCI_mild_liver_disease                    |
-      CCI_moderate_or_severe_liver_disease      |
-      Rx_liver_failure                          |
-      Rx_hepatitis_c,
+      CCI_moderate_or_severe_liver_disease,
[…]

### 2019-08-22 munge/04-candidates.R

@@ -4,7 +4,7 @@
 # Screen all comorbidity measures as well as other factors for
 # predictors to include in combined model
 important_factors <-
-  tibble(preds = c("ECI", "CCI", "Rx", "c_")) %>%
+  tibble(preds = c("ECI", "CCI", "c_")) %>%
   mutate(data = map(preds, getdata)) %>%
   add_row(
     preds = "general",

### 2019-08-22 munge/05-best_model.R

@@ -5,19 +5,18 @@ nms <- paste(setdiff(candidates, "P_Age"), collapse = "|")
 # Object with models to compare and evaluate
 models <-
   tribble(
-    ~name,                ~preds,                                                ~ modavg,
-    "full",                paste0(nms, "|P_Age"),                                TRUE,
-    "full (rcs 2)",        paste0(nms, "|P_Age_ns_[12]"),                        TRUE,
-    "full (rcs 3)",        paste0(nms, "|P_Age_ns_[123]"),                       TRUE,
-    "full-cancer",       paste(setdiff(candidates, "c_cancer"), collapse = "|"), TRUE,
-    "simpl",               paste(setdiff(candidates, "P_ASA"), collapse = "|"),  TRUE,
-    "ASA",                 "P_ASA",                                              FALSE,
-    "CCI",                 "CCI_index_quan_original",                            FALSE,
-    "ECI",                 "ECI_index_sum_all",                                  FALSE,
-    "RxRiskV",             "Rx_index_index",                                     FALSE,
-    "age + sex (main)",    "P_Sex|P_Age",                                        FALSE,
-    "age + sex (rcs 2)",   "P_Sex|P_Age_ns_[12]",                                FALSE,
-    "age + sex (rcs 3)",   "P_Sex|P_Age_ns_[123]",                               FALSE
+    ~name,               ~preds,                                                ~ modavg,
+    "full",              paste0(nms, "|P_Age"),                                TRUE,
+    "full (rcs 2)",      paste0(nms, "|P_Age_ns_[12]"),                        TRUE,
+    "full (rcs 3)",      paste0(nms, "|P_Age_ns_[123]"),                       TRUE,
+    # "full-cancer",       paste(setdiff(candidates, "c_cancer"), collapse = "|"), TRUE,
+    # "simpl",             paste(setdiff(candidates, "P_ASA"),  collapse = "|"), TRUE,
+    "ASA",               "P_ASA",                                              FALSE,
+    "CCI",               "CCI_index_quan_original",                            FALSE,
+    "ECI",               "ECI_index_sum_all",                                  FALSE,
+    "age + sex (main)",  "P_Sex|P_Age",                                        FALSE,
+    "age + sex (rcs 2)", "P_Sex|P_Age_ns_[12]",                                FALSE,
+    "age + sex (rcs 3)", "P_Sex|P_Age_ns_[123]",                               FALSE
   ) %>%
   mutate(
 

### 2019-09-24 reports/article.Rmd
changed everything except report after implementing BRLasso

--- 2019-06-19 reports/article/article.Rmd
@@
 ---
-title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and extrenaly validated model based on Swedish and English register data"
+title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on Swedish and English register data"
 author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Adrian Sayers^5,6^, Szilard Nemes^2,4^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^

### 2019-09-24 docs/Szilard v1/Code/02. Fishing_atcr .R

@@ -1,184 +1,0 @@
-﻿library(dplyr)
-library(pROC)
-anne_pred <- readRDS("Z:\\SHPR\\SZILARD\\Beslutstöd\\Output\\data_with_adverse_events_2016-08-22.rds")
-
-
-
-#
-##  data selection and flow chart
-#       
-
-# 001. OppAr < 2007
-anne_pred <- dplyr::filter(anne_pred, OppAr > 2007) 
-dim(anne_pred)
-# 002. Primary arthros
-anne_pred <- dplyr::filter(anne_pred, DiaGrp == 1) 
-dim(anne_pred)
-# 003. OpNr ==1 
-anne_pred <- dplyr::filter(anne_pred, OpNr == 1) 
-dim(anne_pred)
-# 004. Age between 18 and 100
-anne_pred <- dplyr::filter(anne_pred, (17 < Alder & Alder < 101))
-dim(anne_pred)
-# 005. No resurfacing 
-anne_pred <- dplyr::filter(anne_pred, ProtGrp != 5)
-dim(anne_pred)
-
-#
-## Calculate BMI
-#
-anne_pred <- dplyr::mutate(anne_pred, bmi =  Vikt/(Langd/100)^2)
-sum(is.na(anne_pred$bmi))
-
-# 006. BMI < 51 and existing
-
-anne_pred <- dplyr::filter(anne_pred, !is.na(bmi)) %>%
-             dplyr::filter( bmi  <= 50)
-dim(anne_pred)
-
-# 007. ASA not 4 & 5
-
-anne_pred <- dplyr::filter(anne_pred, ASA < 4) %>%
-             droplevels()
-dim(anne_pred)
-
-
-
-
-
-### -------------- Format the socioeconomic variables -----------------------###
-
-# 001. Education level 
-anne_pred <- dplyr::mutate(anne_pred, Sun2000niva_old = factor(gsub(" ", "", Sun2000niva_old))) %>%
-                  dplyr::mutate(Sun2000niva_old = factor(Sun2000niva_old, 
-                                         levels = c('*', 1, 2, 3, 4, 5, 6, 7), 
-                                         labels = c('NA', 
-                                                    'pre_high_less_9yrs', 'pre_high_9yrs',
-                                                    'high_schol_2yrs', 'high_schol_3yrs', 
-                                                    'univ_less_3yrs', 'univ_3yrs',
-                                                    'graduate'),
-                                         ordered = TRUE)) 
[…]

### 2019-09-24 docs/Szilard v1/Code/04. Summary plots.R

@@ -1,37 +1,0 @@
-﻿
-fit_pred_red <- glm(death90day ~ Kon + Alder + ASA  +
-                              heart_infarct    +
-                              kidney           ,
-                              data = anne_pred2,
-                              family = 'binomial')  
-                              
-                              
-probs <- data.frame(Status = fit_pred_red$model$death90day, Probability = fit_pred_red$fitted)   
-
-
-probs <- dplyr::mutate(probs, Status  = factor(Status, 
-                                    levels = c(0, 1),
-                                    labels = c('Alive', 'Death')))
-
-
-p <- ggplot(probs, aes(Status, Probability)) + 
-          geom_boxplot()                     +
-          geom_boxplot(aes(fill = Status))   +
-          theme_bw()     +
-        geom_blank() +
-    theme(panel.grid.major.x = element_blank(),
-          panel.grid.minor.x = element_blank())                       
-p  
- ggsave(file="Probability.png", dpi = 600) 
-fitAge <- glm(death90day ~  Alder   +
-                            heart_infarct    +
-                            kidney           ,
-                            data = anne_pred2,
-                          family = 'binomial') 
-
-###
-newdat00 <- data.frame(Alder = 65:95, heart_infarct = rep(0, 31), kindey = rep(0, 31))    
-newdat10 <- data.frame(Alder = 65:95, heart_infarct = rep(1, 31), kindey = rep(0, 31))    
-newdat01 <- data.frame(Alder = 65:95, heart_infarct = rep(0, 31), kindey = rep(1, 31))    
-newdat11 <- data.frame(Alder = 65:95, heart_infarct = rep(1, 31), kindey = rep(1, 31))    
-pred00 <-  predict(fitAge, newdat00)                       

### 2019-09-24 docs/Szilard v1/Code/RxRiskV_modified_regexp.R

@@ -1,46 +1,0 @@
-﻿
-
-####  RxRiskV -modified
-
-alcohol_dependence	<- 'N07BB0[1345]'
-allergies	<- 'R0[16]A[CDEX][0125][12356789]'             
-anti_coagulation_therapy	<- 'B01A[AB][01][01234569]|'                              
-anti_platelet_therapy	<- 'B01AC|B01AC[01234679]'                     
-anxiety	<- 'N05BA[01][1246]'
-arrhythmia	<-'C01[AB][ABCD]0[13457]'               
-benign_prostate_hypertrophy	<- 'G04CA0[123]'
-bipolar_disorder	<-'N0[35]A[EGHNX][01][12349]' 
-chronic_heart_failure	 <- 'C0[39][ACD][AC][01][012345679]'
-dementia <- 'N04D[AX]0[1234]'
-depression	<- 'N06A[ABGX][012][012345689]'
-diabetes	<- 'A10[AB][ABCDEFGHX][01][0123456789]'
-end_stage_renal_disease	<- 'B03XA0[123]|A11CC0[1234]'
-epilepsy	<- 'N03A[ABDEGFX][01][0123459]'
-gastric_oesophageal_reflux_disorder	<- 'A02B[ABCDX][015][123456]'
-glaucoma	<- 'S01E[ABCDEX][05][12345]|S01EX'
-gout	<- 'M04A[ABC]01'
-hepatitis_c	<- 'J05A[ABEX][0156][12457]'
-hiv <- 'J05A[EFGRX][016][0123456789]|J05AR'
-hyperkalaemia	<- 'V03AE01'   
-hyperlipidemia <- 'C10[AB][ABCX][01][1234579]'
-hypertension	<- 'C0[239][ABDEKX][ABCX][01][123456789]'
-hyperthyroidism	<- 'H03AA0[12]'
-angina	<- 'C01DA[01][248]'
-ischaemic_heart_disease_hypertension	<- 'C0[78][ACDF][ABG][01][1234567]'
-inflammatory_bowel_disease	<- 'A07E[AC]0[12346]'
-liver_failure <- 	'A06AD11'
-malignancies	<- 'L01[ABCDX][ABCDEX][012345[0123456789]'
-malnutrition	<- 'B05BA[01][03]'
-migraine	<- 'N02C[ACX]0[123456]'
-osteoporosis_pagets	<- 'M05B[ABX]0[1234678]'
[…]

### 2019-09-24 docs/Szilard v1/Code/time between operations.R

@@ -1,181 +1,0 @@
-﻿library(dplyr)
-library(pROC)
-library(tableone)
-anne_pred <- readRDS("Z:\\SHPR\\SZILARD\\Beslutstöd\\Output\\data_with_adverse_events_2016-08-22.rds")
-
-
-
-#
-##  data selection and flow chart
-#       
-
-# 001. OppAr < 2007
-anne_pred <- dplyr::filter(anne_pred, OppAr > 2007) 
-dim(anne_pred)
-# 002. Primary arthros
-anne_pred <- dplyr::filter(anne_pred, DiaGrp == 1) 
-dim(anne_pred)
-# 003. OpNr ==1 
-#anne_pred <- dplyr::filter(anne_pred, OpNr == 1) 
-dim(anne_pred)
-# 004. Age between 18 and 100
-anne_pred <- dplyr::filter(anne_pred, (17 < Alder & Alder < 101))
-dim(anne_pred)
-# 005. No resurfacing 
-anne_pred <- dplyr::filter(anne_pred, ProtGrp != 5)
-dim(anne_pred)
-
-#
-## Calculate BMI
-#
-anne_pred <- dplyr::mutate(anne_pred, bmi =  Vikt/(Langd/100)^2)
-sum(is.na(anne_pred$bmi))
-
-# 006. BMI < 51 and existing
-
-anne_pred <- dplyr::filter(anne_pred, !is.na(bmi)) %>%
-             dplyr::filter( bmi  <= 50)
-dim(anne_pred)
-
-# 007. ASA not 4 & 5
-
-anne_pred <- dplyr::filter(anne_pred, ASA < 4) %>%
-             droplevels()
-dim(anne_pred)
-
-
-
-
-
-### -------------- Format the socioeconomic variables -----------------------###
-
-# 001. Education level 
-anne_pred <- dplyr::mutate(anne_pred, Sun2000niva_old = factor(gsub(" ", "", Sun2000niva_old))) %>%
-                  dplyr::mutate(Sun2000niva_old = factor(Sun2000niva_old, 
-                                         levels = c('*', 1, 2, 3, 4, 5, 6, 7), 
-                                         labels = c('NA', 
-                                                    'pre_high_less_9yrs', 'pre_high_9yrs',
-                                                    'high_schol_2yrs', 'high_schol_3yrs', 
-                                                    'univ_less_3yrs', 'univ_3yrs',
-                                                    'graduate'),
-                                         ordered = TRUE)) 
[…]

### 2019-09-24 docs/Szilard v1/data selection 20160717.R

@@ -1,172 +1,0 @@
-﻿#### 2016-03-18
-
-# Perdiction study Anne G 
-# Data Selection 
-
-
-
-rm(list = ls()); gc()
-
-# Library
-library(dplyr)
-library(tableone)
-#
-
-
-anne_pred <- readRDS("Z:\\SHPR\\SZILARD\\Beslutstöd\\Output\\data_with_adverse_events_2016-08-22.rds")
-
-
-anne_pred  <- dplyr::select(anne_pred, Sida, OpNr, lpnr,
-                                      OppDat, OppAr, DatOpr, OprAr, Doddat, 
-                                      KlinGrp,                                          
-                                      Alder, Kon, 
-                                      AtgGrp, AtgTyp, 
-                                      DiaGrp, 
-                                      ProtGrp, 
-                                      charkat0, ASA, 
-                                      Atgard, OrsGrp,
-                                      Langd, Vikt, 
-                                      elix_index_1years, charlson_index_original_1years, charlson_index_updated_1years, 
-                                      atc_index_modified_1year, atc_index_1year, 
-                                      Sun2000niva_old, Civil)
-
-
-
-#
-##  data selection and flow chart
-#       
-
-# 001. OppAr < 2007
-anne_pred <- dplyr::filter(anne_pred, OppAr > 2007) 
-dim(anne_pred)
-# 002. Primary arthros
-anne_pred <- dplyr::filter(anne_pred, DiaGrp == 1) 
-dim(anne_pred)
-# 003. OpNr ==1 
-anne_pred <- dplyr::filter(anne_pred, OpNr == 1) 
-dim(anne_pred)
-# 004. Age between 18 and 100
-anne_pred <- dplyr::filter(anne_pred, (17 < Alder & Alder < 101))
-dim(anne_pred)
-# 005. No resurfacing 
-anne_pred <- dplyr::filter(anne_pred, ProtGrp != 5)
-dim(anne_pred)
-
-#
-## Calculate BMI
-#
-anne_pred <- dplyr::mutate(anne_pred, bmi =  Vikt/(Langd/100)^2)
-sum(is.na(anne_pred$bmi))
-
-# 006. BMI < 51 and existing
-
-anne_pred <- dplyr::filter(anne_pred, !is.na(bmi)) %>%
-             dplyr::filter( bmi  <= 50)
-dim(anne_pred)
-
-# 007. ASA not 4 & 5
-
[…]

### 2019-09-24 docs/Szilard v1/survival time.r

@@ -1,35 +1,0 @@
-rm(list = ls())
-library(dplyr)
-
-## change the path to your own
-setwd('Z:/SHPR/SZILARD/People/Anne Garland')
-
-anna <- read.table('anna.txt', header = TRUE)
-
-head(anna)
-
-###
-#   Transform the date variables so that R recognizes as date
-#   I assume that the format is yyyy/mm/dd
-anna$OppDat <- as.Date(anna$OppDat, format = "%Y/%m/%d")
-anna$Doddat <- as.Date(anna$Doddat, format = "%Y/%m/%d")
-anna$oprdat <- as.Date(anna$oprdat, format = "%Y/%m/%d")
-
-### sort the data according to 'lopnr' and 'OppDat' operations datum
-anna <- arrange(anna, lopnr, OppDat)
-
-#### add an index telling if its the first or second hip operation
-anna$OppNr <- rep(1:2, dim(anna)[1]/2)
-
-#### calculate the date between opperation and death
-anna$Survtime <- with(anna, as.numeric(Doddat-OppDat))
-
-#### an index telling if the patinet had revisonof the first hip before the
-#### second hip opperation or not
-#### it can take values: NA - there were no revision
-####                     0 - revison after the second hip operation or on the second hip
-####                     1 - fisrt hip revison before the second hip opperation
-anna$Revision <- with(anna, as.numeric(oprdat < OppDat))
-
-anna
-

### 2019-09-24 lib/best_model.R

@@ -1,80 +1,0 @@
-
-# Find best mdoel ---------------------------------------------------------
-
-# Model averaging with model weights based on AIC from bootstrap replicates
-# Note that each BS replicate is taken after the recipi is applied.
-# The downsmapling means that this might be of limited value, wherefore an
-# additional BS is applied below. We use sqrt(B) outer bootstrappinug replicates
-# and sqrt(B) inner replicates
-bsavg <- function(mod_data) {
-  model <- glm(death90f ~ .,
-               binomial, mod_data, x = TRUE, na.action = "na.fail")
-  dr          <- MuMIn::dredge(model)
-  am          <- model.avg(dr, data = mod_data)
-  Weights(am) <- bootWeights(am, R = floor(sqrt(config$N_bots)))
-  am
-}
-
-
-# Fit model based on variable names and average results based on bootstrap
-mod_avg <- function(nms) {
-
-  mod_avg <-
-    df %>%
-    # sqrt(B) outer bootstrap replicates
-    rsample::bootstraps(
-      ceiling(sqrt(config$N_bots)), strata = "death90f", apparent = TRUE) %>%
-    mutate(
-      recipes  = map(splits, prepper, recipe = reci(df), retain = TRUE),
-      # Only keep variables that are matched by the candidate vector names
-      mod_data = map(recipes, juice, death90f, matches(nms)),
-      am       = map(mod_data, bsavg),
-      w        = map(am, ~ enframe(c(Weights(.))))
-    )
-
-  # Average weights over the outer bootstrap replicates
-  w <-
-    mod_avg %>%
-    unnest(w) %>%
-    group_by(name) %>%
-    summarise(w = mean(value))
-
-  # Update weights for the whole training set wit hthe boostatrap averaged weights
-  fit <- mod_avg$am[mod_avg$id == "Apparent"][[1]]
-  Weights(fit) <- w$w
-  fit
-}
-
-
-# Simple glm with result ni list, help function to below
-glml <- function(nms) {
-  lr(juice(prep(reci(df)), death90f, matches(nms)))
-}
-
-
-pred <- function(..., model) {
-  pred_type <- switch(class(model)[1], "_glm" = "prob", averaging = "response")
-  p <- predict(model, ..., type = pred_type)
-  if (inherits(model, "_glm")) p <- 1 - p$.pred_alive
-  p
-}
-
-
-# Evaluate the final model based on ROC-curves and AUC
-evaluate <- function(model, nms, rec) {
[…]

### 2019-09-24 lib/helpers.R

@@ -1,58 +1,0 @@
-con <- shar_linkage_con()
-
-predictors <- c(
-  "P_Age", "P_Sex", "P_BMI", "P_ASA", "P_TypeOfHospital", "P_SurgYear",
-  "education", "civil_status"
-)
-
-# Cahe list of pobjects at once
-cache_all <- Vectorize(cache, "variable")
-
-### Summary function for GLM
-glmCI <- function(model, exponent = FALSE, alpha = 0.05) {
-  co <- coef(model)
-  l  <- qnorm(1 - alpha / 2) * sqrt(diag(summary(model)$cov.scaled))
-
-  # calculuate CI for linear case
-  mat <- tibble::tibble(var = names(co), beta = co, low = co - l, high = co + l)
-
-  # if exponent=TRUE, exponeniate the coefficients and CIs
-  if (exponent) {
-    mat <- mutate_if(mat, is.numeric, exp)
-    names(mat)[names(mat) == "beta"] <-
-      switch(summary(model)$family$link, logit = "OR", log  = "RR")
-  }
-  filter(mat, var != "(Intercept)")
-}
-
-
-# Summarise survfit results
-survEst <- function(fit, times) {
-  x <- as.data.frame(summary(fit, times = times))
-  ind <- c("time", "surv",  "lower", "upper")
-  if (!is.null(fit$strata))
-    ind <- c(ind, "strata")
-  x[ind]
-}
-
-
-# Change table text to labels that can be presented
-prestext <- function(df) {
-  gs <- function(x, pattern, replacement) gsub(pattern, replacement, x)
-  df %>%
-  rename(
-    variable = 1,
-    level = 2
-  ) %>%
-  mutate(
-    level    = tolower(level),
-    variable =
-      variable %>%
-      gs("Rx_index_index", "RxRiskV") %>%
-      gs("P_|c_|Surg|TypeOf|_index_sum_all|_index_quan_original", "") %>%
-      gs("_", " "),
-
-    # Initial capital letter
-    variable = paste0(toupper(substr(variable, 1, 1)), substring(variable, 2))
-  )
-}

### 2019-09-24 lib/modelling.R

@@ -1,78 +1,0 @@
-
-# Find candidate predictors ----------------------------------------------
-
-
-# Logistic regressoin using parsnip
-lr <- function(x) {
-  fit <-
-    logistic_reg() %>%
-    set_engine("glm") %>%
-    fit(death90f ~ ., data = x)
-}
-
-# Identify variables used in the best model according to stepwise regression
-bestfit <- function(x) {
-  vars <- setdiff(names(x), "death90f")
-  best <- stats::step(glm(death90f ~ ., binomial, x), trace = 0)
-
-  tibble(
-    var  = vars,
-    # Fins name matches fom original names (no dummies)
-    incl = map_lgl(vars, ~ any(startsWith(names(coef(best)), .)))
-  )
-}
-
-reci <- function(df, outcome = "death90f", rcs = NULL) {
-  rec <-
-    recipe(as.formula(paste(outcome, "~ .")), df) %>%
-    step_downsample(all_outcomes())
-
-  # Create dummy variables for categorical data
-  if (sum(vapply(df, is.factor, logical(1))) > 1)
-    rec <- rec %>% step_dummy(all_predictors(), all_numeric())
-
-  # RCS for age
-  if (!is.null(rcs) && !is.na(rcs))
-    rec <- rec %>% step_ns(matches("P_Age"), deg_free = rcs)
-  rec
-}
-
-
-# Use bootstrap resampling to find number of times each variable is selected in
-# stepwise regression
-find_predictors <- function(
-  df, rec = reci(df), B = config$N_bots, outcome = "death90f") {
-
-  df %>%
-    rsample::bootstraps(B, strata = outcome) %>%
-    mutate(
-      recipes  = map(splits, prepper, recipe = rec, retain = TRUE),
-      mod_data = map(recipes, juice),
-      included = future_map(mod_data, bestfit)
-    ) %>%
-    unnest(included) %>%
-    spread(var, incl)
-}
-
-# List names of predictors selected at least threshold % of the time
-list_predictors <- function(x, df, outcome = "death90f", thr = .9) {
-
-  winners_all <-
-    x %>%
-    summarise_at(vars(-id), mean) %>%
-    mutate_all(~ . >= thr) %>%
-    c(recursive = TRUE) %>%
-    {names(.)[.]}
-
-  preds    <- setdiff(names(df), outcome)
-  winners  <- preds[map_lgl(preds, ~ any(startsWith(winners_all, .)))]
[…]

### 2019-09-24 munge/04-remove_empty_variables.R

@@ -1,0 +1,88 @@
+# We will not consider any variable without observations in any combination
+# of dead/alive-effected/not affected.
+# This does not exclude any cases. It only exclude potential predictor variables
+
+df0 <-
+  df %>%
+  select(
+    -op_first, -op_last, -stime, -status,
+    -contains("index"),
+    -starts_with("death"), death90f
+  )
+
+y <- df0$death90f
+
+
+# Find logical variables with empty cells for any combination ------------------
+
+comb_lgl <-
+  df0 %>%
+  select_if(is.logical) %>%
+  pivot_longer(everything()) %>%
+  nest(data = c(value)) %>%
+  mutate(
+    data = map(data, add_column, y),
+    data = map(data, count, y, value),
+    noncomplete = map_lgl(data, ~nrow(.) < 4)
+  ) %>%
+  filter(noncomplete) %>%
+  select(-noncomplete) %>%
+  unnest(data) %>%
+  mutate(
+    value = if_else(value, "condition", "condition free")
+  ) %>%
+  unite("value_y", value, y) %>%
+  pivot_wider(names_from = value_y, values_from = n)
+
+cache("comb_lgl")
+
+
+# Same for factor variables -----------------------------------------------
+
+comb_fct_tmp <-
+  df0 %>%
+  select_if(is.factor) %>%
+  select(-starts_with("death"))
+
+n_lvls <-
+  comb_fct_tmp %>%
+  summarise_all(n_distinct) %>%
+  pivot_longer(everything(), values_to = "n_lvls")
+
+# Factor variables not affected
+comb_fct <-
+  comb_fct_tmp %>%
+  pivot_longer(everything()) %>%
+  nest(data = value) %>%
+  left_join(n_lvls, "name") %>%
+  mutate(
+    data        = map(data, add_column, y),
+    data        = map(data, count, y, value) #,
+  ) %>%
+  unnest(data) %>%
+  pivot_wider(names_from = y, values_from = n) %>%
+  mutate(
+    name = if_else(duplicated(name), "", name)
+  ) %>%
+  select(-n_lvls)
+
+
+# df without those variables ----------------------------------------------
+
+df0 <-
+  df0 %>%
+  select(-one_of(comb_lgl$name), death90f) %>%
+  mutate_if(is.logical, as.factor) # needed for recipe
+
+cache("df0")
+
+# Fix model matrix --------------------------------------------------------
+
+df_model <-
[…]

### 2019-09-24 munge/05-BRLasso.R

@@ -1,0 +1,8 @@
+# Perform calculations (time consuming)
+best_coefs_tmp <-
+  tibble(B = seq_len(config$Bmax)) %>%
+  mutate(
+    coefs = future_map(B, ~ selected_vars(df_model, config$N_bots), .progress = TRUE)
+  )
+
+cache("best_coefs_tmp")

### 2019-09-24 src/.extremes.R

@@ -1,40 +1,0 @@
-library(ProjectTemplate)
-load.project()
-
-
-# Vi kikar lite på de mest extrema patietnerna vad gäller ECI.
-
-# Extremes in Elixhauser included in training data but not in test data
-ECI_extremes <-
-  bind_rows(
-    train = df_train,
-    eval = df_test,
-    .id = "dataset"
-  ) %>%
-  count(dataset, ECI_index_sum_all) %>%
-  spread(dataset, n)
-cache("ECI_extremes")
-
-
-# Find the most HEALTHY ECI-person
-healthy <-
-  df_train %>%
-  filter(ECI_index_sum_all == min(ECI_extremes$ECI_index_sum_all))
-# comorbids
-healthy %>%
-  select(starts_with("ECI_"), -ECI_index_sum_all) %>%
-  {names(.)[c(., recursive = TRUE)]}
-# Outcome = alive
-select(healthy, death90f)
-
-
-# Find the most SICK ECI-person
-sick <-
-  df_train %>%
-  filter(ECI_index_sum_all == max(ECI_extremes$ECI_index_sum_all))
-# comorbids
-sick %>%
-  select(starts_with("ECI_"), -ECI_index_sum_all) %>%
-  {names(.)[c(., recursive = TRUE)]}
-# Outcome = alive
-select(sick, death90f)

### 2019-09-24 src/coefs_print.R

@@ -1,27 +1,0 @@
-library(ProjectTemplate)
-load.project()
-
-# Nice text string of coefficients in full model
-cfs       <- names(bestmodelcoefs)[-1]
-
-cfs_dis   <-
-  cfs[grepl("c_|Rx", cfs)] %>%
-  {gsub("c_|Rx_|P_|TRUE|_Female|_X[123]", "", .)} %>%
-  unique() %>%
-  {gsub("_", " ", .)}
-
-cfs_other <-
-  cfs[!grepl("c_|Rx", cfs)] %>%
-  {gsub("c_|Rx_|P_|TRUE|_Female|_X[123]", "", .)} %>%
-  unique()
-
-
-coefs_print <-
-  paste0(
-    paste(cfs_other, collapse = ", "),
-    ", or the precense of ",
-    glue::glue_collapse(cfs_dis, sep = ", ", last = " or "),
-    ", during the year before surgery"
-  )
-
-cache("coefs_print")

### 2019-09-24 src/fig_rocs.R

@@ -1,34 +1,0 @@
-library(ProjectTemplate)
-load.project()
-
-
-# Make data set for plotting
-roc_curves <-
-  rocs %>%
-  select(name, ends_with("roc_curve")) %>%
-  gather("stage", "roc_curve", -name) %>%
-  mutate(
-    roc_curve = map(roc_curve, loesspreds)
-  ) %>%
-  unnest(roc_curve)
-
-cache("roc_curves")
-
-# Plot it!
-ggplot(
-  filter(roc_curves, name != "full"),
-  aes(1 - specificity, sensitivity, col = name)
-) +
-  facet_wrap(~ name) +
-  geom_path(size = 2) +
-  geom_path(
-    aes(1 - specificity, sensitivity),
-    filter(roc_curves, name == "full") %>% select(specificity, sensitivity),
-    size = 2, col = "black"
-  ) +
-  geom_abline(lty = 3) +
-  coord_equal() +
-  theme_minimal() +
-  theme(legend.position = "none")
-
-ggsave("graphs/rocs.png")

### 2019-09-24 src/tab_auc.R

@@ -1,0 +1,13 @@
+library(ProjectTemplate)
+load.project()
+
+
+brlasso_tbl_auc <-
+  all_models %>%
+  arrange(desc(AUC_est)) %>%
+  transmute(
+    Model,
+    AUC   = sprintf("%.2f (%.2f-%.2f)", AUC_est, AUC_lo, AUC_hi)
+  )
+
+cache("brlasso_tbl_auc")

### 2019-09-24 src/table1.R

@@ -15,8 +15,13 @@ dft1 <-
   ) %>%
   select(
     death90f,
-    one_of(predictors),
-    -P_SurgYear,
+    P_Age,
+    P_Sex,
+    P_BMI,
+    P_ASA,
+    P_TypeOfHospital,
+    education,
+    civil_status,
     CCI, ECI,
     starts_with("c_"),
     starts_with("ECI_")
@@ -30,6 +35,8 @@ t1 <-
     data = dft1,
     test = FALSE
   )
+
+cahce("t1")
 
 table1 <-
   t1 %>%
@@ -45,4 +52,4 @@ table1 <-
   ) %>%
   select(what, level, alive, dead)
 
-cache_all(c("t1", "table1"))
+cache("table1")

### 2019-09-24 src/table_full_model_coefs.R

@@ -1,26 +1,0 @@
-library(ProjectTemplate)
-load.project()
-
-table_full_model_coefs <-
-  modsums$modsum[[1]]$coefmat.full %>%
-  as_tibble(rownames = "variables") %>%
-  mutate(
-    coef = Estimate,
-    lo   = Estimate - 1.96 * `Std. Error`,
-    hi   = Estimate + 1.96 * `Std. Error`
-  ) %>%
-  mutate_at(vars(Estimate, lo, hi), exp) %>%
-  transmute(
-    variables = gsub("c_|P_|TRUE|Sex_", "", variables),
-    variables = gsub("ASA_X1", "ASA 1", variables),
-    variables = gsub("ASA_X2", "ASA 2", variables),
-    variables = gsub("ASA_X3", "ASA 3", variables),
-    variables = gsub("_", " ", variables),
-    OR        = Estimate,
-    CI        = sprintf("%.2f-%.2f", lo, hi),
-    p         = ifelse(`Pr(>|z|)` < 0.001, "< 0.001", round(`Pr(>|z|)`, 2))
-  ) %>%
-  filter(variables != "(Intercept)") %>%
-  arrange(variables)
-
-  cache("table_full_model_coefs")

### 2019-09-25 reports/article.Rmd
Automatic flowchart!

@@
 **OBJECTIVE:** Early mortality after total hip arthroplasty (THA) is associated with comorbidity.
-Diagnosis-based tools such as the Charlson or Elixhauser Comorbidity indices, or the prescription-based RxRisk V score, have been used to quantify this.
+Diagnosis-based tools such as the Charlson or Elixhauser Comorbidity indices have been used to quantify this.
 Those instruments are however complex and requires the use of extensive data sets not routinely available in most doctor-patient settings.
@@
 
-**PATIENTS AND METHODS:** We studied `r nrow(df)` patients with THA due to primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` from the Swedish Hip Arthroplasty Register, linked to national the population register, the National Patient Register and the longitudinal integration database for health insurance and labour market studies from Statistics Sweden.
-We used stepwise forward and backward logistic regression in combination with bootstrap averaging (bagging), to develop a prediction model for patient deaths within 90 days after surgery.
+**PATIENTS AND METHODS:** We studied `r nrow(df)` patients with THA due to primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` from the Swedish Hip Arthroplasty Register, linked to the national population register, the National Patient Register and the longitudinal integration database for health insurance and labour market studies from Statistics Sweden.
+We used a bootstrap ranking procedure using a LASSO-type penalty to develop a prediction model for patient deaths within 90 days after surgery.
 Predictive power was assessed by the area under the reciever operating characteristic curve (AUC).
@@
 Best predictive performance for 90-day mortality was found for a model combining `r coefs_print` (AUC = `r with(aucs, AUCtrain_est[name == "full"])`).
-This mdoel was superior to the established but complex comorbidity measures RxRiskV Score (`r with(aucs, AUCtrain_est[name == "RxRiskV"])`), the Charlson comorbidity index (AUC = `r with(aucs, AUCtrain_est[name == "CCI"])`), and the Elixhauser comorbidity score (AUC = `r with(aucs, AUCtrain_est[name == "ECI"])`).
+This mdoel was superior to the established but complex Charlson comorbidity index (AUC = `r with(aucs, AUCtrain_est[name == "CCI"])`), and the Elixhauser comorbidity score (AUC = `r with(aucs, AUCtrain_est[name == "ECI"])`).
 
@@
 The presence of pre-surgery comorbidity is associated with poorer outcome after the insertion of total hip arthroplasty (THA).
-Patients with increased comorbidity an increased risk of early postoperative mortality and revision [@Inacio2015; @Gordon2013], as well as inferior patient-reported outcomes [@Gordon2013; @Hofstede2016].
+Increased comorbidity is associated with an increased risk of early postoperative mortality and revision [@Inacio2015; @Gordon2013], as well as inferior patient-reported outcomes [@Gordon2013; @Hofstede2016].
 In research settings, comorbidity is commonly measured using multi-facetted diagnosis- or prescription-based coding algorithms [@Bozic2013].
@@
 Oftentimes, such datasets can only be created by linking several population-based registries, raising both ethical and practical concerns.
-Also, each of the comorbidity index exist in numerous versions [@Sundararajan2004; @Deyo1992; @Quan2011; @Cleves1997; @Walraven2009].
+Also, each of the comorbidity indices exist in numerous versions [@Sundararajan2004; @Deyo1992; @Quan2011; @Cleves1997; @Walraven2009].
 Interpretation and comparison between different studies is therefore difficult.
@@
 In the context of trauma, prediction tools are common, and it has been possible to reduce the number of variables without losing predictive power [@Gerdin2016].
-No model has so far been broadly accepted for elective THA however [@Manning2016].
+No model has so far been broadly accepted for elective THA however [@Manning2016; @Bulow2017].
 An easily applicable tool with few dimensions is thus needed, both in research and in clinical practice.

### 2019-10-02 reports/article.Rmd
Save changes before making research compendium

@@
 title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on Swedish and English register data"
-author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Adrian Sayers^5,6^, Szilard Nemes^2,4^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
+author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Szilard Nemes^2,4^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^, ... Erik L, ...
 date: '2019-03-12'
@@
 )
-knitr::opts_chunk$set(echo = FALSE, root.dir = '..')
-library(ProjectTemplate)
-load.project()
+knitr::opts_chunk$set(echo = FALSE)
+
+library(tidyverse)
+
+# Load all files
+walk(file.path("../cache", dir("../cache", ".RData")), load, .GlobalEnv)
 ```
@@
 **Correspondence:** 
-AG: anne.l.garland@gmail.com, Phone: +46 498 26 80 00, Postal: Orthopeadic Department, Visby Hospital, St Göransgatan 10, 621 84 Visby, Sweden  
+AG: anne.l.garland@gmail.com, Phone: +46 498 26 80 00, Postal: Orthopeadic Department, Visby Hospital, St Goransgatan 10, 621 84 Visby, Sweden  
 EB: erik.bulow@registercentrum.se  
@@
 
+```{r}
+get_auc <- function(mod) subset(brlasso_tbl_auc, Model == mod, "AUC", drop = TRUE)
+```
+
 **RESLUTS:** The unadjusted cumulative 90-day survival was `r surv90d`.
-Best predictive performance for 90-day mortality was found for a model combining `r coefs_print` (AUC = `r with(aucs, AUCtrain_est[name == "full"])`).
-This mdoel was superior to the established but complex Charlson comorbidity index (AUC = `r with(aucs, AUCtrain_est[name == "CCI"])`), and the Elixhauser comorbidity score (AUC = `r with(aucs, AUCtrain_est[name == "ECI"])`).
+Best predictive performance for 90-day mortality was found for a model combining `r coefs_print` (AUC = `r get_auc("BRLasso any")`).
+This mdoel was superior to the established but complex Charlson comorbidity index (AUC = `r get_auc("CCI")`), and the Elixhauser comorbidity score (AUC = `r get_auc("ECI")`).
+A web calculater to aid estimation of similar probabilities is available at https://erikbulow.shinyapps.io/thamortpred/.
 
@@
 ```{r flowchart, fig.cap = "Flowchart depicting inclusion criteria and number of patients included in the development phase of the model."}
-knitr::include_graphics("../../graphs/fig1.png")
+knitr::include_graphics("../graphs/flowchart.png")
 ```
@@
 Patients recorded in the Swedish Hip Arthroplasty Register (SHAR) with cemented primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` were included in the development phase of the study (Figure \@ref(fig:flowchart)).
-Only the first THA was accounted for in bilaterally operated patients and patients were excluded if a second THA was inserted within 90 days after the first.
-Hence, the ninety-day mortality were not influenced by any subsequent THA surgery.
+Only the last operated hip was accounted for in patients with bilateral THA [@Bulow2019a].
 Follow-up started on the day of surgery and ended at death, emigration, or December 31st `r max(df$P_SurgYear)`, whichever came first.
@@
 
-Age, sex, body mass index (BMI), ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from SHAR, with a completeness of 96-98% [@shpr2017].
+Age, sex, body mass index (BMI), ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from SHAR, with a completeness of 96-98% [@shpr2018].
 Data on education level (low/middle/high) and civil status, were collected from the longitudinal integration database for health insurance and labour market studies from Statistics Sweden [@Ludvigsson2019].
@@
 The register contains all relevant diagnoses coded by ICD-10, as well as admissions and discharge dates for in- and outpatient visits in all private and public hospitals [@Ludvigsson2011].
-We used medical data from the national prescription register where all prescriptions are classified by the Anatomic Therapeutic Chemical classification system (ATC).
 Death dates were linked from the national population register.
 
-Comorbidity was recognized by individual ICD-10 and ATC codes grouped into 17 categories according to CCI [@Charlson1987; @Deyo1992; @Quan2005], 31 categories according to ECI and 42 categories according to the RxRiskV [@Fishman2003; @Sloan2003] classification.
-RxRiskV medications prescribed for inflammatory pain was excluded since those are also sold over-the-counter in grocery stores, yielding questionable completeness in the prescription register.
-Patients with no hospital visits or no prescriptions were assumed to have no comorbidity.
+Comorbidity was recognized by individual ICD-10 grouped into 17 categories according to CCI [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to ECI [@Elixhauser1998; @???Quan2005].
+Patients with no recorded hospital visits during one year before surgery, were assumed to have no comorbidity.
 
-Some comorbidities were identified by more than one classification and some distinct comorbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
-We identified 16 broader categories in addition to 5 standalone ECI classes and 14 RxRiskV classes that were kept unchanged.
-Groups were merged according to clinical relevance as to be recognized in a patient-doctor meeting without access to external register data (see supplementary data).
+```{r tabcategorization}
+knitr::kable(
+  tab_categorization, 
+  caption = "Categorization of individual Charlson (CCI) and Elixhauser (ECI) comorbidities into broader comorbidities.")
+```
+
+Some comorbidities were identified by both CCI and ECI, and some distinct comorbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
+We identified `r nrow(tab_categorization)` broader categories (Table \@ref(tab:tabcategorization)) in addition to 5 standalone ECI classes that were kept unchanged (`r glue::glue_collapse(ECI_keep_text, ", ", last = " and ")`).
+Groups were merged according to clinical relevance as to be recognized in a patient-doctor meeting without access to external register data.
+Comorbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process described below.
 
@@
 
-Further analysis was based on logistic regression since no censoring occoured within the 90 day study period.
+Further analysis were based on logistic regression since no censoring occoured within the 90 day study period.
+We used a modelling procedure described by @Guo2015 as a bootstrap ranking procedure with a logistic LASSO-type penalty.
+Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1. 
+1,000 bootstrap samples were drawn from the initial data set [@Austin2004].
+We used 10-fold cross validation for every boot-strap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic least absolute shrinkage and selection operator (LASSO) model.
+We then only kept the $\lambda$:s minimizing the mean cross-validated deviances.
+Those $\lambda$:s were used to estimate coefficients for each of the 1,000 bootstrap samples.
+Absolute values from those estimates were used as a measure of variable importance.
+Piecewise linear regression was then used to detect a break point at witch a significant drop in variable importance were observed.
+Potential predictors with variable importance above this break point were considered important and kept as model candidates.
+The whole process was repeated ten times.
 
-Univariable models were applied to all predictors individually to estimate odds ratios for descriptive purposes.
+Covariates that were selected each of the ten times were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables.
+We will call this modell "BRL all", where BRL stands for bootstrap ranking LASSO.
+A similar model, including any variable selected at least one out of ten times will be called "BRL any".
+Univariable models with the ASA score, CCI and ECI were used for comparion, as well as a multivariable model with age and sex.
+Each model including age where fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.
 
-All potential predictors were then partioned into four groups where there were assumed to be no colinearity among variables: 1) the 17 combined comorbidities, 2) the 5 ECI comorbidities, 3) the 14 RxRiskV comorbidities, and 4) age, sex, BMI, ASA, type of hospital, year of surgery, education and civil status.
+Each of those models were used to predict the probability of death within 90 days for each patient.
+Sensitivity and specificity were estimated to form reciever operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of predicitve power for each model.
+Models with a lower 95 % confidence limit [@Delong1988] exceeding 0.7, were considered good.
 
-1,000 bootstrap samples were drawn from the initial data set.
-Early mortality is very rare for patients with elective THA however.
-Downsampling stratified on the outcome was therefore applied to each sample to include only as many survivals as deaths.
-This gave us balanced samples and managable computations.
-[@King2001; @Courvoisier2011; @Austin2017a].
+Odds ratios for the model with highest AUC were estimated with 95 % confidence intervals based on interpolations of profile traces [@Ripley2002].
 
-A step-wise linear regression procedure based on the Aikake Information Criteria (AIC) was applied to each bootstrap sample to identify important predictors from each of the four groups of variables [@Sauerbrei1999; @Zellner2004; @Lukacs2010].
-Variables that were kept in the final model for at least 75 % of all bootstrap samples were selected as candidates for further consideration.
-
-A full main effects model were made to include all previously selected candidate predictors.
-Two modified versions were considered as well, where age was modeled by resttricted cubic spline with either 2 or 3 knots.
-Three simplified models with age and sex only, were used for comparison, as well as four univariable regression models with ASA, CCI, ECI and RxRiskV alone.
-
-We then took another 1,000 downsampled bootsrtap samples to estimate coefficients for each model.
-Model averaging based on the information criteria was used for the full main effects model and its variants.
-Hence, an automated model selection procedure evaluating all possible subsets of the included main effects, yielding averaged coeffients estimates based on variable importance.
-
-Each model, now with their coefficient estimates fixed, were then evaluated by their predictive power based on their area under the reciever operation characteristics (ROC) curve (AUC).
-Point estimates as well as 95 % confidence intervals were estimated by emperical bootstrap, once again with 1,000 resamples but this time without downsampling.
-Sensitivity and specificity from each bootstrap sample were subjected to a local polynomial regression fit (loess) to illustrate an averaged reciever operation characteristics (ROC) curve for each model.
-
-We used `r substr(R.version.string, 1, 15)` (R Foundation for Statistical   Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC, MuMIn and rFSA [@Lambert2018].
+We used `r substr(R.version.string, 1, 15)` (R Foundation for Statistical   Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr and pROC.
+We also build an online web calculator using the shiny package.
 
@@
 
-There were `r nrow(df)` patients (Figure \@ref(fig:flowchart)), `r paste(range(df$P_Age), collapse = " - ")` years old, whereof `r round(mean(df$P_Sex == "Female") * 100)` % were female.
+There were `r format(nrow(df), big.mark = ", ")` patients (Figure \@ref(fig:flowchart)), `r paste(range(df$P_Age), collapse = " - ")` years old, whereof `r round(mean(df$P_Sex == "Female") * 100)` % were female.
 `r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` patients died within 90 days and no one was censored before that.
@@
 Some characteristics of the study population are presented in Table \@ref(tab:tab1).
-`r mean(df$Rx_index_index != 0) * 100` % of all patients had at least one pre-surgery comorbidity according to RxRiskV, but only `r mean(df$CCI_index_quan_original != 0) * 100` % and `r mean(df$ECI_index_sum_all  != 0) * 100` % according to ECI and CCI.
+`r mean(df$CCI_index_quan_original != 0) * 100` % of all patients had at least one pre-surgery comorbidity according to CCI, and `r mean(df$ECI_index_sum_all  != 0) * 100` % according to ECI.
 The proportion of patients with ASA class 3 was `r mean(df$P_ASA == 3) * 100` %.
@@
 ```{r tab1}
-prestext(table1) %>% 
-  knitr::kable(caption = "Baseline demographic information on the study population 2008-2012.
-CCI/ECI = Charlson/Elixhauser Comorbidity Index.
-Individual comorbidities without prefixes were based on combinations of different comorbidity classifications.
-Prefix ECI and Rx(RiskV) indicate classification specific comorbidities.")
+knitr::kable(
+  table1, 
+  caption = "Baseline demographics.
+CCI/ECI = Charlson/Elixhauser comorbidity indices.
+Comorbidities prefixed with ECI are defined by the Elixhauser classification.
+Remaining comorbidities are based on the previously described combination of CCI and ECI.
+Comorbidities recorded for at least one patient who survived 90 days, and one who did not, were modelled as potential predictors.")
 ```
 
-Odds ratios from univariable logistic regresison models are presented in Table \@ref(tab:tab2).
-Higher age, ASA, CCI, ECI and RxRiskV, as well as being male or widow(er), or having cancer, gout, kidney disease or heart infarct, increase the risk of death within 90 days after surgery.
-Private hospitals tend to operate healthier patients and the odds of death decrease slightly by year of surgery.
+There were `r nrow(comb_lgl)` comorbidities that were not recorded for any patient who died: `r comb_lgl_text`. Corresponding variables were excluded from the modelling process.
 
-```{r tab2}
-prestext(t2_univariable) %>% 
-  knitr::kable(caption = "Unadjusted odds ratios (OR) for 90-day mortality with 95 % confidence intervals (CI).
-Individual comorbiditities are only presented if selected in at least 75 % of all bootstrap samples (selected).
-Variables included in the final model are marked with 'FM' in model column.")
-```
-
-ASA class was the only variable included in the final models by all stage-wise regression procedures from each of the 1,000 bootstrap samples (100 % in column "selected" in Table \@ref(tab:tab2)).
-Most potential predictors were rarely selected and only `r coefs_print`, where selected at least 75 % of the time (Table \@ref(tab:tab2)).
+There were no observed differences for two versus three knots in models with age modelled by restricted cubic splines.
+We will therefore only present results for three knots.
+No use of splines however improved the observed AUCs compared to simpler main effect models.
+The "BRL any" model had an estimated AUC (95 % CI) of `r get_auc("BRL any")`.
+This model was significantly better than the "BRL all" model with `r get_auc("BRL all")`.
+Univariable models with either ASA, CCI or ECI performed relatively poor.
+The model with age and sex was better but still with an AUC not significantly higher than 0.7 (Table \@ref(tab:auc)).
 
 ```{r auc, warning = FALSE}
-knitr::kable(table_auc, caption = "Area Under the Curve (AUC) as a measure of predictive power for the full model compared to simple models with standard comorbidity measures, as well as a reference model with age and sex.
-Age was included as either a main effect, or in the form of restricted cubic splines (rcs) with either 2 or 3 knots.")
+knitr::kable(
+  brlasso_tbl_auc, 
+  caption = "Area Under the Curve (AUC) as a measure of predictive power for the 'BRL any' model compared to 'BRL all', a model including the independent variables included in all of the bootstrao ranking models, a simpler model with age and sex, as well as univariable models with ASA score and the Charlson (CCI) or Elixhauser (ECI) comorbidity indices.
+Age was included as either a main effect, or in the form of restricted cubic splines (RCS) with three knots.")
 
@@
 
-A model with those covariates, the "full" model, had an out-of-sample AUC of `r aucs$AUCtrain_est[[1]]`.
-The predictive power did not improve with age modeled by restricted cubic splines (Table \@ref(tab:auc) and Figure \@ref(fig:aucci)).
-The full model led to better predictions compared to a simplified model with only age and sex, or any of the conventional comorbidity indices (CCI, ECI and RxRiskV).
-The full model is further described in Table \@ref(tab:fullmodel).
+ROC curves for some of the models are displayed in Figure \@ref(fig:rocs). The "BRL any" model is superior to all other models, although some of the others are intertwined in some areas of the plot.
 
-```{r aucci, fig.cap = "Area Under the Curve (AUC) as a measure of predictive power for the full model compared to simple models with standard comorbidity measures, as well as a reference model with age and sex.
-Age was included as either a main effect, or in the form of restricted cubic splines (rcs) with 2 or 3 knots.
-Confidence intervals based on 1,000 non-paramteric bootstrap replicates."}
+```{r rocs, fig.cap = "ROC curves for some of the models. The 'BRL any' model is distinguished from other models, which are partially over-lapping."}
+knitr::include_graphics("../graphs/brlasso_roc.png")
+```
 
-knitr::include_graphics("../../graphs/auc_ci.png")
+AUCs and 95 % confidence intervals are illustrated for the same models in \@ref(fig:aucci)).
+
+```{r aucci, fig.cap = "Area Under the Curve (AUC) as a measure of predictive power for the 'BRL any' model compared to a simpler model with age and sex, as well as univariable models with ASA score and the Charlson (CCI) or Elixhauser (ECI) comorbidity indices."}
+
+knitr::include_graphics("../graphs/brlasso_auc_ci.png")
+```
+
+The ability of model "BRL any" to distinguish patients who died within 90 days is illustrated in Figure \@ref(fig:sep).
+We can see that patients who did die within 90 days (blue color) had, on average, higher predicted probabilities to do so.
+
+```{r sep, fig.cap = "Patients who died within 90 days (blue) were, on average, estimated to have a higher probability to do so."}
+knitr::include_graphics("../graphs/brlasso_separation_auc.png")
+```
+
+Estimated coefficients and correspondnig odds ratios for the "BRL any" model is presented in Table \@ref(tab:brlany).
+
+```{r brlany}
+
+knitr::kable(
+  rename(brlasso_tbl_coefs, X = X_text),
+  caption = 'Estimated coefficients and odds ratios with 95 % confidence intervals for the "BRL any" model.',
+  escape = FALSE
+)
 
@@
 
-```{r fullmodel}
-knitr::kable(table_full_model_coefs, caption = "Estimated odds ratios for variables in the full model based on full averages from 1,000 bootstrap samples.")
+The resulting "BRL any" model can predict the probability of death ($p$) within 90 days for new patients as: `r coefs_form` where $X_i$ indicate independent variables as notated in table \@ref(tab:brlany).
+Note however that this formula is only valid for patients within the observed age range (`r paste(range(df$P_Age), collapse = " - ")` years).
+
+```{r}
+get_risk <- function(who) preds[preds$id == who, "p", drop = TRUE]
 ```
 
-ROC curves for all models compared to the full model are displayed in Figure \@ref(fig:rocs). Each models ability separate survivors from death are further illustrated in Figure \@ref(fig:sep).
+A `r min(df$P_Age)` yer old woman with ASA = 1 and none of the important comorbidities would have a `r get_risk("baseline")` risk to die.
+If she was instead `r quantile(df$P_Age, .25)` year old (the first age quantlie), her risk would have increased to `r get_risk("healthy_female")`.
+An older and sicker man, `r quantile(df$P_Age, .75)` years old (the third age quantile) with ASA = 3 and a previsous heart condition would have a risk of `r get_risk("sicker_man")`.
+The perhaps unrealistic case of a `r max(df$P_Age)` year old man with ASA = 3 and all listed comorbidities would have a theoretical risk as high as `r get_risk("high_risk")`.
+Note however that this extreme case relies on extrapolation and is highly unreliable, since no such person was actually observed in the data.
 
-```{r rocs, fig.cap = "ROC curves for each model compared to the full main effects model."}
-knitr::include_graphics("../../graphs/rocs.png")
-```
-
-```{r sep, fig.cap = "Separation"}
-knitr::include_graphics("../../graphs/separation_auc.png")
-```
+We have provided a web calculater to aid estimation of similar probabilities (https://erikbulow.shinyapps.io/thamortpred/).
 
@@
 
-We found that a univariable main effects logistic regression model with `r coefs_print` was superior to predict death within 90 days after insertion of elective THA.
-Some covariates are not statistically significant themselves (\@ref(tab:fullmodel)) but are still relevant due to unobserved heteregeneity  [@Mood2010].
+We found that a multivariable main effects logistic regression model with `r coefs_print` was able to predict death within 90 days after insertion of elective THA.
+Some covariates ni the model were not statistically significant by themselves but were still relevant due to unobserved heteregeneity [@Mood2010].
 
@@
 It is however not obvious that this relation must be striclty linear.
-We therefore allowed for a more flexible relation based on restricted cubic splines, but found that a linear relationship was indeed most sufficient.
+We therefore allowed for a more flexible relation based on restricted cubic splines, but found that a linear relationship was equally good.
 
-ASA class was the only covariate chosen by all stepwise regression procedures.
-This might not be surprising due to its definition.
-It is interesting to note however that a univariable model with ASA grade alone, yielded the lowest predictive power of all models compared.
-An out-of-sample point estimate of its AUC was < 0.5 (\@ref(tab:fullmodel) and \@ref(fig:rocs), meaning that predictions based solely on ASA would be worse than random.
-This shows the important of validation, due to the risk of overfitting, as well as the importance of multivariable adjustment.
 ASA class is routinely assessed pre-operatively in most developed countries.
@@
 It has previously been compared to the CCI, but not with respect to mortality after THA [@Whitmore2014; @Kork2015].
-It has been argued that a prescription-based measure would be more reliable than a diagnosed based measure [@Iezzoni1997; @Johnson2006].
-We did confirm that estimation of AUC were more stable for the RxRiskV index compared to CCI and ECI (as seen by the width of the confidence intervals in figure \@ref(fig:aucci)), although point estimates of the latter was marginaly higher.
-This reflects that both coverage and completeness from the prescription register is better compared to the national patient regiseter.
-The first hold records of all medications prescribed, the second only contains information that the doctor deemed  relevant at the patient encounter.
 
@@
 AG and NH initiated the study and managed the ethical review board application.
-EB, AS, SN, and AG performed the statistical analyses.
+EB, EL and SN performed the statistical analyses.
 AG and EB drafted the manuscript.

### 2019-10-02 lib/BR_lasso_coefs.R

@@ -1,0 +1,34 @@
+#' List important variables by importande from bootstrap ranking Lasso
+#'
+#' NOTE! It is important that all varibles are on a standardized scale for the
+#' variable importance to make any sense!
+#' @param df data.frame
+#' @return tibble with important variables and a measure thereof
+BR_lasso_coefs <- function(df) {
+  df %>%
+    rsample::bootstraps(config$N_bots) %>%
+    mutate(
+      data  = map(splits, as_tibble),
+      lasso = map(data, lasso)
+    ) %>%
+    unnest(lasso) %>%
+    group_by(variable) %>%
+    summarise(
+      impor = abs(mean(coef))
+    ) %>%
+    ungroup() %>%
+    filter(impor > 0) %>%
+    arrange(desc(impor)) %>%
+    add_rowindex()
+}
+
+#' @param coefs output from BR_lasso_coefs
+#' @return break point for important variables
+break_p <- function(coefs) {
+  with(
+    coefs,
+    SiZer::piecewise.linear(.row, impor, 1)
+  ) %>%
+  {.$change.point} %>%
+  round()
+}

### 2019-10-02 lib/glmCI.R

@@ -1,26 +1,0 @@
-### Summary function for GLM
-glmCI <- function(model, exponent = FALSE, alpha = 0.05) {
-  co <- coef(model)
-  l  <- qnorm(1 - alpha / 2) * sqrt(diag(summary(model)$cov.scaled))
-
-  # calculuate CI for linear case
-  mat <- tibble::tibble(var = names(co), beta = co, low = co - l, high = co + l)
-
-  # if exponent=TRUE, exponeniate the coefficients and CIs
-  if (exponent) {
-    mat <- mutate_if(mat, is.numeric, exp)
-    names(mat)[names(mat) == "beta"] <-
-      switch(summary(model)$family$link, logit = "OR", log  = "RR")
-  }
-  filter(mat, var != "(Intercept)")
-}
-
-
-# Summarise survfit results
-survEst <- function(fit, times) {
-  x <- as.data.frame(summary(fit, times = times))
-  ind <- c("time", "surv",  "lower", "upper")
-  if (!is.null(fit$strata))
-    ind <- c(ind, "strata")
-  x[ind]
-}

### 2019-10-02 lib/selected_vars.R

@@ -1,34 +1,0 @@
-# Get names of important variables
-selected_vars <- function(df, N_boot = config$N_bots) {
-
-  # Perform lasso regression to bootstrap samples
-  coefs_all <-
-    df %>%
-    rsample::bootstraps(N_boot) %>%
-    mutate(
-      data  = map(splits, as_tibble),
-      lasso = map(data, lasso)
-    ) %>%
-    unnest(lasso) %>%
-    group_by(variable) %>%
-    summarise(
-      impor = abs(mean(coef))
-    ) %>%
-    ungroup() %>%
-    filter(impor > 0) %>%
-    arrange(desc(impor)) %>%
-    add_rowindex()
-
-  # How many variables needed for good model
-  break_p <-
-    with(
-      coefs_all,
-      SiZer::piecewise.linear(.row, impor, 1)
-    ) %>%
-    {.$change.point} %>%
-    round()
-
-  # Keep only important variables
-  coefs_all %>%
-    slice(seq_len(break_p))
-}

### 2019-10-02 munge/02-survdata.R

@@ -51,9 +51,6 @@ df <-
     -P_SurgDate,
     -P_DiaGrp,
     -P_KVA1,
-    -time_between,
-    -new,
-    -bilateral,
     -P_ProstType,
     -contains("index"),
     elix_icd10_index_sum_all,
@@ -62,3 +59,4 @@ df <-
   # Shorter names
   rename_all(~ gsub("charlson_icd10", "CCI", .)) %>%
   rename_all(~ gsub("elix_icd10", "ECI", .))
+

### 2019-10-02 munge/03-compositvariabler.R

@@ -1,81 +1,51 @@
 # Förbered kompisitvariabler som ev kan ersätta enskilda diagnoser
 
+# Fulfix!
+# Fel i coder-paketet döpte "malignancy" till "malingnancy"
+# Ska funka korrekt med senaste coder-version men vill inte köra om just nu!
+df <- rename(df, CCI_malignancy = CCI_malingnancy)
+
+
+# Make mutate statments from Excel-file -----------------------------------
+
+# Define it!
+categorization <-
+  categorization.Blad1 %>%
+  mutate_all(zoo::na.locf) %>%
+  mutate_at(vars(new, old), ~ gsub("/| ", "_", tolower(.))) %>%
+  mutate(
+    new = paste0("c_", new),
+  ) %>%
+  unite("old", from, old) %>%
+  group_by(new) %>%
+  summarise(old = paste(old, collapse = " | ")) %>%
+  ungroup() %>%
+  mutate(
+    new = syms(new),
+    old = parse_exprs(old)
+  )
+
+# Do it!
+c_cols <- bind_cols(pmap(categorization, ~transmute(df, !!.x := !!.y)))
+
+
+# Combine some existing columns with new ones --------------------------------
+
+
+ECI_keep_text <-
+  c("hypothyroidism", "coagulopathy", "obesity", "weight loss",
+    "fluid electrolyte disorders")
+cache("ECI_keep_text")
+ECI_keep_vars <- paste0("ECI_", gsub(" ", "_", ECI_keep_text))
+
 df <-
   df %>%
-  mutate(
-
-    c_heart_condition =
-      ECI_congestive_heart_failure              |
-      CCI_congestive_heart_failure              |
-      ECI_valvular_disease,
-    c_heart_infarct =
-      CCI_myocardial_infarction,
-    c_arrythmia =
-      ECI_cardiac_arrhythmias,
-    c_hypertoni =
-      ECI_hypertension_uncomplicated            |
-      ECI_hypertension_complicated,
-    c_kidney_disease =
-      ECI_renal_failure                         |
-      CCI_renal_disease,
-    c_diabetes =
-      ECI_diabetes_uncomplicated                |
-      ECI_diabetes_complicated                  |
-      CCI_diabetes_without_complication         |
-      CCI_diabetes_complication,
-    c_vascular_disease =
-      ECI_peripheral_vascular_disorder          |
-      CCI_peripheral_vascular_disease           |
-      CCI_cerebrovascular_disease,
-    c_cancer =
[…]

### 2019-10-02 munge/04-remove_empty_variables.R

@@ -36,6 +36,14 @@ comb_lgl <-
 
 cache("comb_lgl")
 
+comb_lgl_text <-
+  comb_lgl$name %>%
+  {gsub("c_", "", .)} %>%
+  {gsub("_", " ", .)} %>%
+  {gsub("aids hiv", "AIDS/HIV", .)} %>%
+  glue_collapse(", ", last = " and ")
+
+cache("comb_lgl_text")
 
 # Same for factor variables -----------------------------------------------
 

### 2019-10-02 munge/06-variables_differnet_BRLasso_models.R

@@ -32,3 +32,4 @@ best_coefs_all <-
   pluck(1)
 
 cache("best_coefs_all")
+

### 2019-10-02 shiny/server.R

@@ -1,31 +1,58 @@
 library(shiny)
-library(ggplot2)
+library(tidyverse)
+source("misc.R")
+
+
+# input <- list(P_Age = 35, P_ASA = 2, checkboxes = c("cns", "cancer")) # EXAMPLE
+
 
 function(input, output) {
 
-   # From all_models$tidy[[3]]$estimate %>% dput()
-  betas <-
-    c(-6.93067750638018,
-      1.45504113519804,
-      0.726409647846406,
-      0.836939634463218,
-      0.718668411163381,
-      0.772009274919897,
-      0.747733943488821)
+  output$textout <-
+    renderUI({
 
-  output$textout <- renderPrint({
+      inp <- vector("list", nrow(coefs))
+      names(inp) <- coefs$coef_name
+      inp <- imap(inp, ~ if (.y %in% names(input)) input[[.y]])
 
-    x <- c(
-      1, input$kidney, input$asa, input$cns, input$heart,
-      (input$age - 72) / 7.8, input$sex
-    )
+      inp[coefs$coef_present %in% input$checkboxes] <- TRUE
+      inp <- map(inp, ~ if (is.null(.)) FALSE else .)
+      inp <-
+        as_tibble(inp) %>%
+        mutate(P_Gender = ifelse(P_Gender == "Male", "Man", "Kvinna")
+      )
 
-    y <- sum(betas * x)
+      p <- (1 - predict(fit_brl_any_lean, inp, "response")) * 100
 
-    p <- exp(y) / (1 + exp(y))
-    p <-  1 - ifelse(is.na(p) & !is.na(y), 1, p)
+      p <-
+        if (p > 99) {
+          "Almost 100 %"
+        } else if (p < 90) {
+          "We do not know!"
+        } else {
+          sprintf("%s %%", prettyNum(p, digits = 0))
+        }
 
-    sprintf("You have a %f %% chance to survive at least 90 days after surgery!", p)
-  })
+      h1(p)
+    })
+
+  output$about <-
+    renderUI({
+
+      doc <- tags$html(
+        tags$head(
+          tags$title('My first page')
+        ),
+        tags$body(
+          h1('My first heading'),
+          p('My first paragraph, with some ',
+            strong('bold'),
+            ' text.'),
+          div(id='myDiv', class='simpleDiv',
+              'Here is a div with some attributes.')
+        )
+      )
+      doc
+    })
 
 }

### 2019-10-02 shiny/ui.R

@@ -1,31 +1,56 @@
 library(shiny)
-library(ggplot2)
+library(shinythemes)
+library(tidyverse)
 
+source("misc.R")
 
 fluidPage(
-  titlePanel("Will you survive 90 days after your surgery?"),
+  titlePanel("Probability to survive at least 90 days after THA surgery?"),
+
+  helpText("Bla bla bla ..."),
 
   sidebarPanel(
     sliderInput(
-      'age',
+      'P_Age',
       'Age',
       min   = 35,
-      max   = 98,
+      max   = 99,
       value = 72,
       step  = 1,
       round = TRUE
     ),
 
-    checkboxInput('sex',    'Male sex'),
-    checkboxInput('asa',    'ASA >= 3'),
-    checkboxInput('kidney', 'Kidney disease?'),
-    checkboxInput('heart',  'Heart condition?'),
-    checkboxInput('cns',    'CNS?')
+    radioButtons(
+      "P_Gender",
+      "Sex",
+      choices = c("Male", "Female"),
+      selected = "Female",
+      inline = TRUE,
+    ),
+
+    radioButtons(
+      "P_ASA",
+      "ASA class",
+      choices = 1:3,
+      selected = 1,
+      inline = TRUE,
+    ),
+
+    checkboxGroupInput(
+      "checkboxes",
+      "Do you have any of the following:",
+      choices = coefs$coef_present[coefs$checkbox],
+    )
+
   ),
 
   mainPanel(
-    textOutput(
-      "textout"
+    tabsetPanel(
+      type = "tabs",
+      tabPanel("Prediction", htmlOutput("textout")),
+      tabPanel("About", htmlOutput("about"))
     )
-  )
+  ),
+
+  theme = shinytheme("superhero")
 )

### 2019-10-02 src/fig_auc.R

@@ -1,16 +1,21 @@
-library(ProjectTemplate)
-load.project()
+suppressMessages({library(ProjectTemplate); load.project()})
 
+digs <- options("digits")
+options(digits = 2)
 
 fig_auc_ci <-
   all_models %>%
+  filter(fig) %>%
   mutate(Model = fct_reorder(Model, AUC_est)) %>%
   ggplot(aes(Model, AUC_est)) +
+  geom_hline(yintercept = .7, color = "darkgreen", linetype = "dashed", size = 2) +
   geom_pointrange(aes(ymin = AUC_lo, ymax = AUC_hi), size = 1.5) +
   coord_flip() +
   theme_minimal() +
   theme(axis.title.y = element_blank()) +
   ylab("AUC with 95 % confidence intervals (De-Long)") +
-  scale_y_continuous(breaks = seq(0, 1, .1))
+  scale_y_continuous(breaks = seq(0, 1, .05))
+
+options(digits = digs$digits)
 
 ggsave("graphs/brlasso_auc_ci.png", fig_auc_ci)

### 2019-10-02 src/tab_define_groups.R

@@ -1,0 +1,11 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+
+tab_categorization <-
+  categorization.Blad1 %>%
+  mutate_all(zoo::na.locf) %>%
+  group_by(new, from) %>%
+  summarise(old = paste(old, collapse = "\n ")) %>%
+  pivot_wider(names_from = from, values_from = old)
+
+cache("tab_categorization")

### 2019-10-02 src/table1.R

@@ -1,5 +1,4 @@
-library(ProjectTemplate)
-load.project()
+suppressMessages({library(ProjectTemplate); load.project()})
 
 fctrs <- c("P_Sex", "P_ASA", "ECI", "CCI")
 
@@ -36,7 +35,7 @@ t1 <-
     test = FALSE
   )
 
-cahce("t1")
+cache("t1")
 
 table1 <-
   t1 %>%

### 2019-10-03 reports/article.Rmd
version of manuscrip supplied to co-athours

@@
 title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on Swedish and English register data"
-author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Szilard Nemes^2,4^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^, ... Erik L, ...
-date: '2019-03-12'
+author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Erik Lenguerrand^5^, Ashley Blom^6^, Mark Wilkinson^7^, Adrian Sayers^8^, Szilard Nemes^2,4^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
+date: '`r Sys.Date()`'
 bibliography: P:/library.bib
@@
 options(
-  digits            = 1,
+  digits            = 0,
   knitr.kable.NA    = '',
@@
 4.	Department of Orthopaedics, Institute of Clinical Sciences, The Sahlgrenska Academy, University of Gothenburg, Gothenburg, Sweden
-5. ...
+5. Translational Health Sciences, Bristol Medical School, University of Bristol, England.
 6. ...
+7. ...
+8. ...
 
 **Correspondence:** 
-AG: anne.l.garland@gmail.com, Phone: +46 498 26 80 00, Postal: Orthopeadic Department, Visby Hospital, St Goransgatan 10, 621 84 Visby, Sweden  
+AG: anne.l.garland@gmail.com, Phone: +46 498 26 80 00, Postal: Orthopedic Department, Visby Hospital, St Goransgatan 10, 621 84 Visby, Sweden  
 EB: erik.bulow@registercentrum.se  
@@
 **OBJECTIVE:** Early mortality after total hip arthroplasty (THA) is associated with comorbidity.
-Diagnosis-based tools such as the Charlson or Elixhauser Comorbidity indices have been used to quantify this.
-Those instruments are however complex and requires the use of extensive data sets not routinely available in most doctor-patient settings.
-We investigated if a simplified model, could substitute those complex indices to predict early mortality after THA.
+Diagnosis-based tools such as the Charlson or Elixhauser comorbidity indices have been used to quantify this.
+Those instruments are however complex and requires extensive data sets not routinely available in most doctor-patient settings.
+We searched for a simplified model to  substitute those complex indices in predicting early mortality after THA.
 
-**PATIENTS AND METHODS:** We studied `r nrow(df)` patients with THA due to primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` from the Swedish Hip Arthroplasty Register, linked to the national population register, the National Patient Register and the longitudinal integration database for health insurance and labour market studies from Statistics Sweden.
-We used a bootstrap ranking procedure using a LASSO-type penalty to develop a prediction model for patient deaths within 90 days after surgery.
-Predictive power was assessed by the area under the reciever operating characteristic curve (AUC).
+**PATIENTS AND METHODS:** We studied `r format(nrow(df), big.mark = ",")` patients with THA due to primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` from the Swedish Hip Arthroplasty Register, linked to the national population register, the National Patient Register and the longitudinal integration database for health insurance and labor market studies from Statistics Sweden.
+We used a bootstrap ranking procedure with logistic regression and a LASSO-type penalty to develop a prediction model for patient deaths within 90 days after surgery.
+Predictive power was assessed by the area under the receiver operating characteristic curve (AUC).
 The final model was applied to British data for external validation.
@@
 **RESLUTS:** The unadjusted cumulative 90-day survival was `r surv90d`.
-Best predictive performance for 90-day mortality was found for a model combining `r coefs_print` (AUC = `r get_auc("BRLasso any")`).
-This mdoel was superior to the established but complex Charlson comorbidity index (AUC = `r get_auc("CCI")`), and the Elixhauser comorbidity score (AUC = `r get_auc("ECI")`).
-A web calculater to aid estimation of similar probabilities is available at https://erikbulow.shinyapps.io/thamortpred/.
+Best predictive performance for 90-day mortality was found for a model combining `r coefs_print` (AUC = `r get_auc("BRL any")`).
+This model was superior to the established but complex Charlson comorbidity index (AUC = `r get_auc("CCI")`), and the Elixhauser comorbidity score (AUC = `r get_auc("ECI")`).
+A web calculator to aid clinical usage was published at https://erikbulow.shinyapps.io/thamortpred/.
 
@@
 
-The presence of pre-surgery comorbidity is associated with poorer outcome after the insertion of total hip arthroplasty (THA).
-Increased comorbidity is associated with an increased risk of early postoperative mortality and revision [@Inacio2015; @Gordon2013], as well as inferior patient-reported outcomes [@Gordon2013; @Hofstede2016].
-In research settings, comorbidity is commonly measured using multi-facetted diagnosis- or prescription-based coding algorithms [@Bozic2013].
-@Inacio2016 studied the ability of three commonly used coding algorithms to predict mortality after THA and total knee arthroplasty: the Charlson (CCI) and Elixhauser (ECI) Comorbidity Indices, as well as RxRiskV.
-These comorbidity measures are quite complex to estimate and are based on the availability of extensive datasets including in- and outpatient data on ICD-codes, or detailed information on drug prescriptions prior to surgery.
-Oftentimes, such datasets can only be created by linking several population-based registries, raising both ethical and practical concerns.
+The presence of pre-surgery comorbidity is associated with poorer outcome after the insertion of total hip arthroplasty (THA) [@Inacio2015], as well as inferior patient-reported outcomes [@Gordon2013; @Hofstede2016].
+In research settings, comorbidity is commonly measured using multi-faceted diagnosis- or prescription-based coding algorithms [@Bozic2013].
+@Inacio2016 studied the ability of the Charlson (CCI) and Elixhauser (ECI) Comorbidity indices to to predict mortality after THA and total knee arthroplasty.
+These comorbidity measures are quite complex to estimate and are based on the availability of extensive data sets including in- and outpatient data on ICD-codes.
+Oftentimes, such data sets can only be created by linking several population-based registries, raising both ethical and practical concerns.
 Also, each of the comorbidity indices exist in numerous versions [@Sundararajan2004; @Deyo1992; @Quan2011; @Cleves1997; @Walraven2009].
 Interpretation and comparison between different studies is therefore difficult.
+
 Comorbidity data have also been used in several universal and arthroplasty-specific risk prediction tools to make risk profiles for individual patients.
@@
 An easily applicable tool with few dimensions is thus needed, both in research and in clinical practice.
-We aimed to find such a model to predict the risk of 90-day postoperative mortality after THA.
+We aimed to find such a model to predict the risk of 90-day postoperative mortality after elective THA.
 
 # Patients and Methods
+
+Patients recorded in the Swedish Hip Arthroplasty Register (SHAR) with cemented primary hip osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` were included in the development phase of the study (Figure \@ref(fig:flowchart)).
+Only the last operated hip was accounted for in patients with bilateral THA [@Bulow2019a].
 
@@
 
-Patients recorded in the Swedish Hip Arthroplasty Register (SHAR) with cemented primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` were included in the development phase of the study (Figure \@ref(fig:flowchart)).
-Only the last operated hip was accounted for in patients with bilateral THA [@Bulow2019a].
-Follow-up started on the day of surgery and ended at death, emigration, or December 31st `r max(df$P_SurgYear)`, whichever came first.
+Data linkage, based on the unique identity numbers assigned to all Swedish residents [@Ludvigsson2009], were used to collect data from a variety of sources, as previously described by @Cnudde2016.
 
-Data linkage, based on the unique identity numbers assigned to all Swedish residents [@Ludvigsson2009], were used to collect data from a variaty of sources, as previsouly described by @Cnudde2016.
-
-Age, sex, body mass index (BMI), ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from SHAR, with a completeness of 96-98% [@shpr2018].
-Data on education level (low/middle/high) and civil status, were collected from the longitudinal integration database for health insurance and labour market studies from Statistics Sweden [@Ludvigsson2019].
+Age, sex, body mass index (BMI), ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from SHAR, with a completeness of 96-98 % [@shpr2018].
+Data on education level (low/middle/high) and civil status (married/un-married/divorced/widow[er]), were collected from the longitudinal integration database for health insurance and labor market studies from Statistics Sweden [@Ludvigsson2019].
 The Swedish National Patient Register was used for comorbidity data during one year before surgery.
@@
 
-Comorbidity was recognized by individual ICD-10 grouped into 17 categories according to CCI [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to ECI [@Elixhauser1998; @???Quan2005].
+Comorbidity was recognized by individual ICD-10 grouped into 17 categories according to CCI [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to ECI [@Elixhauser1998; @Quan2005].
 Patients with no recorded hospital visits during one year before surgery, were assumed to have no comorbidity.
@@
   caption = "Categorization of individual Charlson (CCI) and Elixhauser (ECI) comorbidities into broader comorbidities.")
+
 ```
@@
 Some comorbidities were identified by both CCI and ECI, and some distinct comorbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
-We identified `r nrow(tab_categorization)` broader categories (Table \@ref(tab:tabcategorization)) in addition to 5 standalone ECI classes that were kept unchanged (`r glue::glue_collapse(ECI_keep_text, ", ", last = " and ")`).
+We used those categories to establish `r nrow(tab_categorization)` broader categories (Table \@ref(tab:tabcategorization)) in addition to 5 standalone ECI classes that were kept unchanged (`r glue::glue_collapse(ECI_keep_text, ", ", last = " and ")`).
 Groups were merged according to clinical relevance as to be recognized in a patient-doctor meeting without access to external register data.
 Comorbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process described below.
+The final model was also altered to not include cancer as a predictor.
+Patients with cancer are sometimes treated differently in the clinical setting, introducing difficulties in interpretation of this variable.
 
@@
 
-Further analysis were based on logistic regression since no censoring occoured within the 90 day study period.
-We used a modelling procedure described by @Guo2015 as a bootstrap ranking procedure with a logistic LASSO-type penalty.
+Further analysis were based on logistic regression since no censoring occurred within the 90 day study period.
+We used a modelling procedure described by @Guo2015 as a bootstrap ranking procedure with a logistic least absolute shrinkage and selection operator (LASSO) model.
 Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1. 
 1,000 bootstrap samples were drawn from the initial data set [@Austin2004].
-We used 10-fold cross validation for every boot-strap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic least absolute shrinkage and selection operator (LASSO) model.
-We then only kept the $\lambda$:s minimizing the mean cross-validated deviances.
-Those $\lambda$:s were used to estimate coefficients for each of the 1,000 bootstrap samples.
+We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model.
+We then only kept $\lambda$:s minimizing the mean cross-validated deviances in each sample.
+Those $\lambda$:s were used to estimate model coefficients for each potential predictor.
 Absolute values from those estimates were used as a measure of variable importance.
-Piecewise linear regression was then used to detect a break point at witch a significant drop in variable importance were observed.
+Piece-wise linear regression was then used to detect a break point where a significant drop in variable importance were observed.
 Potential predictors with variable importance above this break point were considered important and kept as model candidates.
@@
 Covariates that were selected each of the ten times were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables.
-We will call this modell "BRL all", where BRL stands for bootstrap ranking LASSO.
+We will call this model "BRL all", where BRL stands for bootstrap ranking LASSO.
 A similar model, including any variable selected at least one out of ten times will be called "BRL any".
-Univariable models with the ASA score, CCI and ECI were used for comparion, as well as a multivariable model with age and sex.
+Univariable models with the ASA score, CCI and ECI were used for comparison, as well as a multivariable model with age and sex.
 Each model including age where fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.
@@
 Each of those models were used to predict the probability of death within 90 days for each patient.
-Sensitivity and specificity were estimated to form reciever operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of predicitve power for each model.
-Models with a lower 95 % confidence limit [@Delong1988] exceeding 0.7, were considered good.
+Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of predicitve power.
+Models with a lower 95 % confidence limit [@Delong1988] above 0.7, were considered good.
 
@@
 We used `r substr(R.version.string, 1, 15)` (R Foundation for Statistical   Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr and pROC.
-We also build an online web calculator using the shiny package.
+We build an online web calculator using the shiny package.
+All R-scripts and necessary software (but no personal data) is available as a live Binder environment (https://mybinder.org/v2/gh/eribul/thamortpred/master?urlpath=rstudio).
+A static archived version is also available at zenodo.org/XXX.
+**Prepared but non-public until paper is accepted/published!**
 
@@
 
-There were `r format(nrow(df), big.mark = ", ")` patients (Figure \@ref(fig:flowchart)), `r paste(range(df$P_Age), collapse = " - ")` years old, whereof `r round(mean(df$P_Sex == "Female") * 100)` % were female.
+We found `r format(nrow(df), big.mark = ", ")` patients (Figure \@ref(fig:flowchart)), `r paste(range(df$P_Age), collapse = " - ")` years old, whereof `r round(mean(df$P_Sex == "Female") * 100)` % were female.
 `r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` patients died within 90 days and no one was censored before that.
-The unadjusted cummulative 90-day survival was `r surv90d`.
+The unadjusted cumulative 90-day survival was `r surv90d`.
 
@@
 Remaining comorbidities are based on the previously described combination of CCI and ECI.
-Comorbidities recorded for at least one patient who survived 90 days, and one who did not, were modelled as potential predictors.")
+Comorbidities recorded for at least one patient who survived 90 days, and one who did not, were modeled as potential predictors.")
 ```
@@
 
-There were no observed differences for two versus three knots in models with age modelled by restricted cubic splines.
+The "BRL any" model included `r coefs_print`. The "BRL all model only included `r coefs_print_all`.
+
+There were no observed differences for two versus three knots in models with age modeled by restricted cubic splines.
 We will therefore only present results for three knots.
@@
 The "BRL any" model had an estimated AUC (95 % CI) of `r get_auc("BRL any")`.
-This model was significantly better than the "BRL all" model with `r get_auc("BRL all")`.
-Univariable models with either ASA, CCI or ECI performed relatively poor.
-The model with age and sex was better but still with an AUC not significantly higher than 0.7 (Table \@ref(tab:auc)).
+The altered version without cancer performed equally good.
+The "BRL any" model was significantly better than the "BRL all" model with AUC `r get_auc("BRL all")`.
+Univariable models with ASA, CCI or ECI performed poorly with AUC lower than, or at least not distinguishable from 0.7.
+The model with age and sex performed better but still with AUC not significantly higher than 0.7 (Table \@ref(tab:auc)).
 
@@
   brlasso_tbl_auc, 
-  caption = "Area Under the Curve (AUC) as a measure of predictive power for the 'BRL any' model compared to 'BRL all', a model including the independent variables included in all of the bootstrao ranking models, a simpler model with age and sex, as well as univariable models with ASA score and the Charlson (CCI) or Elixhauser (ECI) comorbidity indices.
+  caption = "Area Under the Curve (AUC) as a measure of predictive power for the 'BRL any' model compared to 'BRL all', a simpler model with age and sex, as well as univariable models with ASA score and the Charlson (CCI) or Elixhauser (ECI) comorbidity indices.
 Age was included as either a main effect, or in the form of restricted cubic splines (RCS) with three knots.")
@@
 
-ROC curves for some of the models are displayed in Figure \@ref(fig:rocs). The "BRL any" model is superior to all other models, although some of the others are intertwined in some areas of the plot.
+ROC curves for some of the models are displayed in Figure \@ref(fig:rocs). The "BRL any" model is superior to all other models.
 
@@
 
-AUCs and 95 % confidence intervals are illustrated for the same models in \@ref(fig:aucci)).
+AUCs and 95 % confidence intervals are illustrated for the same models in Figure \@ref(fig:aucci)).
 
@@
 
-The ability of model "BRL any" to distinguish patients who died within 90 days is illustrated in Figure \@ref(fig:sep).
-We can see that patients who did die within 90 days (blue color) had, on average, higher predicted probabilities to do so.
+The ability of model "BRL any" to estimate probabilities of death within 90 days is illustrated in Figure \@ref(fig:sep). Patients who died had, on average, higher predicted probabilities to do so.
 
@@
 
-Estimated coefficients and correspondnig odds ratios for the "BRL any" model is presented in Table \@ref(tab:brlany).
+Estimated coefficients and corresponding odds ratios for the "BRL any" model is presented in Table \@ref(tab:brlany).
 
@@
 knitr::kable(
-  rename(brlasso_tbl_coefs, X = X_text),
-  caption = 'Estimated coefficients and odds ratios with 95 % confidence intervals for the "BRL any" model.',
+  select(brlasso_tbl_coefs, term, X = X_math, beta, OR, p),
+  caption = 'Estimated coefficients and odds ratios with 95 % confidence intervals for the "BRL any" model. Notations from the X-column is used in the formula in the disussion section.',
   escape = FALSE
 )
-
 ```
 
+We have provided a web calculator to aid clinical model usage in practice (https://erikbulow.shinyapps.io/thamortpred/).
+
+# Discussion
+
+In this nationwide cohort study we intended to compare the performance of a set of easily accessible data that are routinely collected in daily clinical practice with the complex comorbidity coding algorithms suggested by Charlson and Elixhauser.
+We found that a multivariable main effects logistic regression model with `r coefs_print` was able to make better predictions than either CCI or ECI.
+
 The resulting "BRL any" model can predict the probability of death ($p$) within 90 days for new patients as: `r coefs_form` where $X_i$ indicate independent variables as notated in table \@ref(tab:brlany).
-Note however that this formula is only valid for patients within the observed age range (`r paste(range(df$P_Age), collapse = " - ")` years).
+This formula is valid for patients within the observed age range (`r paste(range(df$P_Age), collapse = " - ")` years).
 
@@
 
-A `r min(df$P_Age)` yer old woman with ASA = 1 and none of the important comorbidities would have a `r get_risk("baseline")` risk to die.
-If she was instead `r quantile(df$P_Age, .25)` year old (the first age quantlie), her risk would have increased to `r get_risk("healthy_female")`.
-An older and sicker man, `r quantile(df$P_Age, .75)` years old (the third age quantile) with ASA = 3 and a previsous heart condition would have a risk of `r get_risk("sicker_man")`.
+For example a `r min(df$P_Age)` yer old woman with ASA = 1 and none of the important comorbidities would only have a `r get_risk("baseline")` risk to die within 90 days after elective THA surgery.
+Another woman, `r quantile(df$P_Age, .25)` years old (the first age quantile), have a higher risk of `r get_risk("healthy_female")`.
+A man, `r quantile(df$P_Age, .75)` years old (the third age quantile) with ASA = 3 and a previous heart condition would have a risk of `r get_risk("sicker_man")` risk.
 The perhaps unrealistic case of a `r max(df$P_Age)` year old man with ASA = 3 and all listed comorbidities would have a theoretical risk as high as `r get_risk("high_risk")`.
-Note however that this extreme case relies on extrapolation and is highly unreliable, since no such person was actually observed in the data.
+Note however that this extreme case relies on extrapolation which is highly unreliable, since no such person was actually observed.
 
-We have provided a web calculater to aid estimation of similar probabilities (https://erikbulow.shinyapps.io/thamortpred/).
+Some covariates in the "BRL any" model were not statistically significant by themselves but were still relevant due to unobserved heterogeneity [@Mood2010].
+Obesity for example is known to be associated with a higher risk of morbidity and all-cause mortality [@Must2000].
+However, previous studies on primary THA cohorts have not indicated a higher risk of mortality in obese patients [@Wallace2014].
+An explanation could be that obese patients selected for THA are comparably healthy and often younger.
 
-# Discussion
+Cancer could also be dropped as a predictor without any loss of predictive power.
+It was also encouraging that socio-demographic factors such as education and civil status, or organizational factors such as type of hospital, did not have a strong enough influence of death to be included in the final model.
 
-We found that a multivariable main effects logistic regression model with `r coefs_print` was able to predict death within 90 days after insertion of elective THA.
-Some covariates ni the model were not statistically significant by themselves but were still relevant due to unobserved heteregeneity [@Mood2010].
-
-It is well known that male sex is associated with earlier deaths.
-It is also well known that the remaining life span will decrease with incresed age.
-It is however not obvious that this relation must be striclty linear.
-We therefore allowed for a more flexible relation based on restricted cubic splines, but found that a linear relationship was equally good.
-
-ASA class is routinely assessed pre-operatively in most developed countries.
+It is known that male sex is associated with earlier deaths and that the remaining life span will decrease with increased age.
+It is less obvious that this relation must be linear.
+We used restricted cubic splines to allow a more flexible relation, but found that a linear relationship was equally good.
+Our model includes ASA class which is routinely assessed pre-operatively in most developed countries.
 It is however known to have a high degree of internal variability [@Haynes1995].
 It has previously been compared to the CCI, but not with respect to mortality after THA [@Whitmore2014; @Kork2015].
+Patients with ASA 4-6 were excluded since those categories describe severe disease, moribund and brain-dead individuals.
+It can be questioned whether such classification is correct for our cohort.
+Comorbidities are also known to influence the outcome after THA [@Inacio2015; @Gordon2013; @Hofstede2016].
+Coding algorithms on the other hand are complex and not used in clinical settings since the administrative burden is too high.
+CCI comprise 1,178 ICD-10 codes and ECI 1,516.
+Therefore, such indices are only used by researchers.
 
-# Further discussion from Garland not yet adjusted to new models
-
-A set of easily accessible data is a better predictor of early mortality after major surgery than complex coding algorithms 
-In this nationwide cohort study we intended to compare the performance of a set of easily accessible data that are routinely collected in daily clinical practice with complex comorbidity coding algorithms (ie CCI, Elixhauser Score and RxRiskV).
-The best predictive strength was found for a relatively simple model including age, gender, presence of cardiac infarction or renal disease during the last 12 months prior to THA surgery, and ASA grade (c=0.81).
-Thios simple model was also better than the above mentioned comorbidity measures at predicting  one-year mortality (Supplementary Figure 1 and Supplementary Table 3).
-Comorbidities are known to influence the outcome after THA [@Inacio2015; @Gordon2013; @Hofstede2016].
-In order to assess the effect of comorbidity on early mortality after THA different coding algorithms have been proposed in research settings.
-The coding algorithms are complex, and hence they demand a merge of information on ICD-ocdes and medication prescriptions from several data sources.
-These coding algorithms are not used in clinical settings since the administrative burden associated with identifying some 30 ICD- or ATC-codes for every patient is not realistic.
-Thus, comorbidity measures based on patient administrative data are only accessible to researchers, but - even then -observational study designs are hampered by the usual limitations such as incompleteness and inaccuracy of coding [@Bozic2013].
-In this present study we found that the prescription-based RxRiskV Score performed better than the diagnosis-based comorbidity comorbidity measures CCI and Elixhauser Score in predicting 90-day mortality.
-The original CCI was somewhat better than the Elixhauser Score in predicting 90-day mortality.
-This differs from earlier findings by Inacio et al.
-where the RxRiskV did not perform as well as the CCI and the Elixhauser Score, and where the c- statistics were generally higher than in our study [@Inacio2016].
-Such dissimilarities could be explained by the facts that our study population was younger, included more women, and that we only included diagnoses and prescriptions registered one year prior to surgery.
-Overall, the predictive strength of all investigated diagnose- or prescription-based comorbidity measures was better than the included dimensions investigated separately.
-To put it differently, in terms of predicting mortality, each comorbidity measure was an improvement over the separate items included in each measure.
-We also found that the ASA classification was better at predicting both 90-day and one-year mortality than the more complex coding algorithms, with a c- statistic of 0.70.
-The ASA classification has been repeatedly compared to the CCI, but no consensus as to which one is superior has been reached, and, to our knowledge, such comparisons have not been performed on THA population [@Whitmore2014; @Kork2015].
-Individuals with an ASA grade of 4 to 6 were excluded from our study since those categories describe severe disease, moribund and brain-dead individuals, and it can be questioned whether the classification is correct, and - if it is - whether these patients should ever have received a THA.
-We thus excluded this very small group.
-Obesity is generally known to be associated with a higher risk of morbidity and all-cause mortality [@Must2000].
-However, previous studies on primary THA cohorts have not indicated a higher risk of mortality in obese patients, a result that is confirmed in our study (Supplementary Table 1) [@Wallace2014].
-An explanation could be that obese patients selected for THA are comparably healthy.
-
-Risk prediction may be useful in the process of patient selection prior to surgery, in the preoperative risk management including a review of current medications, and in perioperative anaesthesia management.
+Risk prediction may be useful in the process of patient selection prior to surgery, in the preoperative risk management including a review of current medications, and in perioperative anesthesia management.
 A number of risk prediction tools of various complexity for adverse outcomes after total joint replacements have been introduced but none has been broadly accepted [@Manning2016].
-In the context of trauma surgery outcome prediction tools are common, and it is seems possible to reduce the number of items without losing predictive power [@Gerdin2016].
+In the context of trauma surgery outcome prediction tools are common, and it seems possible to reduce the number of items without losing predictive power [@Gerdin2016].
 Our results indicate that the risk of early postoperative mortality after THA could be assessed by a relatively simple prediction model.
 
-A strength of this study is its nationwide design with a large cohort of primary THA patients with a reasonable number of events, rendering estimation relatively precise.
-Our sources of data are highly valid, and the proportion of missing data in our cohort was low [@Soderman2000; @shpr2014; @Soderman2001; @Ludvigsson2011].
-Limitations to this study are the potential biases at different levels that are commonly associated with observational data, and the risk of coding errors as expected when dealing with patient administrative data.
-Selection bias is also an issue in this study, since patients who died on the table during attempted THA surgery may not have been reported to the SHAR, but, judging from clinical experience, such events are extremely uncommon in a population of patients scheduled for elective THA surgery for osteoarthritis.
-The much higher expected frequency of fatal events during THA surgery on hip fracture patients was one of the main reasons to exclude this group from the present analyses.
-It is important to distinguish between explanatory observational research and attempts at predicting individual events such as early mortality after surgical interventions.
-The combination of parameters in the best-performing model described in the present study may serve as a predictor of mortality on an individual level, but the described combination of parameters would need to be validated in a different sample of individuals.
-Since this has not yet been done we cannot extrapolate our findings to prediction models in a clinical setting, but aim at performing such additional studies
-Our results indicate that in research on mortality after a very common surgical intervention a less complex comorbidity measure consisting of easily accessible data that are routinely collected in daily clinical practice is superior to some of the commonly used diagnose- or prescription-based coding algorithms.
-It would be interesting to evaluate the ability of our novel set of parameters to predict adverse events and revision rates.
+A strength of this study is the nationwide design with a large cohort of primary THA patients.
+We were able to use exact data linkage by the Swedish unique identity numbers and had no censoring.
+Our data sources are valid with low proportions of missing data [@Soderman2000; @shpr2018; @Soderman2001; @Ludvigsson2011].
 
-Our results derived from a nationwide cohort study indicate that a less data demanding comorbidity measure, the combination of age, gender, presence of heart infarction or renal disease and ASA grade, serves is better at predicting early postoperative mortality after THA than comorbidity measures based on more complex coding algorithms.
+The risk of coding errors might be a limitation to the study, especially so if coding routines might change over time.
+It should also be remembered that the risk model does not study THA as an observed intervention.
+We merely followed the cohort who did already have THA.
+Hence, deaths within 90 days might occur for the patients regardless if THA is inserted or not.
+The proximity in dates however, the maximum of 90 days from THA to death, is an indication that the operation might have influenced the deaths observed.
+The insertion of an elective THA is always preceded by a clinical judgement.
+Hence, no patient with a foreseen death near-by is given THA to begin with.
+We therefore believe that at least a non-significant proportion of deaths within 90 days are related to the THA surgery, or with complications thereafter.
+
+We hope that the supplied web calculator and the transparent reporting of this model might lead to clinical usage that can be part of a pre-surgery discussion between doctors and patients in need of THA.
 
@@
 AG and EB drafted the manuscript.
-All athours edited and finalized the manuscript.
+All authors edited and finalized the manuscript.
 

### 2019-10-03 src/overall_surv.R

@@ -8,7 +8,7 @@ fit <- survfit(Surv(stime, status) ~ 1, df)
 surv90d <-
   with(
     summary(fit, 90),
-    sprintf("%.2f (95 %% CI: %.2f - %.2f)",
+    sprintf("%.1f %% (95 %% CI: %.1f - %.1f)",
     surv * 100, lower * 100, upper * 100)
   )
 cache("surv90d")

### 2019-10-03 src/tab_define_groups.R

@@ -5,7 +5,7 @@ tab_categorization <-
   categorization.Blad1 %>%
   mutate_all(zoo::na.locf) %>%
   group_by(new, from) %>%
-  summarise(old = paste(old, collapse = "\n ")) %>%
+  summarise(old = paste(old, collapse = ", ")) %>%
   pivot_wider(names_from = from, values_from = old)
 
 cache("tab_categorization")

### 2019-10-03 diagnostics/validation_calibration.Rmd
kikat på intern validering och kalibrering

---
title: "Intern validering och kalibrering"
author: "Erik Bulow"
date: '`r Sys.Date()`'
output: 
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
load("../cache/all_models.RData")
```

# Optimism

Mha `rms::validate` gavs mått på optimism gm Bootstrapnig (n = 40 per default, lite lågt kanske men hniner inte köra fler just nu).

# Validering

Subtraheras optimismen från givet AUC ges AUC_est.corr enl följande:

```{r}
all_models %>% 
  select(Model, AUC_est, optimism, AUC_est.corr)
```

Görs samma modifikation på CI-gränserna (DeLong) ges:

```{r}
load("../cache/brlasso_tbl_auc.RData")
brlasso_tbl_auc
```

Här borde man kanske egentligen göra annorlunda (ta kvantiler direkt från Bootstrap-samlingen)? Diskuteras på nätet men tycks inte vara implementerat i `rms` utan kanske i så fall kräver egen programmering.

Grafiskt:

```{r}
knitr::include_graphics("../graphs/brlasso_auc_ci.png")
```

# Kalibrering

`rms::calibrate` ger följande graf. Här framgår att bias-korrektion nite ger så stor effekt i förhållande tlil den totala överskattningen.
Obsrevera att x-skalorna skiljer sig mellan olika modeller, dvs kurvorna ligger närmare för t ex "BRL all" men endast i intervallet 0-0.06.

OBS! Osäker om modellerna med splines (RCS) stämmer då `rms` kodar splines på eget sätt egentligen (vilket jag inte gjorde här).

```{r}
knitr::include_graphics("../graphs/calibration.png")
```

Då jag har lite svårt att avgöra vilkemn avvikelse som kan antas OK kikade jag även på "Calibrarion belts".
Se https://onlinelibrary.wiley.com/doi/full/10.1002/sim.6744

Dessa tar ej hänsyn till optimismen men enl ovan var ju den effekten ganska marginell ändå.
Här tror jag att även RCS-modellerna stämmer. 
Banden blir ju ganska breda ganska snabbt men t ex "BRL any" och "BRL any-cancer" avviker åtm inte med stat sign från den röda linjen (vilket är vad vi önskar).

```{r}
knitr::include_graphics("../graphs/calibration_belt.png")
```

### 2019-10-03 munge/07-compare_models.R

@@ -8,14 +8,19 @@ fns3 <- function(coefs) {
 }
 
 
-glmdf       <- function(...) glm(..., data = df, family = binomial())
+glmdf <- function(f){
+  list(
+    glm = glm(f, data = df, family = binomial()),
+    lrm = rms::lrm(as.formula(f), data = df, x = TRUE, y = TRUE)
+  )
+}
 
 all_models_tmp <-
   tribble(
     ~Model,              ~tab, ~fig, ~fit,
     "BRL all",           TRUE,  FALSE, glmdf(form(best_coefs_all)),
     "BRL any",           TRUE,  TRUE,  glmdf(form(best_coefs_any)),
-    "BRL any-cancer",    TRUE,  FALSE,  glmdf(form(setdiff(best_coefs_any, "c_cancer_TRUE."))),
+    "BRL any-cancer",    TRUE,  FALSE, glmdf(form(setdiff(best_coefs_any, "c_cancer_TRUE."))),
     "BRL any (RCS)",     TRUE,  FALSE, glmdf(fns3(best_coefs_any)),
     "BRL 75%",           FALSE, FALSE, glmdf(form(best_coefs_75)),
     "BRL 75% (RCS)",     FALSE, FALSE, glmdf(fns3(best_coefs_75)),
@@ -24,28 +29,49 @@ all_models_tmp <-
     "ASA",               TRUE,  TRUE,  glmdf(death90f ~ P_ASA),
     "Age and sex",       TRUE,  TRUE,  glmdf(death90f ~ P_Age + P_Gender),
     "Age and sex (RCS)", TRUE,  FALSE, glmdf(death90f ~ splines::ns(P_Age, 3) + P_Gender)
-  )
+  ) %>%
+  mutate(
+    glm = map(fit, pluck, "glm"),
+    lrm = map(fit, pluck, "lrm")
+  ) %>%
+  select(-fit)
 
 
 all_models <-
   all_models_tmp %>%
   mutate(
-    tidy      = map(fit, broom::tidy, conf.int = TRUE, exponentiate = TRUE),
-    AIC       = map_dbl(fit, AIC),
-    pred      = map(fit, predict, type = "response"),
+    tidy      = map(glm, broom::tidy, conf.int = TRUE, exponentiate = TRUE),
+    AIC       = map_dbl(glm, AIC),
+    pred      = map(glm, predict, type = "response"),
     obspred   = map(pred, ~ tibble(pred = ., obs = df$death90f)),
[…]

### 2019-10-03 src/calibration.R

@@ -1,0 +1,21 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+# Simple plots from the rms-package
+png("graphs/calibration.png", 1024, 1024)
+  par(mfrow = c(3, 4))
+  for (i in seq_len(nrow(all_models)))
+    plot(all_models$lrm_calibrate[[i]], main = all_models$Model[[i]])
+dev.off()
+
+# Calibration belts with table and p-value from givitiR
+png("graphs/calibration_belt.png", 1024, 1024)
+  par(mfrow = c(3, 4))
+  for (i in seq_len(nrow(all_models)))
+    plot(
+      all_models$cal_belt[[i]],
+      main = all_models$Model[[i]],
+      xlim = c(0, .15),
+      ylim = c(0, .2)
+    )
+dev.off()
+

### 2019-10-03 src/fig_auc.R

@@ -6,13 +6,23 @@ options(digits = 2)
 fig_auc_ci <-
   all_models %>%
   filter(fig) %>%
-  mutate(Model = fct_reorder(Model, AUC_est)) %>%
-  ggplot(aes(Model, AUC_est)) +
+  select(Model, starts_with("AUC")) %>%
+  pivot_longer(-Model) %>%
+  separate(name, c("AUC","level", "corr")) %>%
+  mutate(corr = ifelse(is.na(corr), "Estimated", "Adjusted for optimism")) %>%
+  pivot_wider(names_from = level) %>%
+  mutate(Model = fct_reorder(Model, est)) %>%
+
+  ggplot(aes(Model, est, group = corr, color = corr)) +
   geom_hline(yintercept = .7, color = "darkgreen", linetype = "dashed", size = 2) +
-  geom_pointrange(aes(ymin = AUC_lo, ymax = AUC_hi), size = 1.5) +
+  geom_pointrange(aes(ymin = lo, ymax = hi), size = 1.5, position = position_dodge(.5)) +
   coord_flip() +
   theme_minimal() +
-  theme(axis.title.y = element_blank()) +
+  theme(
+    axis.title.y = element_blank(),
+    legend.title = element_blank(),
+    legend.position = "bottom"
+  ) +
   ylab("AUC with 95 % confidence intervals (De-Long)") +
   scale_y_continuous(breaks = seq(0, 1, .05))
 

### 2019-10-17 nhsr/nhsr.Rmd
new draft sent to NH, AG, SN and OR.

---
title: "Prediction-modelling from wheat to bread"
author: "Erik Bülow"
institute: "The Swedish Hip Arthroplasty Register"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

background-image: url(https://i.pinimg.com/originals/77/ff/d7/77ffd7a8dbccd96cc579b5554bed07ff.jpg)

---
class: inverse, center, middle

# Get Started

---

### 2019-10-17 reports/article.Rmd
new draft sent to NH, AG, SN and OR.

@@
 ---
-title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on Swedish and English register data"
-author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Erik Lenguerrand^5^, Ashley Blom^6^, Mark Wilkinson^7^, Adrian Sayers^8^, Szilard Nemes^2,4^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
+title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational register data from Sweden, England and Wales"
+author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Erik Lenguerrand^5^, Ashley Blom^6^, Mark Wilkinson^7^, Adrian Sayers^8^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
 date: '`r Sys.Date()`'
 bibliography: P:/library.bib
+csl: the-lancet.csl
 output: 
   bookdown::word_document2:
+    reference_docx: article.dotx
     toc: false
@@
 options(
-  digits            = 0,
+  digits            = 2,
   knitr.kable.NA    = '',
@@
 knitr::opts_chunk$set(echo = FALSE)
-
 library(tidyverse)
@@
 walk(file.path("../cache", dir("../cache", ".RData")), load, .GlobalEnv)
+source("../lib/clean_names.R")
 ```
@@
 
-**Correspondence:** 
-AG: anne.l.garland@gmail.com, Phone: +46 498 26 80 00, Postal: Orthopedic Department, Visby Hospital, St Goransgatan 10, 621 84 Visby, Sweden  
-EB: erik.bulow@registercentrum.se  
-SN: szilard.nemes@registercentrum.se  
-NH: nils.hailer@surgsci.uu.se
+**Correspondence: [OR EB?]** 
+anne.garland@surgsci.uu.se, Phone: +46 498 26 80 00, Postal: Orthopedic Department, Visby Hospital, St Goransgatan 10, 621 84 Visby, Sweden
+
+##### PAGE BREAK
 
@@
 
-**OBJECTIVE:** Early mortality after total hip arthroplasty (THA) is associated with comorbidity.
-Diagnosis-based tools such as the Charlson or Elixhauser comorbidity indices have been used to quantify this.
-Those instruments are however complex and requires extensive data sets not routinely available in most doctor-patient settings.
-We searched for a simplified model to  substitute those complex indices in predicting early mortality after THA.
+**OBJECTIVE:** Shared decision making is essential before total hip arthroplasty (THA) surgery.
+Comorbidities are associated with an increased risk of 90-day morbidity and mortality, but diagnosis code-based instruments such as the Charlson or Elixhauser comorbidity indices are impossible to use in a clinical setting, and the American Society for Anesthesiologists (ASA) grade is imprecise.
+We searched for a simple model to predict early mortality after THA.
 
-**PATIENTS AND METHODS:** We studied `r format(nrow(df), big.mark = ",")` patients with THA due to primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` from the Swedish Hip Arthroplasty Register, linked to the national population register, the National Patient Register and the longitudinal integration database for health insurance and labor market studies from Statistics Sweden.
-We used a bootstrap ranking procedure with logistic regression and a LASSO-type penalty to develop a prediction model for patient deaths within 90 days after surgery.
-Predictive power was assessed by the area under the receiver operating characteristic curve (AUC).
-The final model was applied to British data for external validation.
+**PATIENTS AND METHODS:** We studied `r format(nrow(df), big.mark = ",")` Swedish patients operated with a cemented THA due to primary osteoarthritis and linked data from several national registries in order to collect information on demographics and comorbidities.
+Bootstrap ranking procedures were used together with logistic LASSO regression to develop prediction models for death within 90 days.
+Predictive accuracy was assessed by the areas under the receiver operating characteristic curve (AUC).
+The final model was externally validated on a data-set from the National Joint Registry of England and Wales .
 
 ```{r}
-get_auc <- function(mod) subset(brlasso_tbl_auc, Model == mod, "AUC", drop = TRUE)
+get_auc <- function(mod) {
+  subset(brlasso_tbl_auc, Model == mod, "AUC", drop = TRUE) %>% 
+    {gsub("(", "(95% CI: ",., fixed = TRUE)}
+}
 ```
 
-**RESLUTS:** The unadjusted cumulative 90-day survival was `r surv90d`.
-Best predictive performance for 90-day mortality was found for a model combining `r coefs_print` (AUC = `r get_auc("BRL any")`).
-This model was superior to the established but complex Charlson comorbidity index (AUC = `r get_auc("CCI")`), and the Elixhauser comorbidity score (AUC = `r get_auc("ECI")`).
-A web calculator to aid clinical usage was published at https://erikbulow.shinyapps.io/thamortpred/.
+**RESLUTS:** The unadjusted cumulative 90-day mortality was `r surv90d`.
+Best predictive performance for 90-day mortality was found for a model combining `r coefs_print_reduced` (AUC = `r get_auc("BRL reduced (age as main effect)")`).
+This model was superior to the  Charlson (AUC = `r get_auc("Charlson")`) and Elixhauser (AUC = `r get_auc("Elixhauser")`) comorbidity measures.
+A web calculator to aid clinical usage was published at https://erikbulow.shinyapps.io/thamortpred/ **[Would it be possible to embedd it at shpr.se instead?]**.
 
-**CONCLUSION:** We found a relatively simple prediction model of 90 day mortality after total hip arthroplasty.
-This model requires less data and is easier to calculate compared to previously well-known comorbidity indices.
+**CONCLUSION:** We found a relatively simple prediction model of 90-day mortality after THA. This model requires less data and is easier to use compared to previously well-known comorbidity indices.
+
+##### PAGE BREAK
 
@@
 
-The presence of pre-surgery comorbidity is associated with poorer outcome after the insertion of total hip arthroplasty (THA) [@Inacio2015], as well as inferior patient-reported outcomes [@Gordon2013; @Hofstede2016].
-In research settings, comorbidity is commonly measured using multi-faceted diagnosis- or prescription-based coding algorithms [@Bozic2013].
-@Inacio2016 studied the ability of the Charlson (CCI) and Elixhauser (ECI) Comorbidity indices to to predict mortality after THA and total knee arthroplasty.
-These comorbidity measures are quite complex to estimate and are based on the availability of extensive data sets including in- and outpatient data on ICD-codes.
-Oftentimes, such data sets can only be created by linking several population-based registries, raising both ethical and practical concerns.
-Also, each of the comorbidity indices exist in numerous versions [@Sundararajan2004; @Deyo1992; @Quan2011; @Cleves1997; @Walraven2009].
-Interpretation and comparison between different studies is therefore difficult.
+Shared decision making has evolved into an integral part of patient-physician interactions prior to surgical interventions, and the weighing of risks against benefits is central to this process.
+In terms of gained quality of life and cost-utility, total hip arthroplasty (THA) is an enormously successful procedure, but the risk of adverse events in the short and long term is an essential part of preoperative discussions and various risk prediction models have been developed.[@Price2019]
 
-Comorbidity data have also been used in several universal and arthroplasty-specific risk prediction tools to make risk profiles for individual patients.
-In the context of trauma, prediction tools are common, and it has been possible to reduce the number of variables without losing predictive power [@Gerdin2016].
-No model has so far been broadly accepted for elective THA however [@Manning2016; @Bulow2017].
-An easily applicable tool with few dimensions is thus needed, both in research and in clinical practice.
-We aimed to find such a model to predict the risk of 90-day postoperative mortality after elective THA.
+90-day mortality after THA surgery performed due to osteoarthritis is low, ranging between 0.2% and 0.6% in large joint registry analyses where adjustment for comorbidities is possible.[@Hunt2013; @Garland2017]
+
+The presence of comorbidities is associated with a shorter remaining life span, but the Charlson and Elixhauser comorbidity indices poorly predict mortality after THA and total knee arthroplasty.[@Bulow2017; @Inacio2015] Additionally, these complex comorbidity instruments are based on the availability of extensive data-sets including in- and outpatient data coded by the 10th revision of the international classification of diseases (ICD-10), and each of the comorbidity indices exists in numerous versions.[@Brusselaers2017; @Walraven2009] Interpretation and comparison of different studies is therefore difficult.
+
+Comorbidity data are used to construct universal or arthroplasty-specific risk prediction tools, but no model has so far been broadly accepted to predict mortality after  elective THA.[@Manning2016; @Bulow2017] In order to propel arthroplasty surgery into the age of precision medicine, easily applicable tools to predict short- and long-term complications are urgently needed.
+We aimed to define a model that accurately predicts the risk of 90-day postoperative mortality after THA.
 
@@
 
-Patients recorded in the Swedish Hip Arthroplasty Register (SHAR) with cemented primary hip osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` were included in the development phase of the study (Figure \@ref(fig:flowchart)).
-Only the last operated hip was accounted for in patients with bilateral THA [@Bulow2019a].
+Patients recorded in the Swedish Hip Arthroplasty Register (SHAR) with cemented THA due to osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` were used to develop, and internally validate, the prediction model (Figure \@ref(fig:flowchart)).
+Data from the National Joint Register in England and Wales (NJR) was later used for external validation.
+Only the last operated hip was accounted for in patients with bilateral THA. [@Bulow2019a]
 
-```{r flowchart, fig.cap = "Flowchart depicting inclusion criteria and number of patients included in the development phase of the model."}
-knitr::include_graphics("../graphs/flowchart.png")
-```
+Data linkage, based on the 10-digit identity numbers assigned to all Swedish residents,[@Ludvigsson2009] were used to collect data from a variety of sources, as previously described.[@Cnudde2016]
 
-Data linkage, based on the unique identity numbers assigned to all Swedish residents [@Ludvigsson2009], were used to collect data from a variety of sources, as previously described by @Cnudde2016.
+Age, sex, body mass index (BMI), the American Society for Anesthesiologists (ASA) grade, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, which has a completeness of 96-98%.[@shpr2018] Data on education level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) and civil status (married/un-married/divorced/widow[er]), were collected from the longitudinal integration database for health insurance and labor market studies from Statistics Sweden.[@Ludvigsson2019] The Swedish National Patient Register was used to assess comorbidities during the year preceding index surgery.
+This register contains all relevant diagnoses coded by ICD-10, as well as dates of admission and discharge for in- and outpatient episodes in all private and public hospitals.[@Ludvigsson2011] Death dates were linked from the national population register.
 
-Age, sex, body mass index (BMI), ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from SHAR, with a completeness of 96-98 % [@shpr2018].
-Data on education level (low/middle/high) and civil status (married/un-married/divorced/widow[er]), were collected from the longitudinal integration database for health insurance and labor market studies from Statistics Sweden [@Ludvigsson2019].
-The Swedish National Patient Register was used for comorbidity data during one year before surgery.
-The register contains all relevant diagnoses coded by ICD-10, as well as admissions and discharge dates for in- and outpatient visits in all private and public hospitals [@Ludvigsson2011].
-Death dates were linked from the national population register.
+Comorbidity was defined by individual ICD-10 codes grouped into 17 categories according to Charlson [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to Elixhauser.[@Elixhauser1998; @Quan2005]
 
-Comorbidity was recognized by individual ICD-10 grouped into 17 categories according to CCI [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to ECI [@Elixhauser1998; @Quan2005].
-Patients with no recorded hospital visits during one year before surgery, were assumed to have no comorbidity.
-
-```{r tabcategorization}
-knitr::kable(
-  tab_categorization, 
-  caption = "Categorization of individual Charlson (CCI) and Elixhauser (ECI) comorbidities into broader comorbidities.")
-
-```
-
-Some comorbidities were identified by both CCI and ECI, and some distinct comorbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
-We used those categories to establish `r nrow(tab_categorization)` broader categories (Table \@ref(tab:tabcategorization)) in addition to 5 standalone ECI classes that were kept unchanged (`r glue::glue_collapse(ECI_keep_text, ", ", last = " and ")`).
-Groups were merged according to clinical relevance as to be recognized in a patient-doctor meeting without access to external register data.
-Comorbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process described below.
-The final model was also altered to not include cancer as a predictor.
-Patients with cancer are sometimes treated differently in the clinical setting, introducing difficulties in interpretation of this variable.
+Some comorbidities were identified by both Charlson and Elixhauser, and some distinct comorbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
+We therefore combined individual categories to establish `r nrow(tab_categorization)` broader categories of comorbidity (Table \@ref(tab:tabcategorization)).
+Conditions were merged according to clinical relevance as to be recognized in a patient-doctor meeting without access to external register data.
 
@@
 
-We used the Kaplan-Meier estimator to assess unadjusted cumulative survival.
+We used the Kaplan-Meier estimator to assess unadjusted mortality.
 
 Further analysis were based on logistic regression since no censoring occurred within the 90 day study period.
-We used a modelling procedure described by @Guo2015 as a bootstrap ranking procedure with a logistic least absolute shrinkage and selection operator (LASSO) model.
+We used a modelling procedure with bootstrap ranking and logistic least absolute shrinkage and selection operator (LASSO) model.[@Guo2015]
 Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1.
-1,000 bootstrap samples were drawn from the initial data set [@Austin2004].
-We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model.
+
+Comorbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process.
+1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model.
 We then only kept $\lambda$:s minimizing the mean cross-validated deviances in each sample.
@@
 Absolute values from those estimates were used as a measure of variable importance.
-Piece-wise linear regression was then used to detect a break point where a significant drop in variable importance were observed.
+Piece-wise linear regression was  used to detect a break point where a significant drop in variable importance were observed.
 Potential predictors with variable importance above this break point were considered important and kept as model candidates.
-The whole process was repeated ten times.
+The whole process was repeated 100 times.
 
-Covariates that were selected each of the ten times were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables.
-We will call this model "BRL all", where BRL stands for bootstrap ranking LASSO.
-A similar model, including any variable selected at least one out of ten times will be called "BRL any".
-Univariable models with the ASA score, CCI and ECI were used for comparison, as well as a multivariable model with age and sex.
-Each model including age where fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.
+Covariates that were selected at least one out of the 100 times were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables.
+We will call this model "BRL" for bootstrap ranking LASSO.
+A reduced model with variables chosen at least 33 out of the 100 times, is called "BRL reduced".
+This model, but without cancer as a predictor, was also evaluated, since patients with cancer are sometimes treated differently [NEEDS CLARIFICATION!].
+
+Univariable models with the ASA grade, Charlson or Elixhauser comorbidity indices were used for benchmarking, as well as a multivariable model with age and sex.
+Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.
 
 Each of those models were used to predict the probability of death within 90 days for each patient.
-Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of predicitve power.
-Models with a lower 95 % confidence limit [@Delong1988] above 0.7, were considered good.
+Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of predictive accuracy.
+Models with a lower 95% confidence limit above 0.7, were considered good.
+Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples for internal validation.
+The bias-corrected Somers' $D_{xy}$ rank correlation was used to adjust those intervals for optimism based on 100 re-samples.[@Miller1991] Calibration bands were made to graphically assess model calibration.[@Nattino2016]
 
-Odds ratios for the model with highest AUC were estimated with 95 % confidence intervals based on interpolations of profile traces [@Ripley2002].
+Odds ratios for the final model were estimated with 95% confidence intervals based on interpolations of profile traces.[@Ripley2002]
 
-We used `r substr(R.version.string, 1, 15)` (R Foundation for Statistical   Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr and pROC.
-We build an online web calculator using the shiny package.
-All R-scripts and necessary software (but no personal data) is available as a live Binder environment (https://mybinder.org/v2/gh/eribul/thamortpred/master?urlpath=rstudio).
-A static archived version is also available at zenodo.org/XXX.
-**Prepared but non-public until paper is accepted/published!**
+We also built an online web calculator available at https://erikbulow.shinyapps.io/thamortpred/.
+Patients can use it to easily estimate their own risk of mortality.
+This estimate can be used in a discussion of shared decision making on whether to operate or not.
+
+We used `r substr(R.version.string, 1, 15)` (R Foundation for Statistical   Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC and shiny.
+All R-scripts and necessary configurations (but no personal data) is available at zenodo.org/XXX.
+A linked Binder environment is also available for interactive usage.
 
@@
 Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (271-14 and 360-13).
+**[COI?]**
 
@@
 
-We found `r format(nrow(df), big.mark = ", ")` patients (Figure \@ref(fig:flowchart)), `r paste(range(df$P_Age), collapse = " - ")` years old, whereof `r round(mean(df$P_Sex == "Female") * 100)` % were female.
-`r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` patients died within 90 days and no one was censored before that.
-The unadjusted cumulative 90-day survival was `r surv90d`.
+After application of exclusion criteria, `r format(nrow(df), big.mark = ",")` patients (age `r paste(range(df$P_Age), collapse = " - ")`, `r round(mean(df$P_Sex == "Female") * 100)`% females) were included in the analysis (Figure \@ref(fig:flowchart)).
+`r with(df, sprintf("%d (%.2f%%)", sum(death90), mean(death90) * 100))` patients died within 90 days and no one was censored before that.
+The unadjusted risk of 90 day mortality was `r surv90d`.
 
-Some characteristics of the study population are presented in Table \@ref(tab:tab1).
-`r mean(df$CCI_index_quan_original != 0) * 100` % of all patients had at least one pre-surgery comorbidity according to CCI, and `r mean(df$ECI_index_sum_all  != 0) * 100` % according to ECI.
-The proportion of patients with ASA class 3 was `r mean(df$P_ASA == 3) * 100` %.
-Most individual comorbidities were more common among patients who died.
+Characteristics of the study population are presented in Table \@ref(tab:tab1).
+`r round(mean(df$CCI_index_quan_original != 0) * 100)`% of all patients had at least one pre-surgery comorbidity according to Charlson and `r round(mean(df$ECI_index_sum_all  != 0) * 100)`% according to Elixhauser.
+The proportion of patients with ASA grade 3 was `r round(mean(df$P_ASA == 3) * 100)`%.
+Most comorbidities were more common among patients who died.
 
-```{r tab1}
-knitr::kable(
-  table1, 
-  caption = "Baseline demographics.
-CCI/ECI = Charlson/Elixhauser comorbidity indices.
-Comorbidities prefixed with ECI are defined by the Elixhauser classification.
-Remaining comorbidities are based on the previously described combination of CCI and ECI.
-Comorbidities recorded for at least one patient who survived 90 days, and one who did not, were modeled as potential predictors.")
-```
+There were `r nrow(comb_lgl)` comorbidities that were not recorded for any patient who died: `r comb_lgl_text`, and these variables were thus excluded from the modelling process.
 
-There were `r nrow(comb_lgl)` comorbidities that were not recorded for any patient who died: `r comb_lgl_text`. Corresponding variables were excluded from the modelling process.
+The "BRL" model included `r coefs_print`. The "BRL reduced" model only used `r coefs_print_reduced` (Table \@ref(tab:brlprop)).
 
-The "BRL any" model included `r coefs_print`. The "BRL all model only included `r coefs_print_all`.
+We observed no relevant differences between models based on age described by restricted cubic splines with two versus three knots and therefore only present results for models based on three knots.
+The use of splines did however not improve the accuracy of the models when compared to simpler main effect models.
+The "BRL" model had an estimated AUC of `r get_auc("BRL (age as main effect)")`, and the reduced version performed equally good with AUC `r get_auc("BRL reduced (age as main effect)")`.
+Univariable models with ASA grade, Charlson or Elixhauser performed poorly with an AUC that was lower than 0.7.
+A model based on only age and sex performed slightly better but still with an AUC not statistically significantly above 0.7 (Table \@ref(tab:auc) and Figure \@ref(fig:rocs)).
 
-There were no observed differences for two versus three knots in models with age modeled by restricted cubic splines.
-We will therefore only present results for three knots.
-No use of splines however improved the observed AUCs compared to simpler main effect models.
-The "BRL any" model had an estimated AUC (95 % CI) of `r get_auc("BRL any")`.
-The altered version without cancer performed equally good.
-The "BRL any" model was significantly better than the "BRL all" model with AUC `r get_auc("BRL all")`.
-Univariable models with ASA, CCI or ECI performed poorly with AUC lower than, or at least not distinguishable from 0.7.
-The model with age and sex performed better but still with AUC not significantly higher than 0.7 (Table \@ref(tab:auc)).
+AUCs and 95% confidence intervals, with and without adjustments for optimism, are illustrated for some models in Figure \@ref(fig:aucci)).
 
-```{r auc, warning = FALSE}
-knitr::kable(
-  brlasso_tbl_auc, 
-  caption = "Area Under the Curve (AUC) as a measure of predictive power for the 'BRL any' model compared to 'BRL all', a simpler model with age and sex, as well as univariable models with ASA score and the Charlson (CCI) or Elixhauser (ECI) comorbidity indices.
-Age was included as either a main effect, or in the form of restricted cubic splines (RCS) with three knots.")
+The ability of the "BRL reduced" model to estimate probabilities of death within 90 days is illustrated in Figure \@ref(fig:sep).
+Patients who died had, on average, higher predicted probabilities to do so.
+The model calibration was good for probabilities up to 3%, while higher probabilities might be over-estimated (Figure \@ref(fig:calibration)).
 
-```
-
-ROC curves for some of the models are displayed in Figure \@ref(fig:rocs). The "BRL any" model is superior to all other models.
-
-```{r rocs, fig.cap = "ROC curves for some of the models. The 'BRL any' model is distinguished from other models, which are partially over-lapping."}
-knitr::include_graphics("../graphs/brlasso_roc.png")
-```
-
-AUCs and 95 % confidence intervals are illustrated for the same models in Figure \@ref(fig:aucci)).
-
-```{r aucci, fig.cap = "Area Under the Curve (AUC) as a measure of predictive power for the 'BRL any' model compared to a simpler model with age and sex, as well as univariable models with ASA score and the Charlson (CCI) or Elixhauser (ECI) comorbidity indices."}
-
-knitr::include_graphics("../graphs/brlasso_auc_ci.png")
-```
-
-The ability of model "BRL any" to estimate probabilities of death within 90 days is illustrated in Figure \@ref(fig:sep). Patients who died had, on average, higher predicted probabilities to do so.
-
-```{r sep, fig.cap = "Patients who died within 90 days (blue) were, on average, estimated to have a higher probability to do so."}
-knitr::include_graphics("../graphs/brlasso_separation_auc.png")
-```
-
-Estimated coefficients and corresponding odds ratios for the "BRL any" model is presented in Table \@ref(tab:brlany).
-
-```{r brlany}
-
-knitr::kable(
-  select(brlasso_tbl_coefs, term, X = X_math, beta, OR, p),
-  caption = 'Estimated coefficients and odds ratios with 95 % confidence intervals for the "BRL any" model. Notations from the X-column is used in the formula in the disussion section.',
-  escape = FALSE
-)
-```
-
-We have provided a web calculator to aid clinical model usage in practice (https://erikbulow.shinyapps.io/thamortpred/).
+Estimated coefficients and corresponding odds ratios for the "BRL reduced" model is presented in Table \@ref(tab:brlreduced).
 
@@
 
-In this nationwide cohort study we intended to compare the performance of a set of easily accessible data that are routinely collected in daily clinical practice with the complex comorbidity coding algorithms suggested by Charlson and Elixhauser.
-We found that a multivariable main effects logistic regression model with `r coefs_print` was able to make better predictions than either CCI or ECI.
+We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` was able to better predict 90-day mortality than more complex models such as the Charlson or Elixhauser comorbidity indices.
 
-The resulting "BRL any" model can predict the probability of death ($p$) within 90 days for new patients as: `r coefs_form` where $X_i$ indicate independent variables as notated in table \@ref(tab:brlany).
-This formula is valid for patients within the observed age range (`r paste(range(df$P_Age), collapse = " - ")` years).
+The resulting "BRL reduced" model can predict the probability of death within 90 days as: $p = 1 / [1 +\exp(\beta X)]$ where `r coefs_form`
+This formula is valid for patients within the observed age range (`r paste(range(df$P_Age), collapse = " - ")` years) with observed combinations of diseases.
 
 ```{r}
-get_risk <- function(who) preds[preds$id == who, "p", drop = TRUE]
+get_risk <- function(who) people[people$id == who, "p", drop = TRUE]
 ```
 
-For example a `r min(df$P_Age)` yer old woman with ASA = 1 and none of the important comorbidities would only have a `r get_risk("baseline")` risk to die within 90 days after elective THA surgery.
-Another woman, `r quantile(df$P_Age, .25)` years old (the first age quantile), have a higher risk of `r get_risk("healthy_female")`.
-A man, `r quantile(df$P_Age, .75)` years old (the third age quantile) with ASA = 3 and a previous heart condition would have a risk of `r get_risk("sicker_man")` risk.
-The perhaps unrealistic case of a `r max(df$P_Age)` year old man with ASA = 3 and all listed comorbidities would have a theoretical risk as high as `r get_risk("high_risk")`.
-Note however that this extreme case relies on extrapolation which is highly unreliable, since no such person was actually observed.
+For example a `r min(df$P_Age)` year old woman with ASA grade = 1 and none of the important comorbidities would have a `r get_risk("baseline")` risk to die within 90 days of surgery.
+Another woman, `r quantile(df$P_Age, .25)` years old (the first age quantile), have an elevated risk of `r get_risk("healthy_female")`.
+A `r quantile(df$P_Age, .75)` years old man (the third age quantile) with ASA grade = 3 and a previous heart condition would have a risk of `r get_risk("sicker_man")`.
+If that man was instead 99 years old (the maximum observed age), with cancer as well, his risk would be `r get_risk("sickest_man")`.
+Note however that observed covariate patters with predicted probabilities above 3% were rare.
+Only `r mean(pred_probs > 3) * 100`% (`r sum(pred_probs > 3)` patients) had those (Figure \@ref(fig:sep)).
+Estimated mortality risks above 3% are therefore subject to extrapolation.
+Some risk calculators ignore this problem,[@Bozic2013a] but we think this should be acknowledged.
 
-Some covariates in the "BRL any" model were not statistically significant by themselves but were still relevant due to unobserved heterogeneity [@Mood2010].
-Obesity for example is known to be associated with a higher risk of morbidity and all-cause mortality [@Must2000].
-However, previous studies on primary THA cohorts have not indicated a higher risk of mortality in obese patients [@Wallace2014].
-An explanation could be that obese patients selected for THA are comparably healthy and often younger.
+Obesity in the "BRL reduced" model was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but is relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We note that the proportion of patients with BMI above 30, the definition of obisity (`r mean(df$P_BMI > 30, na.rm = TRUE) * 100`%) is much higher than the proportion of patients with ICD-10 = E66* (obesity) from the national patient register (`r mean(df$c_obesity) * 100`%).
+This proportion corresponds to patients with BMI of at least `r which.min(map_dbl(1:50, ~mean(df$P_BMI > ., na.rm = TRUE)) > mean(df$c_obesity))`.
+Thus, obesity might be indicated in the risk calculator only if the condition is severe.
 
-Cancer could also be dropped as a predictor without any loss of predictive power.
-It was also encouraging that socio-demographic factors such as education and civil status, or organizational factors such as type of hospital, did not have a strong enough influence of death to be included in the final model.
+The "BRL reduced" model without cancer had the same AUC as a model with cancer.
+The AUC value itself is a rank-based and scale-invariant measure that does not account for model calibration.
+Both models were however equally well calibrated as well, for estimated probabilities within the recommended range (0-3%).
+The model without cancer however performed worse for probabilities above 3%.
 
@@
 We used restricted cubic splines to allow a more flexible relation, but found that a linear relationship was equally good.
-Our model includes ASA class which is routinely assessed pre-operatively in most developed countries.
-It is however known to have a high degree of internal variability [@Haynes1995].
-It has previously been compared to the CCI, but not with respect to mortality after THA [@Whitmore2014; @Kork2015].
-Patients with ASA 4-6 were excluded since those categories describe severe disease, moribund and brain-dead individuals.
-It can be questioned whether such classification is correct for our cohort.
-Comorbidities are also known to influence the outcome after THA [@Inacio2015; @Gordon2013; @Hofstede2016].
-Coding algorithms on the other hand are complex and not used in clinical settings since the administrative burden is too high.
-CCI comprise 1,178 ICD-10 codes and ECI 1,516.
-Therefore, such indices are only used by researchers.
+Our model includes ASA grade, which is routinely assessed pre-operatively in most developed countries.
+It is however known to have a high degree of internal variability.[@Haynes1995] It has previously been compared to the Charlson comorbidity index, but not with respect to mortality after THA.[@Whitmore2014; @Kork2015] Patients with ASA grade 4-6 were excluded since those categories describe severe disease, moribund and brain-dead individuals.
+Patients with those conditions are unlikely to get elective THA.
+
+Comorbidity burden is also known to influence the outcome after THA.[@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017] Coding algorithms on the other hand are complex and not used in clinical settings since the administrative burden is too high.
+The Charlson classification comprise 1,178 ICD-10 codes and the Elixhauser 1,516.
+Those indices are therefore only used by researchers.
 
 Risk prediction may be useful in the process of patient selection prior to surgery, in the preoperative risk management including a review of current medications, and in perioperative anesthesia management.
-A number of risk prediction tools of various complexity for adverse outcomes after total joint replacements have been introduced but none has been broadly accepted [@Manning2016].
-In the context of trauma surgery outcome prediction tools are common, and it seems possible to reduce the number of items without losing predictive power [@Gerdin2016].
-Our results indicate that the risk of early postoperative mortality after THA could be assessed by a relatively simple prediction model.
+A number of risk prediction tools of various complexity for adverse events after total joint replacements have been introduced but none has been broadly accepted.[@Manning2016; @Price2019] In the context of trauma surgery, outcome prediction tools are common, and it seems possible to reduce the number of items without losing predictive accuracy.[@Gerdin2016] Our results indicate that the risk of early postoperative mortality after THA could be assessed by a relatively simple prediction model.
 
 A strength of this study is the nationwide design with a large cohort of primary THA patients.
-We were able to use exact data linkage by the Swedish unique identity numbers and had no censoring.
-Our data sources are valid with low proportions of missing data [@Soderman2000; @shpr2018; @Soderman2001; @Ludvigsson2011].
+We were able to use exact data linkage by the Swedish identity numbers and had no censoring.
+Our data sources are valid with low proportions of missing data.[@Soderman2000; @shpr2018; @Soderman2001; @Ludvigsson2011]
 
-The risk of coding errors might be a limitation to the study, especially so if coding routines might change over time.
+The risk of coding errors might be a limitation to the study, especially so if coding routines change over time.
 It should also be remembered that the risk model does not study THA as an observed intervention.
@@
 Hence, deaths within 90 days might occur for the patients regardless if THA is inserted or not.
-The proximity in dates however, the maximum of 90 days from THA to death, is an indication that the operation might have influenced the deaths observed.
+The proximity in time however, the maximum of 90 days from THA to death, is an indication that the operation might have influenced the deaths observed.
 The insertion of an elective THA is always preceded by a clinical judgement.
 Hence, no patient with a foreseen death near-by is given THA to begin with.
-We therefore believe that at least a non-significant proportion of deaths within 90 days are related to the THA surgery, or with complications thereafter.
+We therefore believe that at least a non-significant proportion of deaths within 90 days are related to  THA.
 
@@
 
-AG and NH initiated the study and managed the ethical review board application.
-EB, EL and SN performed the statistical analyses.
+**[ADD NJR CONTRIBUTIONS!]** AG and NH initiated the study and managed the ethical review board application.
+EB developed the statistical model.
+EL and AS performed external validation with data from NJR.
 AG and EB drafted the manuscript.
@@
 
+# Acknowledgement
+
+We would like to thank Szilard Nemes, previous senior statistician at the Swedish Hip Arthroplasty Register, for involvement in planning and interpreting the study.
+
+##### PAGE BREAK
+
+```{r, child = "figs_tabs.Rmd"}
+```
+
+##### PAGE BREAK
+
 # Bibliography

### 2019-10-17 reports/figs_tabs.Rmd
new draft sent to NH, AG, SN and OR.

<!-- FIGURES -->

```{r flowchart, fig.cap = "Flowchart depicting inclusion criteria and number of patients included in the development and internal validation phase of the model."}
knitr::include_graphics("../graphs/flowchart.png")
```

##### PAGE BREAK

```{r rocs, fig.cap = "ROC curves for some of the models. The 'BRL' models are superior but the reduced version works equally well as the more complex version."}
knitr::include_graphics("../graphs/brlasso_roc.png")
```

##### PAGE BREAK

```{r aucci, fig.cap = "Area Under the Curve (AUC) as a measure of predictive accuracy. Only the BRL models have AUC statistically significantly above 0.7. The reduced version works as well as the more complex model. To adjust for optimism does not effect the result."}

knitr::include_graphics("../graphs/brlasso_auc_ci.png")
```

##### PAGE BREAK

```{r sep, fig.cap = "Most patients survived more than 90 days after THA, wherefore the blue bars totally dominate the histogram (upper panel). A standardized density plot however reveals that patients who died within 90 days (pink) were, on average, estimated to have a higher probability to do so (lower panel). There were very few observations with covariate patterns resultnig in death probabilities higher than 3%."}
knitr::include_graphics("../graphs/brlasso_separation_auc.png")
```

##### PAGE BREAK

```{r calibration, fig.cap = "Calibration belts at 80% and 95% (light and dark grey). The predicted probabilities do not deviate from the observed ones by any statistical significance. The risk of over-estimation is however clear above 3%."}
knitr::include_graphics("../graphs/calibration_belt.png")
```

##### PAGE BREAK

<!-- TABLES -->

```{r tabcategorization}
knitr::kable(
  tab_categorization, 
  caption = "Categorization of individual Charlson and Elixhauser comorbidities into broader comorbidities.")

```

##### PAGE BREAK

```{r tab1}
knitr::kable(
  table1, 
  caption = "Characteristics of the study population. Comorbidities recorded for at least one patient who survived 90 days, and one who did not, were modeled as potential predictors.")
```

##### PAGE BREAK

```{r brlprop}
t <- 
  brlasso_tbl_selected %>% 
  mutate(variable = clean_names(variable))

knitr::kable(
  t, 
  caption = "Variables selected by the bootstrap ranking procedure, and therefore kept in the 'BRL' model. Variables chosen at least 33 times out of 100 were kept in the 'BRL reduced' model as well. ASA grade 2 and 3 were kept in both models, since those are usually distinguished anyway. It would however be possible to use a model lumping grade 1 and 2 together.")
```

##### PAGE BREAK

```{r auc, warning = FALSE}
knitr::kable(
  brlasso_tbl_auc, 
  caption = "Area Under the Curve (AUC) as a measure of predictive accuracy for the 'BRL' model, the 'BRL reduced', a model with age and sex, as well as univariable models with ASA grade and the Charlson or Elixhauser comorbidity indices. Age was included as either a main effect, or in the form of restricted cubic splines (RCS) with three knots. 'BRL reduced' was also used without cancer and performed equally well in terms of AUC. The correction for optimism slightly decreased the AUC values, but those changes were marginal.")
```

##### PAGE BREAK

```{r brlreduced}

knitr::kable(
  select(brlasso_tbl_coefs, -math),
  caption = 'Estimated coefficients and odds ratios with 95% confidence intervals for the "BRL reduced" model.'
)
```

### 2019-10-17 munge/05-BRLasso.R

@@ -1,10 +1,12 @@
 # Perform calculations (time consuming)
-best_coefs_tmp <-
-  tibble(B = seq_len(config$Bmax)) %>%
-  mutate(
-    coefs_all = future_map(B, ~ BR_lasso_coefs(df_model), .progress = TRUE),
-    breaks    = map_dbl(coefs_all, break_p),
-    coefs     = map2(coefs_all, breaks, ~slice(.x, seq_len(.y)))
-  )
+system.time(
+  best_coefs_tmp <-
+    tibble(B = seq_len(config$Bmax)) %>%
+    mutate(
+      coefs_all = future_map(B, ~ BR_lasso_coefs(df_model), .progress = TRUE),
+      breaks    = map_dbl(coefs_all, break_p),
+      coefs     = map2(coefs_all, breaks, ~slice(.x, seq_len(.y)))
+    )
+)
 
 cache("best_coefs_tmp")

### 2019-10-17 shiny/clean_names_copy.R

@@ -1,0 +1,28 @@
+clean_names <- function(x, firstupper = TRUE, lvls = TRUE) {
+  x <-
+    gsub("(ECI|[cP])_|_?TRUE.?|_X|TypeOf", "", x) %>%
+    tolower() %>%
+    {gsub("_", " ", .)} %>%
+    {gsub("asa2", "ASA grade = 2", .)} %>%
+    {gsub("asa3", "ASA grade = 3", .)} %>%
+    {gsub("gender ?man", "male sex", .)}
+
+  if (!lvls) {
+    x <- unique(gsub("ASA grade = [23]", "ASA grade", x))
+    x <- gsub("male sex", "sex", x)
+  }
+  if (firstupper) {
+    x <- paste0(toupper(substr(x, 1, 1)), substring(x, 2))
+  }
+
+  x <-
+    ifelse(
+      grepl("bmi|asa|eci|cci", x, ignore.case = TRUE),
+      toupper(x),
+      x
+    )
+
+  x <- ifelse(grepl("aids", x, ignore.case = TRUE), "AIDS/HIV", x)
+  x <- ifelse(grepl("cns", x, ignore.case = TRUE), "CNS disease", x)
+  gsub("GRADE", "grade", x)
+}

### 2019-10-17 shiny/misc.R

@@ -1,14 +1,18 @@
-# file.copy("cache/fit_brl_any_lean.RData", "shiny/fit_brl_any_lean.RData", TRUE)
-load("fit_brl_any_lean.RData")
+# file.copy("cache/fit_brl_reduced_lean.RData", "shiny/fit_brl_reduced_lean.RData", TRUE)
+# file.copy("lib/clean_names.R", "shiny/clean_names_copy.R", TRUE)
+load("fit_brl_reduced_lean.RData") # Absolute path correct for deployment
+source("clean_names_copy.R")
 
 coefs <-
-  tibble(coef = names(coef(fit_brl_any_lean))[-1]) %>%
+  tibble(coef = names(coef(fit_brl_reduced_lean))[-1]) %>%
   mutate(
     checkbox     = grepl("TRUE", coef),
     coef_name    = gsub("(TRUE|Man|[23])$", "", coef),
-    coef_present = gsub("[cP]_|ECI_", "", coef_name),
-    coef_present = gsub("_", " ", coef_present),
-    coef_present = gsub("Gender", "Male sex", coef_present)
+    coef_present =
+      clean_names(coef) %>%
+      {gsub(" = [23]", "", .)} %>%
+      {gsub("Obesity", "(Severe) obesity", .)}
+
   ) %>%
   select(-coef) %>%
-  distinct(.keep_all = TRUE)
+  distinct(coef_name, .keep_all = TRUE)

### 2019-10-17 src/calibration.R

@@ -1,21 +1,61 @@
 suppressMessages({library(ProjectTemplate); load.project()})
 
+all_models_fig <- all_models #filter(all_models, fig)
+
 # Simple plots from the rms-package
-png("graphs/calibration.png", 1024, 1024)
+png("graphs/tmp_calibration.png", 1024, 1024)
   par(mfrow = c(3, 4))
-  for (i in seq_len(nrow(all_models)))
-    plot(all_models$lrm_calibrate[[i]], main = all_models$Model[[i]])
+  for (i in seq_len(nrow(all_models_fig))) {
+    plot(
+      all_models_fig$lrm_calibrate[[i]],
+        main = all_models_fig$Model[[i]],
+        # xlim = c(0, 0.03),
+        # ylim = c(0, 0.03)
+    )
+  }
 dev.off()
 
 # Calibration belts with table and p-value from givitiR
-png("graphs/calibration_belt.png", 1024, 1024)
+png("graphs/tmp_calibration_belt.png", 1024, 1024)
   par(mfrow = c(3, 4))
-  for (i in seq_len(nrow(all_models)))
+  for (i in seq_len(nrow(all_models_fig))) {
     plot(
-      all_models$cal_belt[[i]],
-      main = all_models$Model[[i]],
-      xlim = c(0, .15),
-      ylim = c(0, .2)
+      all_models_fig$cal_belt[[i]],
+      main = all_models_fig$Model[[i]],
+      xlim = c(0, .1),
+      ylim = c(0, .1)
     )
+  }
+dev.off()
+
+
+
+# BRL 33 % model ----------------------------------------------------------
+
+tcks <- seq(.0, .1, .01)
+
+png("graphs/calibration_belt.png", 1024, 1024, pointsize = 36)
+  all_models %>%
+    filter(Model == "BRL reduced (age as main effect)") %>% {
+      plot(
+        pluck(.$cal_belt, 1),
+        xlim = c(0, .06),
+        ylim = c(0, .08),
+        xlab = "Predicted probabilities [%]",
+        ylab = "Observed  probabilities [%]",
+        main = NULL,
+        table = FALSE,
+        polynomialString = FALSE,
+        pvalueString = FALSE,
+        nString = FALSE,
+        mar = c(5, 4, 0, 0) + 0.1,
+        xaxt = "n",
+        yaxt = "n",
+
+      )
[…]

### 2019-10-17 src/fig_separation.R

@@ -1,35 +1,91 @@
 suppressMessages({library(ProjectTemplate); load.project()})
 
-# Only for models with AUC CI lower > 0.7!
-
-fig_separation <-
+obspred <-
   all_models %>%
   select(Model, fig, obspred, AUC_lo) %>%
   unnest(obspred) %>%
-  filter(fig, AUC_lo > .7) %>%
-  rename("Observed survival" = obs) %>%
+  filter(Model == "BRL reduced (age as main effect)") %>%
+  mutate(
+    "Observed survival" =
+      factor(obs, c("dead", "alive"),
+                  c("Died within 90 days", "Survived at least 90 days")
+      )
+  )
+
+pred_probs <- obspred$pred * 100
+cache("pred_probs")
+
+
+# Histogram ---------------------------------------------------------------
+
+fig_separation_hist <-
+  obspred %>%
   ggplot(aes(pred)) +
   geom_histogram(
-    aes(y = ..density.., fill = `Observed survival`),
+    aes(
+      # y = ..density..,
+      fill = `Observed survival`
+    ),
     alpha = .3, position = "identity"
   ) +
   geom_density(aes(col = `Observed survival`)) +
-  # facet_wrap(
-  #   ~ Model,
-  #   #scales = "free",
-  #   labeller = function(l) label_value(l, multi_line = FALSE)
-  # ) +
+  theme_minimal() +
+  theme(
+    legend.position = "none",
+    panel.grid = element_blank(),
+    axis.text.x = element_text(angle = 30),
+    axis.ticks = element_line(size = 1)
+  ) +
+  xlab("") +
+  ylab("Count") +
+  # scale_y_continuous(breaks = NULL) +
+  scale_x_log10(
+    breaks = c(.01, .1, 1, 3, 6, 10) / 100,
+    limits = c(.0001, .1),
+    labels = scales::percent
+  ) +
+  scale_y_continuous(
+    breaks = seq(0, 6000, 1000),
+    labels = function(x) format(x, big.mark = ",")
+  ) +
+  expand_limits(x = 0)
+
+
+# Density -----------------------------------------------------------------
+
+fig_separation_density <-
+  obspred %>%
+  ggplot(aes(pred)) +
+  geom_histogram(
+    aes(
+      y = ..density..,
+      fill = `Observed survival`
+    ),
+    alpha = .3, position = "identity"
+  ) +
+  geom_density(aes(col = `Observed survival`)) +
   theme_minimal() +
[…]

### 2019-10-17 src/overall_surv.R

@@ -8,13 +8,11 @@ fit <- survfit(Surv(stime, status) ~ 1, df)
 surv90d <-
   with(
     summary(fit, 90),
-    sprintf("%.1f %% (95 %% CI: %.1f - %.1f)",
-    surv * 100, lower * 100, upper * 100)
+    sprintf("%.2f %% (95 %% CI: %.2f - %.2f)",
+    (1 - surv) * 100, (1 - upper) * 100, (1 - lower) * 100)
   )
 cache("surv90d")
 
-
-summary(fit, c(c(30, 90) / 365.241, 1:5), scale = 365.241)
 
 p <-
   survminer::ggsurvplot(

### 2019-10-17 src/tab_auc.R

@@ -7,8 +7,7 @@ brlasso_tbl_auc <-
   transmute(
     Model,
     AUC      = sprintf("%.2f (%.2f-%.2f)", AUC_est, AUC_lo, AUC_hi),
-    AUC.corr = sprintf("%.2f (%.2f-%.2f)", AUC_est.corr, AUC_lo.corr, AUC_hi.corr)
-    # good  = if_else(AUC_lo > .7, "*", "")
+    `AUC corrected for optimism` = sprintf("%.2f (%.2f-%.2f)", AUC_est.corr, AUC_lo.corr, AUC_hi.corr)
   )
 
 cache("brlasso_tbl_auc")

### 2019-10-17 src/tab_coefs.R

@@ -1,23 +1,20 @@
 suppressMessages({library(ProjectTemplate); load.project()})
 
-
 # Table to present --------------------------------------------------------
 
 brlasso_tbl_coefs <-
-  all_models %>%
-  filter(Model == "BRL any") %>%
+  filter(all_models, Model == "BRL reduced (age as main effect)") %>%
   select(tidy) %>%
   pluck(1, 1) %>%
   transmute(
-    term = gsub("[cP]_|TRUE", "", term),
+    term = clean_names(term),
     term = gsub("_", " ", term),
-    X_text = c("", sprintf("X_%d", seq_len(n() - 1))),
-    X_math = sprintf("$%s$", X_text) %>% {gsub("$$", "", ., fixed = TRUE)},
+    math = sprintf("\\text{%s} \\\\ ", term) %>% {gsub("\\text{(intercept)} \\\\ ", "", ., fixed = TRUE)},
     beta = log(estimate),
     OR   = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high),
     p    = pvalue_format()(p.value)
   ) %>%
-  mutate_at(vars(OR, p), ~ if_else(term == "(Intercept)", "", .))
+  mutate_at(vars(OR, p), ~ if_else(term == "(intercept)", "", .))
 
 cache("brlasso_tbl_coefs")
 
@@ -30,25 +27,33 @@ set_first <- function(coefs, x) {
   else coefs
 }
 
-coefs_print <-
-  brlasso_tbl_coefs %>%
-  filter(term != "(Intercept)") %>%
-  select(term) %>%
-  pluck(1) %>%
-  {gsub("GenderMan", "sex", .)} %>%
-  {gsub("ECI ", "", .)} %>%
-  {gsub("Age", "age", .)} %>%
-  {gsub("cns", "CNS", .)} %>%
-  {gsub("ASA[23]", "ASA", .)} %>%
-  unique() %>%
-  set_first("ASA") %>%
-  set_first("sex") %>%
-  set_first("age") %>%
-  glue::glue_collapse(", ", last = " and ")
+coefs_print_string <- function(name) {
+  all_vars <-
+    filter(all_models, Model == !!name) %>%
+    {.$tidy[[1]]$term[-1]}
 
+  c_vars     <- all_vars[startsWith(all_vars, "c_")]
 
+  basic_vars <-
+    setdiff(all_vars, c_vars) %>%
[…]

### 2019-10-17 src/table1.R

@@ -1,16 +1,12 @@
 suppressMessages({library(ProjectTemplate); load.project()})
 
-fctrs <- c("P_Sex", "P_ASA", "ECI", "CCI")
-
 dft1 <-
   df %>%
-  rename(
-    CCI = CCI_index_quan_original,
-    ECI = ECI_index_sum_all
-  ) %>%
   mutate(
-    CCI     = replace(CCI, CCI == 4, "4+"),
-    ECI     = replace(ECI, ECI == 3, "3+")
+    Charlson   = replace(CCI_index_quan_original, CCI_index_quan_original == 4, "4+"),
+    Elixhauser = replace(ECI_index_sum_all, ECI_index_sum_all == 3, "3+"),
+    death90f   = factor(death90f, c("dead", "alive"),
+                        c("Died within 90 days", "Survived at least 90 days"))
   ) %>%
   select(
     death90f,
@@ -21,21 +17,28 @@ dft1 <-
     P_TypeOfHospital,
     education,
     civil_status,
-    CCI, ECI,
+    Charlson,
+    Elixhauser,
     starts_with("c_"),
-    starts_with("ECI_")
-  )
+  ) %>%
+  setNames(clean_names(names(.))) %>%
+  rename(`ASA grade` = ASA)
 
 t1 <-
   tableone::CreateTableOne(
-    strata = "death90f",
-    vars = setdiff(names(dft1), "death90f"),
-    factorVars = fctrs,
+    strata = "Death90f",
+    vars = setdiff(names(dft1), "Death90f"),
+    factorVars = c("Sex", "ASA grade", "Elixhauser", "Charlson"),
     data = dft1,
     test = FALSE
   )
 
 cache("t1")
+
+zero <- function(x) {
+  gsub("0 ( 0.0)", "0", x, fixed = TRUE) %>%
+  {gsub("0.0", "<0.1", .)}
+}
 
 table1 <-
   t1 %>%
@@ -44,11 +47,12 @@ table1 <-
   ) %>%
   as_tibble(rownames = "what") %>%
   mutate(
-    what = gsub("= TRUE|P_|c_|TypeOf|P_Surg|index", "", what),
-    what = gsub("_", " ", what),
-    level = ifelse(startsWith(what, " "), what, ""),
-    what = ifelse(level == "", what, "")
+    level = trimws(ifelse(startsWith(what, " "), what, "")),
+    level = paste0(toupper(substr(level, 1, 1)), substring(level, 2)),
[…]

### 2019-10-22 reports/article.Rmd
last version before asking for external validation

@@
 ---
-title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational register data from Sweden, England and Wales"
+title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
 author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Erik Lenguerrand^5^, Ashley Blom^6^, Mark Wilkinson^7^, Adrian Sayers^8^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
@@
 We also built an online web calculator available at https://erikbulow.shinyapps.io/thamortpred/.
-Patients can use it to easily estimate their own risk of mortality.
-This estimate can be used in a discussion of shared decision making on whether to operate or not.
 
@@
 
-The resulting "BRL reduced" model can predict the probability of death within 90 days as: $p = 1 / [1 +\exp(\beta X)]$ where `r coefs_form`
+The resulting "BRL reduced" model can predict the probability of death within 90 days as: $p = 1 / [1 +\exp(-\beta X)]$ where `r coefs_form`
 This formula is valid for patients within the observed age range (`r paste(range(df$P_Age), collapse = " - ")` years) with observed combinations of diseases.
@@
 Another woman, `r quantile(df$P_Age, .25)` years old (the first age quantile), have an elevated risk of `r get_risk("healthy_female")`.
-A `r quantile(df$P_Age, .75)` years old man (the third age quantile) with ASA grade = 3 and a previous heart condition would have a risk of `r get_risk("sicker_man")`.
-If that man was instead 99 years old (the maximum observed age), with cancer as well, his risk would be `r get_risk("sickest_man")`.
+A `r max(df$P_Age)` years old man (the maximum observed age) with ASA grade = 3 and cancer, would have a risk of `r get_risk("sickest_man")`.
 Note however that observed covariate patters with predicted probabilities above 3% were rare.
@@
 
-Obesity in the "BRL reduced" model was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but is relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We note that the proportion of patients with BMI above 30, the definition of obisity (`r mean(df$P_BMI > 30, na.rm = TRUE) * 100`%) is much higher than the proportion of patients with ICD-10 = E66* (obesity) from the national patient register (`r mean(df$c_obesity) * 100`%).
+Obesity in the "BRL reduced" model was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but is relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We note that the proportion of patients with BMI above 30, the definition of obesity (`r mean(df$P_BMI > 30, na.rm = TRUE) * 100`%) is much higher than the proportion of patients with ICD-10 = E66* (obesity) from the National Patient Register (`r mean(df$c_obesity) * 100`%).
 This proportion corresponds to patients with BMI of at least `r which.min(map_dbl(1:50, ~mean(df$P_BMI > ., na.rm = TRUE)) > mean(df$c_obesity))`.
@@
 It is however known to have a high degree of internal variability.[@Haynes1995] It has previously been compared to the Charlson comorbidity index, but not with respect to mortality after THA.[@Whitmore2014; @Kork2015] Patients with ASA grade 4-6 were excluded since those categories describe severe disease, moribund and brain-dead individuals.
-Patients with those conditions are unlikely to get elective THA.
+We suspect that patients with those ASA grades might be misclassified.
+Their true grade could either be lower, or they could have recieved THA due to a femoral hip fracture instead of osteoarthritis.
 

### 2019-10-22 reports/figs_tabs.Rmd
last version before asking for external validation

@@
 
-```{r sep, fig.cap = "Most patients survived more than 90 days after THA, wherefore the blue bars totally dominate the histogram (upper panel).
-A standardized density plot however reveals that patients who died within 90 days (pink) were, on average, estimated to have a higher probability to do so (lower panel).
+```{r sep, fig.cap = "Most patients survived more than 90 days after THA, wherefore the blue bars totally dominate the histogram (upper panel; note the square-root scale on the y-axis, that was necessary to distinguish the pink bars from the horisontal line).
+A normalized density plot however reveals that patients who died within 90 days (pink) were, on average, estimated to have a higher probability to do so (lower panel).
 There were very few observations with covariate patterns resultnig in death probabilities higher than 3%."}
@@
 
-```{r calibration, fig.cap = "Calibration belts at 80% and 95% (light and dark grey).
-The predicted probabilities do not deviate from the observed ones by any statistical significance.
-The risk of over-estimation is however clear above 3%."}
+```{r calibration, fig.cap = "Calibration belts based on deciles of fitted probabilities and loess smoothers at 80% and 95% confidence levels (light and dark grey).
+The predicted probabilities do not deviate much from the observed ones up to approximately 3%.
+A predicted risk of 6% however could correspond to observed probabilities between 2 and 8% (the width of the 95% belt).
+The higher the predicted probability gets, the more it over-estimates the observed probabilities."}
 knitr::include_graphics("../graphs/calibration_belt.png")

### 2019-10-22 munge/01-filter.R

@@ -59,14 +59,28 @@ filters <-
 
     tibble(
       step = 5,
-      excl = "BMI > 50 or missing",
+      excl = "BMI missing",
+      incl = "Total study population",
+      expr = list(quo(!is.na(P_BMI)))
+    ),
+
+    tibble(
+      step = 5,
+      excl = "BMI > 50",
       incl = "Total study population",
       expr = list(quo(P_BMI <= 50))
     ),
 
     tibble(
       step = 5,
-      excl = "ASA = 4, 5 or missing",
+      excl = "ASA missing",
+      incl = "Total study population",
+      expr = list(quo(!is.na(P_ASA)))
+    ),
+
+    tibble(
+      step = 5,
+      excl = "ASA > 3",
       incl = "Total study population",
       expr = list(quo(P_ASA <= 3))
     ),

### 2019-10-22 munge/07-compare_models.R

@@ -9,10 +9,18 @@ fns3 <- function(coefs) {
 
 
 glmdf <- function(f){
-  list(
-    glm = glm(f, data = df, family = binomial()),
-    lrm = rms::lrm(as.formula(f), data = df, x = TRUE, y = TRUE)
-  )
+  glm <-  glm(f, data = df, family = binomial())
+
+  # I tried to use LASSO to avoid over-estimation but found that best penalty-terms were virtually 0 for all models.
+  # I describe this as a sensitivity analysis, but I will not use it any further.
+  lambda <- purrr::possibly(glmnet::cv.glmnet, list(lambda.min = 0))(
+    model.matrix(as.formula(f), df)[, -1, drop = FALSE],
+    df$death90f,
+    family = "binomial"
+  )$lambda.min
+  lrm <-  rms::lrm(as.formula(f), data = df, x = TRUE, y = TRUE, penalty = lambda)
+
+  list(glm = glm, lrm = lrm)
 }
 
 all_models_tmp <-
@@ -23,11 +31,11 @@ all_models_tmp <-
     "BRL reduced without cancer (age as main effect)", TRUE,  FALSE, glmdf(form(setdiff(best_coefs_reduced, "c_cancer_TRUE."))),
     "BRL (age as main effect)",                        TRUE,  TRUE,  glmdf(form(best_coefs)),
     "BRL (age as RCS)",                                TRUE,  FALSE, glmdf(fns3(best_coefs)),
-    "Charlson",                                        TRUE,  TRUE,  glmdf(death90f ~ CCI_index_quan_original),
-    "Elixhauser",                                      TRUE,  TRUE,  glmdf(death90f ~ ECI_index_sum_all),
-    "ASA",                                             TRUE,  TRUE,  glmdf(death90f ~ P_ASA),
-    "Age and sex (age as main effect)",                TRUE,  TRUE,  glmdf(death90f ~ P_Age + P_Gender),
-    "Age and sex (age as RCS)",                        TRUE,  FALSE, glmdf(death90f ~ splines::ns(P_Age, 3) + P_Gender)
+    "Charlson",                                        TRUE,  TRUE,  glmdf("death90f ~ CCI_index_quan_original"),
+    "Elixhauser",                                      TRUE,  TRUE,  glmdf("death90f ~ ECI_index_sum_all"),
[…]

### 2019-10-22 src/fig_separation.R

@@ -16,6 +16,7 @@ pred_probs <- obspred$pred * 100
 cache("pred_probs")
 
 
+
 # Histogram ---------------------------------------------------------------
 
 fig_separation_hist <-
@@ -26,9 +27,8 @@ fig_separation_hist <-
       # y = ..density..,
       fill = `Observed survival`
     ),
-    alpha = .3, position = "identity"
+    alpha = .5, position = "identity"
   ) +
-  geom_density(aes(col = `Observed survival`)) +
   theme_minimal() +
   theme(
     legend.position = "none",
@@ -38,14 +38,13 @@ fig_separation_hist <-
   ) +
   xlab("") +
   ylab("Count") +
-  # scale_y_continuous(breaks = NULL) +
   scale_x_log10(
     breaks = c(.01, .1, 1, 3, 6, 10) / 100,
     limits = c(.0001, .1),
     labels = scales::percent
   ) +
-  scale_y_continuous(
-    breaks = seq(0, 6000, 1000),
+  scale_y_sqrt(
+    breaks = c(0, 100, 500, 1000, 2000, 3000, 4000, 5000), #seq(0, 6000, 1000),
     labels = function(x) format(x, big.mark = ",")
   ) +
   expand_limits(x = 0)
@@ -61,7 +60,7 @@ fig_separation_density <-
       y = ..density..,
       fill = `Observed survival`
     ),
-    alpha = .3, position = "identity"
+    alpha = .5, position = "identity"
   ) +
   geom_density(aes(col = `Observed survival`)) +
   theme_minimal() +
@@ -70,11 +69,12 @@ fig_separation_density <-
     legend.title = element_blank(),
     panel.grid = element_blank(),
     axis.text.x = element_text(angle = 30),
-    axis.ticks = element_line(size = 1)
+    axis.ticks = element_line(size = 1),
+    axis.ticks.y = element_line(color = "white"),
+    axis.text.y = element_text(color = "white", size = 17)     # To align with upper panel
   ) +
   xlab("Predicted probability of death (log-scale)") +
   ylab("Density") +
-  scale_y_continuous(breaks = NULL) +
   scale_x_log10(
     breaks = c(.01, .1, 1, 3, 6, 10) / 100,
     limits = c(.0001, .1),

### 2019-10-22 src/tab_coefs.R

@@ -12,7 +12,7 @@ brlasso_tbl_coefs <-
     math = sprintf("\\text{%s} \\\\ ", term) %>% {gsub("\\text{(intercept)} \\\\ ", "", ., fixed = TRUE)},
     beta = log(estimate),
     OR   = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high),
-    p    = pvalue_format()(p.value)
+    p    = pvalue_format(.01)(p.value)
   ) %>%
   mutate_at(vars(OR, p), ~ if_else(term == "(intercept)", "", .))
 

### 2019-11-13 diagnostics/external_validation_template.Rmd
code template for Erik L to use for external validation

---
title: "External validation"
author: "Erik Bulow"
date: '2019-11-13'
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)
```

# Start

Install and attatch some useful packages

```{r}
pkgs <- c("tidyverse", "doParallel", "pROC", "rms", "givitiR")

# Install
pkgsinst <- setdiff(pkgs, rownames(installed.packages()))
if (length(pkgsinst)) install.packages(pkgsinst)

# Attatch
purrr::walk(pkgs, ~suppressPackageStartupMessages(library(., character.only = TRUE)))
```

Load the exported model to validate (the R object previsously sent, probably with another path).

```{r}
# Load the exported model object
load("../cache/fit_export.RData")
```

Set random seed for reproducability:

```{r}
set.seed(123)
```

# Prepare data

## Inclusion/exlusion

Those were the inclusions/exklusions from Sweden. It might not be necessary to filter out on BMI, hospital and education however. Those variables are not used in the model.

Additional filter to ages 35-99 years to match the Swedish cohort.

```{r}
knitr::include_graphics("../graphs/flowchart.png")
```

# Variables

## Outcome

The outcome variable is boolean (or numeric) indicating wether the patient died of any cause within 90 days after THA (`TRUE`/1) or not (`FALSE`/0). We did not have any censoring in Sweden. I guess we can simply drop cases where status is unknown?

## Predictors

The data to evaluate should look like this:

```{r}
head(fit_export$data)
```

## Baseline variables

- **P_Gender**: Kvinna/Man = Female/Male
- **P_ASA**: level 1-3
- **P_Age**: 35 - 99

## Comorbidities

Comorbidities are identified by Elixhauser and Charlson comorbidity during one year before surgery. This is done by ICD-10 codes in Sweden but other versions exist. If ICD-10-codes are used those might be identified by regular expressions in the table below. Those are based on codes like "C123". Hence no puntctuation (not "C12.3"). The regular expressions might be modified to allow possible other patterns. Regular expressoins below are just combined from differnet groups (separated by "|"). They could be rewritten for clarity. If you don't see any regular expressions, click the small arrow in the upper right corner of the table. The `grepl` function (base R) can be used to identify ICD-10-codes based on the regular expressions.

```{r, echo = FALSE}
load("../cache/categorization.RData")
comorbidities <- 
  categorization %>% 
  map_df(as.character) %>% 
  filter(new %in% names(fit_export$data)[grepl("c_", names(fit_export$data))]) %>% 
  separate(old, letters, sep = "\\|", fill = "right") %>% 
  pivot_longer(letters, values_drop_na = TRUE) %>% 
  mutate(value = trimws(value)) %>% 
  separate(value, c("index", "name"), extra = "merge") %>% 
  mutate(name = gsub("_", " ", name))

CCI <- 
  comorbidities %>% 
  filter(index == "CCI") %>% 
  left_join(coder::charlson_icd10, c(name = "group"))

ECI <- 
  comorbidities %>% 
  filter(index == "ECI") %>% 
  left_join(coder::elix_icd10, c(name = "group"))

comorb_defs <- 
  bind_rows(CCI, ECI) %>% 
  select(new, regex) %>% 
  mutate(regex = sprintf("(%s)", gsub("^", "", regex, fixed = TRUE))) %>% 
  group_by(new) %>% 
  summarise(regex = paste(regex, collapse = "|"))

comorb_defs
```

## Example data

Let's assume we now have some data (I will use the Swedish data just as an example).
```{r}
y <- fit_export$y
X <- fit_export$data
```

# Validation of model as is

```{r}
# Tibble with observed and predicted outcome
obspred <- 
  tibble(
    obs  = y, 
    pred = predict(fit_export, X, type = "response")
  )

# ROC curve
ROC <- pROC::roc(obspred, "obs", "pred", direction = "<")

# Estimate CI for AUC based on bootstrapping
# Use parallel processing to speed up the process
doParallel::registerDoParallel()
AUCci <- pROC::ci.auc(ROC, method = "bootstrap", boot.stratified = FALSE, parallel = TRUE)

# Check calibration. Note that devel should actually be "internal" for this example but I use
# "external", since that's what you will use for the UK data. 
calibration <- givitiR::givitiCalibrationBelt(obspred$obs, obspred$pred, devel = "external")
```

## Results

For this example we had AUC:

```{r}
AUCci
plot(ROC)
```

A calibration belt plot might be illustrated as:

```{r}
plot(calibration, xlim = c(0, 0.03), ylim = c(0, 0.06))
```

# Re-calibrated intercept

Method 2 from table 1 in Steyerberg 2004.

```{r}
Z <- predict(fit_export, X, type = "response")

# Refit the intercept using Z = a + Xb from above as offset
fit2 <- glm(y ~ 1, offset = Z)

# Same calibration and validation as above
obspred2     <- tibble(obs  = y, pred = predict(fit2, type = "response"))
ROC2         <- pROC::roc(obspred2, "obs", "pred", direction = "<")
AUCci2       <- pROC::ci.auc(ROC2, method = "bootstrap", boot.stratified = FALSE, parallel = TRUE)
calibration2 <- givitiR::givitiCalibrationBelt(obspred2$obs, obspred2$pred, devel = "external")
```

## Results

```{r}
AUCci2
plot(ROC2)
plot(calibration2, xlim = c(0, 0.03), ylim = c(0, 0.06))
```

# Re-calibration of intercenpt and calibration slope

Method 3 from table 1 in Steyerberg 2004.

```{r}
fit3         <- glm(y ~ 1 + Z)
obspred3     <- tibble(obs  = y, pred = predict(fit3, type = "response"))
ROC3         <- pROC::roc(obspred3, "obs", "pred", direction = "<")
AUCci3       <- pROC::ci.auc(ROC3, method = "bootstrap", boot.stratified = FALSE, parallel = TRUE)
calibration3 <- givitiR::givitiCalibrationBelt(obspred3$obs, obspred3$pred, devel = "external")
```

## Results

```{r}
AUCci3
plot(ROC3)
plot(calibration3, xlim = c(0, 0.03), ylim = c(0, 0.06))
```

### 2019-12-11 reports/article.Rmd
correct affiliations

@@
 title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
-author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Erik Lenguerrand^5^, Ashley Blom^6^, Mark Wilkinson^7^, Adrian Sayers^8^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
+author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Erik Lenguerrand^5^, Ashley Blom^5^, Mark Wilkinson^6^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
 date: '`r Sys.Date()`'
@@
 4.	Department of Orthopaedics, Institute of Clinical Sciences, The Sahlgrenska Academy, University of Gothenburg, Gothenburg, Sweden
-5. Translational Health Sciences, Bristol Medical School, University of Bristol, England.
-6. ...
-7. ...
-8. ...
+5. Translational Health Sciences, Bristol Medical School, University of Bristol, UK.
+6. Department of Oncology and Metabolism, University of Sheffield, Sheffield, UK.
 
@@
 
-**PATIENTS AND METHODS:** We studied `r format(nrow(df), big.mark = ",")` Swedish patients operated with a cemented THA due to primary osteoarthritis and linked data from several national registries in order to collect information on demographics and comorbidities.
+**PATIENTS AND METHODS:** We studied `r format(nrow(df), big.mark = ",")` Swedish patients operated with a cemented THA due to primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")`, linked data from several national registries in order to collect information on demographics and comorbidities.
 Bootstrap ranking procedures were used together with logistic LASSO regression to develop prediction models for death within 90 days.

### 2019-12-11 diagnostics/external_validation_template.Rmd
include R code for data export in validation template to Erik L

@@
 doParallel::registerDoParallel()
-AUCci <- pROC::ci.auc(ROC, method = "bootstrap", boot.stratified = FALSE, parallel = TRUE)
+AUCci <- 
+  pROC::ci.auc(
+    ROC, 
+    method          = "bootstrap", 
+    boot.stratified = FALSE, 
+    parallel        = TRUE
+  )
 
@@
 # "external", since that's what you will use for the UK data. 
-calibration <- givitiR::givitiCalibrationBelt(obspred$obs, obspred$pred, devel = "external")
+calibration <- 
+  givitiR::givitiCalibrationBelt(
+    obspred$obs, 
+    obspred$pred, 
+    devel = "external"
+  )
 ```
@@
 ```
+
+# Export data to Sweden
+
+## ROC
+
+If it would be OK to export coordinates for ROC-plots I would recommend this code to extract only the minimal data needed:
+
+```{r}
+
+roc_plot_coords <- 
+  data.frame(
+    specificities = ROC$specificities, 
+    sensitivities = ROC$sensitivities
+  )
+
+roc3_plot_coords <- 
+  data.frame(
+    specificities = ROC3$specificities, 
+    sensitivities = ROC3$sensitivities
+  )
+
+```
+
+## AUC with CI
+
+The text output from `AUCci` and `AUCci3` should be enough. Hence, the same character string that gets printed above (but now stored in an object).
+```{r}
+AUCci_print  <- capture.output(AUCci)
+AUCci3_print <- capture.output(AUCci3)
+```
+
+## Export objects
+
+Save objects above to file `export.RData` (in the current working directory).
+
+```{r}
+save(
+  roc_plot_coords,
+  roc3_plot_coords,
+  AUCci_print,
+  AUCci3_print,
+  file = "export.RData"
+)
+```
+
+## Calibration plots
+
+### Help function
+
+This is a simple help function to make a clean calibration belt plot and save it as TIFF (in the curent working directory):
+
+```{r}
+makeplot <- function(x, file_name = deparse(substitute(x))) {
+  tiff(
+    paste0(file_name, ".tiff"), 
+    1024, 1024, pointsize = 36, 
+    compression = "lzw"
+  )
+
+  tcks <- seq(.0, .1, .01)
+
+  plot(
+    x,
+    xlim             = c(0, .06),
+    ylim             = c(0, .08),
+    xlab             = "Predicted probabilities [%]",
+    ylab             = "Observed  probabilities [%]",
+    main             = NULL,
+    table            = FALSE,
+    polynomialString = FALSE,
+    pvalueString     = FALSE,
+    nString          = FALSE,
+    mar              = c(5, 4, 0, 0) + 0.1,
+    xaxt             = "n",
+    yaxt             = "n"
+  )
+  abline(v = .03, lty = "dashed", col = "darkgreen", lwd = 3)
+  axis(1, at = tcks, lab = sprintf("%.0f", tcks * 100), las = TRUE)
+  axis(2, at = tcks, lab = sprintf("%.0f", tcks * 100), las = TRUE)
+
+  dev.off()
+}
+```
+### Make and save figures
+
+```{r}
+makeplot(calibration)
+makeplot(calibration3)
+```
+
+## ROC plots
+
+If it not possible to export data for ROC-plot, here is some code to make a figure and save it as TIFF (in the current working directory) instead:
+
+```{r}
+
+roc_plot_coords %>% 
+  ggplot(aes(1 - specificities, sensitivities)) +
+  geom_path(size = 2) +
+  geom_abline(intercept = 0, slope = 1, color = "grey", linetype = 2) +
+  theme_minimal() +
+  theme(
+    legend.position = c(1, 0),
+    legend.justification = c(1, 0),
+    legend.title = element_blank()
+  )
+
+ggsave(
+  "roc.tiff", 
+  height = 10, 
+  width = 10, 
+  unit = "cm", 
+  dpi = 900, 
+  compression = "lzw"
+)
+```

### 2020-01-21 diagnostics/external_validation_summary.Rmd
incoorporated some results from NJR validation

---
title: "Validation"
author: "Erik Bulow"
date: '2020-01-21'
output: 
  word_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
library(tidyverse)
```

# Flowchart

Patients included from NJR. (Many excluded due to fixation.)

```{r}
knitr::include_graphics("../graphs/flowchart_njr.png")
```

# Summary table

## NJR
```{r}
njr_demography_table <- 
  read_delim(
    "../data/njr_demography_table.csv", ";", 
    escape_double = FALSE, comment = "#", trim_ws = TRUE
  ) %>% 
  mutate_all(~gsub("([^[:space:]])(\\()", "\\1 \\2", ., perl = TRUE)) %>% 
  mutate(
    level = gsub("(ASA )([1-3])(.*)", "\\2", What, perl = TRUE),
    What  = gsub("(ASA)( [1-3])(.*)", "\\1", What, perl = TRUE),
    level = if_else(What == "ASA", level, ""),
    What = if_else(duplicated(What), "", What)
  ) %>% 
  select(What, level, everything())

njr_demography_table
```

## To be compared with SHAR:

```{r}
load("../cache/table1.RData")
table1 %>% 
  mutate(
    what = na_if(what, ""),
    what = zoo::na.locf(what)
    ) %>% 
  filter(
    grepl("^n$|Age|Sex|ASA|Cancer|CNS|Kidney", what)
  )
```

# ROC curve

```{r}
knitr::include_graphics("../graphs/brlasso_roc_validation.png")
```

# AUC

```{r}
knitr::include_graphics("../graphs/brlasso_auc_ci.png")
```

# Calibration

Model as-is overestimates probabilityies (as expected) but a re-calibrated version looks like this:

```{r}
knitr::include_graphics("../graphs/calibration_belt_njr.png")
```

To compare with SHAR:

```{r}
knitr::include_graphics("../graphs/calibration_belt.png")
```

### 2020-01-21 munge/01-filter.R

@@ -1,3 +1,7 @@
+# library(tidyverse)
+# library(DiagrammeR)
+# library(DiagrammeRsvg)
+
 
 # Define filter steps -----------------------------------------------------
 
@@ -6,8 +10,8 @@
 #       Does not need to be unique if several exlusions are combined.
 # incl: Description of included cases
 # excl: Description of cases to exclude
-# expr: quosure defining filter step
-# data: The data set to be filtered. Only required for initial step (first row).
+# expr: quosure defining filter step based on variable names from the data
+# data: The dataset to be filtered. Only required for initial step (first row).
 
 filters <-
   bind_rows(

### 2020-01-21 src/fig_roc.R

@@ -16,3 +16,40 @@ fig_roc <-
   )
 
 ggsave("graphs/brlasso_roc.png", fig_roc)
+
+
+
+# SHAR vs NJR -------------------------------------------------------------
+
+shar_roc <-
+  all_models %>%
+  filter(Model == "BRL reduced (age as main effect)") %>%
+  select(roc_curve) %>%
+  pluck(1, 1)
+
+njr_roc <-
+  tibble(
+    sensitivity = njr_ROC$sensitivities,
+    specificity = njr_ROC$specificities
+  )
+
+val_roc <-
+  bind_rows(
+    SHAR = shar_roc,
+    NJR = njr_roc,
+    .id = "source"
+  )
+
+validation_roc <-
+  val_roc %>%
+  ggplot(aes(1 - specificity, sensitivity, col = source)) +
+  geom_path(size = 2) +
+  geom_abline(intercept = 0, slope = 1, color = "grey", linetype = 2) +
+  theme_minimal() +
+  theme(
+    legend.position = c(1, 0),
+    legend.justification = c(1, 0),
+    legend.title = element_blank()
+  )
+
+ggsave("graphs/brlasso_roc_validation.png", validation_roc)

### 2020-03-09 reports/article.Rmd
corrected version sent to NH, AG, OR.

@@
 title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
-author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Erik Lenguerrand^5^, Ashley Blom^5^, Mark Wilkinson^6^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
+author: Anne Garland`*`^1,2,3^, Erik Bulow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5^, Mark Wilkinson^6^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
 date: '`r Sys.Date()`'
-bibliography: P:/library.bib
+bibliography: library.bib
 csl: the-lancet.csl
@@
 )
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(
+  echo = FALSE,
+  out.width  = "100%", 
+  out.height = "100%"
+)
 library(tidyverse)
@@
 source("../lib/clean_names.R")
+
+shiny_url <- "https://shpr.registercentrum.se/om-registret-1/forskning/prediktion-av-90-dagarsmortalitet/p/SkyeTsTFB"
 ```
@@
 
-**Correspondence: [OR EB?]** 
-anne.garland@surgsci.uu.se, Phone: +46 498 26 80 00, Postal: Orthopedic Department, Visby Hospital, St Goransgatan 10, 621 84 Visby, Sweden
+`*` Authors contributed equally.
+
+**Correspondence:** 
+erik.bulow@vgregion.se. 
+Postal: Svenska Höftprotesregistret, Registercentrum Västra Götaland, SE-413 45, Sweden
 
@@
 
-**OBJECTIVE:** Shared decision making is essential before total hip arthroplasty (THA) surgery.
-Comorbidities are associated with an increased risk of 90-day morbidity and mortality, but diagnosis code-based instruments such as the Charlson or Elixhauser comorbidity indices are impossible to use in a clinical setting, and the American Society for Anesthesiologists (ASA) grade is imprecise.
-We searched for a simple model to predict early mortality after THA.
+**OBJECTIVE:** Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality.
+However, no prediction tools based on European populations are available, diagnosis code-based instruments such as the Charlson co-morbidity index are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) class is imprecise.
+We thus developed a simple model to predict 90-day mortality after elective THA.
 
-**PATIENTS AND METHODS:** We studied `r format(nrow(df), big.mark = ",")` Swedish patients operated with a cemented THA due to primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")`, linked data from several national registries in order to collect information on demographics and comorbidities.
-Bootstrap ranking procedures were used together with logistic LASSO regression to develop prediction models for death within 90 days.
-Predictive accuracy was assessed by the areas under the receiver operating characteristic curve (AUC).
-The final model was externally validated on a data-set from the National Joint Registry of England and Wales .
+**PATIENTS AND METHODS:** We studied `r format(nrow(df), big.mark = ",")` Swedish patients operated with a cemented THA due to primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` and used personal identity numbers to link data from the Swedish Hip Arthroplasty Register, the National Patient Register, the Longitudinal integrated database for health insurance and the population register.
+We used a bootstrap ranking procedure together with logistic LASSO regression to develop a prediction model for death within 90 days after surgery.
+The model was internally validated with bootstraping, and externally validated on a cohort of `r format(N_njr$tot, big.mark = ",")` patients from the National Joint Registry for England and Wales.
+Model performance was assessed by the area under the receiver operation characteristics curve (AUC) for discrimination, and by calibration belt plots.
 
@@
 get_auc <- function(mod) {
-  subset(brlasso_tbl_auc, Model == mod, "AUC", drop = TRUE) %>% 
-    {gsub("(", "(95% CI: ",., fixed = TRUE)}
+  subset(
+    brlasso_tbl_auc, 
+    Model == mod & 
+      `Age as` %in% c(NA, "main effect"), 
+    "AUC", 
+    drop = TRUE
+  ) %>% 
+  {sprintf("(AUC = %s", gsub(" (", ", 95% CI: ",., fixed = TRUE))}
 }
+
 ```
 
-**RESLUTS:** The unadjusted cumulative 90-day mortality was `r surv90d`.
-Best predictive performance for 90-day mortality was found for a model combining `r coefs_print_reduced` (AUC = `r get_auc("BRL reduced (age as main effect)")`).
-This model was superior to the  Charlson (AUC = `r get_auc("Charlson")`) and Elixhauser (AUC = `r get_auc("Elixhauser")`) comorbidity measures.
-A web calculator to aid clinical usage was published at https://erikbulow.shinyapps.io/thamortpred/ **[Would it be possible to embedd it at shpr.se instead?]**.
+**RESLUTS:** The unadjusted cumulative 90-day mortality was `r surv90d` in Sweden and `r njr_surv` in England and Wales.
+We propose a simple main effects model combining `r coefs_print_reduced`.
+This model had good discriminatory ability both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`, rendering it superioir to traditional models based on the ASA class `r get_auc("ASA")`, the Charlson co-morbitiy index `r get_auc("Charlson")`, and the Elixhauser com-morbidity index `r get_auc("Elixhauser")`.
+Our model was well calibrated for predicted probabilities up to 5 %.
+A web calculator to aid clinical usage was published at `r shiny_url`.
 
-**CONCLUSION:** We found a relatively simple prediction model of 90-day mortality after THA. This model requires less data and is easier to use compared to previously well-known comorbidity indices.
+**CONCLUSION:** We found a relatively simple prediction model of 90-day mortality after elective cemented THA based on easily available demographic data and the presence of four defined co-morbidities.
+This may facilitate individualized risk assessment prior to one of the most common surgical interventions.
 
@@
 Shared decision making has evolved into an integral part of patient-physician interactions prior to surgical interventions, and the weighing of risks against benefits is central to this process.
-In terms of gained quality of life and cost-utility, total hip arthroplasty (THA) is an enormously successful procedure, but the risk of adverse events in the short and long term is an essential part of preoperative discussions and various risk prediction models have been developed.[@Price2019]
+In terms of gained quality of life and cost-utility, total hip arthroplasty (THA) is an enormously successful procedure.
+90-day mortality after THA performed due to osteoarthritis is low, ranging from 0.2% to 0.6% in large joint registry studies.[@Hunt2013; @Garland2017] The risk of short-term mortality is nevertheless an essential part of a preoperative discussion between patient and surgeon.
 
-90-day mortality after THA surgery performed due to osteoarthritis is low, ranging between 0.2% and 0.6% in large joint registry analyses where adjustment for comorbidities is possible.[@Hunt2013; @Garland2017]
+Co-morbidity is associated with a shorter remaining life span, but the Charlson and Elixhauser co-morbidity indices poorly predict mortality after THA.[@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017; @Bulow2017; @Inacio2015] Additionally, these complex co-morbidity instruments are based on information from extensive in- or out-patient databases, each measure being defined by 1,178 and 1,516 individual codes respectively from the Swedish adaptation of the international classification of diseases version 10 (ICD-10).
+Each co-morbidity index also exists in numerous versions, making interpretation and comparisons difficult.[@Brusselaers2017; @Walraven2009] The use of these indices is therefore limited to research settings but are impossible to use under clinical circumstances.
 
-The presence of comorbidities is associated with a shorter remaining life span, but the Charlson and Elixhauser comorbidity indices poorly predict mortality after THA and total knee arthroplasty.[@Bulow2017; @Inacio2015] Additionally, these complex comorbidity instruments are based on the availability of extensive data-sets including in- and outpatient data coded by the 10th revision of the international classification of diseases (ICD-10), and each of the comorbidity indices exists in numerous versions.[@Brusselaers2017; @Walraven2009] Interpretation and comparison of different studies is therefore difficult.
+Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery.[@Bozic2013a; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] Due to limited, often single-center based samples, a lack of reported calibration abilities, the absence of external validation and the use of extrapolation there is still no widely accepted model for the prediction of early mortality.[@Manning2016] Importantly, to our knowledge, no reported model has been developed or validated on a European population.
 
-Comorbidity data are used to construct universal or arthroplasty-specific risk prediction tools, but no model has so far been broadly accepted to predict mortality after  elective THA.[@Manning2016; @Bulow2017] In order to propel arthroplasty surgery into the age of precision medicine, easily applicable tools to predict short- and long-term complications are urgently needed.
-We aimed to define a model that accurately predicts the risk of 90-day postoperative mortality after THA.
+We therefore aimed to develop a parsimonious prediction model of 90-day postoperative mortality after THA with internal and external validation of discrimination and calibration, and to compare this to the accuracy of prediction models based on existing co-morbidity measures.
 
@@
 
-Patients recorded in the Swedish Hip Arthroplasty Register (SHAR) with cemented THA due to osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` were used to develop, and internally validate, the prediction model (Figure \@ref(fig:flowchart)).
-Data from the National Joint Register in England and Wales (NJR) was later used for external validation.
-Only the last operated hip was accounted for in patients with bilateral THA. [@Bulow2019a]
+We used data from the Swedish Hip Arthroplasty Register (SHAR) for model derivation and internal validation of discrimination and calibration.
+The best model was then validated externally on patients from England and Wales recorded in the National Joint Registry for England, Wales, Northern Ireland, the Isle of Man and the States of Guernsey (NJR).
+We focused on patients with cemented THA due to osteoarthritis.
+Cementation is the most common fixation technique used in Sweden, thus used as inclusion criteria to increase heterogeneity.
+Only the last operated hip was accounted for in patients with bilateral THA, since those are shown to better resemble unilateral THAs.[@Bulow2019a] (Figure \@ref(fig:flowchart)).
 
-Data linkage, based on the 10-digit identity numbers assigned to all Swedish residents,[@Ludvigsson2009] were used to collect data from a variety of sources, as previously described.[@Cnudde2016]
+## Derivation cohort (Sweden)
 
-Age, sex, body mass index (BMI), the American Society for Anesthesiologists (ASA) grade, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, which has a completeness of 96-98%.[@shpr2018] Data on education level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) and civil status (married/un-married/divorced/widow[er]), were collected from the longitudinal integration database for health insurance and labor market studies from Statistics Sweden.[@Ludvigsson2019] The Swedish National Patient Register was used to assess comorbidities during the year preceding index surgery.
-This register contains all relevant diagnoses coded by ICD-10, as well as dates of admission and discharge for in- and outpatient episodes in all private and public hospitals.[@Ludvigsson2011] Death dates were linked from the national population register.
+The patient cohort was identified from the SHAR.
+The inclusion period started 2008 since this was the first year with the American Society for Anesthesiologists (ASA) class and Body Mass Index (BMI) systematically recorded.
+The last year was 2015, since we had access to co-morbidity from the national patient register up to this point.
+Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009] We used data from a variety of sources, as previously described by Cnudde et al.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98%.[@shpr2018] Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status (married/un-married/divorced/widow[er]), were also collected from LISA.
+The Swedish National Patient Register (NPR) was used to assess co-morbidity during the year preceding index surgery.
+This register contains all relevant diagnoses coded by ICD-10, as well as dates of admission and discharge for in- and outpatient episodes in all private and public hospitals.
+Completeness for NPR is above 99 % and 85-95 % of all diagnostic codes are valid.[@Ludvigsson2011] Death dates were linked from the national population register.
 
-Comorbidity was defined by individual ICD-10 codes grouped into 17 categories according to Charlson [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to Elixhauser.[@Elixhauser1998; @Quan2005]
+Patients with BMI either missing or above 50 were excluded, as were patients with ASA class missing or above III, as well as patients with unknown educational level or type of hospital.
 
-Some comorbidities were identified by both Charlson and Elixhauser, and some distinct comorbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
-We therefore combined individual categories to establish `r nrow(tab_categorization)` broader categories of comorbidity (Table \@ref(tab:tabcategorization)).
-Conditions were merged according to clinical relevance as to be recognized in a patient-doctor meeting without access to external register data.
+## External validation cohort (England and Wales)
 
-## Statistics
+Similar inclusion criteria were applied to the validation cohort from England and Wales.
+We did not exclude patients based on BMI, educational level or civil status however, since those variables were not used in the final model.
+The validation cohort was created by linking NJR to the secondary uses service database from the National Health Service (NHS).
+This database includes co-morbidity recorded in the admitted patient care module of the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Brophy2006; @Burns2012; @Sinha2013; @Simmonds2014; @Thorn2016] Out-patient deaths were linked from the Office of national statistics.
+Data linkage used NHS-numbers, as well as patient names, age, sex and address.[@Sayers2016]
 
-We used the Kaplan-Meier estimator to assess unadjusted mortality.
+## Defining co-morbidity
 
-Further analysis were based on logistic regression since no censoring occurred within the 90 day study period.
-We used a modelling procedure with bootstrap ranking and logistic least absolute shrinkage and selection operator (LASSO) model.[@Guo2015]
-Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1.
+Co-morbidity was defined by individual ICD-10 codes grouped into 17 categories according to Charlson [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to Elixhauser.[@Elixhauser1998; @Quan2005] Some co-morbidities were identified by both Charlson and Elixhauser, and some distinct co-morbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
+We combined individual diagnostics groups to establish `r nrow(tab_categorization)` broader categories of co-morbidity (Table \@ref(tab:tabcategorization)).
+Conditions were merged according to clinical relevance as to be recognized in a patient-physician meeting without access to external register data.
 
-Comorbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process.
+## Statistical analysis for model development
+
+We used the Kaplan-Meier estimator to assess unadjusted mortality after cemented THA as background knowledge.
+Further analysis was based on logistic regression since no censoring occurred within the 90 day study period.
+We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1.
+Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process.
 1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model.
@@
 Those $\lambda$:s were used to estimate model coefficients for each potential predictor.
-Absolute values from those estimates were used as a measure of variable importance.
-Piece-wise linear regression was  used to detect a break point where a significant drop in variable importance were observed.
+The magnitude (absolute values) of those estimates were used as a measure of variable importance.
+Piece-wise linear regression was used to detect a break point where a significant drop in variable importance was observed.
 Potential predictors with variable importance above this break point were considered important and kept as model candidates.
 The whole process was repeated 100 times.
+Covariates that were selected at least once out of the 100 times were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables (main model).
+A reduced model with variables chosen at least 33 out of the 100 times was used as a simpler alternative for comparison.
+This model, without cancer as a predictor, was also evaluated separately, considering that medical indications for THA surgery may look different for patients with cancer compared to cancer free patients.
+Univariable models with the ASA class, Charlson or Elixhauser co-morbidity indices were used for benchmarking, as well as a multivariable model with age and sex.
+Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.
+Odds ratios for the final model were estimated with 95% confidence intervals.[@Ripley2002]
 
-Covariates that were selected at least one out of the 100 times were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables.
-We will call this model "BRL" for bootstrap ranking LASSO.
-A reduced model with variables chosen at least 33 out of the 100 times, is called "BRL reduced".
-This model, but without cancer as a predictor, was also evaluated, since patients with cancer are sometimes treated differently [NEEDS CLARIFICATION!].
+## Statistical analysis for model validation
 
-Univariable models with the ASA grade, Charlson or Elixhauser comorbidity indices were used for benchmarking, as well as a multivariable model with age and sex.
-Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.
+Each model was used to predict the probability of death within 90 days for patients from the SHAR (internal validation).
+Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.[@Fawcett2006] Models with a lower 95% confidence limit above 0.7, were considered good.
+Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples.
+We used the bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples to adjust for optimism.[@Miller1991] Calibration bands were made to graphically assess model calibration, comparing predicted probabilities and observed proportions.[@Nattino2016] The reduced model was then evaluated externally.
+An AUC with 95% confidence interval was calculated for the model as-is.
+Re-calibration of the model intercept was then performed to account for different mortality rates in Sweden compared to England and Wales.
+An updated over-all slope was also calculated to account for country-specific treatment differences.[@Steyerberg2004] Calibration for this re-calibrated model was illustrated in the same calibration belt plot as for the internal calibration.
 
-Each of those models were used to predict the probability of death within 90 days for each patient.
-Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of predictive accuracy.
-Models with a lower 95% confidence limit above 0.7, were considered good.
-Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples for internal validation.
-The bias-corrected Somers' $D_{xy}$ rank correlation was used to adjust those intervals for optimism based on 100 re-samples.[@Miller1991] Calibration bands were made to graphically assess model calibration.[@Nattino2016]
+## Statistical tools
 
-Odds ratios for the final model were estimated with 95% confidence intervals based on interpolations of profile traces.[@Ripley2002]
-
-We also built an online web calculator available at https://erikbulow.shinyapps.io/thamortpred/.
+We built an online web calculator available at `r shiny_url` to be used in clinical practice.
 
 We used `r substr(R.version.string, 1, 15)` (R Foundation for Statistical   Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC and shiny.
-All R-scripts and necessary configurations (but no personal data) is available at zenodo.org/XXX.
+All R-scripts and necessary configurations (but no personal data) is available at zenodo.org/XXX **[Add!]**.
 A linked Binder environment is also available for interactive usage.
@@
 
-Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (271-14 and 360-13).
-**[COI?]**
+Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13).
 
@@
 
-After application of exclusion criteria, `r format(nrow(df), big.mark = ",")` patients (age `r paste(range(df$P_Age), collapse = " - ")`, `r round(mean(df$P_Sex == "Female") * 100)`% females) were included in the analysis (Figure \@ref(fig:flowchart)).
-`r with(df, sprintf("%d (%.2f%%)", sum(death90), mean(death90) * 100))` patients died within 90 days and no one was censored before that.
-The unadjusted risk of 90 day mortality was `r surv90d`.
+## Study participants
 
-Characteristics of the study population are presented in Table \@ref(tab:tab1).
-`r round(mean(df$CCI_index_quan_original != 0) * 100)`% of all patients had at least one pre-surgery comorbidity according to Charlson and `r round(mean(df$ECI_index_sum_all  != 0) * 100)`% according to Elixhauser.
-The proportion of patients with ASA grade 3 was `r round(mean(df$P_ASA == 3) * 100)`%.
-Most comorbidities were more common among patients who died.
+`r format(nrow(df), big.mark = ",")` patients (age `r paste(range(df$P_Age), collapse = " - ")`, `r round(mean(df$P_Sex == "Female") * 100)`% females) were included in the derivation cohort from SHAR (Figure \@ref(fig:flowchart) left panel).
+`r with(df, sprintf("%d (%.2f%%)", sum(death90), mean(death90) * 100))` of those patients died within 90 days and no one was censored before that.
+The unadjusted risk of 90-day mortality was therefore `r surv90d`.
+Further characteristics of the study population are presented in Table \@ref(tab:tab1).
+`r round(mean(df$CCI_index_quan_original != 0) * 100)`% and `r round(mean(df$ECI_index_sum_all  != 0) * 100)`% of the Swedish patients had at least one pre-surgery co-morbidity according to Charlson and Elixhauser.
+The proportion of patients with ASA class III was `r round(mean(df$P_ASA == 3) * 100)`%.
 
-There were `r nrow(comb_lgl)` comorbidities that were not recorded for any patient who died: `r comb_lgl_text`, and these variables were thus excluded from the modelling process.
+In addition, `r format(N_njr$tot, big.mark = ",")` patients were included for the external validation cohort (Figure \@ref(fig:flowchart) right panel).
+Their unadjusted risk of 90 day mortality was higher compared to the derivation cohort `r njr_surv`.
 
-The "BRL" model included `r coefs_print`. The "BRL reduced" model only used `r coefs_print_reduced` (Table \@ref(tab:brlprop)).
+## Model development and internal validation
 
-We observed no relevant differences between models based on age described by restricted cubic splines with two versus three knots and therefore only present results for models based on three knots.
-The use of splines did however not improve the accuracy of the models when compared to simpler main effect models.
-The "BRL" model had an estimated AUC of `r get_auc("BRL (age as main effect)")`, and the reduced version performed equally good with AUC `r get_auc("BRL reduced (age as main effect)")`.
-Univariable models with ASA grade, Charlson or Elixhauser performed poorly with an AUC that was lower than 0.7.
-A model based on only age and sex performed slightly better but still with an AUC not statistically significantly above 0.7 (Table \@ref(tab:auc) and Figure \@ref(fig:rocs)).
+There were `r nrow(comb_lgl)` co-morbidities that were not recorded for any patient who died in the Swedish derivation cohort: `r comb_lgl_text`.
+Those variables were therefore excluded as potential predictors prior to any statistical model derivation.
+The derived main model included `r coefs_print`.
+The reduced model, with covariates included at least 33 out of 100 times, were restricted to `r coefs_print_reduced` (Table \@ref(tab:brlprop)).
 
-AUCs and 95% confidence intervals, with and without adjustments for optimism, are illustrated for some models in Figure \@ref(fig:aucci)).
+There were no differences between models including age as a main effect, or as modeled by restricted cubic splines with either two or three knots.
+We will therefore focus on the simpler models with age as a main effect.
+Similarly, the correction for optimism only affected the third decimals of the AUC confidence intervals.
+We will therefore omit those.
 
-The ability of the "BRL reduced" model to estimate probabilities of death within 90 days is illustrated in Figure \@ref(fig:sep).
+The main and reduced models were no different regarding discriminative power, `r sprintf("%s versus %s", gsub(")", "", get_auc("Main model"), fixed = TRUE), gsub("(", "", get_auc("Reduced model"), fixed = TRUE))`.
+We therefore considered the reduced model as superior due to simplicity.
+Traditional models performed poorly with 95 % confidence intervals not above 0.7: ASA class `r get_auc("ASA")`, the Charlson co-morbitiy index `r get_auc("Charlson")`, and the Elixhauser com-morbidity index `r get_auc("Elixhauser")` (Figure \@ref(fig:rocs) left panel and \@ref(fig:aucci)).
+
+The ability of the reduced model to estimate probabilities of death within 90 days is further illustrated in Figure \@ref(fig:sep).
 Patients who died had, on average, higher predicted probabilities to do so.
-The model calibration was good for probabilities up to 3%, while higher probabilities might be over-estimated (Figure \@ref(fig:calibration)).
+The model calibration was good for estimated probabilities up to 3% and acceptable up to 5%, although with predicted probabilities usually higher then observed proportions of deaths among patients with corresponding covariate patterns (Figure \@ref(fig:calibration)).
 
-Estimated coefficients and corresponding odds ratios for the "BRL reduced" model is presented in Table \@ref(tab:brlreduced).
+Estimated coefficients and corresponding odds ratios for the reduced model are presented in Table \@ref(tab:brlreduced).
+
+To omit cancer from the reduced model did not affect the AUC or calibration for estimated probabilities below 3%.
+It did lead to worse calibration outside this range however, wherefore we recommend including cancer as a predictor.
+
+## External validation
+
+The discriminative ability for the reduced model was not statistically significantly different when applied to the external validation cohort `r get_auc("NJR (Reduced model)")` compared to the internal `r get_auc("Reduced model")` (Figure \@ref(fig:rocs) right panel and Figure \@ref(fig:aucci)).
+Calibration of the re-calibrated model was slightly worse compared to the internal calibration.
+Predicted probabilities between 0.5 % and 1.5 % were lower than observed and estimated 95 % confidence bands were wider.
+Over-all, calibration was still good however for predicted probabilities below 3% and acceptable below 5% (Figure \@ref(fig:calibration)).
 
@@
 
-We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` was able to better predict 90-day mortality than more complex models such as the Charlson or Elixhauser comorbidity indices.
+We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` better discriminated patients who died within 90 days after THA, from patients who survived, compared to traditional models with only ASA class or the Charlson or Elixhauser co-morbidity indices.
 
-The resulting "BRL reduced" model can predict the probability of death within 90 days as: $p = 1 / [1 +\exp(-\beta X)]$ where `r coefs_form`
-This formula is valid for patients within the observed age range (`r paste(range(df$P_Age), collapse = " - ")` years) with observed combinations of diseases.
+The resulting (reduced) model predicts the probability of death within 90 days as: $\hat p = 1 / [1 +\exp(-\hat \beta X)]$ where $\hat \beta$ is the vector of estimated coefficients presented in table \@ref(tab:brlreduced).
+This formula is considered valid for patients aged `r paste(range(df$P_Age), collapse = " - ")` years and for predicted probabilities up to 5%.
 
@@
 
-For example a `r min(df$P_Age)` year old woman with ASA grade = 1 and none of the important comorbidities would have a `r get_risk("baseline")` risk to die within 90 days of surgery.
-Another woman, `r quantile(df$P_Age, .25)` years old (the first age quantile), have an elevated risk of `r get_risk("healthy_female")`.
-A `r max(df$P_Age)` years old man (the maximum observed age) with ASA grade = 3 and cancer, would have a risk of `r get_risk("sickest_man")`.
-Note however that observed covariate patters with predicted probabilities above 3% were rare.
-Only `r mean(pred_probs > 3) * 100`% (`r sum(pred_probs > 3)` patients) had those (Figure \@ref(fig:sep)).
-Estimated mortality risks above 3% are therefore subject to extrapolation.
-Some risk calculators ignore this problem,[@Bozic2013a] but we think this should be acknowledged.
+## Practical usage
 
-Obesity in the "BRL reduced" model was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but is relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We note that the proportion of patients with BMI above 30, the definition of obesity (`r mean(df$P_BMI > 30, na.rm = TRUE) * 100`%) is much higher than the proportion of patients with ICD-10 = E66* (obesity) from the National Patient Register (`r mean(df$c_obesity) * 100`%).
+Our model could be used in clinical practice, either by the formula above, or by a simple web calculator online (`r shiny_url`).
+For example a `r min(df$P_Age)` year old woman with ASA class I and none of the important co-morbidities would have a `r get_risk("baseline")` risk to die within 90 days of surgery.
+Another woman, `r quantile(df$P_Age, .25)` years old (the first age quantile), would have an elevated risk of `r get_risk("healthy_female")`.
+A `r max(df$P_Age)` years old man (the maximum observed age) with ASA class III and cancer, would have a risk of `r get_risk("sickest_man")`.
+Note however that covariate patters with observed probabilities above 5% were rare (`r mean(pred_probs > 5) * 100`%, n = `r sum(pred_probs > 5)`).
+Estimated mortality risks above 5% are therefore subject to extrapolation.
+The observed proportions of deaths for patients with similar characteristics is likely lower (as indicated by Figure \@ref(fig:calibration)).
+Some risk calculators ignore this problem,[@Bozic2013a; @Harris2019] but we think this should be acknowledged.
+
+## Model predictors
+
+Variables in our model were chosen based on statistical properties and not due to clinical relevance.
+Variables and estimated coefficients should therefore not be assigned any exact epidemiological and/or causal meaning per se.[@Breiman2001] We could nevertheless discuss some aspects of the relevant predictors.
+
+Age and sex are well-known predictors of remaining life span.
+It is less obvious that this relation must be linear.
+We used restricted cubic splines to allow a more flexible relation, but found no difference in AUC compared to simpler main effect models.
+ASA class III has the largest estimated coefficient among all predictors, indicating large relative importance.
+This is clinically reasonable since a label of "severe systemic disease" should be based on a relevant patient assessment prior to surgery.
+Conditions were not too severe however since elective surgery was performed after this assessment.
+ASA class is nevertheless known to have a high degree of internal variability.[@Haynes1995] It has previously been compared to the Charlson co-morbidity index, but not with respect to mortality after THA.[@Whitmore2014; @Kork2015] Patients with ASA class IV-VI were excluded since those categories describe severe disease, moribund and brain-dead individuals.
+We suspect that patients with such conditions might have been misclassified.
+Their true class could either be lower, or they could have received THA due to a femoral hip fracture instead of osteoarthritis.
+
+Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We noted that the proportion of patients with BMI above 30, the WHO definition of obesity is much higher (`r mean(df$P_BMI > 30, na.rm = TRUE) * 100`%) than the proportion of patients with ICD-10 = E66 (diagnosed obesity) from the National Patient Register (`r mean(df$c_obesity) * 100`%).
 This proportion corresponds to patients with BMI of at least `r which.min(map_dbl(1:50, ~mean(df$P_BMI > ., na.rm = TRUE)) > mean(df$c_obesity))`.
-Thus, obesity might be indicated in the risk calculator only if the condition is severe.
+Thus, the relevance of obesity might not be the condition itself, but rather all circumstances making the related diagnose appear in the national patient register.
+This might include unmeasured confounders related to patients, doctors, and local hospital guidelines or similar.
 
-The "BRL reduced" model without cancer had the same AUC as a model with cancer.
-The AUC value itself is a rank-based and scale-invariant measure that does not account for model calibration.
-Both models were however equally well calibrated as well, for estimated probabilities within the recommended range (0-3%).
-The model without cancer however performed worse for probabilities above 3%.
+## Strengths and limitations
 
-It is known that male sex is associated with earlier deaths and that the remaining life span will decrease with increased age.
-It is less obvious that this relation must be linear.
-We used restricted cubic splines to allow a more flexible relation, but found that a linear relationship was equally good.
-Our model includes ASA grade, which is routinely assessed pre-operatively in most developed countries.
-It is however known to have a high degree of internal variability.[@Haynes1995] It has previously been compared to the Charlson comorbidity index, but not with respect to mortality after THA.[@Whitmore2014; @Kork2015] Patients with ASA grade 4-6 were excluded since those categories describe severe disease, moribund and brain-dead individuals.
-We suspect that patients with those ASA grades might be misclassified.
-Their true grade could either be lower, or they could have recieved THA due to a femoral hip fracture instead of osteoarthritis.
+A strength of the study is the nationwide design with data from national registers in both Sweden, England and Wales.
+The Swedish registers are valid with low proportions of missing data.[@Soderman2000; @shpr2018; @Soderman2001; @Ludvigsson2011] Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however.[@Spencer2012; @Williams2002; @Brennan2012] A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80%.[@Burns2012] Both in- and outpatient co-morbidity data was available for Sweden, but only in-patient data for England and Wales.
+We did not have acces to any data from primary care or nursing homes in any of the counties.
 
-Comorbidity burden is also known to influence the outcome after THA.[@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017] Coding algorithms on the other hand are complex and not used in clinical settings since the administrative burden is too high.
-The Charlson classification comprise 1,178 ICD-10 codes and the Elixhauser 1,516.
-Those indices are therefore only used by researchers.
+The linkage procedure between NJR and HES was previously described by 
+Smith et al.[@Smith2012] They found that privately founded patients were not included in HES.
+17% of the patients had missing personal data or did not allow linkage, and 6% were not found in HES even if their data was available from NJR.
 
-Risk prediction may be useful in the process of patient selection prior to surgery, in the preoperative risk management including a review of current medications, and in perioperative anesthesia management.
-A number of risk prediction tools of various complexity for adverse events after total joint replacements have been introduced but none has been broadly accepted.[@Manning2016; @Price2019] In the context of trauma surgery, outcome prediction tools are common, and it seems possible to reduce the number of items without losing predictive accuracy.[@Gerdin2016] Our results indicate that the risk of early postoperative mortality after THA could be assessed by a relatively simple prediction model.
+In addition to data validity, the indication for different fixation techniques also differ in Sweden compared to England and Wales.
+Cementation is most commonly used in Sweden, but not so in England and Wales, where relatively younger patients are more likely operated without cement.
+To only include cemented THA therefore implies an over-representation of older and frailer patients from England and Wales compared to Sweden.
+Those patients might be more likely to die within 90 days.
+This was seen in the calibration plot where estimated probabilities below 1.5% underestimated the observed proportion of deaths in the external validation.
 
-A strength of this study is the nationwide design with a large cohort of primary THA patients.
-We were able to use exact data linkage by the Swedish identity numbers and had no censoring.
-Our data sources are valid with low proportions of missing data.[@Soderman2000; @shpr2018; @Soderman2001; @Ludvigsson2011]
+It should be noted that the risk model does not study THA as an observed intervention.
+We merely followed the cohort who already had THA.
+Hence, deaths within 90 days might occur for those patients regardless if THA is inserted or not.
+The proximity in time however, the maximum of 90 days from THA to death, is an indication that the operation might have influenced the observed mortality.
+Insertion of an elective THA is always preceded by a clinical judgement.
+Hence, no patient with a foreseen death near-by is given elective THA.
+We therefore believe that at least a non-significant proportion of deaths within 90 days are related to the THA itself.
 
-The risk of coding errors might be a limitation to the study, especially so if coding routines change over time.
-It should also be remembered that the risk model does not study THA as an observed intervention.
-We merely followed the cohort who did already have THA.
-Hence, deaths within 90 days might occur for the patients regardless if THA is inserted or not.
-The proximity in time however, the maximum of 90 days from THA to death, is an indication that the operation might have influenced the deaths observed.
-The insertion of an elective THA is always preceded by a clinical judgement.
-Hence, no patient with a foreseen death near-by is given THA to begin with.
-We therefore believe that at least a non-significant proportion of deaths within 90 days are related to  THA.
+## Conclusion
 
-We hope that the supplied web calculator and the transparent reporting of this model might lead to clinical usage that can be part of a pre-surgery discussion between doctors and patients in need of THA.
+Our results indicate that the risk of early postoperative mortality after cemented THA can be pre-operatively assessed by a parsimonious prediction model.
+We hope that this model, with its accompanying web calculator, will facilitate a shared decision making between physicians and their patients in need of THA.
 

### 2020-03-09 reports/figs_tabs.Rmd
corrected version sent to NH, AG, OR.

--- 2019-10-17 reports/figs_tabs.Rmd
@@
 
-```{r flowchart, fig.cap = "Flowchart depicting inclusion criteria and number of patients included in the development and internal validation phase of the model."}
+```{r flowchart, fig.cap = "Flowchart depicting inclusion criteria and number of patients.
+Data from the Swedish Hip Arthroplasty Register was used for model derivation and internal validation (left).
+Data from the National Joint Registry in England and Wales was used for external validation (right)."}
 knitr::include_graphics("../graphs/flowchart.png")
@@
 
-```{r rocs, fig.cap = "ROC curves for some of the models. The 'BRL' models are superior but the reduced version works equally well as the more complex version."}
-knitr::include_graphics("../graphs/brlasso_roc.png")
+```{r rocs, fig.cap = "Receiver Operation Characteristics (ROC) curves.
+The area under the ROC curve (AUC) for the reduced model was no different compared to the main model (left panel).
+The reduced model works almost as good for predictions in the National Joint Registry for England and Wales (right panel)."}
+knitr::include_graphics("../graphs/roc.png")
 ```
@@
 
-```{r aucci, fig.cap = "Area Under the Curve (AUC) as a measure of predictive accuracy.
-Only the BRL models have AUC statistically significantly above 0.7.
-The reduced version works as well as the more complex model.
-To adjust for optimism does not effect the result."}
+```{r aucci, fig.cap = "Area Under the Receiver Operation Characteristics Curve (AUC) as a measure of predictive discriminative ability with 95 % bootstrap confidence intervals.
+The reduced model was no different compared to the main model, neither internally, nor externally (overlapping confidence intervals).
+AUC above 0.7 was considered good.
+This was observed for the derived models but not for any other model included for comparison."}
 
-knitr::include_graphics("../graphs/brlasso_auc_ci.png")
+knitr::include_graphics("../graphs/auc_ci.png")
 ```
@@
 
-```{r sep, fig.cap = "Most patients survived more than 90 days after THA, wherefore the blue bars totally dominate the histogram (upper panel).
-A standardized density plot however reveals that patients who died within 90 days (pink) were, on average, estimated to have a higher probability to do so (lower panel).
-There were very few observations with covariate patterns resultnig in death probabilities higher than 3%."}
-knitr::include_graphics("../graphs/brlasso_separation_auc.png")
+```{r sep, fig.cap = "Most patients survived more than 90 days after THA, wherefore the blue bars totally dominate the histogram (upper panel; note the scales).
+A normalized density plot however reveals that patients who died within 90 days were, on average, estimated to have a higher probability to do so (lower panel).
+There were very few observations with covariate patterns resulting in death probabilities higher than 5% (green line).
+Estimated probabilities above this limit are therefore subject to extrapolation."}
+knitr::include_graphics("../graphs/separation.png")
 ```
@@
 
-```{r calibration, fig.cap = "Calibration belts at 80% and 95% (light and dark grey).
-The predicted probabilities do not deviate from the observed ones by any statistical significance.
-The risk of over-estimation is however clear above 3%."}
-knitr::include_graphics("../graphs/calibration_belt.png")
+```{r calibration, fig.cap = "This figure illustrates calibrartion between observed proportions and predicted probabilities with 95 % confidence intervals.
+Deviations above the diagonal line means that the model underpredicts the probability of death within 90 days.
+Such deviation was statistically significant for the external validation for predictions below 1.5%.
+The higher the predicted probability gets, the more they over-estimate the observed proportions of deaths."}
+knitr::include_graphics("../graphs/calibration.png")
 ```
@@
   tab_categorization, 
-  caption = "Categorization of individual Charlson and Elixhauser comorbidities into broader comorbidities.")
+  caption = "Categorization of co-morbdities based on diagnostic groups according to Charlson and Elixhauser. (CNS = central nervous system.)")
 
@@
   table1, 
-  caption = "Characteristics of the study population. Comorbidities recorded for at least one patient who survived 90 days, and one who did not, were modeled as potential predictors.")
+  caption = "Characteristics of the study population.
+Co-morbidities recorded for at least one patient who survived 90 days, and one who did not, were modeled as potential predictors.
+(CNS = central nervous system.)")
 ```
@@
   brlasso_tbl_selected %>% 
-  mutate(variable = clean_names(variable))
+  mutate(
+    variable = clean_names(variable),
+    variable = gsub("= 2", "II", variable),
+    variable = gsub("= 3", "III", variable),
+  )
 
@@
   t, 
-  caption = "Variables selected by the bootstrap ranking procedure, and therefore kept in the 'BRL' model.
-Variables chosen at least 33 times out of 100 were kept in the 'BRL reduced' model as well.
-ASA grade 2 and 3 were kept in both models, since those are usually distinguished anyway.
-It would however be possible to use a model lumping grade 1 and 2 together.")
+  caption = "Variables selected by the bootstrap ranking procedure, and therefore kept in the main model.
+Variables chosen at least 33 times out of 100 were kept in the reduced model as well.
+ASA class II and III were kept in both models, since those are usually distinguished anyway.
+To simplify the models further would be possible by lumping class I and II together.
+Variables chosen more frequently are likely more important predictors.
+This is not necessarily true however since one of several strongly correlated variables might be chosen spuriously.
+(CNS = central nervous system.)")
 ```
 
-##### PAGE BREAK
-
-```{r auc, warning = FALSE}
+<!-- Same info in figure so do not use this table -->
+```{r auc, warning = FALSE, include = FALSE}
 knitr::kable(
-  brlasso_tbl_auc, 
-  caption = "Area Under the Curve (AUC) as a measure of predictive accuracy for the 'BRL' model, the 'BRL reduced', a model with age and sex, as well as univariable models with ASA grade and the Charlson or Elixhauser comorbidity indices.
-Age was included as either a main effect, or in the form of restricted cubic splines (RCS) with three knots.
-'BRL reduced' was also used without cancer and performed equally well in terms of AUC.
-The correction for optimism slightly decreased the AUC values, but those changes were marginal.")
+  brlasso_tbl_auc %>% 
+  filter(
+    `Age as` %in% c(NA, "main effect"),
+    !endsWith(Model, "cancer") 
+  ) %>% 
+  select(Model, AUC) %>% 
+  rename(`AUC (95 % CI)` = AUC),
+  caption = "Area Under the Curve (AUC) as a measure of discriminative ability for the main and reduced models, a model with age and sex, as well as univariable models with ASA class and the Charlson or Elixhauser co-morbidity indices.")
 ```
@@
 ```{r brlreduced}
-
+brlasso_tbl_coefs %>% 
+  select(-math) %>%
+  mutate(
+    term = gsub("= 2", "II", term),
+    term = gsub("= 3", "III", term)
+  ) %>% 
 knitr::kable(
-  select(brlasso_tbl_coefs, -math),
-  caption = 'Estimated coefficients and odds ratios with 95% confidence intervals for the "BRL reduced" model.'
+  caption = 'Estimated coefficients (beta) and odds ratios (OR) with 95% confidence intervals for the reduced model.'
 )

### 2020-03-09 lib/flowchart.R

@@ -1,0 +1,59 @@
+flowchart <- function(desc) {
+
+  nodes <-
+    desc %>%
+    pivot_longer(-step) %>%
+    filter(!is.na(value)) %>%
+    unite("node", step, name, remove = FALSE) %>%
+    add_rowindex() %>%
+    rename(
+      id    = .row,
+      label = value,
+      rank  = step
+    ) %>%
+    mutate(
+      shape     = "rectangle",
+      width     = if_else(startsWith(name, "incl"), 1.5, 2.5),
+      color     = "Black",
+      fillcolor = "White",
+      fontcolor = "Black"
+    )
+
+
+  # Define edges for graph --------------------------------------------------
+
+  incl_edges <-
+    nodes %>%
+    filter(startsWith(name, "incl"))
+
+  excl_edges <-
+    nodes %>%
+    filter(startsWith(name, "excl"))
+
+  edges <-
+    tibble(
+      from = incl_edges$id,
+      to1 = lead(incl_edges$id),
+      to2 = c(excl_edges$id, NA)
+    ) %>%
+    pivot_longer(-from, values_to = "to") %>%
+    select(-name) %>%
+    arrange(from, to) %>%
+    filter(!is.na(to)) %>%
+    {create_edge_df(
+      .$from, .$to,
+      color    = "black",
+      penwidth = 2,
+      len      = 1
+    )}
+
+
+  # Make graph --------------------------------------------------------------
+
+  create_graph(
+    nodes,
+    edges
+  ) %>%
+  add_global_graph_attrs("layout", "dot", "graph") %>% # Apply cluster ranking
+  add_global_graph_attrs("fixedsize", "FALSE", "node") # Nice edge height
+}

### 2020-03-09 munge/04-remove_empty_variables.R

@@ -39,7 +39,8 @@ cache("comb_lgl")
 comb_lgl_text <-
   comb_lgl$name %>%
   clean_names(FALSE) %>%
-  glue_collapse(", ", last = " and ")
+  glue_collapse(", ", last = " and ") %>%
+  {gsub("AIDS/HIV", "acquired immunodeficiency syndrome by the human immunodeficiency virus (AIDS/HIV)", .)}
 
 cache("comb_lgl_text")
 

### 2020-03-09 munge/08-NJR_external_validation.R

@@ -15,3 +15,36 @@ njr_calibration <- cal(obspred);      cache("njr_calibration")
 njr_ROC3         <- roccurve(obspred3); cache("njr_ROC3")
 njr_AUCci3       <- aucci(njr_ROC3);    cache("njr_AUCci3")
 njr_calibration3 <- cal(obspred3);      cache("njr_calibration3")
+
+# Numbers to present
+N_njr <- list(
+  death = length(njr_ROC$cases),
+  surv  = length(njr_ROC$controls)
+  )
+
+N_njr <- within(N_njr, {tot = death + surv})
+
+cache("N_njr")
+
+
+# 90 day survival ---------------------------------------------------------
+
+# We assume there is no censoring and therefore base the estimate on bootstrap
+njr_surv <-
+  tibble(
+    status = c(numeric(N_njr$death) + 1, numeric(N_njr$surv))
+  ) %>%
+  rsample::bootstraps(100) %>%
+  mutate(
+    death = map_dbl(splits, ~ mean(as_tibble(.)$status))
+  ) %>%
+  summarise(
+    est = mean(death),
+    lo = quantile(death, .025),
+    hi = quantile(death, .975),
+    mort = sprintf("%0.2f %% (95 %% CI: %0.2f-%0.2f)", est * 100, lo * 100, hi * 100)
+  ) %>%
+  select(mort) %>%
+  pluck(1, 1)
+
+cache("njr_surv")

### 2020-03-09 src/fig_auc.R

@@ -4,15 +4,21 @@ digs <- options("digits")
 options(digits = 2)
 
 
+# NJR ---------------------------------------------------------------------
+
 njr_AUC <-
   tibble(
-    Model   = "NJR (BRL reduced)",
+    Model   = "Reduced model (NJR)",
     AUC_lo  = njr_AUCci[1],
     AUC_est = njr_AUCci[2],
     AUC_hi  = njr_AUCci[3]
   )
 
-fig_auc_ci <-
+
+
+# Prepare data ------------------------------------------------------------
+
+data_auc_ci <-
   all_models %>%
   filter(fig) %>%
   select(Model, starts_with("AUC")) %>%
@@ -22,26 +28,36 @@ fig_auc_ci <-
   separate(name, c("AUC","level", "corr")) %>%
   mutate(
     corr = case_when(
-      startsWith(Model, "NJR") ~ "Validated",
-      is.na(corr)              ~ "Estimated",
+      Model == "Reduced model (NJR)" ~ "NJR",
+      is.na(corr)              ~ "SHAR",
       TRUE                     ~ "Adjusted for optimism"
-    )
+    ),
+    Model = model_names(Model)
   ) %>%
   pivot_wider(names_from = level) %>%
   mutate(Model = fct_reorder(Model, est)) %>%
-  ggplot(aes(Model, est, group = corr, color = corr)) +
-  geom_hline(yintercept = .7, color = "darkgreen", linetype = "dashed", size = 2) +
-  geom_pointrange(aes(ymin = lo, ymax = hi), size = 1.5, position = position_dodge(.5)) +
+  filter(corr %in% c("NJR", "SHAR"))
+
+annotates <-
+  data_auc_ci %>%
+  pivot_longer(c(lo, est, hi))
+
+# Figure ------------------------------------------------------------------
+
+ggplot(data_auc_ci, aes(Model, est, group = corr, color = corr)) +
+  geom_hline(yintercept = .7, color = "darkgreen", linetype = "dashed", size = 1) +
+  geom_pointrange(aes(ymin = lo, ymax = hi), size = 1) + #, position = position_dodge(.5)) +
+  geom_text(aes(Model, value, label = round(value, 2)), data = annotates, nudge_x = .3, size = 2) +
   coord_flip() +
   theme_minimal() +
   theme(
     axis.title.y = element_blank(),
     legend.title = element_blank(),
-    legend.position = "bottom"
[…]

### 2020-03-09 src/fig_calibration.R

@@ -1,0 +1,46 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+# SHAR (internal calibration) ---------------------------------------------
+
+shar_calibration <-
+  all_models %>%
+  filter(Model == "BRL reduced (age as main effect)") %>%
+  select(cal_belt) %>%
+  pluck(1, 1)
+
+
+# Extract relevant data ---------------------------------------------------
+
+beltplot2tibble <- function(cb) {
+  tibble(
+    x = cb$seqP,
+    L = cb$cbBoundByConfLevel[[1]]$L,
+    U = cb$cbBoundByConfLevel[[1]]$U
+  )
+}
+
+cbs <-
+  bind_rows(
+    `External calibration (NJR)`  = beltplot2tibble(njr_calibration3),
+    `Internal calibration (SHAR)` = beltplot2tibble(shar_calibration),
+    .id = "validation"
+  )
+
+# Make figure -------------------------------------------------------------
+
+ggplot(cbs, aes(x, ymin = L, ymax = U, fill = validation)) +
+  geom_ribbon(alpha = .3) +
+  geom_abline(aes(intercept = 0, slope = 1)) +
+  theme_minimal() +
+  scale_x_continuous(labels = scales::percent_format(1), minor_breaks = NULL) +
+  scale_y_continuous(breaks = seq(.01, .06, 0.01), minor_breaks = NULL, labels = scales::percent_format(1)) +
+  coord_cartesian(c(0, 0.05), c(0, 0.065)) +
+  xlab("Predicted probability") +
+  ylab("Observed proportion") +
+  theme(
+    legend.position = c(1, 0),
+    legend.justification = c(1, 0),
+    legend.title = element_blank()
+  )
+
+ggsave("graphs/calibration.png", height = 10, width = 10 , units = "cm")

### 2020-03-09 src/njr_flowchart.R

@@ -1,0 +1,27 @@
+njr_filters <-
+  tibble(
+    step = 1:5,
+    incl_text = c(
+      "THA 2008-2015\\l (N = 579,798)\\l",
+      "THA due to OA\\l (N = 515,720)\\l",
+      "Patients with OA\\l (N = 460,209)\\l",
+      "Patients with OA\\l and cemented THA\\l (N = 153,477)\\l",
+      "Validation sample\\l (N = 125,428)\\l"
+    ),
+    excl_next = c(
+      "Other diagnoses than OA\\l(N = 64,078)\\l",
+      "First THA for bilateral cases\\l(N = 55,511)\\l",
+      "Exclusion of (N = 306,732):\\l  - Uncemented (N = )\\l  - Hybrid (N = )\\l  - Reverse hybrid (N = )\\l",
+      "Exclusion of (N = 28,049):\\l  - Comorbidity missing (N = 27,181)\\l  - ASA class IV-V (N = 868)\\l",
+      NA
+    )
+  )
+
+flowchart_njr <- flowchart(njr_filters)
+
+
+combine_graphs(flowchart_shar, flowchart_njr) %>%
+  export_graph(
+    "graphs/flowchart.png", "png", width = 1024,
+    title = "SHAR (model development)                                        NJR (external validation)"
+  )

### 2020-03-09 src/tab_auc.R

@@ -1,13 +1,27 @@
 suppressMessages({library(ProjectTemplate); load.project()})
 
+
+njr_AUC <-
+  tibble(
+    Model   = "NJR (BRL reduced) (age as main effect)",
+    AUC_lo  = njr_AUCci[1],
+    AUC_est = njr_AUCci[2],
+    AUC_hi  = njr_AUCci[3]
+  )
+
+
 brlasso_tbl_auc <-
   all_models %>%
   filter(tab) %>%
+  bind_rows(njr_AUC) %>%
   arrange(desc(AUC_est)) %>%
   transmute(
-    Model,
+    Model = model_names(Model, age = TRUE),
     AUC      = sprintf("%.2f (%.2f-%.2f)", AUC_est, AUC_lo, AUC_hi),
-    `AUC corrected for optimism` = sprintf("%.2f (%.2f-%.2f)", AUC_est.corr, AUC_lo.corr, AUC_hi.corr)
-  )
+    #`AUC corrected for optimism` =
+    #  sprintf("%.2f (%.2f-%.2f)", AUC_est.corr, AUC_lo.corr, AUC_hi.corr)
+  ) %>%
+  separate(Model, c("Model", "Age as"), " \\(age as ") %>%
+  mutate(`Age as` = gsub("\\)", "", `Age as`))
 
 cache("brlasso_tbl_auc")

### 2020-03-09 src/tab_coefs.R

@@ -11,10 +11,10 @@ brlasso_tbl_coefs <-
     term = gsub("_", " ", term),
     math = sprintf("\\text{%s} \\\\ ", term) %>% {gsub("\\text{(intercept)} \\\\ ", "", ., fixed = TRUE)},
     beta = log(estimate),
-    OR   = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high),
+    `OR 95 % CI` = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high),
     p    = pvalue_format(.01)(p.value)
   ) %>%
-  mutate_at(vars(OR, p), ~ if_else(term == "(intercept)", "", .))
+  mutate_at(vars(`OR 95 % CI`, p), ~ if_else(term == "(intercept)", "", .))
 
 cache("brlasso_tbl_coefs")
 
@@ -37,16 +37,18 @@ coefs_print_string <- function(name) {
   basic_vars <-
     setdiff(all_vars, c_vars) %>%
     clean_names(firstupper = FALSE, lvls = FALSE) %>%
-    set_first("ASA grade") %>%
+    set_first("ASA class") %>%
     set_first("sex") %>%
     set_first("age")
 
-  c_vars <- clean_names(c_vars, FALSE)
+  c_vars <- clean_names(c_vars, FALSE) %>%
+    {gsub("CNS disease", "disease of the central nervous system", .)} %>%
+    {gsub("obesity", "diagnosed obesity", .)}
 
-  paste(
+  paste0(
     glue::glue_collapse(basic_vars, ", "),
-    "and the precense of",
-    glue::glue_collapse(c_vars, ", ", last = " and ")
+    ", the presence of ",
+    glue::glue_collapse(c_vars, ", ", last = ", and ")
   )
 }
 
@@ -71,6 +73,6 @@ coefs_form <-
   select(coefs) %>%
   pluck(1) %>%
   substr(1, nchar(.) - 4) %>%
-  {sprintf("$$\\begin{aligned} \\beta X = & %s \\end{aligned}.$$", .)}
+  {sprintf("$$\\begin{aligned} \\hat \\beta X = & %s \\end{aligned}.$$", .)}
 
 cache("coefs_form")

### 2020-03-09 src/tab_table1.R

@@ -1,0 +1,81 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+
+# Prepare data ------------------------------------------------------------
+
+dft1 <-
+  df %>%
+  mutate(
+    P_ASA      = factor(P_ASA, as.character(1:3), c("I", "II", "III")),
+    Charlson   = replace(CCI_index_quan_original, CCI_index_quan_original == 4, "4+"),
+    Elixhauser = replace(ECI_index_sum_all, ECI_index_sum_all == 3, "3+"),
+    death90f   = factor(death90f, c("dead", "alive"),
+                        c("Died within 90 days", "Survived at least 90 days"))
+  ) %>%
+  select(
+    death90f,
+    P_Age,
+    P_Sex,
+    P_BMI,
+    P_ASA,
+    P_TypeOfHospital,
+    education,
+    civil_status,
+    Charlson,
+    Elixhauser,
+    starts_with("c_"),
+  ) %>%
+  setNames(clean_names(names(.))) %>%
+  rename(`ASA class` = ASA)
+
+
+
+# Table by survival status ------------------------------------------------
+
+t1 <-
+  tableone::CreateTableOne(
+    strata = "Death90f",
+    vars = setdiff(names(dft1), "Death90f"),
+    factorVars = c("Sex", "ASA class", "Elixhauser", "Charlson"),
+    data = dft1,
+    test = FALSE
+  )
+
+cache("t1")
+
+
+# Add total column --------------------------------------------------------
+
+t1_all <-
+  tableone::CreateTableOne(
+    vars = setdiff(names(dft1), "Death90f"),
+    factorVars = c("Sex", "ASA class", "Elixhauser", "Charlson"),
+    data = dft1,
+    test = FALSE
+  ) %>%
+  print(
+    printToggle = FALSE
+  ) %>%
+  as_tibble()
+
+
+
+# Combine -----------------------------------------------------------------
+
+table1 <-
+  t1 %>%
+  print(
+    printToggle = FALSE
+  ) %>%
+  as_tibble(rownames = "what") %>%
+  add_column(Total = t1_all$Overall) %>%
+  mutate(
+    level = trimws(ifelse(startsWith(what, " "), what, "")),
+    level = paste0(toupper(substr(level, 1, 1)), substring(level, 2)),
+    what = trimws(ifelse(level == "", gsub(" = TRUE", "", what), ""))
+  ) %>%
+  mutate_at(vars(`Died within 90 days`,
[…]

### 2020-03-09 src/table1.R

@@ -1,76 +1,0 @@
-suppressMessages({library(ProjectTemplate); load.project()})
-
-dft1 <-
-  df %>%
-  mutate(
-    Charlson   = replace(CCI_index_quan_original, CCI_index_quan_original == 4, "4+"),
-    Elixhauser = replace(ECI_index_sum_all, ECI_index_sum_all == 3, "3+"),
-    death90f   = factor(death90f, c("dead", "alive"),
-                        c("Died within 90 days", "Survived at least 90 days"))
-  ) %>%
-  select(
-    death90f,
-    P_Age,
-    P_Sex,
-    P_BMI,
-    P_ASA,
-    P_TypeOfHospital,
-    education,
-    civil_status,
-    Charlson,
-    Elixhauser,
-    starts_with("c_"),
-  ) %>%
-  setNames(clean_names(names(.))) %>%
-  rename(`ASA grade` = ASA)
-
-t1 <-
-  tableone::CreateTableOne(
-    strata = "Death90f",
-    vars = setdiff(names(dft1), "Death90f"),
-    factorVars = c("Sex", "ASA grade", "Elixhauser", "Charlson"),
-    data = dft1,
-    test = FALSE
-  )
-
-cache("t1")
-
-
-# Add total column --------------------------------------------------------
-
-t1_all <-
-  tableone::CreateTableOne(
-    vars = setdiff(names(dft1), "Death90f"),
-    factorVars = c("Sex", "ASA grade", "Elixhauser", "Charlson"),
-    data = dft1,
-    test = FALSE
-  ) %>%
-  print(
-    printToggle = FALSE
-  ) %>%
-  as_tibble()
-
-
-
-zero <- function(x) {
-  gsub("0 ( 0.0)", "0", x, fixed = TRUE) %>%
-  {gsub("0.0", "<0.1", .)}
-}
-
-table1 <-
-  t1 %>%
-  print(
-    printToggle = FALSE
-  ) %>%
-  as_tibble(rownames = "what") %>%
-  add_column(Total = t1_all$Overall) %>%
-  mutate(
-    level = trimws(ifelse(startsWith(what, " "), what, "")),
-    level = paste0(toupper(substr(level, 1, 1)), substring(level, 2)),
-    what = trimws(ifelse(level == "", gsub(" = TRUE", "", what), ""))
-  ) %>%
-  mutate_at(vars(`Died within 90 days`,
-                 `Survived at least 90 days`), zero) %>%
[…]

### 2020-03-24 reports/article.Rmd
version after international revision and improved figure resolution.

@@
 date: '`r Sys.Date()`'
-bibliography: library.bib
+bibliography: H:/zotero_lib.bib
 csl: the-lancet.csl
@@
 **OBJECTIVE:** Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality.
-However, no prediction tools based on European populations are available, diagnosis code-based instruments such as the Charlson co-morbidity index are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) class is imprecise.
+However, diagnosis code-based instruments such as the Charlson co-morbidity index are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) class is imprecise.
 We thus developed a simple model to predict 90-day mortality after elective THA.
@@
 We propose a simple main effects model combining `r coefs_print_reduced`.
-This model had good discriminatory ability both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`, rendering it superioir to traditional models based on the ASA class `r get_auc("ASA")`, the Charlson co-morbitiy index `r get_auc("Charlson")`, and the Elixhauser com-morbidity index `r get_auc("Elixhauser")`.
+This model had good discriminatory ability both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`, rendering it superioir to traditional models based on the Charlson `r get_auc("Charlson")`, and the Elixhauser co-morbidity indices `r get_auc("Elixhauser")`.
 Our model was well calibrated for predicted probabilities up to 5 %.
@@
 
-Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery.[@Bozic2013a; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] Due to limited, often single-center based samples, a lack of reported calibration abilities, the absence of external validation and the use of extrapolation there is still no widely accepted model for the prediction of early mortality.[@Manning2016] Importantly, to our knowledge, no reported model has been developed or validated on a European population.
+Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery.[@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] The first prediction model based on a European cohort was the Patient Decision support tool for joint remplacement in the UK (https://jointcalc.shef.ac.uk/).
+Due to limited, often single-center based samples, a lack of reported calibration abilities, the absence of external validation, and the use of extrapolation, there is still room for improved prediction models of early mortality.[@Manning2016]
 
@@
 Cementation is the most common fixation technique used in Sweden, thus used as inclusion criteria to increase heterogeneity.
-Only the last operated hip was accounted for in patients with bilateral THA, since those are shown to better resemble unilateral THAs.[@Bulow2019a] (Figure \@ref(fig:flowchart)).
+Only the last operated hip was accounted for in patients with bilateral THA, since those are shown to better resemble unilateral THAs.[@Bulow2020] (Figure \@ref(fig:flowchart)).
 
@@
 The last year was 2015, since we had access to co-morbidity from the national patient register up to this point.
-Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009] We used data from a variety of sources, as previously described by Cnudde et al.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98%.[@shpr2018] Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status (married/un-married/divorced/widow[er]), were also collected from LISA.
+Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009] We used data from a variety of sources, as previously described by Cnudde et al.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98%.[@Karrholm2018] Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status (married/un-married/divorced/widow[er]), were also collected from LISA.
 The Swedish National Patient Register (NPR) was used to assess co-morbidity during the year preceding index surgery.
@@
 We used the Kaplan-Meier estimator to assess unadjusted mortality after cemented THA as background knowledge.
-Further analysis was based on logistic regression since no censoring occurred within the 90 day study period.
+Further analysis was based on logistic regression, since no censoring occurred in the Swedish cohort within the first 90 days, and we assumed the same was true for patients from England and Wales.
 We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1.
+This way, the magnitude (absolute values) of their estimated coefficients would indicate variable importance on the same scale as the categorical variables.
 Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process.
@@
 We therefore considered the reduced model as superior due to simplicity.
-Traditional models performed poorly with 95 % confidence intervals not above 0.7: ASA class `r get_auc("ASA")`, the Charlson co-morbitiy index `r get_auc("Charlson")`, and the Elixhauser com-morbidity index `r get_auc("Elixhauser")` (Figure \@ref(fig:rocs) left panel and \@ref(fig:aucci)).
+Traditional models performed poorly with 95 % confidence intervals not above 0.7: the Charlson co-morbitiy index `r get_auc("Charlson")`, and the Elixhauser com-morbidity index `r get_auc("Elixhauser")` (Figure \@ref(fig:rocs) left panel and \@ref(fig:aucci)).
 
@@
 
-We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` better discriminated patients who died within 90 days after THA, from patients who survived, compared to traditional models with only ASA class or the Charlson or Elixhauser co-morbidity indices.
+We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` better discriminated patients who died within 90 days after THA, from patients who survived, compared to traditional models with the Charlson or Elixhauser co-morbidity indices.
 
@@
 A strength of the study is the nationwide design with data from national registers in both Sweden, England and Wales.
-The Swedish registers are valid with low proportions of missing data.[@Soderman2000; @shpr2018; @Soderman2001; @Ludvigsson2011] Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however.[@Spencer2012; @Williams2002; @Brennan2012] A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80%.[@Burns2012] Both in- and outpatient co-morbidity data was available for Sweden, but only in-patient data for England and Wales.
+The Swedish registers are valid with low proportions of missing data.[@Soderman2000; @Karrholm2018; @Soderman2001; @Ludvigsson2011] Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however.[@Spencer2012; @Williams2002; @Brennan2012] A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80%.[@Burns2012] Both in- and outpatient co-morbidity data was available for Sweden, but only in-patient data for England and Wales.
 We did not have acces to any data from primary care or nursing homes in any of the counties.
@@
 This was seen in the calibration plot where estimated probabilities below 1.5% underestimated the observed proportion of deaths in the external validation.
+
+We investigated wether the effect of age on mortality was non-linear using splines.
+A similar approach might be relevant for BMI as well, which we only modelled as a linear main effect.
+A U-shaped association between BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been showed for example in England and Wales.[@Mouchti2018] Our recordings of BMI might be imprecise however.[@Stommel2009] Some h○ospitals might report actual measurements but it is unknown to what extent those data are self-estmiated by the patients.
+Therefore, we aimed to avoid overfitting BMI based on too elaborate modelling.
 

### 2020-03-24 munge/08-NJR_external_validation.R

@@ -42,7 +42,7 @@ njr_surv <-
     est = mean(death),
     lo = quantile(death, .025),
     hi = quantile(death, .975),
-    mort = sprintf("%0.2f %% (95 %% CI: %0.2f-%0.2f)", est * 100, lo * 100, hi * 100)
+    mort = sprintf("%0.2f %% (95 %% CI: %0.2f to %0.2f)", est * 100, lo * 100, hi * 100)
   ) %>%
   select(mort) %>%
   pluck(1, 1)

### 2020-03-24 src/fig_auc.R

@@ -44,9 +44,10 @@ annotates <-
 
 # Figure ------------------------------------------------------------------
 
-ggplot(data_auc_ci, aes(Model, est, group = corr, color = corr)) +
-  geom_hline(yintercept = .7, color = "darkgreen", linetype = "dashed", size = 1) +
-  geom_pointrange(aes(ymin = lo, ymax = hi), size = 1) + #, position = position_dodge(.5)) +
+p <-
+  ggplot(data_auc_ci, aes(Model, est, group = corr, color = corr)) +
+  geom_hline(yintercept = .7, color = "darkgreen", linetype = "dashed", size = .5) +
+  geom_pointrange(aes(ymin = lo, ymax = hi), size = .8) + #, position = position_dodge(.5)) +
   geom_text(aes(Model, value, label = round(value, 2)), data = annotates, nudge_x = .3, size = 2) +
   coord_flip() +
   theme_minimal() +
@@ -60,4 +61,5 @@ ggplot(data_auc_ci, aes(Model, est, group = corr, color = corr)) +
 
 options(digits = digs$digits)
 
-ggsave("graphs/auc_ci.png", width = 10, height = 8, units = "cm")
+ggsave("graphs/auc_ci.png", p, width = 10, height = 8, units = "cm")
+ggsave("graphs/auc_ci.tiff", p, width = 10, height = 8, units = "cm", dpi = 1200, compression = "lzw")

### 2020-03-24 src/fig_calibration.R

@@ -28,7 +28,8 @@ cbs <-
 
 # Make figure -------------------------------------------------------------
 
-ggplot(cbs, aes(x, ymin = L, ymax = U, fill = validation)) +
+p <-
+  ggplot(cbs, aes(x, ymin = L, ymax = U, fill = validation)) +
   geom_ribbon(alpha = .3) +
   geom_abline(aes(intercept = 0, slope = 1)) +
   theme_minimal() +
@@ -43,4 +44,5 @@ ggplot(cbs, aes(x, ymin = L, ymax = U, fill = validation)) +
     legend.title = element_blank()
   )
 
-ggsave("graphs/calibration.png", height = 10, width = 10 , units = "cm")
+ggsave("graphs/calibration.png", p, height = 10, width = 10 , units = "cm")
+ggsave("graphs/calibration.tiff", p, height = 10, width = 10 , units = "cm", dpi = 1200, compression = "lzw")

### 2020-03-27 reports/article.Rmd
version with table 2 combining SHAR and NJR. Updated Shiny app.

--- 2020-03-09 reports/article.Rmd
@@
 title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
-author: Anne Garland`*`^1,2,3^, Erik Bulow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5^, Mark Wilkinson^6^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
+author: Anne Garland`*`^1,2,3^, Erik Bülow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5^, Mark Wilkinson^6^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
 date: '`r Sys.Date()`'
-bibliography: library.bib
+bibliography: H:/zotero_lib.bib
 csl: the-lancet.csl
@@
 1.	Department of Orthopaedics, Institute of Surgical Sciences, Uppsala University Hospital, Uppsala, Sweden
-2.	Swedish Hip Arthroplasty Register, Gothenburg, Sweden
+2.	The Swedish Hip Arthroplasty Register, Gothenburg, Sweden
 3.	Department of Orthopaedics, Visby Hospital, Visby, Sweden
@@
 **OBJECTIVE:** Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality.
-However, no prediction tools based on European populations are available, diagnosis code-based instruments such as the Charlson co-morbidity index are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) class is imprecise.
-We thus developed a simple model to predict 90-day mortality after elective THA.
+However, diagnosis code-based co-morbidity instruments such as the Charlson index are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) classification is imprecise.
+We thus developed a parsimonious model to predict 90-day mortality after elective THA.
 
-**PATIENTS AND METHODS:** We studied `r format(nrow(df), big.mark = ",")` Swedish patients operated with a cemented THA due to primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` and used personal identity numbers to link data from the Swedish Hip Arthroplasty Register, the National Patient Register, the Longitudinal integrated database for health insurance and the population register.
-We used a bootstrap ranking procedure together with logistic LASSO regression to develop a prediction model for death within 90 days after surgery.
-The model was internally validated with bootstraping, and externally validated on a cohort of `r format(N_njr$tot, big.mark = ",")` patients from the National Joint Registry for England and Wales.
+**PARTICIPANTS AND METHODS:** We studied `r format(nrow(df), big.mark = ",")` Swedish patients operated with a cemented THA due to primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` and used personal identity numbers for data linkage.
+We used a bootstrap ranking procedure together with logistic least absolute shrinkage and selection operator (LASSO) regression to develop a prediction model for death within 90 days after surgery.
+The model was internally validated by bootstrapping, and externally validated on a cohort of `r format(N_njr$tot, big.mark = ",")` patients from the National Joint Registry for England and Wales.
 Model performance was assessed by the area under the receiver operation characteristics curve (AUC) for discrimination, and by calibration belt plots.
@@
   ) %>% 
-  {sprintf("(AUC = %s", gsub(" (", ", 95% CI: ",., fixed = TRUE))}
+  {sprintf("(AUC = %s", gsub(" (", ", 95 % CI: ",., fixed = TRUE))}
 }
@@
 
-**RESLUTS:** The unadjusted cumulative 90-day mortality was `r surv90d` in Sweden and `r njr_surv` in England and Wales.
+**RESULTS:** The unadjusted cumulative 90-day mortality was `r surv90d` in Sweden and `r njr_surv` in England and Wales.
 We propose a simple main effects model combining `r coefs_print_reduced`.
-This model had good discriminatory ability both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`, rendering it superioir to traditional models based on the ASA class `r get_auc("ASA")`, the Charlson co-morbitiy index `r get_auc("Charlson")`, and the Elixhauser com-morbidity index `r get_auc("Elixhauser")`.
+This model had good discriminatory ability both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`, rendering it superior to traditional models based on the Charlson `r get_auc("Charlson")`, and the Elixhauser `r get_auc("Elixhauser")` co-morbidity indices.
 Our model was well calibrated for predicted probabilities up to 5 %.
@@
 
-**CONCLUSION:** We found a relatively simple prediction model of 90-day mortality after elective cemented THA based on easily available demographic data and the presence of four defined co-morbidities.
+**CONCLUSION:** We developed a parsimonious model to predict 90-day mortality after elective cemented THA based on easily available demographic data and the presence of four defined co-morbidities.
 This may facilitate individualized risk assessment prior to one of the most common surgical interventions.
@@
 In terms of gained quality of life and cost-utility, total hip arthroplasty (THA) is an enormously successful procedure.
-90-day mortality after THA performed due to osteoarthritis is low, ranging from 0.2% to 0.6% in large joint registry studies.[@Hunt2013; @Garland2017] The risk of short-term mortality is nevertheless an essential part of a preoperative discussion between patient and surgeon.
+90-day mortality after THA performed due to osteoarthritis is low, ranging from 0.2 % to 0.6 % in large joint registry studies.[@Hunt2013; @Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.
 
-Co-morbidity is associated with a shorter remaining life span, but the Charlson and Elixhauser co-morbidity indices poorly predict mortality after THA.[@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017; @Bulow2017; @Inacio2015] Additionally, these complex co-morbidity instruments are based on information from extensive in- or out-patient databases, each measure being defined by 1,178 and 1,516 individual codes respectively from the Swedish adaptation of the international classification of diseases version 10 (ICD-10).
-Each co-morbidity index also exists in numerous versions, making interpretation and comparisons difficult.[@Brusselaers2017; @Walraven2009] The use of these indices is therefore limited to research settings but are impossible to use under clinical circumstances.
+Co-morbidity is associated with a shorter remaining life span, but the Charlson and Elixhauser co-morbidity indices poorly predict mortality after THA.[@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017; @Bulow2017; @Inacio2015] Additionally, these complex co-morbidity instruments are based on information from extensive in- or out-patient databases, each measure being defined by 1,178 and 1,516 individual International Classification of Diseases (ICD)-10 codes, respectively.
+Each co-morbidity index also exists in numerous versions, making interpretation and comparisons difficult.[@Brusselaers2017; @Walraven2009] The use of these indices is therefore limited to research settings but they are impossible to use under clinical circumstances.
 
-Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery.[@Bozic2013a; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] Due to limited, often single-center based samples, a lack of reported calibration abilities, the absence of external validation and the use of extrapolation there is still no widely accepted model for the prediction of early mortality.[@Manning2016] Importantly, to our knowledge, no reported model has been developed or validated on a European population.
+Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery.[@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] The first prediction model based on a European cohort was the patient decision support tool for joint remplacement in the UK (https://jointcalc.shef.ac.uk/), which is in the process of validation.
+Due to limited, often single-center based samples, a lack of reported calibration abilities, the absence of external validation, and the use of extrapolation, there is an urgent need for improved prediction models of early mortality.[@Manning2016]
 
@@
 
-# Patients and Methods
+# Participants and Methods
 
@@
 We focused on patients with cemented THA due to osteoarthritis.
-Cementation is the most common fixation technique used in Sweden, thus used as inclusion criteria to increase heterogeneity.
-Only the last operated hip was accounted for in patients with bilateral THA, since those are shown to better resemble unilateral THAs.[@Bulow2019a] (Figure \@ref(fig:flowchart)).
+Cementation is the most common fixation technique used in Sweden, thus used as inclusion criteria to decrease heterogeneity.
+Only the last operated hip was accounted for in patients with bilateral THA, since those are more likely to resemble patients with unilateral THA (Figure \@ref(fig:flowchart)).[@Bulow2020]
 
@@
 The patient cohort was identified from the SHAR.
-The inclusion period started 2008 since this was the first year with the American Society for Anesthesiologists (ASA) class and Body Mass Index (BMI) systematically recorded.
-The last year was 2015, since we had access to co-morbidity from the national patient register up to this point.
-Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009] We used data from a variety of sources, as previously described by Cnudde et al.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98%.[@shpr2018] Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status (married/un-married/divorced/widow[er]), were also collected from LISA.
-The Swedish National Patient Register (NPR) was used to assess co-morbidity during the year preceding index surgery.
+The inclusion period started 2008 since the American Society for Anesthesiologists (ASA) class and Body Mass Index (BMI) were systematically recorded from then on.
+The observation period ended 2015, since we had access to co-morbidity from the national patient register up to this point.
+Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009] We used data from a variety of sources, as previously described.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98 %.[@Karrholm2018] Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status (married/un-married/divorced/widow[er]), were also collected from LISA.
+The Swedish national patient register (NPR) was used to assess co-morbidity during the year preceding index surgery.
 This register contains all relevant diagnoses coded by ICD-10, as well as dates of admission and discharge for in- and outpatient episodes in all private and public hospitals.
@@
 
-Patients with BMI either missing or above 50 were excluded, as were patients with ASA class missing or above III, as well as patients with unknown educational level or type of hospital.
+Patients with missing information on BMI or a measurement above 50 were excluded, as were patients with missing information on ASA class or class IV and above, as well as patients with unknown educational level or type of hospital.
 
@@
 Similar inclusion criteria were applied to the validation cohort from England and Wales.
-We did not exclude patients based on BMI, educational level or civil status however, since those variables were not used in the final model.
-The validation cohort was created by linking NJR to the secondary uses service database from the National Health Service (NHS).
-This database includes co-morbidity recorded in the admitted patient care module of the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Brophy2006; @Burns2012; @Sinha2013; @Simmonds2014; @Thorn2016] Out-patient deaths were linked from the Office of national statistics.
-Data linkage used NHS-numbers, as well as patient names, age, sex and address.[@Sayers2016]
+We excluded no patients based on missing information on BMI, educational level, or civil status however, since those variables were not used in the final model.
+The validation cohort was created by linking NJR data to the secondary uses service database from the National Health Service (NHS).
+This database includes co-morbidity recorded in the admitted patient care module of the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Brophy2006; @Burns2012; @Sinha2013; @Simmonds2014; @Thorn2016] Out-patient deaths were linked from the office of national statistics.
+Data linkage used NHS-numbers, as well as patient name, age, sex and address.[@Sayers2016]
 
@@
 
-Co-morbidity was defined by individual ICD-10 codes grouped into 17 categories according to Charlson [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to Elixhauser.[@Elixhauser1998; @Quan2005] Some co-morbidities were identified by both Charlson and Elixhauser, and some distinct co-morbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
-We combined individual diagnostics groups to establish `r nrow(tab_categorization)` broader categories of co-morbidity (Table \@ref(tab:tabcategorization)).
+Co-morbidity was defined by individual ICD-10 codes grouped into 17 categories according to Charlson [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to Elixhauser.[@Elixhauser1998; @Quan2005] Some co-morbidities were identified in both indices, and some distinct co-morbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
+We combined individual diagnostic groups to establish `r nrow(tab_categorization)` broader categories of co-morbidity (Table \@ref(tab:tabcategorization)).
 Conditions were merged according to clinical relevance as to be recognized in a patient-physician meeting without access to external register data.
@@
 
-We used the Kaplan-Meier estimator to assess unadjusted mortality after cemented THA as background knowledge.
-Further analysis was based on logistic regression since no censoring occurred within the 90 day study period.
+We used the Kaplan-Meier estimator to assess unadjusted mortality after cemented THA.
+Further analysis was based on logistic regression, since no loss to follow up occurred in the Swedish cohort within the first 90 days, and we assumed the same was true for patients from England and Wales.
 We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1.
+Thus, the magnitude (absolute values) of their estimated coefficients would indicate variable importance on the same scale as the categorical variables.
 Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process.
@@
 The magnitude (absolute values) of those estimates were used as a measure of variable importance.
-Piece-wise linear regression was used to detect a break point where a significant drop in variable importance was observed.
-Potential predictors with variable importance above this break point were considered important and kept as model candidates.
+Piece-wise linear regression was used to detect a breaking point where a significant drop in variable importance was observed.
+Potential predictors with variable importance above this breaking point were considered important and kept as model candidates.
 The whole process was repeated 100 times.
@@
 A reduced model with variables chosen at least 33 out of the 100 times was used as a simpler alternative for comparison.
-This model, without cancer as a predictor, was also evaluated separately, considering that medical indications for THA surgery may look different for patients with cancer compared to cancer free patients.
-Univariable models with the ASA class, Charlson or Elixhauser co-morbidity indices were used for benchmarking, as well as a multivariable model with age and sex.
+This model, without cancer as a predictor, was also evaluated separately, considering that medical indications for THA surgery may be different for patients with cancer compared to patients without malignancies.
+Univariable models with the ASA class, Charlson or Elixhauser co-morbidity indices were used for benchmarking, as well as a multivariable model including only age and sex.
 Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.
-Odds ratios for the final model were estimated with 95% confidence intervals.[@Ripley2002]
+Odds ratios for the final model were estimated with 95 % confidence intervals.[@Ripley2002]
 
@@
 Each model was used to predict the probability of death within 90 days for patients from the SHAR (internal validation).
-Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.[@Fawcett2006] Models with a lower 95% confidence limit above 0.7, were considered good.
+Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.[@Fawcett2006] Models with a lower 95 % confidence limit above 0.7, were considered good.
 Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples.
 We used the bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples to adjust for optimism.[@Miller1991] Calibration bands were made to graphically assess model calibration, comparing predicted probabilities and observed proportions.[@Nattino2016] The reduced model was then evaluated externally.
-An AUC with 95% confidence interval was calculated for the model as-is.
+An AUC with 95 % confidence interval was calculated for the model as-is.
 Re-calibration of the model intercept was then performed to account for different mortality rates in Sweden compared to England and Wales.
@@
 
-Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13).
+Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13) and **XXXX in the UK (RSC2017/21 ???)**.
+Informed consent was not mandatory according to the Swedish patient data law from 2009.
+
+# Indirect Patient and Public Involvement
+
+There is a patient representative at SHAR to ensure that the patients' interests are respected in planned research.
 
@@
 
-`r format(nrow(df), big.mark = ",")` patients (age `r paste(range(df$P_Age), collapse = " - ")`, `r round(mean(df$P_Sex == "Female") * 100)`% females) were included in the derivation cohort from SHAR (Figure \@ref(fig:flowchart) left panel).
-`r with(df, sprintf("%d (%.2f%%)", sum(death90), mean(death90) * 100))` of those patients died within 90 days and no one was censored before that.
+`r format(nrow(df), big.mark = ",")` patients (age `r paste(range(df$P_Age), collapse = " - ")`, `r round(mean(df$P_Sex == "Female") * 100)` % females) were included in the derivation cohort from SHAR (Figure \@ref(fig:flowchart) left panel).
+`r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` of those patients died within 90 days and no one was censored before that.
 The unadjusted risk of 90-day mortality was therefore `r surv90d`.
 Further characteristics of the study population are presented in Table \@ref(tab:tab1).
-`r round(mean(df$CCI_index_quan_original != 0) * 100)`% and `r round(mean(df$ECI_index_sum_all  != 0) * 100)`% of the Swedish patients had at least one pre-surgery co-morbidity according to Charlson and Elixhauser.
-The proportion of patients with ASA class III was `r round(mean(df$P_ASA == 3) * 100)`%.
+`r round(mean(df$CCI_index_quan_original != 0) * 100)` % and `r round(mean(df$ECI_index_sum_all  != 0) * 100)` % of the Swedish patients had at least one pre-surgery co-morbidity according to Charlson and Elixhauser.
+The proportion of patients with ASA class III was `r round(mean(df$P_ASA == 3) * 100)` %.
 
-In addition, `r format(N_njr$tot, big.mark = ",")` patients were included for the external validation cohort (Figure \@ref(fig:flowchart) right panel).
-Their unadjusted risk of 90 day mortality was higher compared to the derivation cohort `r njr_surv`.
+In addition, `r format(N_njr$tot, big.mark = ",")` patients were included for the external validation cohort from England and Wales (Figure \@ref(fig:flowchart) right panel).
+Their unadjusted risk of 90 day mortality was `r njr_surv`, which was higher compared to the derivation cohort from Sweden.
 
@@
 
-There were `r nrow(comb_lgl)` co-morbidities that were not recorded for any patient who died in the Swedish derivation cohort: `r comb_lgl_text`.
+```{r}
+spell <- function(x)
+  c("one", "two", "three", "four", "five", "six", 
+    "seven", "eight", "nine", "ten", "eleven", "twelve")[x]
+```
+
+There were `r spell(nrow(comb_lgl))` co-morbidities that were not recorded for any patient who died in the Swedish derivation cohort: `r comb_lgl_text`.
 Those variables were therefore excluded as potential predictors prior to any statistical model derivation.
 The derived main model included `r coefs_print`.
-The reduced model, with covariates included at least 33 out of 100 times, were restricted to `r coefs_print_reduced` (Table \@ref(tab:brlprop)).
+The reduced model, with covariates included at least 33 out of 100 times, was restricted to `r coefs_print_reduced` (Table \@ref(tab:brlprop)).
 
-There were no differences between models including age as a main effect, or as modeled by restricted cubic splines with either two or three knots.
-We will therefore focus on the simpler models with age as a main effect.
-Similarly, the correction for optimism only affected the third decimals of the AUC confidence intervals.
-We will therefore omit those.
+There were no differences between models including age as a main effect, compared to those were age was modeled by restricted cubic splines with either two or three knots.
+We therefore focused on the more parsimonious models with age as a main effect.
+Similarly, the correction for optimism only affected the third decimals of the AUC confidence intervals and was therefore omitted.
 
-The main and reduced models were no different regarding discriminative power, `r sprintf("%s versus %s", gsub(")", "", get_auc("Main model"), fixed = TRUE), gsub("(", "", get_auc("Reduced model"), fixed = TRUE))`.
-We therefore considered the reduced model as superior due to simplicity.
-Traditional models performed poorly with 95 % confidence intervals not above 0.7: ASA class `r get_auc("ASA")`, the Charlson co-morbitiy index `r get_auc("Charlson")`, and the Elixhauser com-morbidity index `r get_auc("Elixhauser")` (Figure \@ref(fig:rocs) left panel and \@ref(fig:aucci)).
+```{r}
+half_paran <- function(x) 
+  gsub(
+    "(\\(AUC = )(0.\\d*)(, )(95 % CI: 0.\\d{2} to 0.\\d{2}\\))", 
+    "\\2 \\(\\4", 
+    get_auc(x), 
+    perl = TRUE
+  )
+```
+
+The main and reduced models were no different regarding discriminative power `r sprintf("%s versus %s", gsub(")", "", get_auc("Main model"), fixed = TRUE), gsub("(", "", get_auc("Reduced model"), fixed = TRUE))`.
+We therefore considered the reduced model as superior due to its simplicity.
+Traditional models performed poorly with 95 % confidence intervals not above 0.7: The Charlson co-morbitiy index had an AUC of `r half_paran("Charlson")` and the Elixhauser co-morbidity an AUC of `r gsub("\\)", ";", half_paran("Elixhauser"))` Figure \@ref(fig:rocs) left panel and Figure \@ref(fig:aucci)).
 
 The ability of the reduced model to estimate probabilities of death within 90 days is further illustrated in Figure \@ref(fig:sep).
-Patients who died had, on average, higher predicted probabilities to do so.
-The model calibration was good for estimated probabilities up to 3% and acceptable up to 5%, although with predicted probabilities usually higher then observed proportions of deaths among patients with corresponding covariate patterns (Figure \@ref(fig:calibration)).
+Model calibration was good for estimated probabilities up to 3 % and acceptable up to 5 %, although with predicted probabilities usually higher than observed proportions (Figure \@ref(fig:calibration)).
 
-Estimated coefficients and corresponding odds ratios for the reduced model are presented in Table \@ref(tab:brlreduced).
+Estimated model coefficients and corresponding odds ratios for the reduced model are presented in Table \@ref(tab:brlreduced).
 
-To omit cancer from the reduced model did not affect the AUC or calibration for estimated probabilities below 3%.
-It did lead to worse calibration outside this range however, wherefore we recommend including cancer as a predictor.
+Omitting cancer from the reduced model did not affect the AUC or calibration for estimated probabilities below 3 %, but calibration outside this range deteriorated, and we thus retained cancer as an important predictor.
 
@@
 
-The discriminative ability for the reduced model was not statistically significantly different when applied to the external validation cohort `r get_auc("NJR (Reduced model)")` compared to the internal `r get_auc("Reduced model")` (Figure \@ref(fig:rocs) right panel and Figure \@ref(fig:aucci)).
-Calibration of the re-calibrated model was slightly worse compared to the internal calibration.
-Predicted probabilities between 0.5 % and 1.5 % were lower than observed and estimated 95 % confidence bands were wider.
-Over-all, calibration was still good however for predicted probabilities below 3% and acceptable below 5% (Figure \@ref(fig:calibration)).
+The discriminative ability for the reduced model was not statistically significantly different when applied to the external validation cohort `r get_auc("NJR (Reduced model)")` compared to the internal validation `r get_auc("Reduced model")` (Figure \@ref(fig:rocs) right panel and Figure \@ref(fig:aucci)).
+Calibration of the re-calibrated model was slightly inferior compared to the internal calibration.
+Predicted probabilities between 0.5 % and 1.5 % were lower than observed proportions and the estimated 95 % confidence bands were wider.
+Over-all, calibration was still good however for predicted probabilities below 3 % and acceptable below 5 % (Figure \@ref(fig:calibration)).
 
@@
 
-We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` better discriminated patients who died within 90 days after THA, from patients who survived, compared to traditional models with only ASA class or the Charlson or Elixhauser co-morbidity indices.
+## Principal findings
 
-The resulting (reduced) model predicts the probability of death within 90 days as: $\hat p = 1 / [1 +\exp(-\hat \beta X)]$ where $\hat \beta$ is the vector of estimated coefficients presented in table \@ref(tab:brlreduced).
-This formula is considered valid for patients aged `r paste(range(df$P_Age), collapse = " - ")` years and for predicted probabilities up to 5%.
+We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` was considerably better at predicting death within 90 days after THA when compared to traditional models based on the Charlson or Elixhauser co-morbidity indices.
+
+The resulting model predicted the probability of death within 90 days as: $\hat p = 1 / [1 +\exp(-\hat \beta X)]$ where $\hat \beta$ is the vector of estimated coefficients presented in table \@ref(tab:brlreduced).
+This formula is considered valid for patients aged `r paste(range(df$P_Age), collapse = " - ")` years and for predicted probabilities up to 5 %.
 
@@
 
-## Practical usage
+## Clinical usage
 
-Our model could be used in clinical practice, either by the formula above, or by a simple web calculator online (`r shiny_url`).
-For example a `r min(df$P_Age)` year old woman with ASA class I and none of the important co-morbidities would have a `r get_risk("baseline")` risk to die within 90 days of surgery.
-Another woman, `r quantile(df$P_Age, .25)` years old (the first age quantile), would have an elevated risk of `r get_risk("healthy_female")`.
-A `r max(df$P_Age)` years old man (the maximum observed age) with ASA class III and cancer, would have a risk of `r get_risk("sickest_man")`.
-Note however that covariate patters with observed probabilities above 5% were rare (`r mean(pred_probs > 5) * 100`%, n = `r sum(pred_probs > 5)`).
-Estimated mortality risks above 5% are therefore subject to extrapolation.
+Our model could be used in clinical practice, either by the formula above, or by accessing a simple web calculator online (`r shiny_url`).
+For example, a `r min(df$P_Age)` year old woman with ASA class I and none of the important co-morbidities would have a `r get_risk("baseline")` risk to die within 90 days of surgery.
+A woman, `r quantile(df$P_Age, .25)` years old, i.e.
+the first age quantile, would have an elevated risk of `r get_risk("healthy_female")`.
+A `r max(df$P_Age)` year old man, i.e.
+the maximum observed age, with ASA class III and cancer, would have a risk of `r get_risk("sickest_man")`.
+Note, however, that covariate patters with observed probabilities above 5 % were rare (`r mean(pred_probs > 5) * 100` %, n = `r sum(pred_probs > 5)`), and estimated mortality risks above 5 % are therefore subject to extrapolation.
 The observed proportions of deaths for patients with similar characteristics is likely lower (as indicated by Figure \@ref(fig:calibration)).
@@
 
-## Model predictors
+Variables in our model were chosen based on statistical properties and not due to clinical relevance.
+Variables and estimated coefficients should therefore not be assigned any exact epidemiological and/or causal meaning per se.[@Breiman2001]
 
-Variables in our model were chosen based on statistical properties and not due to clinical relevance.
-Variables and estimated coefficients should therefore not be assigned any exact epidemiological and/or causal meaning per se.[@Breiman2001] We could nevertheless discuss some aspects of the relevant predictors.
-
-Age and sex are well-known predictors of remaining life span.
-It is less obvious that this relation must be linear.
+Age and sex are well-known predictors of remaining life span, but it is less obvious that this relation must be linear.
 We used restricted cubic splines to allow a more flexible relation, but found no difference in AUC compared to simpler main effect models.
@@
 This is clinically reasonable since a label of "severe systemic disease" should be based on a relevant patient assessment prior to surgery.
-Conditions were not too severe however since elective surgery was performed after this assessment.
 ASA class is nevertheless known to have a high degree of internal variability.[@Haynes1995] It has previously been compared to the Charlson co-morbidity index, but not with respect to mortality after THA.[@Whitmore2014; @Kork2015] Patients with ASA class IV-VI were excluded since those categories describe severe disease, moribund and brain-dead individuals.
 We suspect that patients with such conditions might have been misclassified.
-Their true class could either be lower, or they could have received THA due to a femoral hip fracture instead of osteoarthritis.
 
-Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We noted that the proportion of patients with BMI above 30, the WHO definition of obesity is much higher (`r mean(df$P_BMI > 30, na.rm = TRUE) * 100`%) than the proportion of patients with ICD-10 = E66 (diagnosed obesity) from the National Patient Register (`r mean(df$c_obesity) * 100`%).
-This proportion corresponds to patients with BMI of at least `r which.min(map_dbl(1:50, ~mean(df$P_BMI > ., na.rm = TRUE)) > mean(df$c_obesity))`.
-Thus, the relevance of obesity might not be the condition itself, but rather all circumstances making the related diagnose appear in the national patient register.
-This might include unmeasured confounders related to patients, doctors, and local hospital guidelines or similar.
+Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We noted that the proportion of patients with BMI above 30, the WHO definition of obesity was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) in the Swedish cohort than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the national patient register (`r round(mean(df$c_obesity) * 100, 1)` %).
+We believe this might be due to the fact that BMI is systematically recorded in the SHAR, but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.
 
@@
 
-A strength of the study is the nationwide design with data from national registers in both Sweden, England and Wales.
-The Swedish registers are valid with low proportions of missing data.[@Soderman2000; @shpr2018; @Soderman2001; @Ludvigsson2011] Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however.[@Spencer2012; @Williams2002; @Brennan2012] A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80%.[@Burns2012] Both in- and outpatient co-morbidity data was available for Sweden, but only in-patient data for England and Wales.
-We did not have acces to any data from primary care or nursing homes in any of the counties.
+A strength of this study is its nationwide design with data from both Sweden, England and Wales.
+The Swedish registers are valid with low proportions of missing data.[@Soderman2000; @Karrholm2018; @Soderman2001; @Ludvigsson2011] Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however.[@Spencer2012; @Williams2002; @Brennan2012] A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80 %.[@Burns2012] Both in- and outpatient co-morbidity data were available for Sweden, but only in-patient data for England and Wales.
+We did not have access to any data from primary care or nursing homes in any of the counties.
 
-The linkage procedure between NJR and HES was previously described by 
-Smith et al.[@Smith2012] They found that privately founded patients were not included in HES.
-17% of the patients had missing personal data or did not allow linkage, and 6% were not found in HES even if their data was available from NJR.
+The linkage procedure between NJR and HES has been previously described,[@Smith2012] where it was noted that privately funded patients were not included in HES.
+17 % of the patients had missing personal data or did not allow linkage, and 6 % were not found in HES although their data were available from the NJR.
 
-In addition to data validity, the indication for different fixation techniques also differ in Sweden compared to England and Wales.
-Cementation is most commonly used in Sweden, but not so in England and Wales, where relatively younger patients are more likely operated without cement.
+In addition to data validity, the indications for choosing between different fixation techniques differ in Sweden compared to England and Wales.
+Cementation is the most commonly used fixation principle in Sweden but not so in England and Wales, where younger patients are more likely to be operated with cementless implants.
 To only include cemented THA therefore implies an over-representation of older and frailer patients from England and Wales compared to Sweden.
-Those patients might be more likely to die within 90 days.
-This was seen in the calibration plot where estimated probabilities below 1.5% underestimated the observed proportion of deaths in the external validation.
+Those patients might be more likely to die within 90 days, and this assumption would be in agreement with the calibration plot where estimated probabilities below 1.5 % underestimated the observed proportion of deaths in the external validation cohort from the NJR.
+
+We investigated whether the effect of age on mortality was non-linear using restricted cubic splines.
+A similar approach might be relevant for BMI, which we only modelled as a linear main effect.
+A U-shaped association between mortality and BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales.[@Mouchti2018] Our recordings of BMI might be imprecise however.[@Stommel2009] Some hospitals might report actual measurements but it is unknown to what extent those data are self-estimated by the patients or estimated by health care personnel.
+Therefore, we aimed to avoid overfitting BMI based on too elaborate transformations.
 
@@
 Hence, deaths within 90 days might occur for those patients regardless if THA is inserted or not.
-The proximity in time however, the maximum of 90 days from THA to death, is an indication that the operation might have influenced the observed mortality.
-Insertion of an elective THA is always preceded by a clinical judgement.
-Hence, no patient with a foreseen death near-by is given elective THA.
-We therefore believe that at least a non-significant proportion of deaths within 90 days are related to the THA itself.
+The proximity in time however, the maximum of 90 days from THA to death, is an indication that the operation might be the main cause of death for the majority of deceased patients.
 
@@
 Our results indicate that the risk of early postoperative mortality after cemented THA can be pre-operatively assessed by a parsimonious prediction model.
-We hope that this model, with its accompanying web calculator, will facilitate a shared decision making between physicians and their patients in need of THA.
+We hope that this model, with its accompanying web calculator, will facilitate shared decision-making between physicians and their patients in need of THA.
 
@@
 
-**[ADD NJR CONTRIBUTIONS!]** AG and NH initiated the study and managed the ethical review board application.
-EB developed the statistical model.
-EL and AS performed external validation with data from NJR.
-AG and EB drafted the manuscript.
-All authors edited and finalized the manuscript.
+**[ARE THOSE CORRECT?]** AG and NH initiated the study and managed the ethical review board application in Sweden.
+AB and MW conceptualized the external validation and managed the ethical review board application in the UK.
+EB developed the statistical model. EL and AS performed external validation with data from NJR. AG and EB drafted, and all authors edited and finalized the manuscript.
+
+# Competing interests statement
+
+All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi_disclosure.pdf and declare **[... SUMMARIZE STATEMANT AFTER ALL COI:s RETURNED]**
+
+# Transparency Statement
+
+NH, the principal investigator of this study, affirms that this is an accurate report with no relevant data or aspects omitted.
+
+# Data sharing statement
+
+Data is available upon request to the principal investigator.
 
@@
 
+# Summary box
+
+What is already known:
+
+- Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality.
+- Available diagnosis code-based co-morbidity instruments are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) classification is imprecise.
+
+What this study adds:
+
+- Our results indicate that the risk of early postoperative mortality after cemented THA can be pre-operatively assessed by a parsimonious prediction model. 
+- The presented model, with its accompanying web calculator, could facilitate a shared decision-making between physicians and their patients in need of THA.
+
 ##### PAGE BREAK
 
-```{r, child = "figs_tabs.Rmd"}
+```{r, child = "figs.Rmd"}
+```
+
+```{r, child = "tabs.Rmd"}
 ```

### 2020-03-27 reports/figs.Rmd
version with table 2 combining SHAR and NJR. Updated Shiny app.

<!-- FIGURES -->

```{r flowchart, fig.cap = "Flowchart depicting inclusion criteria and number of patients. Data from the Swedish Hip Arthroplasty Register were used for model derivation and internal validation (left). Data from the National Joint Registry in England and Wales were used for external validation (right)."}
knitr::include_graphics("../graphs/flowchart.png")
```

##### PAGE BREAK

```{r rocs, fig.cap = "Receiver Operation Characteristics (ROC) curves. The area under the ROC curve (AUC) for the reduced model derived on data from the SHAR was very similar to the main model (left panel). External validation of this model, on data from the National Joint Registry for England and Wales, produced a similar curve as well (right panel)."}
knitr::include_graphics("../graphs/roc.png")
```

##### PAGE BREAK

```{r aucci, fig.cap = "Area Under the Receiver Operation Characteristics Curve (AUC) as a measure of predictive discriminative ability with 95 % bootstrap confidence intervals. AUC above 0.7 were considered good. The reduced model was similar to the main model, and not statistically significantly inferior when used with external data from the NJR."}

knitr::include_graphics("../graphs/auc_ci.png")
```

##### PAGE BREAK

```{r sep, fig.cap = "The vast majority of patients survived more than 90 days after THA, and therefore the blue bars dominate the histogram (upper panel; note the scales). A normalized density plot however reveals that patients who died within 90 days were, on average, estimated to have a higher probability to do so (lower panel). There were very few observations with covariate patterns resulting in death probabilities higher than 5 % (green line). Estimated probabilities above this limit are therefore subject to extrapolation."}
knitr::include_graphics("../graphs/separation.png")
```

##### PAGE BREAK

```{r calibration, fig.cap = "This figure illustrates calibration between observed proportions and predicted probabilities with 95 % confidence intervals. Deviations above the diagonal line indicated that the model under-estimated the probability of death within 90 days. Such deviation was statistically significant for the external validation for predictions below 1.5 %. In contrast, with higher predicted probabilities, the model over-estimated the observed proportions of deaths."}
knitr::include_graphics("../graphs/calibration.png")
```

### 2020-03-27 reports/tabs.Rmd
version with table 2 combining SHAR and NJR. Updated Shiny app.

--- 2020-03-09 reports/figs_tabs.Rmd
@@
-<!-- FIGURES -->
-
-```{r flowchart, fig.cap = "Flowchart depicting inclusion criteria and number of patients.
-Data from the Swedish Hip Arthroplasty Register was used for model derivation and internal validation (left).
-Data from the National Joint Registry in England and Wales was used for external validation (right)."}
-knitr::include_graphics("../graphs/flowchart.png")
-```
-
 ##### PAGE BREAK
 
-```{r rocs, fig.cap = "Receiver Operation Characteristics (ROC) curves.
-The area under the ROC curve (AUC) for the reduced model was no different compared to the main model (left panel).
-The reduced model works almost as good for predictions in the National Joint Registry for England and Wales (right panel)."}
-knitr::include_graphics("../graphs/roc.png")
-```
-
-##### PAGE BREAK
-
-```{r aucci, fig.cap = "Area Under the Receiver Operation Characteristics Curve (AUC) as a measure of predictive discriminative ability with 95 % bootstrap confidence intervals.
-The reduced model was no different compared to the main model, neither internally, nor externally (overlapping confidence intervals).
-AUC above 0.7 was considered good.
-This was observed for the derived models but not for any other model included for comparison."}
-
-knitr::include_graphics("../graphs/auc_ci.png")
-```
-
-##### PAGE BREAK
-
-```{r sep, fig.cap = "Most patients survived more than 90 days after THA, wherefore the blue bars totally dominate the histogram (upper panel; note the scales).
-A normalized density plot however reveals that patients who died within 90 days were, on average, estimated to have a higher probability to do so (lower panel).
-There were very few observations with covariate patterns resulting in death probabilities higher than 5% (green line).
-Estimated probabilities above this limit are therefore subject to extrapolation."}
-knitr::include_graphics("../graphs/separation.png")
-```
-
-##### PAGE BREAK
-
-```{r calibration, fig.cap = "This figure illustrates calibrartion between observed proportions and predicted probabilities with 95 % confidence intervals.
-Deviations above the diagonal line means that the model underpredicts the probability of death within 90 days.
-Such deviation was statistically significant for the external validation for predictions below 1.5%.
-The higher the predicted probability gets, the more they over-estimate the observed proportions of deaths."}
-knitr::include_graphics("../graphs/calibration.png")
-```
-
-##### PAGE BREAK
-
-<!-- TABLES -->
-
 ```{r tabcategorization}
+tab_categorization %>%
+  ungroup() %>% 
+  mutate(
+    `Comorbidities by groups` = 
+      gsub("Heart infarct", "Myocardial infarction", `Comorbidities by groups`)
+  ) %>% 
   knitr::kable(
-  tab_categorization, 
-  caption = "Categorization of co-morbdities based on diagnostic groups according to Charlson and Elixhauser. (CNS = central nervous system.)")
+    caption = "Categorization of co-morbidities based on diagnostic groups according to Charlson and Elixhauser. (CNS = central nervous system.)")
 
@@
 knitr::kable(
-  table1, 
-  caption = "Characteristics of the study population.
-Co-morbidities recorded for at least one patient who survived 90 days, and one who did not, were modeled as potential predictors.
-(CNS = central nervous system.)")
+  mutate(table1_combined_tot, what = gsub("Heart infarct", "Myocardial infarction", what)), 
+  caption = "Characteristics of the study population in the model derivation cohort (SHAR) and external validation cohort (NJR).
+Educational levels were classified as low (up to 9 years), middle (10-12 years) and high (at least 12 years).
+BMI = Body mass index.
+ASA = American Society for Anesthesiologists classification.
+CNS = central nervous system.")
 ```
@@
 
-<!-- Same info in figure so do not use this table -->
-```{r auc, warning = FALSE, include = FALSE}
-knitr::kable(
-  brlasso_tbl_auc %>% 
-  filter(
-    `Age as` %in% c(NA, "main effect"),
-    !endsWith(Model, "cancer") 
-  ) %>% 
-  select(Model, AUC) %>% 
-  rename(`AUC (95 % CI)` = AUC),
-  caption = "Area Under the Curve (AUC) as a measure of discriminative ability for the main and reduced models, a model with age and sex, as well as univariable models with ASA class and the Charlson or Elixhauser co-morbidity indices.")
-```
-
 ##### PAGE BREAK
@@
 knitr::kable(
-  caption = 'Estimated coefficients (beta) and odds ratios (OR) with 95% confidence intervals for the reduced model.'
+  caption = 'Estimated coefficients (beta) and odds ratios (OR) with 95 % confidence intervals for the reduced model.'
 )

### 2020-03-27 shiny/clean_names_copy.R

@@ -24,5 +24,16 @@ clean_names <- function(x, firstupper = TRUE, lvls = TRUE) {
 
   x <- ifelse(grepl("aids", x, ignore.case = TRUE), "AIDS/HIV", x)
   x <- ifelse(grepl("cns", x, ignore.case = TRUE), "CNS disease", x)
-  gsub("GRADE", "grade", x)
+  gsub("GRADE", "class", x, ignore.case = TRUE)
+}
+
+
+model_names <- function(x, age = FALSE) {
+  if (!age)
+    x <- gsub(" \\(age as main effect\\)", "", x)
+
+  x <- gsub("ASA", "ASA class", x)
+
+  gsub("BRL reduced", "Reduced model", x) %>%
+    {gsub("BRL", "Main model", .)}
 }

### 2020-03-27 shiny/ui.R

@@ -1,15 +1,26 @@
 library(shiny)
 library(shinythemes)
+library(shinyBS)
 library(tidyverse)
+library(flexdashboard)
 
 source("misc.R")
 
 fluidPage(
-  titlePanel("Probability to survive at least 90 days after THA surgery?"),
 
-  helpText("Bla bla bla ..."),
+  navbarPage(
+    tags$style(HTML(".navbar-default .navbar-nav > li > a {color:white;}")),
+    tabPanel("Svenska", icon = icon("stripe-s"), value = "se"),
+    tabPanel("English", icon = icon("flag"), value = "en"),
+    id = "lang"
+  ),
+
+  titlePanel("90-day survival after total hip replacement"),
+
+  helpText(textOutput("text_intro")),
 
   sidebarPanel(
+    p("Please provide some personal details so we can estimate your probability:"),
     sliderInput(
       'P_Age',
       'Age',
@@ -20,6 +31,14 @@ fluidPage(
       round = TRUE
     ),
 
+    bsPopover(
+      "P_Age",
+      "Age at surgery",
+      "How old will you be at the time of your planned hip replacement?",
+      placement = "right",
+      options = list(container = "body")
+    ),
+
     radioButtons(
       "P_Gender",
       "Sex",
@@ -28,29 +47,60 @@ fluidPage(
       inline = TRUE,
     ),
 
-    radioButtons(
+    bsPopover(
+      "P_Gender",
+      "Legal gender",
+      "What is your legal gender? (Our model was based on legal gender identified by the third digit of Swedish personal identity numbers.)",
+      placement = "right",
+      options = list(container = "body")
+    ),
+
+
+   radioButtons(
       "P_ASA",
-      "ASA grade",
-      choices = 1:3,
+      tagList("ASA class", a("(?)", href = "https://www.asahq.org/standards-and-guidelines/asa-physical-status-classification-system")),
+      choices = c(I = 1, II = 2, III = 3),
       selected = 1,
       inline = TRUE,
+    ),
+    bsPopover(
+      "P_ASA",
+      "ASA Class",
+      "Do you have: normal health (I), a mild systemic disease (II), or a severe systemic disease (III)? Your physician might help you with this assessment.",
+      placement = "right",
+      options = list(container = "body")
     ),
 
     checkboxGroupInput(
       "checkboxes",
       "Do you have any of the following:",
       choices = coefs$coef_present[coefs$checkbox],
-    )
+    ),
+
+    bsPopover(
[…]

### 2020-03-27 src/tab_define_groups.R

@@ -1,7 +1,5 @@
 suppressMessages({library(ProjectTemplate); load.project()})
 
-firstupper <-
-
 tab_categorization <-
   categorization.Blad1 %>%
   mutate_all(zoo::na.locf) %>%
@@ -12,6 +10,8 @@ tab_categorization <-
     `Comorbidities by groups` = new,
     Charlson = CCI,
     Elixhauser = ECI
-  )
+  ) %>%
+  ungroup() %>%
+  mutate_all(~gsub("Aids/hiv", "AIDS/HIV", .))
 
 cache("tab_categorization")

### 2020-03-30 reports/article.Rmd
Ethical details. Correct affiliations and minor edits.

@@
 title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
-author: Anne Garland`*`^1,2,3^, Erik Bülow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5^, Mark Wilkinson^6^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
+author: Anne Garland`*`^1,2,3^, Erik Bülow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5,6^, J. Mark Wilkinson^7^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
 date: '`r Sys.Date()`'
@@
 5. Translational Health Sciences, Bristol Medical School, University of Bristol, UK.
-6. Department of Oncology and Metabolism, University of Sheffield, Sheffield, UK.
+6. The National Institute of Health Research Biomedical Research Centre Bristol
+7. Department of Oncology and Metabolism, University of Sheffield, Sheffield, UK.
 
@@
 
-Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery.[@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] The first prediction model based on a European cohort was the patient decision support tool for joint remplacement in the UK (https://jointcalc.shef.ac.uk/), which is in the process of validation.
-Due to limited, often single-center based samples, a lack of reported calibration abilities, the absence of external validation, and the use of extrapolation, there is an urgent need for improved prediction models of early mortality.[@Manning2016]
+Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery.[@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] The first prediction model based on a European cohort was the patient decision support tool for joint remplacement in the UK (https://jointcalc.shef.ac.uk/).
+However, it remains unclear to what level the depth of co-morbidity characterization influences the accuracy of  mortality prediction models.
 
-We therefore aimed to develop a parsimonious prediction model of 90-day postoperative mortality after THA with internal and external validation of discrimination and calibration, and to compare this to the accuracy of prediction models based on existing co-morbidity measures.
+Here we aimed to develop a parsimonious prediction model of 90-day postoperative mortality after THA with internal and external validation of discrimination and calibration, and to compare this to the accuracy of prediction models based on existing co-morbidity measures.[@Manning2016]
 
@@
 
-Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13) and **XXXX in the UK (RSC2017/21 ???)**.
-Informed consent was not mandatory according to the Swedish patient data law from 2009.
+Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13) and by the NJR in the UK (RSC2017/21).
+Informed consent was not mandatory according to the Swedish patient data law from 2009, and the UK law for pseudonymised data.
 
@@
 
-**[ARE THOSE CORRECT?]** AG and NH initiated the study and managed the ethical review board application in Sweden.
-AB and MW conceptualized the external validation and managed the ethical review board application in the UK.
+AG and NH initiated the study and managed the ethical review board application in Sweden. AB and JMW conceptualized the external validation and managed the ethical review board application in the UK. 
 EB developed the statistical model. EL and AS performed external validation with data from NJR. AG and EB drafted, and all authors edited and finalized the manuscript.
+
+# Exclusive licence
+
+We grant a worldwide licence to the publishers and its licensees in perpetuity, in all forms, formats and media (whether known now or created in the future), to i) publish, reproduce, distribute, display and store the Contribution, ii) translate the Contribution into other languages, create adaptations, reprints, include within collections and create summaries, extracts and/or, abstracts of the Contribution and convert or allow conversion into any format including without limitation audio, iii) create any other derivative work(s) based in whole or part on the on the Contribution, iv) to exploit all subsidiary rights to exploit all subsidiary rights that currently exist or as may exist in the future in the Contribution, v) the inclusion of electronic links from the Contribution to third party material where-ever it may be located; and, vi) licence any third party to do any or all of the above.
 
@@
 
-All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi_disclosure.pdf and declare **[... SUMMARIZE STATEMANT AFTER ALL COI:s RETURNED]**
+All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi_disclosure.pdf and declare no financial relationships with any organisations that might have an interest in the submitted work in the previous three years, and no other relationships or activities that could appear to have influenced the submitted work.
 

### 2020-04-01 reports/article.Rmd
Add Zenodo repo
Updated Shiny app with help texts in English and Swedish.

@@
 
-**OBJECTIVE:** Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality.
-However, diagnosis code-based co-morbidity instruments such as the Charlson index are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) classification is imprecise.
-We thus developed a parsimonious model to predict 90-day mortality after elective THA.
+**OBJECTIVES** We developed and validated a statistical model to predict 90-day mortality after total hip arthroplasty (THA) due to osteoarthritis.
 
-**PARTICIPANTS AND METHODS:** We studied `r format(nrow(df), big.mark = ",")` Swedish patients operated with a cemented THA due to primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` and used personal identity numbers for data linkage.
-We used a bootstrap ranking procedure together with logistic least absolute shrinkage and selection operator (LASSO) regression to develop a prediction model for death within 90 days after surgery.
-The model was internally validated by bootstrapping, and externally validated on a cohort of `r format(N_njr$tot, big.mark = ",")` patients from the National Joint Registry for England and Wales.
-Model performance was assessed by the area under the receiver operation characteristics curve (AUC) for discrimination, and by calibration belt plots.
+**DESIGN** A multi-national observational registry study with prospective data collection and data linkage.
+We used a bootstrap ranking procedure together with logistic LASSO-regression, evaluated by the area under the receiver operation characteristics curve (AUC), and calibration belt plots.
+
+**SETTING** Patients operated `r paste(range(df$P_SurgYear), collapse = " - ")` and identified through the Swedish Hip Arthroplasty Register, including all private and public hospitals in Sweden, for model derivation and internal validation.
+Similar patients from England and Wales,  recorded in the National Joint Register, for external model validation.
+
+```{r}
+shar_tot <- nrow(df)
+shar_female <- df$P_Sex == "Female"
+age_shar <- sprintf("%.0f [%.0f - %.0f]", mean(df$P_Age), min(df$P_Age), max(df$P_Age))
+sex_shar <- 
+  sprintf(
+    "%.0f %% [%s of %s]", 
+    mean(shar_female) * 100, 
+    format(sum(shar_female), big.mark = ","), 
+    format(shar_tot, big.mark = ",")
+)
+deaths_shar <- 
+  sprintf(
+    "%.02f %% [%.0f of %s]", 
+    mean(df$death90) * 100, 
+    sum(df$death90), 
+    format(nrow(df), big.mark = ",")
+  )
+
+shar_summary <- sprintf(
+  "(mean [range] age %s; %s females; %s deaths within 90 days)",
+  age_shar, sex_shar, deaths_shar
+)
+
+extr <- function(col, reg) 
+  subset(table1_combined_tot, startsWith(what, col), reg, drop = TRUE)
+extr_perc <- function(...) 
+  gsub("([\\d,]*)( )(\\()(\\d*)(\\.\\d*)(\\))", "\\4 \\% \\[\\1 of", extr(...), perl = TRUE)
+
+njr_age_mean <- as.numeric(gsub("(.*) (.*)", "\\1", extr("Age", "NJR"), perl = TRUE))
+# Same range by design (to match Swedish cohort)
+njr_age <- sprintf("%.0f [%.0f - %.0f]", njr_age_mean, min(df$P_Age), max(df$P_Age))
+njr_sex <- sprintf("%s %s]", extr_perc("Sex", "NJR"), format(N_njr$tot, big.mark = ","))
+njr_deaths <- 
+  sprintf(
+    "%.02f %% [%.0f of %s]", 
+    (N_njr$death / N_njr$tot) * 100, 
+    N_njr$death, 
+    format(N_njr$tot, big.mark = ",")
+  )
+njr_summary <- sprintf(
+  "(mean [range] age %s; %s females; %s deaths within 90 days)",
+  njr_age, njr_sex, njr_deaths
+)
+```
+
+**PARTICIPANTS** `r format(nrow(df), big.mark = ",")` patients with cemented total hip arthroplasty (THA) due to osteoarthritis, and ASA class I-III, for model derivation and internal validation `r shar_summary`.
+An additional `r format(N_njr$tot, big.mark = ",")` patients for external validation `r njr_summary`.
+
+**MAIN OUTCOME MEASURES** Predicted probability of death due to any reason within 90 days after THA.
 
@@
 
-**RESULTS:** The unadjusted cumulative 90-day mortality was `r surv90d` in Sweden and `r njr_surv` in England and Wales.
-We propose a simple main effects model combining `r coefs_print_reduced`.
-This model had good discriminatory ability both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`, rendering it superior to traditional models based on the Charlson `r get_auc("Charlson")`, and the Elixhauser `r get_auc("Elixhauser")` co-morbidity indices.
-Our model was well calibrated for predicted probabilities up to 5 %.
+**RESULTS** A simple main effects model combining `r coefs_print_reduced`, with good internal discrimination `r get_auc("Reduced model")`, as well as externally `r get_auc("NJR (Reduced model)")`.
+This model was superior to traditional models based on the Charlson `r get_auc("Charlson")`, and the Elixhauser `r get_auc("Elixhauser")` co-morbidity indices.
+The model was well calibrated for predicted probabilities up to 5 %.
 A web calculator to aid clinical usage was published at `r shiny_url`.
@@
 This may facilitate individualized risk assessment prior to one of the most common surgical interventions.
+
+##### PAGE BREAK
+
+# Summary box
+
+What is already known:
+
+- Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality.
+- Available diagnosis code-based co-morbidity instruments are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) classification is imprecise.
+
+What this study adds:
+
+- Our results indicate that the risk of early postoperative mortality after cemented THA can be pre-operatively assessed by a parsimonious prediction model. 
+- The presented model, with its accompanying web calculator, could facilitate a shared decision-making between physicians and their patients in need of THA.
 
@@
 We used `r substr(R.version.string, 1, 15)` (R Foundation for Statistical   Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC and shiny.
-All R-scripts and necessary configurations (but no personal data) is available at zenodo.org/XXX **[Add!]**.
+All R-scripts and necessary configurations (but no personal data) is available at http://doi.org/10.5281/zenodo.3732852.
 A linked Binder environment is also available for interactive usage.
@@
 
+##### PAGE BREAK
+
 # Contribution of authors
 
-AG and NH initiated the study and managed the ethical review board application in Sweden. AB and JMW conceptualized the external validation and managed the ethical review board application in the UK. 
+AG and NPH initiated the study and managed the ethical review board application in Sweden.
+AB and JMW conceptualized the external validation and managed the ethical review board application in the UK.
 EB developed the statistical model. EL and AS performed external validation with data from NJR. AG and EB drafted, and all authors edited and finalized the manuscript.
@@
 
-We would like to thank Szilard Nemes, previous senior statistician at the Swedish Hip Arthroplasty Register, for involvement in planning and interpreting the study.
-
-# Summary box
-
-What is already known:
-
-- Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality.
-- Available diagnosis code-based co-morbidity instruments are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) classification is imprecise.
-
-What this study adds:
-
-- Our results indicate that the risk of early postoperative mortality after cemented THA can be pre-operatively assessed by a parsimonious prediction model. 
-- The presented model, with its accompanying web calculator, could facilitate a shared decision-making between physicians and their patients in need of THA.
+We thank the patients and staff of all the hospitals in England, Wales and Northern Ireland who have contributed data to the National Joint Registry.
+We are grateful to the Healthcare Quality Improvement Partnership (HQIP), the NJR Research Committee and staff at the NJR Centre for facilitating this work.
+We also thank all contributers to the Swedish Hip Arthroplasty Register (SHAR), especially Szilard Nemes, previous senior statistician, for his involvement in the study.
+The authors have conformed to the NJR's standard protocol for data access and publication.
+The views expressed represent those of the authors and do not necessarily reflect those of the National Joint Registry Steering Committee or the Healthcare Quality Improvement Partnership (HQIP) who do not vouch for how the information is presented.
 

### 2020-04-01 shiny/misc.R

@@ -1,8 +1,13 @@
+
+# Load data ---------------------------------------------------------------
+
 # file.copy("cache/fit_brl_reduced_lean.RData", "shiny/fit_brl_reduced_lean.RData", TRUE)
 # file.copy("lib/clean_names.R", "shiny/clean_names_copy.R", TRUE)
 load("fit_brl_reduced_lean.RData") # Absolute path correct for deployment
 source("clean_names_copy.R")
 
+
+# Table of coeficient names form the mdoel -------------------------------------
 coefs <-
   tibble(coef = names(coef(fit_brl_reduced_lean))[-1]) %>%
   mutate(
@@ -11,15 +16,25 @@ coefs <-
     coef_present =
       clean_names(coef) %>%
       {gsub(" = [23]", "", .)} %>%
-      {gsub("Obesity", "Diagnosed obesity", .)}
+      {gsub("Obesity", "Diagnosed obesity", .)},
+    coef_present_se =
+      # Add Swedish disease names to present if Swedish selected
+      case_when(
+        coef_present == "CNS disease" ~ "Neurologisk sjukdom",
+        coef_present == "Kidney disease" ~ "Njursjukdom",
+        coef_present == "Diagnosed obesity" ~ "Diagnostiserad fetma",
+        TRUE ~ coef_present
+      )
 
   ) %>%
   select(-coef) %>%
   distinct(coef_name, .keep_all = TRUE)
 
+# Odds ratios
 OR <- exp(coef(fit_brl_reduced_lean))
 
-# Make tibble with patient data
+
+# Make tibble with patient data ------------------------------------------------
 pred <- function(input) {
   inp <- vector("list", nrow(coefs))
   names(inp) <- coefs$coef_name
@@ -30,7 +45,15 @@ pred <- function(input) {
   patdata <-
     map(inp, ~ if (is.null(.)) FALSE else .) %>%
     as_tibble(inp) %>%
-    mutate(P_Gender = ifelse(P_Gender == "Male", "Man", "Kvinna"))
+    mutate(P_Gender = ifelse(P_Gender == "\u2642", "Man", "Kvinna"))
 
   (1 - predict(fit_brl_reduced_lean, patdata, "response")) * 100
+}
+
+
+# Get text by language and id --------------------------------------------------
+texts <- readxl::read_excel("texts.xlsx")
+
+get_text <- function(lang, id) {
+  texts[texts$id == id, lang, drop = TRUE]
 }

### 2020-04-01 shiny/server.R

@@ -1,38 +1,123 @@
 library(shiny)
 library(tidyverse)
 library(flexdashboard)
+library(shinyBS)
 source("misc.R")
 
 
 # input <- list(P_Age = 35, P_ASA = 2, checkboxes = c("cns", "cancer")) # EXAMPLE
 
 
-function(input, output) {
-  make_pred <- reactive(pred(input))
+function(input, output, session) {
 
-  output$text_intro <-
-    renderText({
-      if (input$lang == "se") {
-       "Har du höftledsartros och överväger operation med en (cementerad) totalprotes?
-        Detta är en standardiserad och väldigt säker operation.
-        Av de som opereras i Sverige är det bara 0.03 % som avlider inom 90 dagar
-        (dvs 3 av 10 000 patienter). Med ytterligare information om just dig kan
-        vi ge en mer anpassad skattning för just din sannolikhet."
-      } else {
-        "Do you have osteoarthritis of the hip? Do you consider a (cemented) total hip replacement?
-        This is a secure and standardized procedure.
-        The average risk to die within 90 days is 0.03 % in Sweden (3 deaths for every 10,000 patients).
-        You can use this tool to get a more individualized estimate."
-      }
-    })
+
+# Intro texts -------------------------------------------------------------
+
+  output$text_title     <- renderText({get_text(input$lang, "title")})
+  output$text_intro     <- renderText({get_text(input$lang, "intro")})
+  output$text_details   <- renderText({get_text(input$lang, "details")})
+  output$text_tab_p     <- renderText({get_text(input$lang, "tab_p")})
+  output$text_tab_about <- renderText({get_text(input$lang, "tab_about")})
+  output$text_your_prob <- renderText({get_text(input$lang, "your_prob")})
+
+
+# Input titles ------------------------------------------------------------
+
+  # Select co-morbidities
+  check_names <- reactive({
+    if (input$lang == "en") {
+      coefs$coef_present[coefs$checkbox]
+    } else {
+      coefs$coef_present_se[coefs$checkbox]
+    }
+  })
+
+  TIP <- reactiveValues()
+  observe({
+    TIP$age_title      <- get_text(input$lang, "help_age_title")
+    TIP$age_content    <- get_text(input$lang, "help_age_content")
+    TIP$sex_title      <- get_text(input$lang, "help_sex_title")
[…]

### 2020-04-01 shiny/ui.R

@@ -8,6 +8,9 @@ source("misc.R")
 
 fluidPage(
 
+  # https://github.com/ebailey78/shinyBS/issues/88
+  singleton(tags$head(tags$script(src = "pop_patch.js"))),
+
   navbarPage(
     tags$style(HTML(".navbar-default .navbar-nav > li > a {color:white;}")),
     tabPanel("Svenska", icon = icon("stripe-s"), value = "se"),
@@ -15,75 +18,15 @@ fluidPage(
     id = "lang"
   ),
 
-  titlePanel("90-day survival after total hip replacement"),
-
-  helpText(textOutput("text_intro")),
+  titlePanel(textOutput("text_title")),
+  helpText(p(textOutput("text_intro"))),
 
   sidebarPanel(
-    p("Please provide some personal details so we can estimate your probability:"),
-    sliderInput(
-      'P_Age',
-      'Age',
-      min   = 35,
-      max   = 99,
-      value = 72,
-      step  = 1,
-      round = TRUE
-    ),
-
-    bsPopover(
-      "P_Age",
-      "Age at surgery",
-      "How old will you be at the time of your planned hip replacement?",
-      placement = "right",
-      options = list(container = "body")
-    ),
-
-    radioButtons(
-      "P_Gender",
-      "Sex",
-      choices = c("Male", "Female"),
-      selected = "Female",
-      inline = TRUE,
-    ),
-
-    bsPopover(
-      "P_Gender",
-      "Legal gender",
-      "What is your legal gender? (Our model was based on legal gender identified by the third digit of Swedish personal identity numbers.)",
-      placement = "right",
-      options = list(container = "body")
-    ),
-
-
-   radioButtons(
-      "P_ASA",
-      tagList("ASA class", a("(?)", href = "https://www.asahq.org/standards-and-guidelines/asa-physical-status-classification-system")),
-      choices = c(I = 1, II = 2, III = 3),
-      selected = 1,
-      inline = TRUE,
-    ),
-    bsPopover(
-      "P_ASA",
-      "ASA Class",
-      "Do you have: normal health (I), a mild systemic disease (II), or a severe systemic disease (III)? Your physician might help you with this assessment.",
-      placement = "right",
-      options = list(container = "body")
-    ),
-
-    checkboxGroupInput(
-      "checkboxes",
-      "Do you have any of the following:",
-      choices = coefs$coef_present[coefs$checkbox],
-    ),
-
-    bsPopover(
[…]

### 2020-04-01 shiny/www/pop_patch.js

@@ -1,0 +1,13 @@
+shinyBS.addTooltip = function(id, type, opts) {
+  var $id = shinyBS.getTooltipTarget(id);
+  var dopts = {html: true};
+  opts = $.extend(opts, dopts);
+
+  if(type == "tooltip") {
+    $id.tooltip("destroy");
+    setTimeout(function() {$id.tooltip(opts);},200);
+  } else if(type == "popover") {
+    $id.popover("destroy");
+    setTimeout(function() {$id.popover(opts);},200);
+  }
+};

### 2020-04-01 src/tab_table1_njr.R

@@ -1,57 +1,0 @@
-library(ProjectTemplate)
-load.project()
-
-# First part with pre-computed variables ----------------------------------
-
-t1_njr_1 <-
-  readxl::read_excel(
-    "data/descrip9odays mortality.xlsx",
-    range = "B2:F16",
-    col_names = c("what", "level", "Died within 90 days", "Survived at least 90 days", "Total")
-  ) %>%
-  mutate(
-    level = replace(level, zoo::na.locf(what) == "ASA", c("I", "II", "III"))
-  ) %>%
-  mutate_all(coalesce, "") %>%
-  # Empty row before ASA
-  add_row(
-    what = "ASA class (%)",
-    level = "",
-    `Died within 90 days` = "",
-    `Survived at least 90 days` = "",
-    Total = "",
-    .after = 3           ### WARNING! Bad practice!
-  ) %>%
-  # Empty row before Charlson
-  add_row(
-    what = "Elixhauser (%)",
-    level = "",
-    `Died within 90 days` = "",
-    `Survived at least 90 days` = "",
-    Total = "",
-    .after = 7           ### WARNING! Bad practice!
-  ) %>%
-  # Empty row before Elixhauser
-  add_row(
-    what = "Charlson (%)",
-    level = "",
-    `Died within 90 days` = "",
-    `Survived at least 90 days` = "",
-    Total = "",
-    .after = 13           ### WARNING! Bad practice!
-  ) %>%
-  mutate(
-    what = gsub("(ASA|Charlson|Elixhauser)$", "", what),
-    what = case_when(
-      what == "age" ~ "Age (mean (SD))",
-      what == "female" ~ "Sex = Female (%)",
-      TRUE ~ what
-    )
-  ) %>%
-  mutate_all(~ gsub("([0-9])(\\()", "\\1 \\2", ., perl = TRUE)) %>%
-  slice(1:7, 14:18, 8:13) ### WARNING! Bad practice!
-
-cache("t1_njr_1")
-
-
-

### 2020-04-15 reports/article.Rmd
modified abstract /NH

@@
 
-**OBJECTIVES** We developed and validated a statistical model to predict 90-day mortality after total hip arthroplasty (THA) due to osteoarthritis.
+**OBJECTIVES** To develop and validate a personalized model to predict 90-day mortality after total hip arthroplasty (THA).
 
-**DESIGN** A multi-national observational registry study with prospective data collection and data linkage.
-We used a bootstrap ranking procedure together with logistic LASSO-regression, evaluated by the area under the receiver operation characteristics curve (AUC), and calibration belt plots.
+**DESIGN** Prospective multi-national population-based study.
 
-**SETTING** Patients operated `r paste(range(df$P_SurgYear), collapse = " - ")` and identified through the Swedish Hip Arthroplasty Register, including all private and public hospitals in Sweden, for model derivation and internal validation.
-Similar patients from England and Wales,  recorded in the National Joint Register, for external model validation.
+**SETTING** Swedish Hip Arthroplasty Registry (SHAR) data for model derivation and internal validation, and data from the National Joint Registry (NJR) for England and Wales for external validation.
 
@@
 
-**PARTICIPANTS** `r format(nrow(df), big.mark = ",")` patients with cemented total hip arthroplasty (THA) due to osteoarthritis, and ASA class I-III, for model derivation and internal validation `r shar_summary`.
-An additional `r format(N_njr$tot, big.mark = ",")` patients for external validation `r njr_summary`.
+**PARTICIPANTS** `r format(nrow(df), big.mark = ",")` patients with cemented total hip arthroplasty (THA) due to osteoarthritis recorded in the SHAR, and `r format(N_njr$tot, big.mark = ",")` similarly selected patients recorded in the NJR.
 
-**MAIN OUTCOME MEASURES** Predicted probability of death due to any reason within 90 days after THA.
+**MAIN OUTCOME MEASURES** Predicted probability of death for any reason within 90 days after primary THA, estimated using a bootstrap ranking procedure together with a least absolute shrinkage and selection operator (LASSO) logistic regression model and evaluated by areas under the curve (AUC).
 
@@
 
-**RESULTS** A simple main effects model combining `r coefs_print_reduced`, with good internal discrimination `r get_auc("Reduced model")`, as well as externally `r get_auc("NJR (Reduced model)")`.
-This model was superior to traditional models based on the Charlson `r get_auc("Charlson")`, and the Elixhauser `r get_auc("Elixhauser")` co-morbidity indices.
-The model was well calibrated for predicted probabilities up to 5 %.
-A web calculator to aid clinical usage was published at `r shiny_url`.
+**RESULTS**
 
-**CONCLUSION:** We developed a parsimonious model to predict 90-day mortality after elective cemented THA based on easily available demographic data and the presence of four defined co-morbidities.
-This may facilitate individualized risk assessment prior to one of the most common surgical interventions.
+We found a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
+The model had good discrimination, both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`.
+This model was superior to traditional models based on the Charlson `r get_auc("Charlson")` and Elixhauser `r get_auc("Elixhauser")` co-morbidity indices.
+The model was well calibrated for predicted probabilities up to 5%.
+
+**CONCLUSION:** We developed a personalized model to predict 90-day mortality after elective cemented THA based on easily available demographic data and the presence of four defined co-morbidities.
+This may facilitate individualized risk assessment prior to one of the most common surgical interventions, and a web-based calculator to aid clinical decision-making was published (`r shiny_url`).
 

### 2020-05-06 reports/article.Rmd
updated athours and affiliations

@@
 title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
-author: Anne Garland`*`^1,2,3^, Erik Bülow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5,6^, J. Mark Wilkinson^7^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
+author: Anne Garland, PhD`*`^1,2,3^, Erik Bülow, MSc`*`^2,4^, Erik Lenguerrand, MSc^5^, Prof Ashley Blom, PhD^5,6^, Prof J.
+Mark Wilkinson, PhD^7^, Adrian Sayers, MSc^5^, Prof Ola Rolfson, PhD^2,4^, Prof Nils P.
+Hailer, PhD^1,2^
 date: '`r Sys.Date()`'
@@
 
-# Affiliation and contact
+`*` Authors contributed equally.
+
+## Affiliations
 
@@
 
-`*` Authors contributed equally.
+## Addrasses
 
-**Correspondence:** 
-erik.bulow@vgregion.se. 
-Postal: Svenska Höftprotesregistret, Registercentrum Västra Götaland, SE-413 45, Sweden
+AG:
+EB: 
+EL:
+AB:
+JMW:
+AS:
+OR:
+NPH:
+
+## Correspondence 
+Erik Bülow  
++46 70 08 234 28  
+erik.bulow@vgregion.se  
+Postal: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45, Sweden
 

### 2020-05-07 reports/article.Rmd
New abstract for LR. Word limit of 300 words. Fix #7.

@@
 title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
-author: Anne Garland, PhD`*`^1,2,3^, Erik Bülow, MSc`*`^2,4^, Erik Lenguerrand, MSc^5^, Prof Ashley Blom, PhD^5,6^, Prof J.
+author: Anne Garland, PhD`*`^1,2,3^, Erik B?low, MSc`*`^2,4^, Erik Lenguerrand, MSc^5^, Prof Ashley Blom, PhD^5,6^, Prof J.
 Mark Wilkinson, PhD^7^, Adrian Sayers, MSc^5^, Prof Ola Rolfson, PhD^2,4^, Prof Nils P.
@@
 
-AG:
-EB: 
-EL:
-AB:
-JMW:
-AS:
-OR:
-NPH:
+AG: Andersrogatan 5, SE-621 45 Visby, SWEDEN
+EB/OR: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45 Gothenbutg, SWEDEN 
+EL/AB/AS: Musculoskeletal Research Unit, University of Bristol, Beacon House, Queens Road, Bristol, BS8 1QU, UK
+JMW: Department of Oncology and Metabolism, The Medical School, Beech Hill Road, Sheffield, S10 2RX, UK
+NPH: Akademiska sjukhuset, 751 85 Uppsala, SWEDEN
 
 ## Correspondence 
-Erik Bülow  
-+46 70 08 234 28  
+Erik B?low  
++ 46 70 08 234 28  
 erik.bulow@vgregion.se  
-Postal: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45, Sweden
+Postal: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45 Gothenburg, SWEDEN
 
 ##### PAGE BREAK
-
-# Abstract
-
-**OBJECTIVES** To develop and validate a personalized model to predict 90-day mortality after total hip arthroplasty (THA).
-
-**DESIGN** Prospective multi-national population-based study.
-
-**SETTING** Swedish Hip Arthroplasty Registry (SHAR) data for model derivation and internal validation, and data from the National Joint Registry (NJR) for England and Wales for external validation.
 
@@
 )
-```
 
-**PARTICIPANTS** `r format(nrow(df), big.mark = ",")` patients with cemented total hip arthroplasty (THA) due to osteoarthritis recorded in the SHAR, and `r format(N_njr$tot, big.mark = ",")` similarly selected patients recorded in the NJR.
-
-**MAIN OUTCOME MEASURES** Predicted probability of death for any reason within 90 days after primary THA, estimated using a bootstrap ranking procedure together with a least absolute shrinkage and selection operator (LASSO) logistic regression model and evaluated by areas under the curve (AUC).
-
-```{r}
 get_auc <- function(mod) {
@@
 
-**RESULTS**
+# Abstract
 
-We found a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
+**Background: ** Mortality within 90 days after total hip arthroplasty (THA) due to osteoarthritis is low, but the risk of short-term mortality is nevertheless essential to discuss preoperative between patients and surgeons.
+We aimed to develop a parsimonious prediction model with internal and external validation of discrimination and calibration, and to compare it to prediction models based on existing co-morbidity indices.
+**Methods: **We performed a prospective multi-national population-based registry study.
+We included `r format(nrow(df), big.mark = ",")` patients with cemented THA due to osteoarthritis from the Swedish Hip Arthroplasty Register for model derivation and internal validation, as well as `r format(N_njr$tot, big.mark = ",")` patients from the National Joint Register from England and Wales for external model validation.
+A model was developed using a bootstrap ranking procedure with a least absolute shrinkage and selection operator (LASSO) logistic regression model combined with piecewise linear regression.
+Discriminative ability was evaluated by the area under the receiver operating characteristics curve (AUC).
+Calibration belt plots were used for model calibration.
+**Finindgs: ** We found a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
 The model had good discrimination, both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`.
@@
 The model was well calibrated for predicted probabilities up to 5%.
-
-**CONCLUSION:** We developed a personalized model to predict 90-day mortality after elective cemented THA based on easily available demographic data and the presence of four defined co-morbidities.
-This may facilitate individualized risk assessment prior to one of the most common surgical interventions, and a web-based calculator to aid clinical decision-making was published (`r shiny_url`).
+**Interpretation: ** We developed a model of 90-day mortality after THA based on age, sex, and the presence of four defined co-morbidities.
+This may facilitate individualized risk assessment prior to one of the most common surgical interventions.
+We published a web-calculator to aid clinical decision-making (`r shiny_url`).
+**Funding: ** We had no external funding.
 

### 2020-05-07 reports/article.Rmd
removed some sections in the end required by BMJ (but hopefully not by LR).

@@
 
-# Exclusive licence
-
-We grant a worldwide licence to the publishers and its licensees in perpetuity, in all forms, formats and media (whether known now or created in the future), to i) publish, reproduce, distribute, display and store the Contribution, ii) translate the Contribution into other languages, create adaptations, reprints, include within collections and create summaries, extracts and/or, abstracts of the Contribution and convert or allow conversion into any format including without limitation audio, iii) create any other derivative work(s) based in whole or part on the on the Contribution, iv) to exploit all subsidiary rights to exploit all subsidiary rights that currently exist or as may exist in the future in the Contribution, v) the inclusion of electronic links from the Contribution to third party material where-ever it may be located; and, vi) licence any third party to do any or all of the above.
-
-# Competing interests statement
-
-All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi_disclosure.pdf and declare no financial relationships with any organisations that might have an interest in the submitted work in the previous three years, and no other relationships or activities that could appear to have influenced the submitted work.
-
-# Transparency Statement
-
-NH, the principal investigator of this study, affirms that this is an accurate report with no relevant data or aspects omitted.
-
-# Data sharing statement
-
-Data is available upon request to the principal investigator.
-
 # Acknowledgement

### 2020-05-11 reports/article.Rmd
fix decimal points. Fix #11.

@@
 title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
-author: Anne Garland, PhD`*`^1,2,3^, Erik B?low, MSc`*`^2,4^, Erik Lenguerrand, MSc^5^, Prof Ashley Blom, PhD^5,6^, Prof J.
+author: Anne Garland, PhD`*`^1,2,3^, Erik Bulow, MSc`*`^2,4^, Erik Lenguerrand, MSc^5^, Prof Ashley Blom, PhD^5,6^, Prof J.
 Mark Wilkinson, PhD^7^, Adrian Sayers, MSc^5^, Prof Ola Rolfson, PhD^2,4^, Prof Nils P.
@@
   digits            = 2,
+  OutDec            = "\xB7",
   knitr.kable.NA    = '',
@@
 ## Correspondence 
-Erik B?low  
+Erik Bulow  
 + 46 70 08 234 28  
@@
 ```{r}
+# Change to ugly decimal point as required by LR
+ud <- function(x) gsub("\\.", "\xB7", x)
+
 shar_tot <- nrow(df)
 shar_female <- df$P_Sex == "Female"
-age_shar <- sprintf("%.0f [%.0f - %.0f]", mean(df$P_Age), min(df$P_Age), max(df$P_Age))
+age_shar <- sprintf("%.0f [%.0f - %.0f]", mean(df$P_Age), min(df$P_Age), max(df$P_Age)) %>% ud()
 sex_shar <- 
@@
     mean(shar_female) * 100, 
-    format(sum(shar_female), big.mark = ","), 
-    format(shar_tot, big.mark = ",")
-)
+    format(sum(shar_female), big.mark = ",", decimal.mark = "\xB7"), 
+    format(shar_tot, big.mark = ",", decimal.mark = "\xB7")
+) %>% ud()
 deaths_shar <- 
@@
     format(nrow(df), big.mark = ",")
-  )
+  ) %>% ud()
 
@@
   age_shar, sex_shar, deaths_shar
-)
+) %>% ud()
 
@@
 extr_perc <- function(...) 
-  gsub("([\\d,]*)( )(\\()(\\d*)(\\.\\d*)(\\))", "\\4 \\% \\[\\1 of", extr(...), perl = TRUE)
+  gsub("([\\d,]*)( )(\\()(\\d*)(\\.\\d*)(\\))", "\\4 \\% \\[\\1 of", extr(...), perl = TRUE) %>% ud()
 
@@
 # Same range by design (to match Swedish cohort)
-njr_age <- sprintf("%.0f [%.0f - %.0f]", njr_age_mean, min(df$P_Age), max(df$P_Age))
-njr_sex <- sprintf("%s %s]", extr_perc("Sex", "NJR"), format(N_njr$tot, big.mark = ","))
+njr_age <- sprintf("%.0f [%.0f - %.0f]", njr_age_mean, min(df$P_Age), max(df$P_Age)) %>% ud()
+njr_sex <- sprintf("%s %s]", extr_perc("Sex", "NJR"), format(N_njr$tot, big.mark = ",")) %>% ud()
 njr_deaths <- 
@@
     format(N_njr$tot, big.mark = ",")
-  )
+  ) %>% ud()
 njr_summary <- sprintf(
@@
   njr_age, njr_sex, njr_deaths
-)
+) %>% ud()
 
@@
   ) %>% 
-  {sprintf("(AUC = %s", gsub(" (", ", 95 % CI: ",., fixed = TRUE))}
+  {sprintf("(AUC = %s", gsub(" (", ", 95 % CI: ",., fixed = TRUE))} %>% 
+  ud()
 }
@@
 We aimed to develop a parsimonious prediction model with internal and external validation of discrimination and calibration, and to compare it to prediction models based on existing co-morbidity indices.
+
 **Methods: **We performed a prospective multi-national population-based registry study.
@@
 Calibration belt plots were used for model calibration.
+
 **Finindgs: ** We found a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
@@
 The model was well calibrated for predicted probabilities up to 5%.
+
 **Interpretation: ** We developed a model of 90-day mortality after THA based on age, sex, and the presence of four defined co-morbidities.
@@
 We published a web-calculator to aid clinical decision-making (`r shiny_url`).
+
 **Funding: ** We had no external funding.
@@
 In terms of gained quality of life and cost-utility, total hip arthroplasty (THA) is an enormously successful procedure.
-90-day mortality after THA performed due to osteoarthritis is low, ranging from 0.2 % to 0.6 % in large joint registry studies.[@Hunt2013; @Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.
+90-day mortality after THA performed due to osteoarthritis is low, ranging from `r 0.2` % to `r 0.6` % in large joint registry studies.[@Hunt2013; @Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.
 
@@
 Each model was used to predict the probability of death within 90 days for patients from the SHAR (internal validation).
-Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.[@Fawcett2006] Models with a lower 95 % confidence limit above 0.7, were considered good.
+Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.[@Fawcett2006] Models with a lower 95 % confidence limit above `r 0.7`, were considered good.
 Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples.
@@
 `r format(nrow(df), big.mark = ",")` patients (age `r paste(range(df$P_Age), collapse = " - ")`, `r round(mean(df$P_Sex == "Female") * 100)` % females) were included in the derivation cohort from SHAR (Figure \@ref(fig:flowchart) left panel).
-`r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` of those patients died within 90 days and no one was censored before that.
-The unadjusted risk of 90-day mortality was therefore `r surv90d`.
+`r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100)) %>% ud()` of those patients died within 90 days and no one was censored before that.
+The unadjusted risk of 90-day mortality was therefore `r ud(surv90d)`.
 Further characteristics of the study population are presented in Table \@ref(tab:tab1).
@@
 In addition, `r format(N_njr$tot, big.mark = ",")` patients were included for the external validation cohort from England and Wales (Figure \@ref(fig:flowchart) right panel).
-Their unadjusted risk of 90 day mortality was `r njr_surv`, which was higher compared to the derivation cohort from Sweden.
+Their unadjusted risk of 90 day mortality was `r ud(njr_surv)`, which was higher compared to the derivation cohort from Sweden.
 
@@
 Calibration of the re-calibrated model was slightly inferior compared to the internal calibration.
-Predicted probabilities between 0.5 % and 1.5 % were lower than observed proportions and the estimated 95 % confidence bands were wider.
+Predicted probabilities between `r 0.5` % and `r 1.5` % were lower than observed proportions and the estimated 95 % confidence bands were wider.
 Over-all, calibration was still good however for predicted probabilities below 3 % and acceptable below 5 % (Figure \@ref(fig:calibration)).
@@
 
-Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We noted that the proportion of patients with BMI above 30, the WHO definition of obesity was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) in the Swedish cohort than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the national patient register (`r round(mean(df$c_obesity) * 100, 1)` %).
+Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE) %>% ud()`) but was still relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We noted that the proportion of patients with BMI above 30, the WHO definition of obesity was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) in the Swedish cohort than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the national patient register (`r round(mean(df$c_obesity) * 100, 1)` %).
 We believe this might be due to the fact that BMI is systematically recorded in the SHAR, but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.
@@
 To only include cemented THA therefore implies an over-representation of older and frailer patients from England and Wales compared to Sweden.
-Those patients might be more likely to die within 90 days, and this assumption would be in agreement with the calibration plot where estimated probabilities below 1.5 % underestimated the observed proportion of deaths in the external validation cohort from the NJR.
+Those patients might be more likely to die within 90 days, and this assumption would be in agreement with the calibration plot where estimated probabilities below `r 1.5` % underestimated the observed proportion of deaths in the external validation cohort from the NJR.
 
@@
 A similar approach might be relevant for BMI, which we only modelled as a linear main effect.
-A U-shaped association between mortality and BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales.[@Mouchti2018] Our recordings of BMI might be imprecise however.[@Stommel2009] Some hospitals might report actual measurements but it is unknown to what extent those data are self-estimated by the patients or estimated by health care personnel.
+A U-shaped association between mortality and BMI categorized as underweight (below `r 18.5` kg/m^2^), normal (`r 8.5` to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales.[@Mouchti2018] Our recordings of BMI might be imprecise however.[@Stommel2009] Some hospitals might report actual measurements but it is unknown to what extent those data are self-estimated by the patients or estimated by health care personnel.
 Therefore, we aimed to avoid overfitting BMI based on too elaborate transformations.

### 2020-05-11 reports/tabs.Rmd
fix decimal points. Fix #11.

@@
 ```{r tab1}
+table1_combined_tot %>% 
+  mutate(what = gsub("Heart infarct", "Myocardial infarction", what)) %>% 
+  mutate_all(~gsub("\\.", "\xB7", .)) %>% 
   knitr::kable(
-  mutate(table1_combined_tot, what = gsub("Heart infarct", "Myocardial infarction", what)), 
     caption = "Characteristics of the study population in the model derivation cohort (SHAR) and external validation cohort (NJR).
@@
 brlasso_tbl_coefs %>% 
+   mutate_all(~gsub("\\.", "\xB7", .)) %>% 
   select(-math) %>%

### 2020-05-11 reports/article.Rmd
adjusted CI to LR format. Fix #10.

@@
   ) %>% 
-  {sprintf("(AUC = %s", gsub(" (", ", 95 % CI: ",., fixed = TRUE))} %>% 
+  {sprintf("(AUC = %s", gsub(" (", ", 95 % CI ",., fixed = TRUE))} %>% 
   ud()

### 2020-05-11 src/tab_auc.R

@@ -17,7 +17,7 @@ brlasso_tbl_auc <-
   arrange(desc(AUC_est)) %>%
   transmute(
     Model = model_names(Model, age = TRUE),
-    AUC      = sprintf("%.2f (%.2f to %.2f)", AUC_est, AUC_lo, AUC_hi),
+    AUC      = sprintf("%.2f (%.2f - %.2f)", AUC_est, AUC_lo, AUC_hi),
     #`AUC corrected for optimism` =
     #  sprintf("%.2f (%.2f-%.2f)", AUC_est.corr, AUC_lo.corr, AUC_hi.corr)
   ) %>%

### 2020-05-11 src/tab_coefs.R

@@ -10,7 +10,7 @@ brlasso_tbl_coefs <-
     term = clean_names(term),
     term = gsub("_", " ", term),
     math = sprintf("\\text{%s} \\\\ ", term) %>% {gsub("\\text{(intercept)} \\\\ ", "", ., fixed = TRUE)},
-    beta = log(estimate),
+    beta = round(log(estimate), 2),
     `OR 95 % CI` = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high),
     p    = pvalue_format(.01)(p.value)
   ) %>%

### 2020-05-11 reports/article.Rmd
Cut down to 30 references. Fix #6.

@@
 In terms of gained quality of life and cost-utility, total hip arthroplasty (THA) is an enormously successful procedure.
-90-day mortality after THA performed due to osteoarthritis is low, ranging from `r 0.2` % to `r 0.6` % in large joint registry studies.[@Hunt2013; @Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.
+90-day mortality after THA performed due to osteoarthritis is low, ranging from `r 0.2` % to `r 0.6` % in large joint registry studies.[@Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.
 
 Co-morbidity is associated with a shorter remaining life span, but the Charlson and Elixhauser co-morbidity indices poorly predict mortality after THA.[@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017; @Bulow2017; @Inacio2015] Additionally, these complex co-morbidity instruments are based on information from extensive in- or out-patient databases, each measure being defined by 1,178 and 1,516 individual International Classification of Diseases (ICD)-10 codes, respectively.
-Each co-morbidity index also exists in numerous versions, making interpretation and comparisons difficult.[@Brusselaers2017; @Walraven2009] The use of these indices is therefore limited to research settings but they are impossible to use under clinical circumstances.
+Each co-morbidity index also exists in numerous versions, making interpretation and comparisons difficult.[@Brusselaers2017] The use of these indices is therefore limited to research settings but they are impossible to use under clinical circumstances.
 
@@
 The observation period ended 2015, since we had access to co-morbidity from the national patient register up to this point.
-Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009] We used data from a variety of sources, as previously described.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98 %.[@Karrholm2018] Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status (married/un-married/divorced/widow[er]), were also collected from LISA.
+Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration.
+We used data from a variety of sources, as previously described.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98 %.[@Karrholm2018] Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status (married/un-married/divorced/widow[er]), were also collected from LISA.
 The Swedish national patient register (NPR) was used to assess co-morbidity during the year preceding index surgery.
@@
 The validation cohort was created by linking NJR data to the secondary uses service database from the National Health Service (NHS).
-This database includes co-morbidity recorded in the admitted patient care module of the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Brophy2006; @Burns2012; @Sinha2013; @Simmonds2014; @Thorn2016] Out-patient deaths were linked from the office of national statistics.
+This database includes co-morbidity recorded in the admitted patient care module of the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Thorn2016] Out-patient deaths were linked from the office of national statistics.
 Data linkage used NHS-numbers, as well as patient name, age, sex and address.[@Sayers2016]
@@
 
-Co-morbidity was defined by individual ICD-10 codes grouped into 17 categories according to Charlson [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to Elixhauser.[@Elixhauser1998; @Quan2005] Some co-morbidities were identified in both indices, and some distinct co-morbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
+Co-morbidity was defined by individual ICD-10 codes grouped into 17 categories according to Charlson and 31 categories according to Elixhauser.[@Quan2005] Some co-morbidities were identified in both indices, and some distinct co-morbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
 We combined individual diagnostic groups to establish `r nrow(tab_categorization)` broader categories of co-morbidity (Table \@ref(tab:tabcategorization)).
@@
 Further analysis was based on logistic regression, since no loss to follow up occurred in the Swedish cohort within the first 90 days, and we assumed the same was true for patients from England and Wales.
-We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1.
+We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015] Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1.
 Thus, the magnitude (absolute values) of their estimated coefficients would indicate variable importance on the same scale as the categorical variables.
 Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process.
-1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model.
+1,000 bootstrap samples were drawn from the observed data set.
+We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model.
 We then only kept $\lambda$:s minimizing the mean cross-validated deviances in each sample.
@@
 Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.
-Odds ratios for the final model were estimated with 95 % confidence intervals.[@Ripley2002]
+Odds ratios for the final model were estimated with 95 % confidence intervals.
 
@@
 Each model was used to predict the probability of death within 90 days for patients from the SHAR (internal validation).
-Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.[@Fawcett2006] Models with a lower 95 % confidence limit above `r 0.7`, were considered good.
+Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.
+Models with a lower 95 % confidence limit above `r 0.7`, were considered good.
 Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples.
@@
 Variables in our model were chosen based on statistical properties and not due to clinical relevance.
-Variables and estimated coefficients should therefore not be assigned any exact epidemiological and/or causal meaning per se.[@Breiman2001]
+Variables and estimated coefficients should therefore not be assigned any exact epidemiological and/or causal meaning per se.
 
@@
 This is clinically reasonable since a label of "severe systemic disease" should be based on a relevant patient assessment prior to surgery.
-ASA class is nevertheless known to have a high degree of internal variability.[@Haynes1995] It has previously been compared to the Charlson co-morbidity index, but not with respect to mortality after THA.[@Whitmore2014; @Kork2015] Patients with ASA class IV-VI were excluded since those categories describe severe disease, moribund and brain-dead individuals.
+ASA class is nevertheless known to have a high degree of internal variability.[@Haynes1995] It has been compared to the Charlson co-morbidity index, but not with respect to mortality after THA.
+Patients with ASA class IV-VI were excluded since those categories describe severe disease, moribund and brain-dead individuals.
 We suspect that patients with such conditions might have been misclassified.
 
-Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE) %>% ud()`) but was still relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We noted that the proportion of patients with BMI above 30, the WHO definition of obesity was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) in the Swedish cohort than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the national patient register (`r round(mean(df$c_obesity) * 100, 1)` %).
+Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE) %>% ud()`) but was still relevant as a predictor due to unobserved heterogeneity.
+We noted that the proportion of patients with BMI above 30, the WHO definition of obesity was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) in the Swedish cohort than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the national patient register (`r round(mean(df$c_obesity) * 100, 1)` %).
 We believe this might be due to the fact that BMI is systematically recorded in the SHAR, but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.
@@
 A strength of this study is its nationwide design with data from both Sweden, England and Wales.
-The Swedish registers are valid with low proportions of missing data.[@Soderman2000; @Karrholm2018; @Soderman2001; @Ludvigsson2011] Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however.[@Spencer2012; @Williams2002; @Brennan2012] A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80 %.[@Burns2012] Both in- and outpatient co-morbidity data were available for Sweden, but only in-patient data for England and Wales.
+The Swedish registers are valid with low proportions of missing data.[@Karrholm2018; @Ludvigsson2011] Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however.
+A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80 %.[@Burns2012] Both in- and outpatient co-morbidity data were available for Sweden, but only in-patient data for England and Wales.
 We did not have access to any data from primary care or nursing homes in any of the counties.
@@
 A similar approach might be relevant for BMI, which we only modelled as a linear main effect.
-A U-shaped association between mortality and BMI categorized as underweight (below `r 18.5` kg/m^2^), normal (`r 8.5` to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales.[@Mouchti2018] Our recordings of BMI might be imprecise however.[@Stommel2009] Some hospitals might report actual measurements but it is unknown to what extent those data are self-estimated by the patients or estimated by health care personnel.
+A U-shaped association between mortality and BMI categorized as underweight (below `r 18.5` kg/m^2^), normal (`r 8.5` to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales.[@Mouchti2018] Our recordings of BMI might be imprecise however.
+Some hospitals report actual measurements but it is unknown to what extent those data are self-estimated by the patients or by health care personnel.
 Therefore, we aimed to avoid overfitting BMI based on too elaborate transformations.

### 2020-05-13 reports/article.Rmd
Adjustments after NH:s comments. Fix #17.

@@
 Mark Wilkinson, PhD^7^, Adrian Sayers, MSc^5^, Prof Ola Rolfson, PhD^2,4^, Prof Nils P.
-Hailer, PhD^1,2^
+Hailer, MD^1,2^
 date: '`r Sys.Date()`'
@@
 
-## Addrasses
+## Addresses
 
-AG: Andersrogatan 5, SE-621 45 Visby, SWEDEN  
-EB/OR: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45 Gothenbutg, SWEDEN   
-EL/AB/AS: Musculoskeletal Research Unit, University of Bristol, Beacon House, Queens Road, Bristol, BS8 1QU, UK  
-JMW: Department of Oncology and Metabolism, The Medical School, Beech Hill Road, Sheffield, S10 2RX, UK  
-NPH: Akademiska sjukhuset, 751 85 Uppsala, SWEDEN
+- AG: Andersrogatan 5, SE-621 45 Visby, SWEDEN
+- EB/OR: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45 Gothenbutg, SWEDEN
+- EL/AB/AS: Musculoskeletal Research Unit, University of Bristol, Beacon House, Queens Road, Bristol, BS8 1QU, UK
+- JMW: Department of Oncology and Metabolism, The Medical School, Beech Hill Road, Sheffield, S10 2RX, UK
+- NPH: Akademiska sjukhuset, 751 85 Uppsala, SWEDEN
 
 ## Correspondence 
-Erik Bulow  
-+ 46 70 08 234 28  
-erik.bulow@vgregion.se  
-Postal: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45 Gothenburg, SWEDEN
+Erik Bulow. + 46 70 08 234 28. erik.bulow@vgregion.se. Postal: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45 Gothenburg, SWEDEN
 
@@
 
-**Background: ** Mortality within 90 days after total hip arthroplasty (THA) due to osteoarthritis is low, but the risk of short-term mortality is nevertheless essential to discuss preoperative between patients and surgeons.
-We aimed to develop a parsimonious prediction model with internal and external validation of discrimination and calibration, and to compare it to prediction models based on existing co-morbidity indices.
+**Background: ** Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality.
+However, diagnosis code-based instruments such as the Charlson co-morbidity index are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) class is imprecise.
+We thus developed a simple model to predict 90-day mortality after elective THA.
 
@@
 
-**Finindgs: ** We found a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
+**Findings: ** We propose a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
 The model had good discrimination, both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`.
@@
 
-**Interpretation: ** We developed a model of 90-day mortality after THA based on age, sex, and the presence of four defined co-morbidities.
+**Interpretation: ** We developed a parsimonious model to predict 90-day mortality after elective cemented THA based on easily available demographic data and the presence of four defined co-morbidities.
 This may facilitate individualized risk assessment prior to one of the most common surgical interventions.
@@
 
-# Summary box
+# Research in context
 
-What is already known:
+## Evidence before this study
 
-- Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality.
-- Available diagnosis code-based co-morbidity instruments are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) classification is imprecise.
+We performed an initial ad hoc literature review based on Google Scholar, personal recommendations and relevant references.
+We then used 2Dsearch (https://2dsearch.com accessed 2020-05-11) to build a search query for a more systematic review: `(femur OR hip OR bone OR bones OR osteoarthritis OR rheumatoid OR rheumatism) AND (prognostic OR prediction OR forecast OR forecasting OR forecasts OR predict OR predicted OR predicting OR predictions OR predicts OR projections) AND (death OR mortality OR deaths OR survival OR survive) AND (replacement OR replaced OR replacements OR arthroplasty OR "Arthroplasty, Replacement" OR prosthesis OR implant OR prostheses OR prosthetic OR prosthetics) AND (calibration OR calibrate OR calibrated OR validation OR validate OR validated OR validating OR validity OR precision OR accuracy OR accurate OR precise OR "external validation" OR evaluation OR verification OR internal) AND (build OR building OR construct OR create OR develop OR establish OR train OR derive OR derivation) AND (models OR modelling OR "statistical model" OR modeling) AND (comorbidity OR comorbid OR co-morbid OR co-occurring OR diagnoses OR co-morbidity OR Elixhauser OR Charlson OR "comorbidity index" OR "comorbidity score" OR "Rx Risk V" OR ASA)`.
+We used no time or language constraints and used this query in PubMed (https://pubmed.gov accessed 2020-05-11), and identified 143 articles, whereof one was a duplicate.
+We performed an article screening with Rayyan (https://rayyan.qcri.org/ accessed 2020-05-12).
+131 articles were excluded based on titles, and an additional 6 after reading the abstracts.
+5 articles remained, whereof 3 were already considered during the ad hoc review process.
+Research on cadavers, canines, and patients with diagnoses other than elective hip disorders (including hip fractures) were excluded during the screening process, as were prediction models with outcomes other than mortality.
+In addition, we used Open Grey (https://opengrey.eu accessed 2020-05-12) with the query above, as well as a simpler query (`hip AND prediction`).
+We found 9 items, whereof none was considered relevant.
+The same queries were used with Epistemonikos (https://epistemonikos.org accessed 2020-05-12).
+No items were found using the longer query, but 33 systematic reviews based on the shorter.
+30 where excluded after title screening, and the remaining 3 after screening of abstracts.
 
-What this study adds:
+In summary, several prediction models have been developed for short-term mortality in patients operated with total hip arthroplasty (THA).
+Most of them are based on small sample sizes, make use of too elaborate predictors based on data that are unavailable in a clinical setting, only consider in-hospital deaths, are inadequately internally or externally validated or not validated at all, or have no national population coverage.
+Only few model descriptions are accompanied by a practical tool for use in clinical practice (often only nomograms or a mathematical formula are supplied).
+Some relevant models are further described in the main text.
+Meta-analysis is most relevant for pooling individual effect estimates or measures of predictive power for a model validated on different cohorts.
+Hence, no relevant meta-analysis was applicable to our study.
 
-- Our results indicate that the risk of early postoperative mortality after cemented THA can be pre-operatively assessed by a parsimonious prediction model. 
-- The presented model, with its accompanying web calculator, could facilitate a shared decision-making between physicians and their patients in need of THA.
+## Added value of this study
+
+Our study is the first to develop and externally validate a parsimonious prediction model for early mortality after THA based on two European cohorts, and to simultaneously provide a simple web-calculator for use in clinical practice.
+
+## Implications of all the available evidence
+
+Prediction models must consider all aspects of the relevant cohorts, including different health care systems, population demographics, and treatment traditions.
+Our tool was derived on a Swedish cohort and then validated on a similar cohort from England and Wales.
+We believe that the model might have broader clinical relevance and is suitable for clinical usage.
 
@@
 
-Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery.[@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] The first prediction model based on a European cohort was the patient decision support tool for joint remplacement in the UK (https://jointcalc.shef.ac.uk/).
+Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery.[@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] The first prediction model based on a European cohort was the patient decision support tool for joint replacement in the UK (https://jointcalc.shef.ac.uk/).
 However, it remains unclear to what level the depth of co-morbidity characterization influences the accuracy of  mortality prediction models.
@@
 Predicted probabilities between `r 0.5` % and `r 1.5` % were lower than observed proportions and the estimated 95 % confidence bands were wider.
-Over-all, calibration was still good however for predicted probabilities below 3 % and acceptable below 5 % (Figure \@ref(fig:calibration)).
+Over-all, calibration was good for predicted probabilities below 3 % and acceptable below 5 % (Figure \@ref(fig:calibration)).
 
@@
 
-We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` was considerably better at predicting death within 90 days after THA when compared to traditional models based on the Charlson or Elixhauser co-morbidity indices.
+We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` was considerably better at predicting death within 90 days after THA than traditional models based on the Charlson or Elixhauser co-morbidity indices.
 
@@
 Our model could be used in clinical practice, either by the formula above, or by accessing a simple web calculator online (`r shiny_url`).
-For example, a `r min(df$P_Age)` year old woman with ASA class I and none of the important co-morbidities would have a `r get_risk("baseline")` risk to die within 90 days of surgery.
+For example, a `r min(df$P_Age)`-year-old woman with ASA class I and none of the important co-morbidities would have a `r get_risk("baseline")` risk to die within 90 days of surgery.
 A woman, `r quantile(df$P_Age, .25)` years old, i.e.
 the first age quantile, would have an elevated risk of `r get_risk("healthy_female")`.
-A `r max(df$P_Age)` year old man, i.e.
+A `r max(df$P_Age)`-year-old man, i.e.
 the maximum observed age, with ASA class III and cancer, would have a risk of `r get_risk("sickest_man")`.
@@
 Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE) %>% ud()`) but was still relevant as a predictor due to unobserved heterogeneity.
-We noted that the proportion of patients with BMI above 30, the WHO definition of obesity was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) in the Swedish cohort than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the national patient register (`r round(mean(df$c_obesity) * 100, 1)` %).
+We noted that the proportion of patients registered in the SHAR with BMI above 30, the WHO definition of obesity, was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the Swedish national patient register (`r round(mean(df$c_obesity) * 100, 1)` %).
 We believe this might be due to the fact that BMI is systematically recorded in the SHAR, but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.
@@
 We are grateful to the Healthcare Quality Improvement Partnership (HQIP), the NJR Research Committee and staff at the NJR Centre for facilitating this work.
-We also thank all contributers to the Swedish Hip Arthroplasty Register (SHAR), especially Szilard Nemes, previous senior statistician, for his involvement in the study.
+We also thank all contributors to the Swedish Hip Arthroplasty Register (SHAR), especially Szilard Nemes, previous senior statistician, for his involvement in the study.
 The authors have conformed to the NJR's standard protocol for data access and publication.

### 2020-05-13 reports/tabs.Rmd
Adjustments after NH:s comments. Fix #17.

@@
     variable = gsub("= 3", "III", variable),
+    variable = gsub("Heart infarct", "Myocardial infarction", variable)
   )

### 2020-05-23 reports/article.Rmd
Minor formatting. Fix #23.

--- 2020-05-11 reports/article.Rmd
@@
 title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
-author: Anne Garland, PhD`*`^1,2,3^, Erik Bulow, MSc`*`^2,4^, Erik Lenguerrand, MSc^5^, Prof Ashley Blom, PhD^5,6^, Prof J.
-Mark Wilkinson, PhD^7^, Adrian Sayers, MSc^5^, Prof Ola Rolfson, PhD^2,4^, Prof Nils P.
-Hailer, PhD^1,2^
+author: Anne Garland`*`^1,2,3^, Erik Bülow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5,6^, J. Mark Wilkinson^7^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
 date: '`r Sys.Date()`'
@@
   digits            = 2,
-  OutDec            = "\xB7",
   knitr.kable.NA    = '',
@@
 
-`*` Authors contributed equally.
-
-## Affiliations
+# Affiliation and contact
 
@@
 
-## Addrasses
+`*` Authors contributed equally.
 
-AG: Andersrogatan 5, SE-621 45 Visby, SWEDEN  
-EB/OR: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45 Gothenbutg, SWEDEN   
-EL/AB/AS: Musculoskeletal Research Unit, University of Bristol, Beacon House, Queens Road, Bristol, BS8 1QU, UK  
-JMW: Department of Oncology and Metabolism, The Medical School, Beech Hill Road, Sheffield, S10 2RX, UK  
-NPH: Akademiska sjukhuset, 751 85 Uppsala, SWEDEN
+**Correspondence:** 
+erik.bulow@vgregion.se. 
+Postal: Svenska Höftprotesregistret, Registercentrum Västra Götaland, SE-413 45, Sweden
 
-## Correspondence 
-Erik Bulow  
-+ 46 70 08 234 28  
-erik.bulow@vgregion.se  
-Postal: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45 Gothenburg, SWEDEN
+# Abstract
 
-##### PAGE BREAK
+**OBJECTIVES** To develop and validate a personalized model to predict 90-day mortality after total hip arthroplasty (THA).
+
+**DESIGN** Prospective multi-national population-based study.
+
+**SETTING** Swedish Hip Arthroplasty Registry (SHAR) data for model derivation and internal validation, and data from the National Joint Registry (NJR) for England and Wales for external validation.
 
 ```{r}
-# Change to ugly decimal point as required by LR
-ud <- function(x) gsub("\\.", "\xB7", x)
-
 shar_tot <- nrow(df)
 shar_female <- df$P_Sex == "Female"
-age_shar <- sprintf("%.0f [%.0f - %.0f]", mean(df$P_Age), min(df$P_Age), max(df$P_Age)) %>% ud()
+age_shar <- sprintf("%.0f [%.0f - %.0f]", mean(df$P_Age), min(df$P_Age), max(df$P_Age))
 sex_shar <- 
@@
     mean(shar_female) * 100, 
-    format(sum(shar_female), big.mark = ",", decimal.mark = "\xB7"), 
-    format(shar_tot, big.mark = ",", decimal.mark = "\xB7")
-) %>% ud()
+    format(sum(shar_female), big.mark = ","), 
+    format(shar_tot, big.mark = ",")
+)
 deaths_shar <- 
@@
     format(nrow(df), big.mark = ",")
-  ) %>% ud()
+  )
 
@@
   age_shar, sex_shar, deaths_shar
-) %>% ud()
+)
 
@@
 extr_perc <- function(...) 
-  gsub("([\\d,]*)( )(\\()(\\d*)(\\.\\d*)(\\))", "\\4 \\% \\[\\1 of", extr(...), perl = TRUE) %>% ud()
+  gsub("([\\d,]*)( )(\\()(\\d*)(\\.\\d*)(\\))", "\\4 \\% \\[\\1 of", extr(...), perl = TRUE)
 
@@
 # Same range by design (to match Swedish cohort)
-njr_age <- sprintf("%.0f [%.0f - %.0f]", njr_age_mean, min(df$P_Age), max(df$P_Age)) %>% ud()
-njr_sex <- sprintf("%s %s]", extr_perc("Sex", "NJR"), format(N_njr$tot, big.mark = ",")) %>% ud()
+njr_age <- sprintf("%.0f [%.0f - %.0f]", njr_age_mean, min(df$P_Age), max(df$P_Age))
+njr_sex <- sprintf("%s %s]", extr_perc("Sex", "NJR"), format(N_njr$tot, big.mark = ","))
 njr_deaths <- 
@@
     format(N_njr$tot, big.mark = ",")
-  ) %>% ud()
+  )
 njr_summary <- sprintf(
@@
   njr_age, njr_sex, njr_deaths
-) %>% ud()
+)
+```
 
+**PARTICIPANTS** `r format(nrow(df), big.mark = ",")` patients with cemented total hip arthroplasty (THA) due to osteoarthritis recorded in the SHAR, and `r format(N_njr$tot, big.mark = ",")` similarly selected patients recorded in the NJR.
+
+**MAIN OUTCOME MEASURES** Predicted probability of death for any reason within 90 days after primary THA, estimated using a bootstrap ranking procedure together with a least absolute shrinkage and selection operator (LASSO) logistic regression model and evaluated by areas under the curve (AUC).
+
+```{r}
 get_auc <- function(mod) {
@@
   ) %>% 
-  {sprintf("(AUC = %s", gsub(" (", ", 95 % CI ",., fixed = TRUE))} %>% 
-  ud()
+  {sprintf("(AUC = %s", gsub(" (", ", 95 % CI: ",., fixed = TRUE))}
 }
@@
 
-# Abstract
+**RESULTS**
 
-**Background: ** Mortality within 90 days after total hip arthroplasty (THA) due to osteoarthritis is low, but the risk of short-term mortality is nevertheless essential to discuss preoperative between patients and surgeons.
-We aimed to develop a parsimonious prediction model with internal and external validation of discrimination and calibration, and to compare it to prediction models based on existing co-morbidity indices.
-
-**Methods: **We performed a prospective multi-national population-based registry study.
-We included `r format(nrow(df), big.mark = ",")` patients with cemented THA due to osteoarthritis from the Swedish Hip Arthroplasty Register for model derivation and internal validation, as well as `r format(N_njr$tot, big.mark = ",")` patients from the National Joint Register from England and Wales for external model validation.
-A model was developed using a bootstrap ranking procedure with a least absolute shrinkage and selection operator (LASSO) logistic regression model combined with piecewise linear regression.
-Discriminative ability was evaluated by the area under the receiver operating characteristics curve (AUC).
-Calibration belt plots were used for model calibration.
-
-**Finindgs: ** We found a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
+We found a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
 The model had good discrimination, both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`.
@@
 
-**Interpretation: ** We developed a model of 90-day mortality after THA based on age, sex, and the presence of four defined co-morbidities.
-This may facilitate individualized risk assessment prior to one of the most common surgical interventions.
-We published a web-calculator to aid clinical decision-making (`r shiny_url`).
-
-**Funding: ** We had no external funding.
-
-##### PAGE BREAK
-
-# Summary box
-
-What is already known:
-
-- Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality.
-- Available diagnosis code-based co-morbidity instruments are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) classification is imprecise.
-
-What this study adds:
-
-- Our results indicate that the risk of early postoperative mortality after cemented THA can be pre-operatively assessed by a parsimonious prediction model. 
-- The presented model, with its accompanying web calculator, could facilitate a shared decision-making between physicians and their patients in need of THA.
-
-##### PAGE BREAK
+**CONCLUSION:** We developed a personalized model to predict 90-day mortality after elective cemented THA based on easily available demographic data and the presence of four defined co-morbidities.
+This may facilitate individualized risk assessment prior to one of the most common surgical interventions, and a web-based calculator to aid clinical decision-making was published (`r shiny_url`).
 
@@
 In terms of gained quality of life and cost-utility, total hip arthroplasty (THA) is an enormously successful procedure.
-90-day mortality after THA performed due to osteoarthritis is low, ranging from `r 0.2` % to `r 0.6` % in large joint registry studies.[@Hunt2013; @Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.
+90-day mortality after THA performed due to osteoarthritis is low, ranging from 0.2 % to 0.6 % in large joint registry studies.[@Hunt2013; @Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.
 
@@
 Each model was used to predict the probability of death within 90 days for patients from the SHAR (internal validation).
-Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.[@Fawcett2006] Models with a lower 95 % confidence limit above `r 0.7`, were considered good.
+Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.[@Fawcett2006] Models with a lower 95 % confidence limit above 0.7, were considered good.
 Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples.
@@
 `r format(nrow(df), big.mark = ",")` patients (age `r paste(range(df$P_Age), collapse = " - ")`, `r round(mean(df$P_Sex == "Female") * 100)` % females) were included in the derivation cohort from SHAR (Figure \@ref(fig:flowchart) left panel).
-`r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100)) %>% ud()` of those patients died within 90 days and no one was censored before that.
-The unadjusted risk of 90-day mortality was therefore `r ud(surv90d)`.
+`r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` of those patients died within 90 days and no one was censored before that.
+The unadjusted risk of 90-day mortality was therefore `r surv90d`.
 Further characteristics of the study population are presented in Table \@ref(tab:tab1).
@@
 In addition, `r format(N_njr$tot, big.mark = ",")` patients were included for the external validation cohort from England and Wales (Figure \@ref(fig:flowchart) right panel).
-Their unadjusted risk of 90 day mortality was `r ud(njr_surv)`, which was higher compared to the derivation cohort from Sweden.
+Their unadjusted risk of 90 day mortality was `r njr_surv`, which was higher compared to the derivation cohort from Sweden.
 
@@
 Calibration of the re-calibrated model was slightly inferior compared to the internal calibration.
-Predicted probabilities between `r 0.5` % and `r 1.5` % were lower than observed proportions and the estimated 95 % confidence bands were wider.
+Predicted probabilities between 0.5 % and 1.5 % were lower than observed proportions and the estimated 95 % confidence bands were wider.
 Over-all, calibration was still good however for predicted probabilities below 3 % and acceptable below 5 % (Figure \@ref(fig:calibration)).
@@
 
-Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE) %>% ud()`) but was still relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We noted that the proportion of patients with BMI above 30, the WHO definition of obesity was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) in the Swedish cohort than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the national patient register (`r round(mean(df$c_obesity) * 100, 1)` %).
+Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We noted that the proportion of patients with BMI above 30, the WHO definition of obesity was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) in the Swedish cohort than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the national patient register (`r round(mean(df$c_obesity) * 100, 1)` %).
 We believe this might be due to the fact that BMI is systematically recorded in the SHAR, but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.
@@
 To only include cemented THA therefore implies an over-representation of older and frailer patients from England and Wales compared to Sweden.
-Those patients might be more likely to die within 90 days, and this assumption would be in agreement with the calibration plot where estimated probabilities below `r 1.5` % underestimated the observed proportion of deaths in the external validation cohort from the NJR.
+Those patients might be more likely to die within 90 days, and this assumption would be in agreement with the calibration plot where estimated probabilities below 1.5 % underestimated the observed proportion of deaths in the external validation cohort from the NJR.
 
@@
 A similar approach might be relevant for BMI, which we only modelled as a linear main effect.
-A U-shaped association between mortality and BMI categorized as underweight (below `r 18.5` kg/m^2^), normal (`r 8.5` to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales.[@Mouchti2018] Our recordings of BMI might be imprecise however.[@Stommel2009] Some hospitals might report actual measurements but it is unknown to what extent those data are self-estimated by the patients or estimated by health care personnel.
+A U-shaped association between mortality and BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales.[@Mouchti2018] Our recordings of BMI might be imprecise however.[@Stommel2009] Some hospitals might report actual measurements but it is unknown to what extent those data are self-estimated by the patients or estimated by health care personnel.
 Therefore, we aimed to avoid overfitting BMI based on too elaborate transformations.
@@
 
-##### PAGE BREAK
-
 # Contribution of authors
@@
 EB developed the statistical model. EL and AS performed external validation with data from NJR. AG and EB drafted, and all authors edited and finalized the manuscript.
+
+# Exclusive licence
+
+We grant a worldwide licence to the publishers and its licensees in perpetuity, in all forms, formats and media (whether known now or created in the future), to i) publish, reproduce, distribute, display and store the Contribution, ii) translate the Contribution into other languages, create adaptations, reprints, include within collections and create summaries, extracts and/or, abstracts of the Contribution and convert or allow conversion into any format including without limitation audio, iii) create any other derivative work(s) based in whole or part on the on the Contribution, iv) to exploit all subsidiary rights to exploit all subsidiary rights that currently exist or as may exist in the future in the Contribution, v) the inclusion of electronic links from the Contribution to third party material where-ever it may be located; and, vi) licence any third party to do any or all of the above.
+
+# Competing interests statement
+
+All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi_disclosure.pdf and declare no financial relationships with any organisations that might have an interest in the submitted work in the previous three years, and no other relationships or activities that could appear to have influenced the submitted work.
+
+# Transparency Statement
+
+NH, the principal investigator of this study, affirms that this is an accurate report with no relevant data or aspects omitted.
+
+# Data sharing statement
+
+Data is available upon request to the principal investigator.
 
@@
 
-##### PAGE BREAK
-
 ```{r, child = "figs.Rmd"}
@@
 
-##### PAGE BREAK
-
 # Bibliography

### 2020-05-23 reports/article.Rmd
Change ref style. Fix #26.

@@
 bibliography: H:/zotero_lib.bib
-csl: the-lancet.csl
+csl: bmc-medicine.csl
 output: 
@@
 In terms of gained quality of life and cost-utility, total hip arthroplasty (THA) is an enormously successful procedure.
-90-day mortality after THA performed due to osteoarthritis is low, ranging from 0.2 % to 0.6 % in large joint registry studies.[@Hunt2013; @Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.
+90-day mortality after THA performed due to osteoarthritis is low, ranging from 0.2 % to 0.6 % in large joint registry studies [@Hunt2013; @Garland2017].
+The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.
 
-Co-morbidity is associated with a shorter remaining life span, but the Charlson and Elixhauser co-morbidity indices poorly predict mortality after THA.[@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017; @Bulow2017; @Inacio2015] Additionally, these complex co-morbidity instruments are based on information from extensive in- or out-patient databases, each measure being defined by 1,178 and 1,516 individual International Classification of Diseases (ICD)-10 codes, respectively.
-Each co-morbidity index also exists in numerous versions, making interpretation and comparisons difficult.[@Brusselaers2017; @Walraven2009] The use of these indices is therefore limited to research settings but they are impossible to use under clinical circumstances.
+Co-morbidity is associated with a shorter remaining life span, but the Charlson and Elixhauser co-morbidity indices poorly predict mortality after THA [@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017; @Bulow2017].
+Additionally, these complex co-morbidity instruments are based on information from extensive in- or out-patient databases, each measure being defined by 1,178 and 1,516 individual International Classification of Diseases (ICD)-10 codes, respectively.
+Each co-morbidity index also exists in numerous versions, making interpretation and comparisons difficult [@Brusselaers2017; @Walraven2009].
+The use of these indices is therefore limited to research settings but they are impossible to use under clinical circumstances.
 
-Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery.[@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] The first prediction model based on a European cohort was the patient decision support tool for joint remplacement in the UK (https://jointcalc.shef.ac.uk/).
+Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery [@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019].
+The first prediction model based on a European cohort was the patient decision support tool for joint remplacement in the UK (https://jointcalc.shef.ac.uk/).
 However, it remains unclear to what level the depth of co-morbidity characterization influences the accuracy of  mortality prediction models.
 
-Here we aimed to develop a parsimonious prediction model of 90-day postoperative mortality after THA with internal and external validation of discrimination and calibration, and to compare this to the accuracy of prediction models based on existing co-morbidity measures.[@Manning2016]
+Here we aimed to develop a parsimonious prediction model of 90-day postoperative mortality after THA with internal and external validation of discrimination and calibration, and to compare this to the accuracy of prediction models based on existing co-morbidity measures [@Manning2016].
 
@@
 Cementation is the most common fixation technique used in Sweden, thus used as inclusion criteria to decrease heterogeneity.
-Only the last operated hip was accounted for in patients with bilateral THA, since those are more likely to resemble patients with unilateral THA (Figure \@ref(fig:flowchart)).[@Bulow2020]
+Only the last operated hip was accounted for in patients with bilateral THA, since those are more likely to resemble patients with unilateral THA (Figure \@ref(fig:flowchart)) [@Bulow2020].
 
@@
 The observation period ended 2015, since we had access to co-morbidity from the national patient register up to this point.
-Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009] We used data from a variety of sources, as previously described.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98 %.[@Karrholm2018] Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status (married/un-married/divorced/widow[er]), were also collected from LISA.
+Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration [@Ludvigsson2009].
+We used data from a variety of sources, as previously described [@Cnudde2016].
+Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98 % [@Karrholm2018].
+Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden [@Ludvigsson2019].
+Civil status (married/un-married/divorced/widow[er]), were also collected from LISA.
 The Swedish national patient register (NPR) was used to assess co-morbidity during the year preceding index surgery.
 This register contains all relevant diagnoses coded by ICD-10, as well as dates of admission and discharge for in- and outpatient episodes in all private and public hospitals.
-Completeness for NPR is above 99 % and 85-95 % of all diagnostic codes are valid.[@Ludvigsson2011] Death dates were linked from the national population register.
+Completeness for NPR is above 99 % and 85-95 % of all diagnostic codes are valid [@Ludvigsson2011].
+Death dates were linked from the national population register.
 
@@
 The validation cohort was created by linking NJR data to the secondary uses service database from the National Health Service (NHS).
-This database includes co-morbidity recorded in the admitted patient care module of the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Brophy2006; @Burns2012; @Sinha2013; @Simmonds2014; @Thorn2016] Out-patient deaths were linked from the office of national statistics.
-Data linkage used NHS-numbers, as well as patient name, age, sex and address.[@Sayers2016]
+This database includes co-morbidity recorded in the admitted patient care module of the hospital episode statistics (HES) database in England, and the patient episode database for Wales [@Brophy2006; @Burns2012; @Sinha2013; @Simmonds2014; @Thorn2016].
+Out-patient deaths were linked from the office of national statistics.
+Data linkage used NHS-numbers, as well as patient name, age, sex and address [@Sayers2016].
 
@@
 
-Co-morbidity was defined by individual ICD-10 codes grouped into 17 categories according to Charlson [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to Elixhauser.[@Elixhauser1998; @Quan2005] Some co-morbidities were identified in both indices, and some distinct co-morbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
+Co-morbidity was defined by individual ICD-10 codes grouped into 17 categories according to Charlson [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to Elixhauser [@Elixhauser1998; @Quan2005].
+Some co-morbidities were identified in both indices, and some distinct co-morbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
 We combined individual diagnostic groups to establish `r nrow(tab_categorization)` broader categories of co-morbidity (Table \@ref(tab:tabcategorization)).
@@
 Further analysis was based on logistic regression, since no loss to follow up occurred in the Swedish cohort within the first 90 days, and we assumed the same was true for patients from England and Wales.
-We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1.
+We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO) [@Guo2015; @Baranowski2020].
+Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1.
 Thus, the magnitude (absolute values) of their estimated coefficients would indicate variable importance on the same scale as the categorical variables.
 Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process.
-1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model.
+1,000 bootstrap samples were drawn from the observed data set [@Austin2004].
+We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model.
 We then only kept $\lambda$:s minimizing the mean cross-validated deviances in each sample.
@@
 Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.
-Odds ratios for the final model were estimated with 95 % confidence intervals.[@Ripley2002]
+Odds ratios for the final model were estimated with 95 % confidence intervals [@Ripley2002].
 
@@
 Each model was used to predict the probability of death within 90 days for patients from the SHAR (internal validation).
-Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.[@Fawcett2006] Models with a lower 95 % confidence limit above 0.7, were considered good.
+Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability [@Fawcett2006].
+Models with a lower 95 % confidence limit above 0.7, were considered good.
 Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples.
-We used the bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples to adjust for optimism.[@Miller1991] Calibration bands were made to graphically assess model calibration, comparing predicted probabilities and observed proportions.[@Nattino2016] The reduced model was then evaluated externally.
+We used the bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples to adjust for optimism [@Miller1991].
+Calibration bands were made to graphically assess model calibration, comparing predicted probabilities and observed proportions [@Nattino2016].
+The reduced model was then evaluated externally.
 An AUC with 95 % confidence interval was calculated for the model as-is.
 Re-calibration of the model intercept was then performed to account for different mortality rates in Sweden compared to England and Wales.
-An updated over-all slope was also calculated to account for country-specific treatment differences.[@Steyerberg2004] Calibration for this re-calibrated model was illustrated in the same calibration belt plot as for the internal calibration.
+An updated over-all slope was also calculated to account for country-specific treatment differences [@Steyerberg2004].
+Calibration for this re-calibrated model was illustrated in the same calibration belt plot as for the internal calibration.
 
@@
 The observed proportions of deaths for patients with similar characteristics is likely lower (as indicated by Figure \@ref(fig:calibration)).
-Some risk calculators ignore this problem,[@Bozic2013a; @Harris2019] but we think this should be acknowledged.
+Some risk calculators ignore this problem [@Bozic2013a; @Harris2019], but we think this should be acknowledged.
 
 Variables in our model were chosen based on statistical properties and not due to clinical relevance.
-Variables and estimated coefficients should therefore not be assigned any exact epidemiological and/or causal meaning per se.[@Breiman2001]
+Variables and estimated coefficients should therefore not be assigned any exact epidemiological and/or causal meaning per se [@Breiman2001].
 
@@
 This is clinically reasonable since a label of "severe systemic disease" should be based on a relevant patient assessment prior to surgery.
-ASA class is nevertheless known to have a high degree of internal variability.[@Haynes1995] It has previously been compared to the Charlson co-morbidity index, but not with respect to mortality after THA.[@Whitmore2014; @Kork2015] Patients with ASA class IV-VI were excluded since those categories describe severe disease, moribund and brain-dead individuals.
+ASA class is nevertheless known to have a high degree of internal variability [@Haynes1995].
+It has previously been compared to the Charlson co-morbidity index, but not with respect to mortality after THA [@Whitmore2014; @Kork2015].
+Patients with ASA class IV-VI were excluded since those categories describe severe disease, moribund and brain-dead individuals.
 We suspect that patients with such conditions might have been misclassified.
 
-Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We noted that the proportion of patients with BMI above 30, the WHO definition of obesity was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) in the Swedish cohort than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the national patient register (`r round(mean(df$c_obesity) * 100, 1)` %).
+Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity [@Mood2010].
+We noted that the proportion of patients with BMI above 30, the WHO definition of obesity was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) in the Swedish cohort than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the national patient register (`r round(mean(df$c_obesity) * 100, 1)` %).
 We believe this might be due to the fact that BMI is systematically recorded in the SHAR, but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.
@@
 A strength of this study is its nationwide design with data from both Sweden, England and Wales.
-The Swedish registers are valid with low proportions of missing data.[@Soderman2000; @Karrholm2018; @Soderman2001; @Ludvigsson2011] Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however.[@Spencer2012; @Williams2002; @Brennan2012] A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80 %.[@Burns2012] Both in- and outpatient co-morbidity data were available for Sweden, but only in-patient data for England and Wales.
+The Swedish registers are valid with low proportions of missing data [@Soderman2000; @Karrholm2018; @Soderman2001; @Ludvigsson2011].
+Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however [@Spencer2012; @Williams2002; @Brennan2012].
+A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80 % [@Burns2012].
+Both in- and outpatient co-morbidity data were available for Sweden, but only in-patient data for England and Wales.
 We did not have access to any data from primary care or nursing homes in any of the counties.
 
-The linkage procedure between NJR and HES has been previously described,[@Smith2012] where it was noted that privately funded patients were not included in HES.
+The linkage procedure between NJR and HES has been previously described [@Smith2012], where it was noted that privately funded patients were not included in HES.
 17 % of the patients had missing personal data or did not allow linkage, and 6 % were not found in HES although their data were available from the NJR.
@@
 A similar approach might be relevant for BMI, which we only modelled as a linear main effect.
-A U-shaped association between mortality and BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales.[@Mouchti2018] Our recordings of BMI might be imprecise however.[@Stommel2009] Some hospitals might report actual measurements but it is unknown to what extent those data are self-estimated by the patients or estimated by health care personnel.
+A U-shaped association between mortality and BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales [@Mouchti2018].
+Our recordings of BMI might be imprecise however [@Stommel2009].
+Some hospitals might report actual measurements but it is unknown to what extent those data are self-estimated by the patients or estimated by health care personnel.
 Therefore, we aimed to avoid overfitting BMI based on too elaborate transformations.

### 2020-05-23 reports/article.Rmd
Formal declarations. Fix #25.

@@
 A linked Binder environment is also available for interactive usage.
-
-## Ethical approval
-
-Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13) and by the NJR in the UK (RSC2017/21).
-Informed consent was not mandatory according to the Swedish patient data law from 2009, and the UK law for pseudonymised data.
+ta law from 2009, and the UK law for pseudonymised data.
 
@@
 
-# Contribution of authors
+# Declarations
+
+## Ethics approval and consent to participate
+
+Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13) and by the NJR in the UK (RSC2017/21).
+Informed consent was not mandatory according to the Swedish patient da
+
+## Consent for publication
+
+All athours give consent to publish the article.
+
+## Availability of data and materials
+
+The datasets generated and/or analysed during the current study are not publicly available due to the European general data protection regulation (GDPR) but are available from the corresponding author on reasonable request.
+
+## Competing interests
+
+The authors declare that they have no competing interests.
+
+## Funding
+
+Not applicable.
+
+# Authors' contributions
 
@@
 
-# Exclusive licence
-
-We grant a worldwide licence to the publishers and its licensees in perpetuity, in all forms, formats and media (whether known now or created in the future), to i) publish, reproduce, distribute, display and store the Contribution, ii) translate the Contribution into other languages, create adaptations, reprints, include within collections and create summaries, extracts and/or, abstracts of the Contribution and convert or allow conversion into any format including without limitation audio, iii) create any other derivative work(s) based in whole or part on the on the Contribution, iv) to exploit all subsidiary rights to exploit all subsidiary rights that currently exist or as may exist in the future in the Contribution, v) the inclusion of electronic links from the Contribution to third party material where-ever it may be located; and, vi) licence any third party to do any or all of the above.
-
-# Competing interests statement
-
-All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi_disclosure.pdf and declare no financial relationships with any organisations that might have an interest in the submitted work in the previous three years, and no other relationships or activities that could appear to have influenced the submitted work.
-
-# Transparency Statement
-
-NH, the principal investigator of this study, affirms that this is an accurate report with no relevant data or aspects omitted.
-
-# Data sharing statement
-
-Data is available upon request to the principal investigator.
-
-# Acknowledgement
+# Acknowledgements
 

### 2020-05-23 reports/article.Rmd
updated abstract. Fix #24.

@@
 **Correspondence:** 
-erik.bulow@vgregion.se. 
+erik.bulow@gu.se. 
 Postal: Svenska Höftprotesregistret, Registercentrum Västra Götaland, SE-413 45, Sweden
-
-# Abstract
-
-**OBJECTIVES** To develop and validate a personalized model to predict 90-day mortality after total hip arthroplasty (THA).
-
-**DESIGN** Prospective multi-national population-based study.
-
-**SETTING** Swedish Hip Arthroplasty Registry (SHAR) data for model derivation and internal validation, and data from the National Joint Registry (NJR) for England and Wales for external validation.
 
@@
 shar_female <- df$P_Sex == "Female"
-age_shar <- sprintf("%.0f [%.0f - %.0f]", mean(df$P_Age), min(df$P_Age), max(df$P_Age))
+
+age_shar <- 
+  sprintf("%.0f [%.0f - %.0f]", mean(df$P_Age), min(df$P_Age), max(df$P_Age))
+
 sex_shar <- 
@@
     mean(shar_female) * 100, 
-    format(sum(shar_female), big.mark = ","), 
-    format(shar_tot, big.mark = ",")
+    format(sum(shar_female), big.mark = ",", decimal.mark = "\xB7"), 
+    format(shar_tot, big.mark = ",", decimal.mark = "\xB7")
   )
+
 deaths_shar <- 
@@
   subset(table1_combined_tot, startsWith(what, col), reg, drop = TRUE)
+
 extr_perc <- function(...) 
   gsub("([\\d,]*)( )(\\()(\\d*)(\\.\\d*)(\\))", "\\4 \\% \\[\\1 of", extr(...), perl = TRUE)
+njr_age_mean <- as.numeric(gsub("(.*) (.*)", "\\1", extr("Age", "NJR"), perl = TRUE))
 
-njr_age_mean <- as.numeric(gsub("(.*) (.*)", "\\1", extr("Age", "NJR"), perl = TRUE))
 # Same range by design (to match Swedish cohort)
@@
 )
-```
-
-**PARTICIPANTS** `r format(nrow(df), big.mark = ",")` patients with cemented total hip arthroplasty (THA) due to osteoarthritis recorded in the SHAR, and `r format(N_njr$tot, big.mark = ",")` similarly selected patients recorded in the NJR.
-
-**MAIN OUTCOME MEASURES** Predicted probability of death for any reason within 90 days after primary THA, estimated using a bootstrap ranking procedure together with a least absolute shrinkage and selection operator (LASSO) logistic regression model and evaluated by areas under the curve (AUC).
-
-```{r}
 get_auc <- function(mod) {
@@
   ) %>% 
-  {sprintf("(AUC = %s", gsub(" (", ", 95 % CI: ",., fixed = TRUE))}
+  {sprintf("(AUC = %s", gsub(" (", ", 95 % CI ",., fixed = TRUE))}
 }
-
 ```
 
-**RESULTS**
+# Abstract
 
-We found a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
+**Background:** Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality.
+However, diagnosis code-based instruments such as the Charlson co-morbidity index are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) class is imprecise.
+We thus developed a simple model to predict 90-day mortality after elective THA.
+
+**Methods:** We performed a prospective multi-national population-based registry study.
+We included `r format(nrow(df), big.mark = ",")` patients with cemented THA due to osteoarthritis from the Swedish Hip Arthroplasty Register for model derivation and internal validation, as well as `r format(N_njr$tot, big.mark = ",")` patients from the National Joint Register from England and Wales for external model validation.
+A model was developed using a bootstrap ranking procedure with a least absolute shrinkage and selection operator (LASSO) logistic regression model combined with piecewise linear regression.
+Discriminative ability was evaluated by the area under the receiver operating characteristics curve (AUC).
+Calibration belt plots were used for model calibration.
+
+**Results:** We propose a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
 The model had good discrimination, both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`.
@@
 
-**CONCLUSION:** We developed a personalized model to predict 90-day mortality after elective cemented THA based on easily available demographic data and the presence of four defined co-morbidities.
-This may facilitate individualized risk assessment prior to one of the most common surgical interventions, and a web-based calculator to aid clinical decision-making was published (`r shiny_url`).
+**Conclusions:** We developed a parsimonious model to predict 90-day mortality after elective cemented THA based on easily available demographic data and the presence of four defined co-morbidities.
+This may facilitate individualized risk assessment prior to one of the most common surgical interventions.
+We published a web-calculator to aid clinical decision-making (`r shiny_url`).
+
+**Keywords:** total hip arthroplasty, prediction model, orthopaedics, epidemiology, web calculator, validation.
 

### 2020-06-11 reports/article.Rmd
adjust refs to BJJ. Fix #41.

@@
 title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
-author: Anne Garland`*`^1,2,3^, Erik Bülow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5,6^, J. Mark Wilkinson^7^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
+author: Anne Garland, PhD`*`^1,2,3^, Erik Bulow, MSc`*`^2,4^, Erik Lenguerrand, MSc^5^, Prof Ashley Blom, PhD^5,6^, Prof J.
+Mark Wilkinson, PhD^7^, Adrian Sayers, MSc^5^, Prof Ola Rolfson, PhD^2,4^, Prof Nils P.
+Hailer, MD^1,2^
 date: '`r Sys.Date()`'
 bibliography: H:/zotero_lib.bib
-csl: bmc-medicine.csl
+csl: bjj.csl
 output: 
@@
 
-`*` Authors contributed equally.
+## Addresses
 
-**Correspondence:** 
-erik.bulow@gu.se. 
-Postal: Svenska Höftprotesregistret, Registercentrum Västra Götaland, SE-413 45, Sweden
+- AG: Andersrogatan 5, SE-621 45 Visby, SWEDEN
+- EB/OR: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45 Gothenbutg, SWEDEN
+- EL/AB/AS: Musculoskeletal Research Unit, University of Bristol, Beacon House, Queens Road, Bristol, BS8 1QU, UK
+- JMW: Department of Oncology and Metabolism, The Medical School, Beech Hill Road, Sheffield, S10 2RX, UK
+- NPH: Akademiska sjukhuset, 751 85 Uppsala, SWEDEN
+
+## Correspondence 
+Erik Bulow. + 46 70 08 234 28. erik.bulow@vgregion.se. Postal: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45 Gothenburg, SWEDEN
+
+##### PAGE BREAK
 
 ```{r}
+# Change to ugly decimal point as required by LR
+ud <- function(x) gsub("\\.", "\xB7", x)
+
 shar_tot <- nrow(df)
 shar_female <- df$P_Sex == "Female"
-
-age_shar <- 
-  sprintf("%.0f [%.0f - %.0f]", mean(df$P_Age), min(df$P_Age), max(df$P_Age))
-
+age_shar <- sprintf("%.0f [%.0f - %.0f]", mean(df$P_Age), min(df$P_Age), max(df$P_Age)) %>% ud()
 sex_shar <- 
@@
     format(shar_tot, big.mark = ",", decimal.mark = "\xB7")
-  )
-
+) %>% ud()
 deaths_shar <- 
@@
   subset(table1_combined_tot, startsWith(what, col), reg, drop = TRUE)
+extr_perc <- function(...) 
+  gsub("([\\d,]*)( )(\\()(\\d*)(\\.\\d*)(\\))", "\\4 \\% \\[\\1 of", extr(...), perl = TRUE) %>% ud()
 
-extr_perc <- function(...) 
-  gsub("([\\d,]*)( )(\\()(\\d*)(\\.\\d*)(\\))", "\\4 \\% \\[\\1 of", extr(...), perl = TRUE)
 njr_age_mean <- as.numeric(gsub("(.*) (.*)", "\\1", extr("Age", "NJR"), perl = TRUE))
-
 # Same range by design (to match Swedish cohort)
-njr_age <- sprintf("%.0f [%.0f - %.0f]", njr_age_mean, min(df$P_Age), max(df$P_Age))
-njr_sex <- sprintf("%s %s]", extr_perc("Sex", "NJR"), format(N_njr$tot, big.mark = ","))
+njr_age <- sprintf("%.0f [%.0f - %.0f]", njr_age_mean, min(df$P_Age), max(df$P_Age)) %>% ud()
+njr_sex <- sprintf("%s %s]", extr_perc("Sex", "NJR"), format(N_njr$tot, big.mark = ",")) %>% ud()
 njr_deaths <- 
@@
   njr_age, njr_sex, njr_deaths
-)
+) %>% ud()
+
 get_auc <- function(mod) {
@@
   ) %>% 
-  {sprintf("(AUC = %s", gsub(" (", ", 95 % CI ",., fixed = TRUE))}
+  {sprintf("(AUC = %s", gsub(" (", ", 95 % CI ",., fixed = TRUE))} %>% 
+  ud()
 }
+
 ```
@@
 
-**Background:** Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality.
+**Background: ** Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality.
 However, diagnosis code-based instruments such as the Charlson co-morbidity index are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) class is imprecise.
@@
 
-**Methods:** We performed a prospective multi-national population-based registry study.
+**Methods: **We performed a prospective multi-national population-based registry study.
 We included `r format(nrow(df), big.mark = ",")` patients with cemented THA due to osteoarthritis from the Swedish Hip Arthroplasty Register for model derivation and internal validation, as well as `r format(N_njr$tot, big.mark = ",")` patients from the National Joint Register from England and Wales for external model validation.
@@
 
-**Results:** We propose a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
+**Findings: ** We propose a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
 The model had good discrimination, both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`.
@@
 
-**Conclusions:** We developed a parsimonious model to predict 90-day mortality after elective cemented THA based on easily available demographic data and the presence of four defined co-morbidities.
+**Interpretation: ** We developed a parsimonious model to predict 90-day mortality after elective cemented THA based on easily available demographic data and the presence of four defined co-morbidities.
 This may facilitate individualized risk assessment prior to one of the most common surgical interventions.
@@
 
-**Keywords:** total hip arthroplasty, prediction model, orthopaedics, epidemiology, web calculator, validation.
+**Funding: ** We had no external funding.
+
+##### PAGE BREAK
+
+# Research in context
+
+## Evidence before this study
+
+We performed an initial ad hoc literature review based on Google Scholar, personal recommendations and relevant references.
+We then used 2Dsearch (https://2dsearch.com accessed 2020-05-11) to build a search query for a more systematic review: `(femur OR hip OR bone OR bones OR osteoarthritis OR rheumatoid OR rheumatism) AND (prognostic OR prediction OR forecast OR forecasting OR forecasts OR predict OR predicted OR predicting OR predictions OR predicts OR projections) AND (death OR mortality OR deaths OR survival OR survive) AND (replacement OR replaced OR replacements OR arthroplasty OR "Arthroplasty, Replacement" OR prosthesis OR implant OR prostheses OR prosthetic OR prosthetics) AND (calibration OR calibrate OR calibrated OR validation OR validate OR validated OR validating OR validity OR precision OR accuracy OR accurate OR precise OR "external validation" OR evaluation OR verification OR internal) AND (build OR building OR construct OR create OR develop OR establish OR train OR derive OR derivation) AND (models OR modelling OR "statistical model" OR modeling) AND (comorbidity OR comorbid OR co-morbid OR co-occurring OR diagnoses OR co-morbidity OR Elixhauser OR Charlson OR "comorbidity index" OR "comorbidity score" OR "Rx Risk V" OR ASA)`.
+We used no time or language constraints and used this query in PubMed (https://pubmed.gov accessed 2020-05-11), and identified 143 articles, whereof one was a duplicate.
+We performed an article screening with Rayyan (https://rayyan.qcri.org/ accessed 2020-05-12).
+131 articles were excluded based on titles, and an additional 6 after reading the abstracts.
+5 articles remained, whereof 3 were already considered during the ad hoc review process.
+Research on cadavers, canines, and patients with diagnoses other than elective hip disorders (including hip fractures) were excluded during the screening process, as were prediction models with outcomes other than mortality.
+In addition, we used Open Grey (https://opengrey.eu accessed 2020-05-12) with the query above, as well as a simpler query (`hip AND prediction`).
+We found 9 items, whereof none was considered relevant.
+The same queries were used with Epistemonikos (https://epistemonikos.org accessed 2020-05-12).
+No items were found using the longer query, but 33 systematic reviews based on the shorter.
+30 where excluded after title screening, and the remaining 3 after screening of abstracts.
+
+In summary, several prediction models have been developed for short-term mortality in patients operated with total hip arthroplasty (THA).
+Most of them are based on small sample sizes, make use of too elaborate predictors based on data that are unavailable in a clinical setting, only consider in-hospital deaths, are inadequately internally or externally validated or not validated at all, or have no national population coverage.
+Only few model descriptions are accompanied by a practical tool for use in clinical practice (often only nomograms or a mathematical formula are supplied).
+Some relevant models are further described in the main text.
+Meta-analysis is most relevant for pooling individual effect estimates or measures of predictive power for a model validated on different cohorts.
+Hence, no relevant meta-analysis was applicable to our study.
+
+## Added value of this study
+
+Our study is the first to develop and externally validate a parsimonious prediction model for early mortality after THA based on two European cohorts, and to simultaneously provide a simple web-calculator for use in clinical practice.
+
+## Implications of all the available evidence
+
+Prediction models must consider all aspects of the relevant cohorts, including different health care systems, population demographics, and treatment traditions.
+Our tool was derived on a Swedish cohort and then validated on a similar cohort from England and Wales.
+We believe that the model might have broader clinical relevance and is suitable for clinical usage.
+
+##### PAGE BREAK
 
@@
 In terms of gained quality of life and cost-utility, total hip arthroplasty (THA) is an enormously successful procedure.
-90-day mortality after THA performed due to osteoarthritis is low, ranging from 0.2 % to 0.6 % in large joint registry studies [@Hunt2013; @Garland2017].
-The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.
+90-day mortality after THA performed due to osteoarthritis is low, ranging from `r 0.2` % to `r 0.6` % in large joint registry studies.[@Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.
 
-Co-morbidity is associated with a shorter remaining life span, but the Charlson and Elixhauser co-morbidity indices poorly predict mortality after THA [@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017; @Bulow2017].
-Additionally, these complex co-morbidity instruments are based on information from extensive in- or out-patient databases, each measure being defined by 1,178 and 1,516 individual International Classification of Diseases (ICD)-10 codes, respectively.
-Each co-morbidity index also exists in numerous versions, making interpretation and comparisons difficult [@Brusselaers2017; @Walraven2009].
-The use of these indices is therefore limited to research settings but they are impossible to use under clinical circumstances.
+Co-morbidity is associated with a shorter remaining life span, but the Charlson and Elixhauser co-morbidity indices poorly predict mortality after THA.[@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017; @Bulow2017; @Inacio2015] Additionally, these complex co-morbidity instruments are based on information from extensive in- or out-patient databases, each measure being defined by 1,178 and 1,516 individual International Classification of Diseases (ICD)-10 codes, respectively.
+Each co-morbidity index also exists in numerous versions, making interpretation and comparisons difficult.[@Brusselaers2017] The use of these indices is therefore limited to research settings but they are impossible to use under clinical circumstances.
 
-Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery [@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019].
-The first prediction model based on a European cohort was the patient decision support tool for joint remplacement in the UK (https://jointcalc.shef.ac.uk/).
+Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery.[@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] The first prediction model based on a European cohort was the patient decision support tool for joint replacement in the UK (https://jointcalc.shef.ac.uk/).
 However, it remains unclear to what level the depth of co-morbidity characterization influences the accuracy of  mortality prediction models.
 
-Here we aimed to develop a parsimonious prediction model of 90-day postoperative mortality after THA with internal and external validation of discrimination and calibration, and to compare this to the accuracy of prediction models based on existing co-morbidity measures [@Manning2016].
+Here we aimed to develop a parsimonious prediction model of 90-day postoperative mortality after THA with internal and external validation of discrimination and calibration, and to compare this to the accuracy of prediction models based on existing co-morbidity measures.[@Manning2016]
 
@@
 Cementation is the most common fixation technique used in Sweden, thus used as inclusion criteria to decrease heterogeneity.
-Only the last operated hip was accounted for in patients with bilateral THA, since those are more likely to resemble patients with unilateral THA (Figure \@ref(fig:flowchart)) [@Bulow2020].
+Only the last operated hip was accounted for in patients with bilateral THA, since those are more likely to resemble patients with unilateral THA (Figure \@ref(fig:flowchart)).[@Bulow2020]
 
@@
 The observation period ended 2015, since we had access to co-morbidity from the national patient register up to this point.
-Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration [@Ludvigsson2009].
-We used data from a variety of sources, as previously described [@Cnudde2016].
-Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98 % [@Karrholm2018].
-Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden [@Ludvigsson2019].
-Civil status (married/un-married/divorced/widow[er]), were also collected from LISA.
+Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration.
+We used data from a variety of sources, as previously described.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98 %.[@Karrholm2018] Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status (married/un-married/divorced/widow[er]), were also collected from LISA.
 The Swedish national patient register (NPR) was used to assess co-morbidity during the year preceding index surgery.
 This register contains all relevant diagnoses coded by ICD-10, as well as dates of admission and discharge for in- and outpatient episodes in all private and public hospitals.
-Completeness for NPR is above 99 % and 85-95 % of all diagnostic codes are valid [@Ludvigsson2011].
-Death dates were linked from the national population register.
+Completeness for NPR is above 99 % and 85-95 % of all diagnostic codes are valid.[@Ludvigsson2011] Death dates were linked from the national population register.
 
@@
 The validation cohort was created by linking NJR data to the secondary uses service database from the National Health Service (NHS).
-This database includes co-morbidity recorded in the admitted patient care module of the hospital episode statistics (HES) database in England, and the patient episode database for Wales [@Brophy2006; @Burns2012; @Sinha2013; @Simmonds2014; @Thorn2016].
-Out-patient deaths were linked from the office of national statistics.
-Data linkage used NHS-numbers, as well as patient name, age, sex and address [@Sayers2016].
+This database includes co-morbidity recorded in the admitted patient care module of the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Thorn2016] Out-patient deaths were linked from the office of national statistics.
+Data linkage used NHS-numbers, as well as patient name, age, sex and address.[@Sayers2016]
 
@@
 
-Co-morbidity was defined by individual ICD-10 codes grouped into 17 categories according to Charlson [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to Elixhauser [@Elixhauser1998; @Quan2005].
-Some co-morbidities were identified in both indices, and some distinct co-morbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
+Co-morbidity was defined by individual ICD-10 codes grouped into 17 categories according to Charlson and 31 categories according to Elixhauser.[@Quan2005] Some co-morbidities were identified in both indices, and some distinct co-morbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
 We combined individual diagnostic groups to establish `r nrow(tab_categorization)` broader categories of co-morbidity (Table \@ref(tab:tabcategorization)).
@@
 Further analysis was based on logistic regression, since no loss to follow up occurred in the Swedish cohort within the first 90 days, and we assumed the same was true for patients from England and Wales.
-We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO) [@Guo2015; @Baranowski2020].
-Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1.
+We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015] Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1.
 Thus, the magnitude (absolute values) of their estimated coefficients would indicate variable importance on the same scale as the categorical variables.
 Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process.
-1,000 bootstrap samples were drawn from the observed data set [@Austin2004].
+1,000 bootstrap samples were drawn from the observed data set.
 We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model.
@@
 Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.
-Odds ratios for the final model were estimated with 95 % confidence intervals [@Ripley2002].
+Odds ratios for the final model were estimated with 95 % confidence intervals.
 
@@
 Each model was used to predict the probability of death within 90 days for patients from the SHAR (internal validation).
-Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability [@Fawcett2006].
-Models with a lower 95 % confidence limit above 0.7, were considered good.
+Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.
+Models with a lower 95 % confidence limit above `r 0.7`, were considered good.
 Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples.
-We used the bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples to adjust for optimism [@Miller1991].
-Calibration bands were made to graphically assess model calibration, comparing predicted probabilities and observed proportions [@Nattino2016].
-The reduced model was then evaluated externally.
+We used the bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples to adjust for optimism.[@Miller1991] Calibration bands were made to graphically assess model calibration, comparing predicted probabilities and observed proportions.[@Nattino2016] The reduced model was then evaluated externally.
 An AUC with 95 % confidence interval was calculated for the model as-is.
 Re-calibration of the model intercept was then performed to account for different mortality rates in Sweden compared to England and Wales.
-An updated over-all slope was also calculated to account for country-specific treatment differences [@Steyerberg2004].
-Calibration for this re-calibrated model was illustrated in the same calibration belt plot as for the internal calibration.
+An updated over-all slope was also calculated to account for country-specific treatment differences.[@Steyerberg2004] Calibration for this re-calibrated model was illustrated in the same calibration belt plot as for the internal calibration.
 
@@
 A linked Binder environment is also available for interactive usage.
-ta law from 2009, and the UK law for pseudonymised data.
+
+## Ethical approval
+
+Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13) and by the NJR in the UK (RSC2017/21).
+Informed consent was not mandatory according to the Swedish patient data law from 2009, and the UK law for pseudonymised data.
 
@@
 Calibration of the re-calibrated model was slightly inferior compared to the internal calibration.
-Predicted probabilities between 0.5 % and 1.5 % were lower than observed proportions and the estimated 95 % confidence bands were wider.
-Over-all, calibration was still good however for predicted probabilities below 3 % and acceptable below 5 % (Figure \@ref(fig:calibration)).
+Predicted probabilities between `r 0.5` % and `r 1.5` % were lower than observed proportions and the estimated 95 % confidence bands were wider.
+Over-all, calibration was good for predicted probabilities below 3 % and acceptable below 5 % (Figure \@ref(fig:calibration)).
 
@@
 The observed proportions of deaths for patients with similar characteristics is likely lower (as indicated by Figure \@ref(fig:calibration)).
-Some risk calculators ignore this problem [@Bozic2013a; @Harris2019], but we think this should be acknowledged.
+Some risk calculators ignore this problem,[@Bozic2013a; @Harris2019] but we think this should be acknowledged.
 
 Variables in our model were chosen based on statistical properties and not due to clinical relevance.
-Variables and estimated coefficients should therefore not be assigned any exact epidemiological and/or causal meaning per se [@Breiman2001].
+Variables and estimated coefficients should therefore not be assigned any exact epidemiological and/or causal meaning per se.
 
@@
 This is clinically reasonable since a label of "severe systemic disease" should be based on a relevant patient assessment prior to surgery.
-ASA class is nevertheless known to have a high degree of internal variability [@Haynes1995].
-It has previously been compared to the Charlson co-morbidity index, but not with respect to mortality after THA [@Whitmore2014; @Kork2015].
+ASA class is nevertheless known to have a moderate or high degree of internal variability.[@Sankar2004] It has been compared to the Charlson co-morbidity index, but not with respect to mortality after THA.
 Patients with ASA class IV-VI were excluded since those categories describe severe disease, moribund and brain-dead individuals.
@@
 
-Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity [@Mood2010].
-We noted that the proportion of patients with BMI above 30, the WHO definition of obesity was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) in the Swedish cohort than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the national patient register (`r round(mean(df$c_obesity) * 100, 1)` %).
+Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE) %>% ud()`) but was still relevant as a predictor due to unobserved heterogeneity.
+We noted that the proportion of patients registered in the SHAR with BMI above 30, the WHO definition of obesity, was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the Swedish national patient register (`r round(mean(df$c_obesity) * 100, 1)` %).
 We believe this might be due to the fact that BMI is systematically recorded in the SHAR, but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.
@@
 A strength of this study is its nationwide design with data from both Sweden, England and Wales.
-The Swedish registers are valid with low proportions of missing data [@Soderman2000; @Karrholm2018; @Soderman2001; @Ludvigsson2011].
-Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however [@Spencer2012; @Williams2002; @Brennan2012].
-A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80 % [@Burns2012].
-Both in- and outpatient co-morbidity data were available for Sweden, but only in-patient data for England and Wales.
+The Swedish registers are valid with low proportions of missing data.[@Karrholm2018; @Ludvigsson2011] Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however.
+A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80 %.[@Burns2012] Both in- and outpatient co-morbidity data were available for Sweden, but only in-patient data for England and Wales.
 We did not have access to any data from primary care or nursing homes in any of the counties.
 
-The linkage procedure between NJR and HES has been previously described [@Smith2012], where it was noted that privately funded patients were not included in HES.
+The linkage procedure between NJR and HES has been previously described,[@Smith2012] where it was noted that privately funded patients were not included in HES.
 17 % of the patients had missing personal data or did not allow linkage, and 6 % were not found in HES although their data were available from the NJR.
@@
 A similar approach might be relevant for BMI, which we only modelled as a linear main effect.
-A U-shaped association between mortality and BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales [@Mouchti2018].
-Our recordings of BMI might be imprecise however [@Stommel2009].
-Some hospitals might report actual measurements but it is unknown to what extent those data are self-estimated by the patients or estimated by health care personnel.
+A U-shaped association between mortality and BMI categorized as underweight (below `r 18.5` kg/m^2^), normal (`r 8.5` to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales.[@Mouchti2018] Our recordings of BMI might be imprecise however.
+Some hospitals report actual measurements but it is unknown to what extent those data are self-estimated by the patients or by health care personnel.
 Therefore, we aimed to avoid overfitting BMI based on too elaborate transformations.
@@
 
-# Declarations
+##### PAGE BREAK
 
-## Ethics approval and consent to participate
-
-Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13) and by the NJR in the UK (RSC2017/21).
-Informed consent was not mandatory according to the Swedish patient da
-
-## Consent for publication
-
-All athours give consent to publish the article.
-
-## Availability of data and materials
-
-The datasets generated and/or analysed during the current study are not publicly available due to the European general data protection regulation (GDPR) but are available from the corresponding author on reasonable request.
-
-## Competing interests
-
-The authors declare that they have no competing interests.
-
-## Funding
-
-Not applicable.
-
-# Authors' contributions
+# Contribution of authors
 
@@
 
-# Acknowledgements
+# Acknowledgement
 
@@
 We are grateful to the Healthcare Quality Improvement Partnership (HQIP), the NJR Research Committee and staff at the NJR Centre for facilitating this work.
-We also thank all contributers to the Swedish Hip Arthroplasty Register (SHAR), especially Szilard Nemes, previous senior statistician, for his involvement in the study.
+We also thank all contributors to the Swedish Hip Arthroplasty Register (SHAR), especially Szilard Nemes, previous senior statistician, for his involvement in the study.
 The authors have conformed to the NJR's standard protocol for data access and publication.
 The views expressed represent those of the authors and do not necessarily reflect those of the National Joint Registry Steering Committee or the Healthcare Quality Improvement Partnership (HQIP) who do not vouch for how the information is presented.
+
+##### PAGE BREAK
 
@@
 
+##### PAGE BREAK
+
 # Bibliography

### 2020-06-11 reports/article.Rmd
removed research box required by LR but not by BJJ.

@@
 **Funding: ** We had no external funding.
-
-##### PAGE BREAK
-
-# Research in context
-
-## Evidence before this study
-
-We performed an initial ad hoc literature review based on Google Scholar, personal recommendations and relevant references.
-We then used 2Dsearch (https://2dsearch.com accessed 2020-05-11) to build a search query for a more systematic review: `(femur OR hip OR bone OR bones OR osteoarthritis OR rheumatoid OR rheumatism) AND (prognostic OR prediction OR forecast OR forecasting OR forecasts OR predict OR predicted OR predicting OR predictions OR predicts OR projections) AND (death OR mortality OR deaths OR survival OR survive) AND (replacement OR replaced OR replacements OR arthroplasty OR "Arthroplasty, Replacement" OR prosthesis OR implant OR prostheses OR prosthetic OR prosthetics) AND (calibration OR calibrate OR calibrated OR validation OR validate OR validated OR validating OR validity OR precision OR accuracy OR accurate OR precise OR "external validation" OR evaluation OR verification OR internal) AND (build OR building OR construct OR create OR develop OR establish OR train OR derive OR derivation) AND (models OR modelling OR "statistical model" OR modeling) AND (comorbidity OR comorbid OR co-morbid OR co-occurring OR diagnoses OR co-morbidity OR Elixhauser OR Charlson OR "comorbidity index" OR "comorbidity score" OR "Rx Risk V" OR ASA)`.
-We used no time or language constraints and used this query in PubMed (https://pubmed.gov accessed 2020-05-11), and identified 143 articles, whereof one was a duplicate.
-We performed an article screening with Rayyan (https://rayyan.qcri.org/ accessed 2020-05-12).
-131 articles were excluded based on titles, and an additional 6 after reading the abstracts.
-5 articles remained, whereof 3 were already considered during the ad hoc review process.
-Research on cadavers, canines, and patients with diagnoses other than elective hip disorders (including hip fractures) were excluded during the screening process, as were prediction models with outcomes other than mortality.
-In addition, we used Open Grey (https://opengrey.eu accessed 2020-05-12) with the query above, as well as a simpler query (`hip AND prediction`).
-We found 9 items, whereof none was considered relevant.
-The same queries were used with Epistemonikos (https://epistemonikos.org accessed 2020-05-12).
-No items were found using the longer query, but 33 systematic reviews based on the shorter.
-30 where excluded after title screening, and the remaining 3 after screening of abstracts.
-
-In summary, several prediction models have been developed for short-term mortality in patients operated with total hip arthroplasty (THA).
-Most of them are based on small sample sizes, make use of too elaborate predictors based on data that are unavailable in a clinical setting, only consider in-hospital deaths, are inadequately internally or externally validated or not validated at all, or have no national population coverage.
-Only few model descriptions are accompanied by a practical tool for use in clinical practice (often only nomograms or a mathematical formula are supplied).
-Some relevant models are further described in the main text.
-Meta-analysis is most relevant for pooling individual effect estimates or measures of predictive power for a model validated on different cohorts.
-Hence, no relevant meta-analysis was applicable to our study.
-
-## Added value of this study
-
-Our study is the first to develop and externally validate a parsimonious prediction model for early mortality after THA based on two European cohorts, and to simultaneously provide a simple web-calculator for use in clinical practice.
-
-## Implications of all the available evidence
-
-Prediction models must consider all aspects of the relevant cohorts, including different health care systems, population demographics, and treatment traditions.
-Our tool was derived on a Swedish cohort and then validated on a similar cohort from England and Wales.
-We believe that the model might have broader clinical relevance and is suitable for clinical usage.
 

### 2020-06-11 reports/article.Rmd
Abstract for BJJ. Fix #36.

@@
 
-**Background: ** Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality.
-However, diagnosis code-based instruments such as the Charlson co-morbidity index are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) class is imprecise.
-We thus developed a simple model to predict 90-day mortality after elective THA.
+**Aims: ** To develop and externally validate a parsimonious statistical prediction model of 90-day mortality after elective Total Hip Arthroplasty (THA), and to provide a web-calculator for clinical usage.
 
-**Methods: **We performed a prospective multi-national population-based registry study.
-We included `r format(nrow(df), big.mark = ",")` patients with cemented THA due to osteoarthritis from the Swedish Hip Arthroplasty Register for model derivation and internal validation, as well as `r format(N_njr$tot, big.mark = ",")` patients from the National Joint Register from England and Wales for external model validation.
+**Patients and methods: ** We included `r format(nrow(df), big.mark = ",")` patients with cemented THA due to osteoarthritis from the Swedish Hip Arthroplasty Register for model derivation and internal validation, as well as `r format(N_njr$tot, big.mark = ",")` patients from the National Joint Register from England and Wales for external model validation.
 A model was developed using a bootstrap ranking procedure with a least absolute shrinkage and selection operator (LASSO) logistic regression model combined with piecewise linear regression.
@@
 
-**Findings: ** We propose a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
+**Results: ** We propose a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
 The model had good discrimination, both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`.
@@
 
-**Interpretation: ** We developed a parsimonious model to predict 90-day mortality after elective cemented THA based on easily available demographic data and the presence of four defined co-morbidities.
-This may facilitate individualized risk assessment prior to one of the most common surgical interventions.
+**Concusion: ** We developed a parsimonious model which may facilitate individualized risk assessment prior to one of the most common surgical interventions.
 We published a web-calculator to aid clinical decision-making (`r shiny_url`).
 
-**Funding: ** We had no external funding.
+**Clinical relevance:**
+
+- We developed a parsimonious model for risk assessment of 90-day mortality after elective THA. 
+- The model considers age, sex, ASA class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
+- We published an on-line web-calculator (`r shiny_url`).
 

### 2020-06-11 reports/article.Rmd
Ändrat presentation av numeriska värden. Fix #35.

@@
 
-`r format(nrow(df), big.mark = ",")` patients (age `r paste(range(df$P_Age), collapse = " - ")`, `r round(mean(df$P_Sex == "Female") * 100)` % females) were included in the derivation cohort from SHAR (Figure \@ref(fig:flowchart) left panel).
+`r format(nrow(df), big.mark = ",")` patients (mean age `r mean(df$P_Age)` years, range `r paste(range(df$P_Age), collapse = " - ")` and `r round(mean(df$P_Sex == "Female") * 100)` % females) were included in the derivation cohort from SHAR (Figure \@ref(fig:flowchart) left panel).
 `r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100)) %>% ud()` of those patients died within 90 days and no one was censored before that.
@@
 This is clinically reasonable since a label of "severe systemic disease" should be based on a relevant patient assessment prior to surgery.
-ASA class is nevertheless known to have a moderate or high degree of internal variability.[@Sankar2004] It has been compared to the Charlson co-morbidity index, but not with respect to mortality after THA.
+ASA class is nevertheless known to have a moderate or high degree of internal variability.[@Sankar2014] It has been compared to the Charlson co-morbidity index, but not with respect to mortality after THA.
 Patients with ASA class IV-VI were excluded since those categories describe severe disease, moribund and brain-dead individuals.
@@
 Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE) %>% ud()`) but was still relevant as a predictor due to unobserved heterogeneity.
-We noted that the proportion of patients registered in the SHAR with BMI above 30, the WHO definition of obesity, was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the Swedish national patient register (`r round(mean(df$c_obesity) * 100, 1)` %).
+We noted that the proportion of patients registered in the SHAR with BMI above 30, the WHO definition of obesity, was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %, n = `r format(sum(df$P_BMI > 30 & !is.na(df$P_BMI)), big.mark = ",")`) than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the Swedish national patient register (`r round(mean(df$c_obesity) * 100, 1)` %, , n = `r format(sum(df$c_obesity & !is.na(df$c_obesity)), big.mark = ",")`).
 We believe this might be due to the fact that BMI is systematically recorded in the SHAR, but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.

### 2020-06-11 reports/article.Rmd
revert to normal decimal points. Fix #.

@@
 ```{r}
-# Change to ugly decimal point as required by LR
-ud <- function(x) gsub("\\.", "\xB7", x)
-
 shar_tot <- nrow(df)
 shar_female <- df$P_Sex == "Female"
-age_shar <- sprintf("%.0f [%.0f - %.0f]", mean(df$P_Age), min(df$P_Age), max(df$P_Age)) %>% ud()
+age_shar <- sprintf("%.0f [%.0f - %.0f]", mean(df$P_Age), min(df$P_Age), max(df$P_Age))
 sex_shar <- 
@@
   njr_age, njr_sex, njr_deaths
-) %>% ud()
+)
 
@@
   ) %>% 
-  {sprintf("(AUC = %s", gsub(" (", ", 95 % CI ",., fixed = TRUE))} %>% 
-  ud()
+  {sprintf("(AUC = %s", gsub(" (", ", 95 % CI ",., fixed = TRUE))}
 }
@@
 `r format(nrow(df), big.mark = ",")` patients (mean age `r mean(df$P_Age)` years, range `r paste(range(df$P_Age), collapse = " - ")` and `r round(mean(df$P_Sex == "Female") * 100)` % females) were included in the derivation cohort from SHAR (Figure \@ref(fig:flowchart) left panel).
-`r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100)) %>% ud()` of those patients died within 90 days and no one was censored before that.
-The unadjusted risk of 90-day mortality was therefore `r ud(surv90d)`.
+`r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` of those patients died within 90 days and no one was censored before that.
+The unadjusted risk of 90-day mortality was therefore `r surv90d`.
 Further characteristics of the study population are presented in Table \@ref(tab:tab1).
@@
 
-Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE) %>% ud()`) but was still relevant as a predictor due to unobserved heterogeneity.
+Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity.
 We noted that the proportion of patients registered in the SHAR with BMI above 30, the WHO definition of obesity, was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %, n = `r format(sum(df$P_BMI > 30 & !is.na(df$P_BMI)), big.mark = ",")`) than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the Swedish national patient register (`r round(mean(df$c_obesity) * 100, 1)` %, , n = `r format(sum(df$c_obesity & !is.na(df$c_obesity)), big.mark = ",")`).

### 2020-06-11 reports/tabs.Rmd
revert to normal decimal points. Fix #.

--- 2020-03-27 reports/tabs.Rmd
@@
 ```{r tab1}
+table1_combined_tot %>% 
+  mutate(what = gsub("Heart infarct", "Myocardial infarction", what)) %>% 
 knitr::kable(
-  mutate(table1_combined_tot, what = gsub("Heart infarct", "Myocardial infarction", what)), 
     caption = "Characteristics of the study population in the model derivation cohort (SHAR) and external validation cohort (NJR).

### 2020-06-11 reports/article.Rmd
Removed details not needed.

@@
 title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
-author: Anne Garland, PhD`*`^1,2,3^, Erik Bulow, MSc`*`^2,4^, Erik Lenguerrand, MSc^5^, Prof Ashley Blom, PhD^5,6^, Prof J.
-Mark Wilkinson, PhD^7^, Adrian Sayers, MSc^5^, Prof Ola Rolfson, PhD^2,4^, Prof Nils P.
-Hailer, MD^1,2^
+author: Anne Garland`*`^1,2,3^, Erik Bulow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5,6^, J. Mark Wilkinson^7^, Adrian Sayers^5^, Ola Rolfson2,4^, Nils P. Hailer^1,2^
 date: '`r Sys.Date()`'
@@
 7. Department of Oncology and Metabolism, University of Sheffield, Sheffield, UK.
-
-## Addresses
-
-- AG: Andersrogatan 5, SE-621 45 Visby, SWEDEN
-- EB/OR: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45 Gothenbutg, SWEDEN
-- EL/AB/AS: Musculoskeletal Research Unit, University of Bristol, Beacon House, Queens Road, Bristol, BS8 1QU, UK
-- JMW: Department of Oncology and Metabolism, The Medical School, Beech Hill Road, Sheffield, S10 2RX, UK
-- NPH: Akademiska sjukhuset, 751 85 Uppsala, SWEDEN
 
@@
 
-Here we aimed to develop a parsimonious prediction model of 90-day postoperative mortality after THA with internal and external validation of discrimination and calibration, and to compare this to the accuracy of prediction models based on existing co-morbidity measures.[@Manning2016]
+We aimed to develop a parsimonious prediction model of 90-day postoperative mortality after THA with internal and external validation of discrimination and calibration, and to compare this to the accuracy of prediction models based on existing co-morbidity measures.[@Manning2016]
 

### 2020-06-11 reports/article.Rmd
number of patients in MM. Fix #38.

@@
 Patients with missing information on BMI or a measurement above 50 were excluded, as were patients with missing information on ASA class or class IV and above, as well as patients with unknown educational level or type of hospital.
+`r format(nrow(df), big.mark = ",")` patients (mean age `r mean(df$P_Age)` years, range `r paste(range(df$P_Age), collapse = " - ")` and `r round(mean(df$P_Sex == "Female") * 100)` % females) were included (Figure \@ref(fig:flowchart) left panel).
 
@@
 We excluded no patients based on missing information on BMI, educational level, or civil status however, since those variables were not used in the final model.
+`r format(N_njr$tot, big.mark = ",")` patients were included (Figure \@ref(fig:flowchart) right panel).
 The validation cohort was created by linking NJR data to the secondary uses service database from the National Health Service (NHS).
@@
 Odds ratios for the final model were estimated with 95 % confidence intervals.
+$p$-values below 0.05 were considered statistically significant.
 
@@
 
-`r format(nrow(df), big.mark = ",")` patients (mean age `r mean(df$P_Age)` years, range `r paste(range(df$P_Age), collapse = " - ")` and `r round(mean(df$P_Sex == "Female") * 100)` % females) were included in the derivation cohort from SHAR (Figure \@ref(fig:flowchart) left panel).
-`r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` of those patients died within 90 days and no one was censored before that.
+`r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` of the patients from SHAR died within 90 days and no one was censored before that.
 The unadjusted risk of 90-day mortality was therefore `r surv90d`.
@@
 
-In addition, `r format(N_njr$tot, big.mark = ",")` patients were included for the external validation cohort from England and Wales (Figure \@ref(fig:flowchart) right panel).
-Their unadjusted risk of 90 day mortality was `r njr_surv`, which was higher compared to the derivation cohort from Sweden.
+Patients from NJR had an unadjusted risk of 90 day mortality of `r njr_surv`, which was higher compared to the derivation cohort from Sweden.
 

### 2020-06-11 reports/article.Rmd
removed some redundant tble data from te text. Fix #43.

@@
 title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
-author: Anne Garland`*`^1,2,3^, Erik Bulow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5,6^, J. Mark Wilkinson^7^, Adrian Sayers^5^, Ola Rolfson2,4^, Nils P. Hailer^1,2^
+author: Anne Garland`*`^1,2,3^, Erik Bulow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5,6^, J. Mark Wilkinson^7^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
 date: '`r Sys.Date()`'
@@
 Further characteristics of the study population are presented in Table \@ref(tab:tab1).
-`r round(mean(df$CCI_index_quan_original != 0) * 100)` % and `r round(mean(df$ECI_index_sum_all  != 0) * 100)` % of the Swedish patients had at least one pre-surgery co-morbidity according to Charlson and Elixhauser.
-The proportion of patients with ASA class III was `r round(mean(df$P_ASA == 3) * 100)` %.
-
 Patients from NJR had an unadjusted risk of 90 day mortality of `r njr_surv`, which was higher compared to the derivation cohort from Sweden.
@@
 Calibration of the re-calibrated model was slightly inferior compared to the internal calibration.
-Predicted probabilities between `r 0.5` % and `r 1.5` % were lower than observed proportions and the estimated 95 % confidence bands were wider.
+Predicted probabilities between 0.5 % and 1.5 % were lower than observed proportions and the estimated 95 % confidence bands were wider.
 Over-all, calibration was good for predicted probabilities below 3 % and acceptable below 5 % (Figure \@ref(fig:calibration)).

### 2020-06-11 reports/article.Rmd
prepared figure formats. Fix #32. Fix #42.

@@
 Cementation is the most common fixation technique used in Sweden, thus used as inclusion criteria to decrease heterogeneity.
-Only the last operated hip was accounted for in patients with bilateral THA, since those are more likely to resemble patients with unilateral THA (Figure \@ref(fig:flowchart)).[@Bulow2020]
+Only the last operated hip was accounted for in patients with bilateral THA, since those are more likely to resemble patients with unilateral THA (Figure \@ref(fig:flowchartshar)).[@Bulow2020]
 
@@
 Patients with missing information on BMI or a measurement above 50 were excluded, as were patients with missing information on ASA class or class IV and above, as well as patients with unknown educational level or type of hospital.
-`r format(nrow(df), big.mark = ",")` patients (mean age `r mean(df$P_Age)` years, range `r paste(range(df$P_Age), collapse = " - ")` and `r round(mean(df$P_Sex == "Female") * 100)` % females) were included (Figure \@ref(fig:flowchart) left panel).
+`r format(nrow(df), big.mark = ",")` patients (mean age `r mean(df$P_Age)` years, range `r paste(range(df$P_Age), collapse = " - ")` and `r round(mean(df$P_Sex == "Female") * 100)` % females) were included (Figure \@ref(fig:flowchartshar)).
 
@@
 We excluded no patients based on missing information on BMI, educational level, or civil status however, since those variables were not used in the final model.
-`r format(N_njr$tot, big.mark = ",")` patients were included (Figure \@ref(fig:flowchart) right panel).
+`r format(N_njr$tot, big.mark = ",")` patients were included (Figure \@ref(fig:flowchartnjr)).
 The validation cohort was created by linking NJR data to the secondary uses service database from the National Health Service (NHS).
@@
 We therefore considered the reduced model as superior due to its simplicity.
-Traditional models performed poorly with 95 % confidence intervals not above 0.7: The Charlson co-morbitiy index had an AUC of `r half_paran("Charlson")` and the Elixhauser co-morbidity an AUC of `r gsub("\\)", ";", half_paran("Elixhauser"))` Figure \@ref(fig:rocs) left panel and Figure \@ref(fig:aucci)).
+Traditional models performed poorly with 95 % confidence intervals not above 0.7: The Charlson co-morbitiy index had an AUC of `r half_paran("Charlson")` and the Elixhauser co-morbidity an AUC of `r gsub("\\)", ";", half_paran("Elixhauser"))` Figure \@ref(fig:rocvalidation) and Figure \@ref(fig:aucci)).
 
-The ability of the reduced model to estimate probabilities of death within 90 days is further illustrated in Figure \@ref(fig:sep).
+The ability of the reduced model to estimate probabilities of death within 90 days is further illustrated in Figure \@ref(fig:separationhist) and \@ref(fig:separationdensity).
 Model calibration was good for estimated probabilities up to 3 % and acceptable up to 5 %, although with predicted probabilities usually higher than observed proportions (Figure \@ref(fig:calibration)).
@@
 
-The discriminative ability for the reduced model was not statistically significantly different when applied to the external validation cohort `r get_auc("NJR (Reduced model)")` compared to the internal validation `r get_auc("Reduced model")` (Figure \@ref(fig:rocs) right panel and Figure \@ref(fig:aucci)).
+The discriminative ability for the reduced model was not statistically significantly different when applied to the external validation cohort `r get_auc("NJR (Reduced model)")` compared to the internal validation `r get_auc("Reduced model")` (Figure \@ref(fig:rocvalidation) and Figure \@ref(fig:aucci)).
 Calibration of the re-calibrated model was slightly inferior compared to the internal calibration.

### 2020-06-11 reports/figs.Rmd
fix decimal points. Fix #11.
prepared figure formats. Fix #32. Fix #42.

@@
 
-```{r flowchart, fig.cap = "Flowchart depicting inclusion criteria and number of patients.
-Data from the Swedish Hip Arthroplasty Register were used for model derivation and internal validation (left).
-Data from the National Joint Registry in England and Wales were used for external validation (right)."}
-knitr::include_graphics("../graphs/flowchart.png")
+```{r flowchartshar, fig.cap = "Flowchart depicting inclusion criteria and number of patients from the Swedish derivation cohort."}
+# knitr::include_graphics("../graphs/flowchart_shar.ps")
+knitr::include_graphics("../graphs/images.png")
 ```
 
-##### PAGE BREAK
+```{r flowchartnjr, fig.cap = "Flowchart depicting inclusion criteria and number of patients for the external validation cohort from England and Wales."}
+# knitr::include_graphics("../graphs/flowchart_njr.ps")
+knitr::include_graphics("../graphs/images.png")
 
-```{r rocs, fig.cap = "Receiver Operation Characteristics (ROC) curves.
-The area under the ROC curve (AUC) for the reduced model derived on data from the SHAR was very similar to the main model (left panel).
-External validation of this model, on data from the National Joint Registry for England and Wales, produced a similar curve as well (right panel)."}
-knitr::include_graphics("../graphs/roc.png")
 ```
 
-##### PAGE BREAK
+```{r rocderivation, fig.cap = "Receiver Operation Characteristics (ROC) curves.
+The area under the ROC curve (AUC) for the reduced model derived on data from the SHAR was very similar to the main model"}
+# knitr::include_graphics("../graphs/roc_derivation.eps")
+knitr::include_graphics("../graphs/images.png")
+
+```
+
+```{r rocvalidation, fig.cap = "Receiver Operation Characteristics (ROC) curves.
+There was no difference when applying the reduced model to the external validation cohort from England and Wales ,compared to the Swedish derivation cohort."}
+# knitr::include_graphics("../graphs/roc_validation.eps")
+knitr::include_graphics("../graphs/images.png")
+```
 
@@
 
-knitr::include_graphics("../graphs/auc_ci.png")
+# knitr::include_graphics("../graphs/auc_ci.eps")
+knitr::include_graphics("../graphs/images.png")
 ```
 
-##### PAGE BREAK
-
-```{r sep, fig.cap = "The vast majority of patients survived more than 90 days after THA, and therefore the blue bars dominate the histogram (upper panel; note the scales).
-A normalized density plot however reveals that patients who died within 90 days were, on average, estimated to have a higher probability to do so (lower panel).
+```{r separationhist, fig.cap = "The vast majority of patients survived more than 90 days after THA, and therefore the blue bars dominate the histogram (note the scales).
 There were very few observations with covariate patterns resulting in death probabilities higher than 5 % (green line).
 Estimated probabilities above this limit are therefore subject to extrapolation."}
-knitr::include_graphics("../graphs/separation.png")
+# knitr::include_graphics("../graphs/separation_hist.tiff")
+knitr::include_graphics("../graphs/images.png")
 ```
 
-##### PAGE BREAK
+```{r separationdens, fig.cap = "A normalized density plot reveals that patients who died within 90 days were, on average, estimated to have a higher probability to do so.
+There were very few observations with covariate patterns resulting in death probabilities higher than 5 % (green line).
+Estimated probabilities above this limit are therefore subject to extrapolation."}
+# knitr::include_graphics("../graphs/separation_dens.tiff")
+knitr::include_graphics("../graphs/images.png")
+
+```
 
@@
 In contrast, with higher predicted probabilities, the model over-estimated the observed proportions of deaths."}
-knitr::include_graphics("../graphs/calibration.png")
+# knitr::include_graphics("../graphs/calibration.tiff")
+knitr::include_graphics("../graphs/images.png")
+
 ```

### 2020-06-11 src/fig_auc.R

@@ -63,3 +63,4 @@ options(digits = digs$digits)
 
 ggsave("graphs/auc_ci.png", p, width = 10, height = 8, units = "cm")
 ggsave("graphs/auc_ci.tiff", p, width = 10, height = 8, units = "cm", dpi = 1200, compression = "lzw")
+ggsave("graphs/auc_ci.eps", p, width = 10, height = 8, units = "cm")

### 2020-06-11 reports/article.Rmd
Adopted to word limit. Fix #31.

@@
 
-**Results: ** We propose a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
-The model had good discrimination, both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`.
+**Results: ** A main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity had good discrimination, both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`.
 This model was superior to traditional models based on the Charlson `r get_auc("Charlson")` and Elixhauser `r get_auc("Elixhauser")` co-morbidity indices.
@@
 Shared decision making has evolved into an integral part of patient-physician interactions prior to surgical interventions, and the weighing of risks against benefits is central to this process.
-In terms of gained quality of life and cost-utility, total hip arthroplasty (THA) is an enormously successful procedure.
-90-day mortality after THA performed due to osteoarthritis is low, ranging from `r 0.2` % to `r 0.6` % in large joint registry studies.[@Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.
+90-day mortality after THA performed due to osteoarthritis is low, ranging from 0.2 % to 0.6 %.[@Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.
 
 Co-morbidity is associated with a shorter remaining life span, but the Charlson and Elixhauser co-morbidity indices poorly predict mortality after THA.[@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017; @Bulow2017; @Inacio2015] Additionally, these complex co-morbidity instruments are based on information from extensive in- or out-patient databases, each measure being defined by 1,178 and 1,516 individual International Classification of Diseases (ICD)-10 codes, respectively.
-Each co-morbidity index also exists in numerous versions, making interpretation and comparisons difficult.[@Brusselaers2017] The use of these indices is therefore limited to research settings but they are impossible to use under clinical circumstances.
+The use of these indices is therefore limited to research settings.
 
-Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery.[@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] The first prediction model based on a European cohort was the patient decision support tool for joint replacement in the UK (https://jointcalc.shef.ac.uk/).
+Pre-operative co-morbidity data have been used for prediction models of early mortality and adverse events such as surgical site infection, in the context of THA surgery.[@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] The first prediction model based on a European cohort was developed in the UK (https://jointcalc.shef.ac.uk/).
 However, it remains unclear to what level the depth of co-morbidity characterization influences the accuracy of mortality prediction models.
 
-We aimed to develop a parsimonious prediction model of 90-day postoperative mortality after THA with internal and external validation of discrimination and calibration, and to compare this to the accuracy of prediction models based on existing co-morbidity measures.[@Manning2016]
+We aimed to develop a parsimonious prediction model of 90-day mortality after THA with internal and external validation of discrimination and calibration, and to compare this to the accuracy of prediction models based on existing co-morbidity measures.[@Manning2016]
 
@@
 
-We used data from the Swedish Hip Arthroplasty Register (SHAR) for model derivation and internal validation of discrimination and calibration.
-The best model was then validated externally on patients from England and Wales recorded in the National Joint Registry for England, Wales, Northern Ireland, the Isle of Man and the States of Guernsey (NJR).
+We used data from the Swedish Hip Arthroplasty Register (SHAR) for model derivation and internal validation.
+The model was then validated externally on patients from England and Wales recorded in the National Joint Registry for England, Wales, Northern Ireland, the Isle of Man and the States of Guernsey (NJR).
 We focused on patients with cemented THA due to osteoarthritis.
 Cementation is the most common fixation technique used in Sweden, thus used as inclusion criteria to decrease heterogeneity.
-Only the last operated hip was accounted for in patients with bilateral THA, since those are more likely to resemble patients with unilateral THA (Figure \@ref(fig:flowchartshar)).[@Bulow2020]
+Only the last operated hip was accounted for in patients with bilateral THA (Figure \@ref(fig:flowchartshar)).[@Bulow2020]
 
@@
 
-The patient cohort was identified from the SHAR.
-The inclusion period started 2008 since the American Society for Anesthesiologists (ASA) class and Body Mass Index (BMI) were systematically recorded from then on.
+The inclusion period started 2008 since the American Society for Anesthesiologists (ASA) class and Body Mass Index (BMI) were systematically recorded in SHAR from then on.
 The observation period ended 2015, since we had access to co-morbidity from the national patient register up to this point.
 Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration.
-We used data from a variety of sources, as previously described.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98 %.[@Karrholm2018] Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status (married/un-married/divorced/widow[er]), were also collected from LISA.
-The Swedish national patient register (NPR) was used to assess co-morbidity during the year preceding index surgery.
-This register contains all relevant diagnoses coded by ICD-10, as well as dates of admission and discharge for in- and outpatient episodes in all private and public hospitals.
+The data linkage has been described previously.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital and year of surgery were collected from the SHAR, with a completeness of 96-98 %.[@Karrholm2018] Data on educational level were recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status, were also collected from LISA and co-morbidity during the year preceding index surgery from the National Patient Register (NPR).
+NPR contains diagnoses coded by ICD-10 from in- and outpatient episodes in all private and public hospitals.
 Completeness for NPR is above 99 % and 85-95 % of all diagnostic codes are valid.[@Ludvigsson2011] Death dates were linked from the national population register.
@@
 
-Similar inclusion criteria were applied to the validation cohort from England and Wales.
+Similar inclusion criteria were applied to the validation cohort.
 We excluded no patients based on missing information on BMI, educational level, or civil status however, since those variables were not used in the final model.
 `r format(N_njr$tot, big.mark = ",")` patients were included (Figure \@ref(fig:flowchartnjr)).
-The validation cohort was created by linking NJR data to the secondary uses service database from the National Health Service (NHS).
-This database includes co-morbidity recorded in the admitted patient care module of the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Thorn2016] Out-patient deaths were linked from the office of national statistics.
-Data linkage used NHS-numbers, as well as patient name, age, sex and address.[@Sayers2016]
+NJR data was linked to the secondary uses service database from the National Health Service (NHS) with co-morbidity from the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Thorn2016] Out-patient deaths were linked from the office of national statistics based on NHS-numbers, as well as patient name, age, sex and address.[@Sayers2016]
 
@@
 
-Co-morbidity was defined by individual ICD-10 codes grouped into 17 categories according to Charlson and 31 categories according to Elixhauser.[@Quan2005] Some co-morbidities were identified in both indices, and some distinct co-morbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol).
+Co-morbidity was defined by individual ICD-10 codes grouped into 17 categories by Charlson and 31 by Elixhauser.[@Quan2005] Some conditions were identified by both indices, and some were closely related and therefore merged (such as hypertension with and without complications, or abuse of either drugs or alcohol).
 We combined individual diagnostic groups to establish `r nrow(tab_categorization)` broader categories of co-morbidity (Table \@ref(tab:tabcategorization)).
-Conditions were merged according to clinical relevance as to be recognized in a patient-physician meeting without access to external register data.
 
@@
 
-We used the Kaplan-Meier estimator to assess unadjusted mortality after cemented THA.
-Further analysis was based on logistic regression, since no loss to follow up occurred in the Swedish cohort within the first 90 days, and we assumed the same was true for patients from England and Wales.
-We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015] Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1.
-Thus, the magnitude (absolute values) of their estimated coefficients would indicate variable importance on the same scale as the categorical variables.
+We used Kaplan--Meier and a procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015] Age and BMI were normalized to have mean = 0 and standard deviation = 1.
 Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process.
 1,000 bootstrap samples were drawn from the observed data set.
-We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model.
-We then only kept $\lambda$:s minimizing the mean cross-validated deviances in each sample.
-Those $\lambda$:s were used to estimate model coefficients for each potential predictor.
+We used 10-fold cross validation for every bootstrap sample with a range of potential penalty values ($\lambda$:s) in a logistic LASSO model.
+We  kept $\lambda$:s minimizing the mean cross-validated deviances and those were used to estimate model coefficients for each potential predictor.
 The magnitude (absolute values) of those estimates were used as a measure of variable importance.
 Piece-wise linear regression was used to detect a breaking point where a significant drop in variable importance was observed.
-Potential predictors with variable importance above this breaking point were considered important and kept as model candidates.
+Potential predictors with variable importance above this breaking point were kept as model candidates.
 The whole process was repeated 100 times.
-Covariates that were selected at least once out of the 100 times were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables (main model).
-A reduced model with variables chosen at least 33 out of the 100 times was used as a simpler alternative for comparison.
-This model, without cancer as a predictor, was also evaluated separately, considering that medical indications for THA surgery may be different for patients with cancer compared to patients without malignancies.
-Univariable models with the ASA class, Charlson or Elixhauser co-morbidity indices were used for benchmarking, as well as a multivariable model including only age and sex.
+Covariates that were selected at least once were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables (main model).
+A reduced model with variables chosen at least 33 out of the 100 times were used as a simpler alternative.
+This model was also evaluated without cancer as a predictor, considering that medical indications for THA surgery may be different for patients with cancer or without cancer.
+Univariable models with the ASA class, Charlson or Elixhauser co-morbidity indices were used for benchmarking, as well as a model including only age and sex.
 Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.
-Odds ratios for the final model were estimated with 95 % confidence intervals.
-$p$-values below 0.05 were considered statistically significant.
+Odds ratios for the final model were estimated with 95 % confidence intervals and $p$-values below 0.05 were considered statistically significant.
 
@@
 Each model was used to predict the probability of death within 90 days for patients from the SHAR (internal validation).
-Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.
-Models with a lower 95 % confidence limit above `r 0.7`, were considered good.
-Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples.
+Receiver operating characteristic (ROC) curves were drawn and the area under those curves (AUC) were used as a measure of discriminative ability.
+Models with a lower bootstrap-based 95 % confidence limit above 0.7, were considered good.
 We used the bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples to adjust for optimism.[@Miller1991] Calibration bands were made to graphically assess model calibration, comparing predicted probabilities and observed proportions.[@Nattino2016] The reduced model was then evaluated externally.
-An AUC with 95 % confidence interval was calculated for the model as-is.
+An AUC with 95 % CI was calculated for the model as-is.
 Re-calibration of the model intercept was then performed to account for different mortality rates in Sweden compared to England and Wales.
-An updated over-all slope was also calculated to account for country-specific treatment differences.[@Steyerberg2004] Calibration for this re-calibrated model was illustrated in the same calibration belt plot as for the internal calibration.
+An updated over-all slope was also calculated to account for country-specific treatment differences.[@Steyerberg2004]
 
@@
 
-We built an online web calculator available at `r shiny_url` to be used in clinical practice.
+We built an online web calculator  (`r shiny_url`) to be used in clinical practice.
 
@@
 
-# Indirect Patient and Public Involvement
-
-There is a patient representative at SHAR to ensure that the patients' interests are respected in planned research.
-
 # Results
@@
 
-`r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` of the patients from SHAR died within 90 days and no one was censored before that.
-The unadjusted risk of 90-day mortality was therefore `r surv90d`.
-Further characteristics of the study population are presented in Table \@ref(tab:tab1).
-Patients from NJR had an unadjusted risk of 90 day mortality of `r njr_surv`, which was higher compared to the derivation cohort from Sweden.
+`r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` of the patients from SHAR died within 90 days (`r surv90d` according to Kaplan--Meier) and no one was censored before that.
+Patients from NJR had an unadjusted risk of 90 day mortality of `r njr_surv`.
+Further cohort characteristics are presented in Table \@ref(tab:tab1).
 
@@
 
-There were `r spell(nrow(comb_lgl))` co-morbidities that were not recorded for any patient who died in the Swedish derivation cohort: `r comb_lgl_text`.
-Those variables were therefore excluded as potential predictors prior to any statistical model derivation.
+There were `r spell(nrow(comb_lgl))` co-morbidities not recorded for any patient who died in the Swedish derivation cohort and therefore not considered as potential predictors: `r comb_lgl_text`.
 The derived main model included `r coefs_print`.
-The reduced model, with covariates included at least 33 out of 100 times, was restricted to `r coefs_print_reduced` (Table \@ref(tab:brlprop)).
+The reduced model was restricted to `r coefs_print_reduced` (Table \@ref(tab:brlprop)).
 
-There were no differences between models including age as a main effect, compared to those were age was modeled by restricted cubic splines with either two or three knots.
+There were no differences between models including age as a main effect or as restricted cubic splines.
 We therefore focused on the more parsimonious models with age as a main effect.
@@
 
-The main and reduced models were no different regarding discriminative power `r sprintf("%s versus %s", gsub(")", "", get_auc("Main model"), fixed = TRUE), gsub("(", "", get_auc("Reduced model"), fixed = TRUE))`.
+The main and reduced models were no different regarding discriminative ability `r sprintf("%s versus %s", gsub(")", "", get_auc("Main model"), fixed = TRUE), gsub("(", "", get_auc("Reduced model"), fixed = TRUE))`.
 We therefore considered the reduced model as superior due to its simplicity.
@@
 The ability of the reduced model to estimate probabilities of death within 90 days is further illustrated in Figure \@ref(fig:separationhist) and \@ref(fig:separationdensity).
-Model calibration was good for estimated probabilities up to 3 % and acceptable up to 5 %, although with predicted probabilities usually higher than observed proportions (Figure \@ref(fig:calibration)).
-
+Model calibration was good for estimated probabilities up to 3 % and acceptable up to 5 % (Figure \@ref(fig:calibration)).
 Estimated model coefficients and corresponding odds ratios for the reduced model are presented in Table \@ref(tab:brlreduced).
-
 Omitting cancer from the reduced model did not affect the AUC or calibration for estimated probabilities below 3 %, but calibration outside this range deteriorated, and we thus retained cancer as an important predictor.
@@
 
-## Principal findings
-
-We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` was considerably better at predicting death within 90 days after THA than traditional models based on the Charlson or Elixhauser co-morbidity indices.
-
+We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` was better at predicting death within 90 days after THA than traditional models based on the Charlson or Elixhauser co-morbidity indices.
 The resulting model predicted the probability of death within 90 days as: $\hat p = 1 / [1 +\exp(-\hat \beta X)]$ where $\hat \beta$ is the vector of estimated coefficients presented in table \@ref(tab:brlreduced).
-This formula is considered valid for patients aged `r paste(range(df$P_Age), collapse = " - ")` years and for predicted probabilities up to 5 %.
+This formula is considered valid for patients aged `r paste(range(df$P_Age), collapse = "--")` years and for predicted probabilities up to 5 %.
 
@@
 
-Our model could be used in clinical practice, either by the formula above, or by accessing a simple web calculator online (`r shiny_url`).
-For example, a `r min(df$P_Age)`-year-old woman with ASA class I and none of the important co-morbidities would have a `r get_risk("baseline")` risk to die within 90 days of surgery.
-A woman, `r quantile(df$P_Age, .25)` years old, i.e.
-the first age quantile, would have an elevated risk of `r get_risk("healthy_female")`.
-A `r max(df$P_Age)`-year-old man, i.e.
-the maximum observed age, with ASA class III and cancer, would have a risk of `r get_risk("sickest_man")`.
-Note, however, that covariate patters with observed probabilities above 5 % were rare (`r mean(pred_probs > 5) * 100` %, n = `r sum(pred_probs > 5)`), and estimated mortality risks above 5 % are therefore subject to extrapolation.
-The observed proportions of deaths for patients with similar characteristics is likely lower (as indicated by Figure \@ref(fig:calibration)).
+Our model could be used in clinical practice, either by the formula, or by the web calculator (`r shiny_url`).
+A `r min(df$P_Age)`-year-old woman with ASA class I and none of the important co-morbidities would have a `r get_risk("baseline")` risk to die within 90 days of surgery.
+A woman, `r quantile(df$P_Age, .25)` years old, would have an elevated risk of `r get_risk("healthy_female")`.
+A `r max(df$P_Age)`-year-old man with ASA class III and cancer, would have a risk of `r get_risk("sickest_man")`.
+Note, however, that covariate patters with observed probabilities above 5 % were rare (`r mean(pred_probs > 5) * 100` %, n = `r sum(pred_probs > 5)`).
+Risks above 5 % are therefore subject to extrapolation.
+The true proportions are likely lower (as indicated by Figure \@ref(fig:calibration)).
 Some risk calculators ignore this problem,[@Bozic2013a; @Harris2019] but we think this should be acknowledged.
 
-Variables in our model were chosen based on statistical properties and not due to clinical relevance.
-Variables and estimated coefficients should therefore not be assigned any exact epidemiological and/or causal meaning per se.
-
-Age and sex are well-known predictors of remaining life span, but it is less obvious that this relation must be linear.
-We used restricted cubic splines to allow a more flexible relation, but found no difference in AUC compared to simpler main effect models.
+Variables in our model were chosen based on statistical properties and should therefore not be assigned any exact epidemiological and/or causal meaning per se.
+Nevertheless, age and sex are well-known predictors of remaining life span.
 ASA class III has the largest estimated coefficient among all predictors, indicating large relative importance.
 This is clinically reasonable since a label of "severe systemic disease" should be based on a relevant patient assessment prior to surgery.
-ASA class is nevertheless known to have a moderate or high degree of internal variability.[@Sankar2014] It has been compared to the Charlson co-morbidity index, but not with respect to mortality after THA.
-Patients with ASA class IV-VI were excluded since those categories describe severe disease, moribund and brain-dead individuals.
-We suspect that patients with such conditions might have been misclassified.
 
-Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity.
-We noted that the proportion of patients registered in the SHAR with BMI above 30, the WHO definition of obesity, was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %, n = `r format(sum(df$P_BMI > 30 & !is.na(df$P_BMI)), big.mark = ",")`) than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the Swedish national patient register (`r round(mean(df$c_obesity) * 100, 1)` %, , n = `r format(sum(df$c_obesity & !is.na(df$c_obesity)), big.mark = ",")`).
-We believe this might be due to the fact that BMI is systematically recorded in the SHAR, but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.
+Obesity was not statistically significant ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity.
+We noted that the proportion of patients in the SHAR with BMI above 30, the WHO definition of obesity, was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %, n = `r format(sum(df$P_BMI > 30 & !is.na(df$P_BMI)), big.mark = ",")`) than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the NPR (`r round(mean(df$c_obesity) * 100, 1)` %, n = `r format(sum(df$c_obesity & !is.na(df$c_obesity)), big.mark = ",")`).
+BMI is systematically recorded in the SHAR, but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.
 
@@
 
-A strength of this study is its nationwide design with data from both Sweden, England and Wales.
-The Swedish registers are valid with low proportions of missing data.[@Karrholm2018; @Ludvigsson2011] Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however.
-A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80 %.[@Burns2012] Both in- and outpatient co-morbidity data were available for Sweden, but only in-patient data for England and Wales.
-We did not have access to any data from primary care or nursing homes in any of the counties.
-
-The linkage procedure between NJR and HES has been previously described,[@Smith2012] where it was noted that privately funded patients were not included in HES.
+The nationwide design with data from both Sweden, England and Wales is a strength.
+The Swedish registers are valid with low proportions of missing data.[@Karrholm2018; @Ludvigsson2011] Some concerns have been raised regarding the HES database from UK, but a systematic review found that the overall median diagnostic accuracy (comparing ICD-codes to individual case notes) was 80 %.[@Burns2012] Both in- and outpatient co-morbidity data were available for Sweden, but only in-patient data for England and Wales.
+The linkage between NJR and HES has been previously described,[@Smith2012] where it was noted that privately funded patients were not included in HES.
 17 % of the patients had missing personal data or did not allow linkage, and 6 % were not found in HES although their data were available from the NJR.
 
-In addition to data validity, the indications for choosing between different fixation techniques differ in Sweden compared to England and Wales.
-Cementation is the most commonly used fixation principle in Sweden but not so in England and Wales, where younger patients are more likely to be operated with cementless implants.
-To only include cemented THA therefore implies an over-representation of older and frailer patients from England and Wales compared to Sweden.
-Those patients might be more likely to die within 90 days, and this assumption would be in agreement with the calibration plot where estimated probabilities below `r 1.5` % underestimated the observed proportion of deaths in the external validation cohort from the NJR.
+The indications for different fixation techniques differ between the cphorts.
+Cementation is commonly used in Sweden but not so in England and Wales, where younger patients are more likely to be operated with cementless implants.
+To only include cemented THA therefore implies an over-representation of older and frailer patients from England and Wales.
+Those patients might be more likely to die within 90 days, and this assumption would be in agreement with the calibration plot where estimated probabilities below 1.5 % underestimated the observed proportion of deaths in the external validation cohort.
 
 We investigated whether the effect of age on mortality was non-linear using restricted cubic splines.
-A similar approach might be relevant for BMI, which we only modelled as a linear main effect.
-A U-shaped association between mortality and BMI categorized as underweight (below `r 18.5` kg/m^2^), normal (`r 8.5` to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales.[@Mouchti2018] Our recordings of BMI might be imprecise however.
+A similar approach might be relevant for BMIt.
+A U-shaped association between mortality and BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been reported.[@Mouchti2018] Our recordings of BMI might be imprecise however.
 Some hospitals report actual measurements but it is unknown to what extent those data are self-estimated by the patients or by health care personnel.
@@
 We merely followed the cohort who already had THA.
-Hence, deaths within 90 days might occur for those patients regardless if THA is inserted or not.
+Deaths within 90 days might occur for those patients regardless if THA is inserted or not.
 The proximity in time however, the maximum of 90 days from THA to death, is an indication that the operation might be the main cause of death for the majority of deceased patients.
@@
 Our results indicate that the risk of early postoperative mortality after cemented THA can be pre-operatively assessed by a parsimonious prediction model.
-We hope that this model, with its accompanying web calculator, will facilitate shared decision-making between physicians and their patients in need of THA.
-
-##### PAGE BREAK
+We hope that this model, with its accompanying web calculator, will facilitate shared decision-making.
 

### 2020-08-24 reports/article.Rmd
Affiliations etc. Fix #47

@@
 
-1.	Department of Orthopaedics, Institute of Surgical Sciences, Uppsala University Hospital, Uppsala, Sweden
-2.	The Swedish Hip Arthroplasty Register, Gothenburg, Sweden
-3.	Department of Orthopaedics, Visby Hospital, Visby, Sweden
-4.	Department of Orthopaedics, Institute of Clinical Sciences, The Sahlgrenska Academy, University of Gothenburg, Gothenburg, Sweden
-5. Translational Health Sciences, Bristol Medical School, University of Bristol, UK.
-6. The National Institute of Health Research Biomedical Research Centre Bristol
-7. Department of Oncology and Metabolism, University of Sheffield, Sheffield, UK.
+1.	Department of Orthopaedics, Institute of Surgical Sciences, Uppsala University Hospital, Uppsala, Sweden (postal: Akademiska sjukhuset, ingång 61, 6 tr, SE-751 85 Uppsala, Sweden)
+2.	The Swedish Hip Arthroplasty Register, Gothenburg, Sweden (postal: Registercentrum Västra Götaland, SE-413 45 Gothenburg, Sweden)
+3.	Department of Orthopaedics, Visby Hospital, Visby, Sweden (postal: S:t Göransgatan 5, SE-621 55 Visby, Sweden)
+4.	Department of Orthopaedics, Institute of Clinical Sciences, The Sahlgrenska Academy, University of Gothenburg, Gothenburg, Sweden (postal: Avdelningen för ortopedi, Mölndals sjukhus, Göteborgsvägen 31, SE-431 80 Mölndal, Sweden)
+5. Translational Health Sciences, Bristol Medical School, University of Bristol, UK (postal: Bristol Medical School, University of Bristol, First Floor, 5 Tyndall Avenue
+Bristol, BS8 1UD, United Kingdom).
+6.
+The National Institute of Health Research Biomedical Research Centre Bristol (postal: NIHR Biomedical Research Centre Bristol, University Hospitals Bristol and Weston NHS Foundation Trust, University of Bristol, Oakfield House, Oakfield Grove,
+Bristol BS8 2BN, United Kingdom)
+7. Department of Oncology and Metabolism, University of Sheffield, Sheffield, UK. (postal: The Medical School, The University of Sheffield, Beech Hill Road, Sheffield, S10 2RX, UNited Kingdom)
 
@@
 Erik Bulow. + 46 70 08 234 28. erik.bulow@vgregion.se. Postal: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45 Gothenburg, SWEDEN
+
+## Titles
+
+**AG:** MD, PhD, orthopaedic surgeon  
+**EB:** MSc (PhD after 2020-09-18), statistician  
+**EL:** PhD, statistician, Senior Research Fellow  
+**AB:** MB, ChB, FRCS, MD, PhD, Head of Bristol Medical School, Professor of Orthopaedic Surgery, orthopaedic surgeon  
+**MJW:** MB, ChB, PhD, FRCS, professor of orthopaedics, orthopaedic surgeon  
+**AS:** MSc, statistician, senior research fellow  
+**OR:** MD, PhD, orthopaedic surgeon, professor of orthopaedics, head of department, registry director of the Swedish Hip Arthroplasty Register  
+**NPH:** MD, PhD, orthopaedic surgeon, professor of orthopaedics, research leader
+
+# Contribution of authors
+
+AG and NPH initiated the study and managed the ethical review board application in Sweden.
+AB and JMW conceptualized the external validation and managed the ethical review board application in the UK.
+EB developed the statistical model.
+EL and AS performed external validation with data from NJR.
+AG and EB drafted, and all authors edited and finalized the manuscript.
 
@@
 
-The ability of the reduced model to estimate probabilities of death within 90 days is further illustrated in Figure \@ref(fig:separationhist) and \@ref(fig:separationdensity).
+The ability of the reduced model to estimate probabilities of death within 90 days is further illustrated in Figure \@ref(fig:separationhist) and \@ref(fig:separationdens).
 Model calibration was good for estimated probabilities up to 3 % and acceptable up to 5 % (Figure \@ref(fig:calibration)).
@@
 We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` was better at predicting death within 90 days after THA than traditional models based on the Charlson or Elixhauser co-morbidity indices.
-The resulting model predicted the probability of death within 90 days as: $\hat p = 1 / [1 +\exp(-\hat \beta X)]$ where $\hat \beta$ is the vector of estimated coefficients presented in table \@ref(tab:brlreduced).
+The resulting model predicted the probability of death within 90 days as: $\hat p = 1 / [1 +\exp(-\hat \beta x)]$ where $\hat \beta$ is the vector of estimated coefficients (including intercept) presented in table \@ref(tab:brlreduced).
 This formula is considered valid for patients aged `r paste(range(df$P_Age), collapse = "--")` years and for predicted probabilities up to 5 %.
@@
 A `r max(df$P_Age)`-year-old man with ASA class III and cancer, would have a risk of `r get_risk("sickest_man")`.
-Note, however, that covariate patters with observed probabilities above 5 % were rare (`r mean(pred_probs > 5) * 100` %, n = `r sum(pred_probs > 5)`).
+Note, however, that covariate patterns with observed probabilities above 5 % were rare (`r mean(pred_probs > 5) * 100` %, n = `r sum(pred_probs > 5)`).
 Risks above 5 % are therefore subject to extrapolation.
@@
 Variables in our model were chosen based on statistical properties and should therefore not be assigned any exact epidemiological and/or causal meaning per se.
-Nevertheless, age and sex are well-known predictors of remaining life span.
-ASA class III has the largest estimated coefficient among all predictors, indicating large relative importance.
+Nevertheless, age and sex are well-known predictors of remaining life span.~~[@Gibbs2020; @Shohat2020]~~ ASA class III has the largest estimated coefficient among all predictors, indicating large relative importance.
 This is clinically reasonable since a label of "severe systemic disease" should be based on a relevant patient assessment prior to surgery.
@@
 
-The indications for different fixation techniques differ between the cphorts.
+The indications for different fixation techniques differ between the cohorts.
 Cementation is commonly used in Sweden but not so in England and Wales, where younger patients are more likely to be operated with cementless implants.
@@
 We investigated whether the effect of age on mortality was non-linear using restricted cubic splines.
-A similar approach might be relevant for BMIt.
+A similar approach might be relevant for BMI.
 A U-shaped association between mortality and BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been reported.[@Mouchti2018] Our recordings of BMI might be imprecise however.
@@
 We hope that this model, with its accompanying web calculator, will facilitate shared decision-making.
-
-# Contribution of authors
-
-AG and NPH initiated the study and managed the ethical review board application in Sweden.
-AB and JMW conceptualized the external validation and managed the ethical review board application in the UK.
-EB developed the statistical model. EL and AS performed external validation with data from NJR. AG and EB drafted, and all authors edited and finalized the manuscript.
 

### 2020-08-24 reports/article.Rmd
Fix #50

@@
 
+~~It should be noted that short-term mortality after elective THA is rare.
+As such, it might not be the most immediate issue for patients making an informed decision on whether to operate or not.
+Other pertinent outcomes includes adverse events after surgery such as prosthetic joint infections or dislocation.
+Our next step is therefore to extend the web calculator to such events as well.~~
+
 ## Strengths and limitations

### 2020-08-25 reports/article.Rmd
Stat figs. Fix #52.

@@
 
-We used Kaplan--Meier and a procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015] Age and BMI were normalized to have mean = 0 and standard deviation = 1.
+We used a statistical procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO; ~~figure \@ref(fig:br1) and \@ref(fig:br2)~~).[@Guo2015] Age and BMI were normalized to have mean = 0 and standard deviation = 1.
 Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process.

### 2020-08-25 reports/figs.Rmd
Stat figs. Fix #52.

@@
 
+```
+
+(ref:br1) ~~Outer steps of the bootstrap ranking procedure.
+All included patients (A) were re-sampled with replacements 100 times (B).
+For each bootstrap sample, the inner process (C; as depicted in figure \@ref(fig:br2)) was applied.
+The final output was a table with all potential predictors ranked by the number of times they got selected (D).
+Potential predictors selected at least once were included in the main model, and each predictor selected at least 33 out of the 100 times, were included in a reduced model as well.~~
+
+```{r br1, fig.cap = "(ref:br1)", out.width = "\\columnwidth"}
+# knitr::include_graphics("../graphs/br1.pdf")
+knitr::include_graphics("../graphs/images.png")
+```
+
+(ref:br2) ~~Inner steps of the bootstrap ranking procedure.
+For each bootstrap sample (1), 100 new bootstrap samples were created (2).
+Logistic LASSO-regression was applied within each sample (3).
+Absolute values of the estimated coefficient values (whereof some shrunk to 0 by the LASSO), were averaged as a measure of variable importance (4).
+Variables with their estimated variable importance above an estimated break-point from linear piece-wise regression (5) were kept as potential predictors (6).~~
+
+```{r br2, fig.cap = "(ref:br2)", out.width = "\\columnwidth"}
+# knitr::include_graphics("../graphs/br2.pdf")
+knitr::include_graphics("../graphs/images.png")
 ```

### 2020-08-25 reports/article.Rmd
Age distributuion. Fix #51.

@@
 Patients from NJR had an unadjusted risk of 90 day mortality of `r njr_surv`.
-Further cohort characteristics are presented in Table \@ref(tab:tab1).
+~~The median age of patients who died within 90 days were `r age_median` years (range `r age_range`) with distribution in Table \@ref(tab:agedist).~~ Further cohort characteristics are presented in Table \@ref(tab:tab1).
 
@@
 The resulting model predicted the probability of death within 90 days as: $\hat p = 1 / [1 +\exp(-\hat \beta x)]$ where $\hat \beta$ is the vector of estimated coefficients (including intercept) presented in table \@ref(tab:brlreduced).
-This formula is considered valid for patients aged `r paste(range(df$P_Age), collapse = "--")` years and for predicted probabilities up to 5 %.
+This formula is considered valid for patients aged ~~`r age_range`~~ years and for predicted probabilities up to 5 %.
 
@@
 Our model could be used in clinical practice, either by the formula, or by the web calculator (`r shiny_url`).
-A `r min(df$P_Age)`-year-old woman with ASA class I and none of the important co-morbidities would have a `r get_risk("baseline")` risk to die within 90 days of surgery.
+A ~~`r substr(age_range, 1, 2)`~~-year-old woman with ASA class I and none of the important co-morbidities would have a ~~`r get_risk("baseline")`~~ risk to die within 90 days of surgery.
 A woman, `r quantile(df$P_Age, .25)` years old, would have an elevated risk of `r get_risk("healthy_female")`.
-A `r max(df$P_Age)`-year-old man with ASA class III and cancer, would have a risk of `r get_risk("sickest_man")`.
+A ~~`r substr(age_range, 4, 5)`~~-year-old man with ASA class III and cancer, would have a risk of ~~`r get_risk("sickest_man")`~~.
 Note, however, that covariate patterns with observed probabilities above 5 % were rare (`r mean(pred_probs > 5) * 100` %, n = `r sum(pred_probs > 5)`).

### 2020-08-25 reports/tabs.Rmd
Age distributuion. Fix #51.
describe conditions. Fix #53

@@
 ##### PAGE BREAK
+
+(ref:comorbidity) Categorization of comorbidities based on diagnostic groups according to the Charlson and Elixhauser comorbidity indices.
+~~Each sub-condition was firsdt classified based on data from the Swedish National patient register and from the Hospital Episodes statistics registry in England and Wales, coded by the International classification of diseases veriosn 10 (ICD-10), with categorization defined in table 1 and 2 from Quan et al.[Quan2005]~~.
+(CNS = central nervous system.)
 
@@
   knitr::kable(
-    caption = "Categorization of co-morbidities based on diagnostic groups according to Charlson and Elixhauser. (CNS = central nervous system.)")
+    caption = "(ref:comorbidity)")
 
@@
 CNS = central nervous system.")
+```
+
+##### PAGE BREAK
+
+(ref:agedist) ~~Age distribution among patients who died (range `r age_range`years).~~
+```{r agedist}
+knitr::kable(
+  tab_agedist,
+  caption = "(ref:agedist)"
+)
+
 ```

### 2020-08-25 munge/02-survdata.R
limitation of cement. Fix #54.

@@ -1,7 +1,7 @@
 # Add important variables to data
 
-df <-
-  df_shpr %>%
+prepare_df <- function(df) {
+  df %>%
   mutate_if(is.logical, coalesce, FALSE) %>%
   mutate_at(vars(contains("index")), coalesce, 0) %>%
   mutate(
@@ -44,8 +44,8 @@ df <-
   select(
     -P_AcetCupCemMix,
     -P_FemStemCemMix,
-    -LopNr,
-    -P_Side,
+  #  -LopNr,
+  #  -P_Side,
     -opnr,
     -DateOfDeath,
     -P_SurgDate,
@@ -59,4 +59,7 @@ df <-
   # Shorter names
   rename_all(~ gsub("charlson_icd10", "CCI", .)) %>%
   rename_all(~ gsub("elix_icd10", "ECI", .))
+}
 
+df <- prepare_df(df_shpr) %>% select(-LopNr, -P_Side)
+df_ignore_fixation <- prepare_df(df_shpr_ignore_fixation)

### 2020-08-27 reports/article.Rmd
limitation of cement. Fix #54.
Clarify diagnosed obesoty. FIx #55.

@@
 
-Obesity was not statistically significant ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity.
+~~Diagnosed~~ obesity was not statistically significant ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity.
 We noted that the proportion of patients in the SHAR with BMI above 30, the WHO definition of obesity, was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %, n = `r format(sum(df$P_BMI > 30 & !is.na(df$P_BMI)), big.mark = ",")`) than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the NPR (`r round(mean(df$c_obesity) * 100, 1)` %, n = `r format(sum(df$c_obesity & !is.na(df$c_obesity)), big.mark = ",")`).
@@
 Those patients might be more likely to die within 90 days, and this assumption would be in agreement with the calibration plot where estimated probabilities below 1.5 % underestimated the observed proportion of deaths in the external validation cohort.
+~~Possible adaptations of the model to other settings or countries should therfore always be preceeded by additional external validation with data from that target population.~~
 

### 2020-08-27 reports/article.Rmd
Fix tables for #55.

@@
 
-**Results: ** A main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity had good discrimination, both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`.
+**Results: ** A main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and ~~diagnosed~~ obesity had good discrimination, both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`.
 This model was superior to traditional models based on the Charlson `r get_auc("Charlson")` and Elixhauser `r get_auc("Elixhauser")` co-morbidity indices.
@@
 - We developed a parsimonious model for risk assessment of 90-day mortality after elective THA. 
-- The model considers age, sex, ASA class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity.
+- The model considers age, sex, ASA class, the presence of cancer, diseases of the central nervous system, kidney disease, and ~~diagnosed~~ obesity.
 - We published an on-line web-calculator (`r shiny_url`).

### 2020-08-27 reports/tabs.Rmd
Fix tables for #55.

@@
     `Comorbidities by groups` = 
-      gsub("Heart infarct", "Myocardial infarction", `Comorbidities by groups`)
+      gsub("Heart infarct", "Myocardial infarction", `Comorbidities by groups`),
+    `Comorbidities by groups` = 
+      gsub("Obesity", "~~Diagnosed~~ obesity", `Comorbidities by groups`)
   ) %>% 
@@
 table1_combined_tot %>% 
-  mutate(what = gsub("Heart infarct", "Myocardial infarction", what)) %>% 
+  mutate(
+    what = gsub("Heart infarct", "Myocardial infarction", what),
+    what = gsub("Obesity", "~~Diagnosed~~ obesity", what)
+  ) %>% 
 knitr::kable(
@@
     variable = gsub("= 3", "III", variable),
-    variable = gsub("Heart infarct", "Myocardial infarction", variable)
+    variable = gsub("Heart infarct", "Myocardial infarction", variable),
+    variable = gsub("Obesity", "~~Diagnosed~~ obesity", variable)
   )
@@
     term = gsub("= 2", "II", term),
-    term = gsub("= 3", "III", term)
+    term = gsub("= 3", "III", term),
+    term = gsub("Obesity", "~~Diagnosed~~ besity", term)
   ) %>% 

### 2020-08-27 reports/article.Rmd
Clarify excl of uncemented. Fix #56

@@
 We focused on patients with cemented THA due to osteoarthritis.
-Cementation is the most common fixation technique used in Sweden, thus used as inclusion criteria to decrease heterogeneity.
+~~We thus excluded patients where either the femoral stem and/or the acetabular cup were uncemented (uncemented, hybrids and reverse hybrids).~~ Cementation is the most common fixation technique used in Sweden, thus used as inclusion criteria to decrease heterogeneity.
 Only the last operated hip was accounted for in patients with bilateral THA (Figure \@ref(fig:flowchartshar)).[@Bulow2020]

### 2020-08-27 reports/article.Rmd
Relative survival. Fix #57.

@@
 The proximity in time however, the maximum of 90 days from THA to death, is an indication that the operation might be the main cause of death for the majority of deceased patients.
+~~This was also indicated by a stydy performed by Cnudde et al.
+who showed that the relative survival for Swedish patients with THA was above 100 % within a year after surgery.
+Hence, patients with elective THA live longer on average compared to the general population of the same age, sex, and year of birth.
+An elevated risk of death (a relative survival below 100 %) was seen, however, for Swedish patients with cemented THA during the first 90 days after surgery, indicating that this event is in itself associated with those deaths.[@Cnudde2018b]~~
 

### 2020-08-27 reports/article.Rmd
Completeness and coverage. Fix #58

@@
 
-The inclusion period started 2008 since the American Society for Anesthesiologists (ASA) class and Body Mass Index (BMI) were systematically recorded in SHAR from then on.
+~~All public and private Swedish hospitals report their THAs to SHAR, yielding a coverage of 100 %, with an individual patient completeness of 98 %.[@Karrholm2018]~~ Our inclusion period started 2008 since the American Society for Anesthesiologists (ASA) class and Body Mass Index (BMI) were systematically recorded in SHAR from then on.
 The observation period ended 2015, since we had access to co-morbidity from the national patient register up to this point.
@@
 `r format(N_njr$tot, big.mark = ",")` patients were included (Figure \@ref(fig:flowchartnjr)).
-NJR data was linked to the secondary uses service database from the National Health Service (NHS) with co-morbidity from the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Thorn2016] Out-patient deaths were linked from the office of national statistics based on NHS-numbers, as well as patient name, age, sex and address.[@Sayers2016]
+
+~~Reporting to NJR is mandatory for all hospitals run by the National health service (NHS) in England and Wales.[@NationalJointRegistryNJR] The register coverage is thus 100&nbsp;% for such hospitals, although privately founded THAs are not included.
+Patient participation is based on informed consent.
+Consent rates varies slightly between years and regions but were 92.3&nbsp;% in both England and Wales in 2018.[@TheNJREditorialBoard]~~ NJR data was linked to the secondary uses service database from the National Health Service (NHS) with co-morbidity from the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Thorn2016] Out-patient deaths were linked from the office of national statistics based on NHS-numbers, as well as patient name, age, sex and address.[@Sayers2016]
 

### 2020-09-03 reports/article.Rmd
response from NH and EB

--- 2020-08-27 reports/article.Rmd
@@
 title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
-author: Anne Garland`*`^1,2,3^, Erik Bulow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5,6^, J. Mark Wilkinson^7^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
+author: Anne Garland,`*`^,1,2,3^ Erik Bülow,`*`^,2,4^ Erik Lenguerrand,^5^ Ashley Blom,^5,6^ J. Mark Wilkinson,^7^ Adrian Sayers,^5^ Ola Rolfson,^2,4^ Nils P. Hailer^1,2^
 date: '`r Sys.Date()`'
@@
 
-shiny_url <- "https://shpr.registercentrum.se/om-registret-1/forskning/prediktion-av-90-dagarsmortalitet/p/SkyeTsTFB"
+shiny_url <- "https://erikbulow.shinyapps.io/thamortpred/"
+# "https://shpr.registercentrum.se/om-registret-1/forskning/prediktion-av-90-dagarsmortalitet/p/SkyeTsTFB"
 ```
@@
 
-1.	Department of Orthopaedics, Institute of Surgical Sciences, Uppsala University Hospital, Uppsala, Sweden (postal: Akademiska sjukhuset, ingång 61, 6 tr, SE-751 85 Uppsala, Sweden)
-2.	The Swedish Hip Arthroplasty Register, Gothenburg, Sweden (postal: Registercentrum Västra Götaland, SE-413 45 Gothenburg, Sweden)
-3.	Department of Orthopaedics, Visby Hospital, Visby, Sweden (postal: S:t Göransgatan 5, SE-621 55 Visby, Sweden)
-4.	Department of Orthopaedics, Institute of Clinical Sciences, The Sahlgrenska Academy, University of Gothenburg, Gothenburg, Sweden (postal: Avdelningen för ortopedi, Mölndals sjukhus, Göteborgsvägen 31, SE-431 80 Mölndal, Sweden)
-5. Translational Health Sciences, Bristol Medical School, University of Bristol, UK (postal: Bristol Medical School, University of Bristol, First Floor, 5 Tyndall Avenue
-Bristol, BS8 1UD, United Kingdom).
+1.	Department of Orthopaedics, Institute of Surgical Sciences, Uppsala University Hospital, Uppsala, Sweden *(postal: Akademiska sjukhuset, ingång 61, 6 tr, SE-751 85 Uppsala, Sweden)*
+2.	The Swedish Hip Arthroplasty Register, Gothenburg, Sweden *(postal: Registercentrum Västra Götaland, SE-413 45 Gothenburg, Sweden)*
+3.	Department of Orthopaedics, Visby Hospital, Visby, Sweden *(postal: S:t Göransgatan 5, SE-621 55 Visby, Sweden)*
+4.	Department of Orthopaedics, Institute of Clinical Sciences, The Sahlgrenska Academy, University of Gothenburg, Gothenburg, Sweden *(postal: Avdelningen för ortopedi, Mölndals sjukhus, Göteborgsvägen 31, SE-431 80 Mölndal, Sweden)*
+5.
+Translational Health Sciences, Bristol Medical School, University of Bristol, UK *(postal: Bristol Medical School, University of Bristol, First Floor, 5 Tyndall Avenue, Bristol, BS8 1UD, United Kingdom)*.
 6.
-The National Institute of Health Research Biomedical Research Centre Bristol (postal: NIHR Biomedical Research Centre Bristol, University Hospitals Bristol and Weston NHS Foundation Trust, University of Bristol, Oakfield House, Oakfield Grove,
-Bristol BS8 2BN, United Kingdom)
-7. Department of Oncology and Metabolism, University of Sheffield, Sheffield, UK. (postal: The Medical School, The University of Sheffield, Beech Hill Road, Sheffield, S10 2RX, UNited Kingdom)
+The National Institute of Health Research Biomedical Research Centre Bristol *(postal: NIHR Biomedical Research Centre Bristol, University Hospitals Bristol and Weston NHS Foundation Trust, University of Bristol, Oakfield House, Oakfield Grove, Bristol BS8 2BN, United Kingdom)*
+7. Department of Oncology and Metabolism, University of Sheffield, Sheffield, UK. *(postal: The Medical School, The University of Sheffield, Beech Hill Road, Sheffield, S10 2RX, UNited Kingdom)*
 
 ## Correspondence 
-Erik Bulow. + 46 70 08 234 28. erik.bulow@vgregion.se. Postal: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45 Gothenburg, SWEDEN
+Erik Bülow. + 46 70 08 234 28. erik.bulow@gu.se. Postal: Svenska Höftprotesregistret, Registercentrum Västra Götaland, SE-413 45 Gothenburg, SWEDEN
 
@@
 **AG:** MD, PhD, orthopaedic surgeon  
-**EB:** MSc (PhD after 2020-09-18), statistician  
-**EL:** PhD, statistician, Senior Research Fellow  
+**EB:** MSc (will be PhD after 2020-09-18), statistician  
+**EL:** PhD, statistician, Senior research fellow  
 **AB:** MB, ChB, FRCS, MD, PhD, Head of Bristol Medical School, Professor of Orthopaedic Surgery, orthopaedic surgeon  
-**MJW:** MB, ChB, PhD, FRCS, professor of orthopaedics, orthopaedic surgeon  
+**MJW:** MB, ChB, MD, PhD, FRCS, professor of orthopaedics, orthopaedic surgeon  
 **AS:** MSc, statistician, senior research fellow  
@@
 Discriminative ability was evaluated by the area under the receiver operating characteristics curve (AUC).
-Calibration belt plots were used for model calibration.
+Calibration belt plots were used to assess model calibration.
 
-**Results: ** A main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and ~~diagnosed~~ obesity had good discrimination, both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`.
-This model was superior to traditional models based on the Charlson `r get_auc("Charlson")` and Elixhauser `r get_auc("Elixhauser")` co-morbidity indices.
+**Results: ** A main effects model combining age, sex, American Society for Anaesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and ~~diagnosed~~ obesity had good discrimination, both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`.
+This model was superior to traditional models based on the Charlson `r get_auc("Charlson")` and Elixhauser `r get_auc("Elixhauser")` comorbidity indices.
 The model was well calibrated for predicted probabilities up to 5%.
 
-**Concusion: ** We developed a parsimonious model which may facilitate individualized risk assessment prior to one of the most common surgical interventions.
+**Conclusion: ** We developed a parsimonious model which may facilitate individualized risk assessment prior to one of the most common surgical interventions.
 We published a web-calculator to aid clinical decision-making (`r shiny_url`).
@@
 
-Co-morbidity is associated with a shorter remaining life span, but the Charlson and Elixhauser co-morbidity indices poorly predict mortality after THA.[@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017; @Bulow2017; @Inacio2015] Additionally, these complex co-morbidity instruments are based on information from extensive in- or out-patient databases, each measure being defined by 1,178 and 1,516 individual International Classification of Diseases (ICD)-10 codes, respectively.
+Comorbidity is associated with a shorter remaining life span, but the Charlson and Elixhauser comorbidity indices poorly predict mortality after THA.[@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017; @Bulow2017; @Inacio2015] Additionally, these complex comorbidity instruments are based on information from extensive in- or out-patient databases, each measure being defined by 1,178 and 1,516 individual International Classification of Diseases (ICD)-10 codes, respectively.
 The use of these indices is therefore limited to research settings.
 
-Pre-operative co-morbidity data have been used for prediction models of early mortality and adverse events such as surgical site infection, in the context of THA surgery.[@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] The first prediction model based on a European cohort was developed in the UK (https://jointcalc.shef.ac.uk/).
-However, it remains unclear to what level the depth of co-morbidity characterization influences the accuracy of mortality prediction models.
+Pre-operative comorbidity data have been used for prediction models of early mortality and adverse events such as surgical site infection, in the context of THA surgery.[@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] The first prediction model based on a European cohort was developed in the UK (https://jointcalc.shef.ac.uk/).
+However, it remains unclear to what level the depth of comorbidity characterization influences the accuracy of mortality prediction models.
 
-We aimed to develop a parsimonious prediction model of 90-day mortality after THA with internal and external validation of discrimination and calibration, and to compare this to the accuracy of prediction models based on existing co-morbidity measures.[@Manning2016]
+We aimed to develop a parsimonious prediction model of 90-day mortality after THA with internal and external validation of discrimination and calibration, and to compare this to the accuracy of prediction models based on existing comorbidity measures.[@Manning2016]
 
@@
 
-The inclusion period started 2008 since the American Society for Anesthesiologists (ASA) class and Body Mass Index (BMI) were systematically recorded in SHAR from then on.
-The observation period ended 2015, since we had access to co-morbidity from the national patient register up to this point.
+~~All public and private Swedish hospitals report their THAs to SHAR, yielding a coverage of 100 %, with an individual patient completeness of 98 %.[@Karrholm2018]~~ Our inclusion period started 2008 since the American Society for Anaesthesiologists (ASA) class and Body Mass Index (BMI) were systematically recorded in SHAR from then on.
+The observation period ended 2015, since we had access to comorbidity from the national patient register up to this point.
 Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration.
-The data linkage has been described previously.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital and year of surgery were collected from the SHAR, with a completeness of 96-98 %.[@Karrholm2018] Data on educational level were recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status, were also collected from LISA and co-morbidity during the year preceding index surgery from the National Patient Register (NPR).
+This data linkage has been described previously.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital and year of surgery were collected from the SHAR.
+Data on educational level were recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labour market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status, were also collected from LISA and comorbidity during the year preceding index surgery from the National Patient Register (NPR).
 NPR contains diagnoses coded by ICD-10 from in- and outpatient episodes in all private and public hospitals.
@@
 Patients with missing information on BMI or a measurement above 50 were excluded, as were patients with missing information on ASA class or class IV and above, as well as patients with unknown educational level or type of hospital.
-`r format(nrow(df), big.mark = ",")` patients (mean age `r mean(df$P_Age)` years, range `r paste(range(df$P_Age), collapse = " - ")` and `r round(mean(df$P_Sex == "Female") * 100)` % females) were included (Figure \@ref(fig:flowchartshar)).
+`r format(nrow(df), big.mark = ",")` patients (mean age `r round(mean(df$P_Age))` years, range `r paste(range(df$P_Age), collapse = " - ")` and `r round(mean(df$P_Sex == "Female") * 100)` % females) were included (Figure \@ref(fig:flowchartshar)).
 
@@
 `r format(N_njr$tot, big.mark = ",")` patients were included (Figure \@ref(fig:flowchartnjr)).
-NJR data was linked to the secondary uses service database from the National Health Service (NHS) with co-morbidity from the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Thorn2016] Out-patient deaths were linked from the office of national statistics based on NHS-numbers, as well as patient name, age, sex and address.[@Sayers2016]
+
+~~Reporting to NJR is mandatory for all hospitals run by the National Health Service (NHS) in England and Wales.[@NationalJointRegistryNJR] The register coverage is thus 100&nbsp;% for such hospitals, although privately founded THAs are not included.
+Patient participation is based on informed consent.
+Consent rates varies slightly between years and regions but were 92.3&nbsp;% in both England and Wales in 2018.[@TheNJREditorialBoard]~~ NJR data was linked to the secondary uses service database from the National Health Service (NHS) with comorbidity from the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Thorn2016] Out-patient deaths were linked from the office of national statistics based on NHS-numbers, as well as patient name, age, sex and address.[@Sayers2016]
 
@@
 
-Co-morbidity was defined by individual ICD-10 codes grouped into 17 categories by Charlson and 31 by Elixhauser.[@Quan2005] Some conditions were identified by both indices, and some were closely related and therefore merged (such as hypertension with and without complications, or abuse of either drugs or alcohol).
-We combined individual diagnostic groups to establish `r nrow(tab_categorization)` broader categories of co-morbidity (Table \@ref(tab:tabcategorization)).
+Comorbidity was defined by individual ICD-10 codes grouped into 17 categories by Charlson and 31 by Elixhauser.[@Quan2005] Some conditions were identified by both indices, and some were closely related and therefore merged (such as hypertension with and without complications, or abuse of either drugs or alcohol).
+We combined individual diagnostic groups to establish `r nrow(tab_categorization)` broader categories of comorbidity (Table \@ref(tab:tabcategorization)).
 
@@
 We used a statistical procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO; ~~figure \@ref(fig:br1) and \@ref(fig:br2)~~).[@Guo2015] Age and BMI were normalized to have mean = 0 and standard deviation = 1.
-Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process.
-1,000 bootstrap samples were drawn from the observed data set.
+Comorbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process.
+100 bootstrap samples were drawn from the observed data set.
 We used 10-fold cross validation for every bootstrap sample with a range of potential penalty values ($\lambda$:s) in a logistic LASSO model.
 We  kept $\lambda$:s minimizing the mean cross-validated deviances and those were used to estimate model coefficients for each potential predictor.
-The magnitude (absolute values) of those estimates were used as a measure of variable importance.
+The means of the magnitude (absolute values) of those estimates were used as measures of variable importance.
 Piece-wise linear regression was used to detect a breaking point where a significant drop in variable importance was observed.
@@
 This model was also evaluated without cancer as a predictor, considering that medical indications for THA surgery may be different for patients with cancer or without cancer.
-Univariable models with the ASA class, Charlson or Elixhauser co-morbidity indices were used for benchmarking, as well as a model including only age and sex.
+Univariable models with the ASA class, as well as the Charlson or Elixhauser comorbidity indices were used for benchmarking, as well as a model including only age and sex.
 Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.
@@
 
-`r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` of the patients from SHAR died within 90 days (`r surv90d` according to Kaplan--Meier) and no one was censored before that.
-Patients from NJR had an unadjusted risk of 90 day mortality of `r njr_surv`.
-~~The median age of patients who died within 90 days were `r age_median` years (range `r age_range`) with distribution in Table \@ref(tab:agedist).~~ Further cohort characteristics are presented in Table \@ref(tab:tab1).
+`r sprintf("%d [%s]", sum(df$death90), surv90d)` of the patients from SHAR died within 90 days and no one was censored before that.
+~~The median age of patients who died were `r age_median` years (range `r age_range`) with distribution in Table \@ref(tab:agedist).~~ Patients from NJR had an unadjusted risk of 90 day mortality of `r njr_surv`.
+Further cohort characteristics are presented in Table \@ref(tab:tab1).
 
@@
 
-There were `r spell(nrow(comb_lgl))` co-morbidities not recorded for any patient who died in the Swedish derivation cohort and therefore not considered as potential predictors: `r comb_lgl_text`.
-The derived main model included `r coefs_print`.
+There were `r spell(nrow(comb_lgl))` comorbidities not recorded for any patient who died in the Swedish derivation cohort and therefore not considered as potential predictors: `r comb_lgl_text`.
+The derived main model included `r gsub("anemia", "anaemia", coefs_print)`.
 The reduced model was restricted to `r coefs_print_reduced` (Table \@ref(tab:brlprop)).
@@
 We therefore considered the reduced model as superior due to its simplicity.
-Traditional models performed poorly with 95 % confidence intervals not above 0.7: The Charlson co-morbitiy index had an AUC of `r half_paran("Charlson")` and the Elixhauser co-morbidity an AUC of `r gsub("\\)", ";", half_paran("Elixhauser"))` Figure \@ref(fig:rocvalidation) and Figure \@ref(fig:aucci)).
+Traditional models performed poorly with 95 % confidence intervals not above 0.7: The Charlson comorbidity index had an AUC of `r half_paran("Charlson")` and the Elixhauser comorbidity an AUC of `r gsub("\\)", ";", half_paran("Elixhauser"))` Figure \@ref(fig:rocvalidation) and Figure \@ref(fig:aucci)).
 
@@
 
-We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` was better at predicting death within 90 days after THA than traditional models based on the Charlson or Elixhauser co-morbidity indices.
+We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` was better at predicting death within 90 days after THA than traditional models based on the Charlson or Elixhauser comorbidity indices.
 The resulting model predicted the probability of death within 90 days as: $\hat p = 1 / [1 +\exp(-\hat \beta x)]$ where $\hat \beta$ is the vector of estimated coefficients (including intercept) presented in table \@ref(tab:brlreduced).
@@
 Our model could be used in clinical practice, either by the formula, or by the web calculator (`r shiny_url`).
-A ~~`r substr(age_range, 1, 2)`~~-year-old woman with ASA class I and none of the important co-morbidities would have a ~~`r get_risk("baseline")`~~ risk to die within 90 days of surgery.
+A ~~`r substr(age_range, 1, 2)`~~-year-old woman with ASA class I and none of the important comorbidities would have a ~~`r get_risk("baseline")`~~ risk to die within 90 days of surgery.
 A woman, `r quantile(df$P_Age, .25)` years old, would have an elevated risk of `r get_risk("healthy_female")`.
@@
 Variables in our model were chosen based on statistical properties and should therefore not be assigned any exact epidemiological and/or causal meaning per se.
-Nevertheless, age and sex are well-known predictors of remaining life span.~~[@Gibbs2020; @Shohat2020]~~ ASA class III has the largest estimated coefficient among all predictors, indicating large relative importance.
+Nevertheless, age and sex are well-known predictors of remaining life span.~~[@Gibbs2020]~~ ASA class III has the largest estimated coefficient among all predictors, indicating large relative importance.
 This is clinically reasonable since a label of "severe systemic disease" should be based on a relevant patient assessment prior to surgery.
@@
 We noted that the proportion of patients in the SHAR with BMI above 30, the WHO definition of obesity, was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %, n = `r format(sum(df$P_BMI > 30 & !is.na(df$P_BMI)), big.mark = ",")`) than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the NPR (`r round(mean(df$c_obesity) * 100, 1)` %, n = `r format(sum(df$c_obesity & !is.na(df$c_obesity)), big.mark = ",")`).
-BMI is systematically recorded in the SHAR, but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.
+BMI is systematically recorded in the SHAR ~~with only around 5 % missing data,~~ but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.
 
@@
 As such, it might not be the most immediate issue for patients making an informed decision on whether to operate or not.
-Other pertinent outcomes includes adverse events after surgery such as prosthetic joint infections or dislocation.
+Other important outcomes includes adverse events after surgery such as prosthetic joint infections or dislocation.
 Our next step is therefore to extend the web calculator to such events as well.~~
@@
 The nationwide design with data from both Sweden, England and Wales is a strength.
-The Swedish registers are valid with low proportions of missing data.[@Karrholm2018; @Ludvigsson2011] Some concerns have been raised regarding the HES database from UK, but a systematic review found that the overall median diagnostic accuracy (comparing ICD-codes to individual case notes) was 80 %.[@Burns2012] Both in- and outpatient co-morbidity data were available for Sweden, but only in-patient data for England and Wales.
+The Swedish registers are valid with low proportions of missing data.[@Karrholm2018; @Ludvigsson2011] Some concerns have been raised regarding the HES database from UK, but a systematic review found that the overall median diagnostic accuracy (comparing ICD-codes to individual case notes) was 80 %.[@Burns2012] Both in- and outpatient comorbidity data were available for Sweden, but only in-patient data for England and Wales.
 The linkage between NJR and HES has been previously described,[@Smith2012] where it was noted that privately funded patients were not included in HES.
@@
 Those patients might be more likely to die within 90 days, and this assumption would be in agreement with the calibration plot where estimated probabilities below 1.5 % underestimated the observed proportion of deaths in the external validation cohort.
-~~Possible adaptations of the model to other settings or countries should therfore always be preceeded by additional external validation with data from that target population.~~
+~~We tried to also validate the derived model on the Swedish cohort of patients with uncemented, hybrid or reverse hybrid THAs.
+This was not possible for each group separately since we had too few deaths in each cohort.
+It was numerically possible for all patients combined, however, for which we had 28 deaths in total.
+The 95 % CI for the AUC was very wide, however (0.54--0.76), and therefore of less relevance.
+Possible adaptations of the model to other settings or countries should always be preceded by additional external validation with data from that target population.~~
 
@@
 A similar approach might be relevant for BMI.
-A U-shaped association between mortality and BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been reported.[@Mouchti2018] Our recordings of BMI might be imprecise however.
+A U-shaped association between mortality and BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or overweight/obese (above 25 kg/m^2^) has been reported.[@Mouchti2018] Our recordings of BMI might be imprecise however.
 Some hospitals report actual measurements but it is unknown to what extent those data are self-estimated by the patients or by health care personnel.
 Therefore, we aimed to avoid overfitting BMI based on too elaborate transformations.
+~~To not include BMI in the external validation is, however, beneficial, since the completeness of this variable has traditionaly been low; 60 % in 2010 but since then increasing up to 80 % in 2018.
+Instead, the related ICD-10 code was used to capture diagnosed obesity from NPR and HES respectivelly.~~
 
@@
 The proximity in time however, the maximum of 90 days from THA to death, is an indication that the operation might be the main cause of death for the majority of deceased patients.
+~~This was also indicated by a study performed by Cnudde et al.
+who showed that the relative survival for Swedish patients with THA was above 100 % within a year after surgery.
+Hence, patients with elective THA live longer on average compared to the general population of the same age, sex, and year of birth.
+An elevated risk of death (a relative survival below 100 %) was seen, however, for cemented THA during the first 90 days after surgery, indicating that the event itself is associated with increased short-term mortality.[@Cnudde2018b]~~
 

### 2020-09-03 reports/figs.Rmd
response from NH and EB

@@
 The final output was a table with all potential predictors ranked by the number of times they got selected (D).
-Potential predictors selected at least once were included in the main model, and each predictor selected at least 33 out of the 100 times, were included in a reduced model as well.~~
+Potential predictors selected at least once were included in the main model, and each predictor selected at least 33 out of the 100 times, were included in a reduced model as well.[@Bulow2020a]~~
 
@@
 Absolute values of the estimated coefficient values (whereof some shrunk to 0 by the LASSO), were averaged as a measure of variable importance (4).
-Variables with their estimated variable importance above an estimated break-point from linear piece-wise regression (5) were kept as potential predictors (6).~~
+Variables with their estimated variable importance above an estimated break-point from linear piece-wise regression (5) were kept as potential predictors (6).[@Bulow2020a]~~
 
@@
 ```{r rocvalidation, fig.cap = "Receiver Operation Characteristics (ROC) curves.
-There was no difference when applying the reduced model to the external validation cohort from England and Wales ,compared to the Swedish derivation cohort."}
+There was no difference when applying the reduced model to the external validation cohort from England and Wales, compared to the Swedish derivation cohort."}
 # knitr::include_graphics("../graphs/roc_validation.eps")

### 2020-09-03 reports/tabs.Rmd
response from NH and EB

--- 2020-08-25 reports/tabs.Rmd
@@
 (ref:comorbidity) Categorization of comorbidities based on diagnostic groups according to the Charlson and Elixhauser comorbidity indices.
-~~Each sub-condition was firsdt classified based on data from the Swedish National patient register and from the Hospital Episodes statistics registry in England and Wales, coded by the International classification of diseases veriosn 10 (ICD-10), with categorization defined in table 1 and 2 from Quan et al.[Quan2005]~~.
-(CNS = central nervous system.)
+~~Each sub-condition was first classified based on data from the Swedish National patient register and from the Hospital Episodes statistics registry in England and Wales, coded by the International classification of diseases version 10 (ICD-10), with categorization defined in table 1 and 2 from Quan et al.[@Quan2005]~~ (CNS = central nervous system.)
 
@@
     `Comorbidities by groups` = 
-      gsub("Heart infarct", "Myocardial infarction", `Comorbidities by groups`)
+      gsub("Heart infarct", "Myocardial infarction", `Comorbidities by groups`),
+    `Comorbidities by groups` = 
+      gsub("Obesity", "~~Diagnosed~~ obesity", `Comorbidities by groups`)
   ) %>% 
+  mutate_all(~ gsub("nemia", "naemia", .)) %>% 
+  mutate_all(~ gsub("tumor", "tumour", .)) %>% 
   knitr::kable(
@@
 
-```
-
-##### PAGE BREAK
-
-```{r tab1}
-table1_combined_tot %>% 
-  mutate(what = gsub("Heart infarct", "Myocardial infarction", what)) %>% 
-knitr::kable(
-    caption = "Characteristics of the study population in the model derivation cohort (SHAR) and external validation cohort (NJR).
-Educational levels were classified as low (up to 9 years), middle (10-12 years) and high (at least 12 years).
-BMI = Body mass index.
-ASA = American Society for Anesthesiologists classification.
-CNS = central nervous system.")
 ```
@@
 
+```{r tab1}
+table1_combined_tot %>% 
+  mutate(
+    what = gsub("Heart infarct", "Myocardial infarction", what),
+    what = gsub("Obesity", "~~Diagnosed~~ obesity", what),
+    what = gsub("Anemia", "Anaemia", what)
+  ) %>% 
+knitr::kable(
+    caption = "Characteristics of the study population in the model derivation cohort (SHAR) and external validation cohort (NJR).
+Educational levels were classified as low (up to 9 years), middle (10-12 years) and high (at least 12 years).
+BMI = Body mass index.
+ASA = American Society for Anaesthesiologists classification.
+CNS = central nervous system.")
+```
+
+##### PAGE BREAK
+
 ```{r brlprop}
@@
     variable = gsub("= 3", "III", variable),
-    variable = gsub("Heart infarct", "Myocardial infarction", variable)
+    variable = gsub("Heart infarct", "Myocardial infarction", variable),
+    variable = gsub("Obesity", "~~Diagnosed~~ obesity", variable),
+    variable = gsub("Anemia", "Anaemia", variable)
   )
@@
     term = gsub("= 2", "II", term),
-    term = gsub("= 3", "III", term)
+    term = gsub("= 3", "III", term),
+    term = gsub("Obesity", "~~Diagnosed~~ obesity", term)
   ) %>% 

### 2020-09-05 reports/article.Rmd
updates according to NHs suggestions. Fix #60

@@
 
-shiny_url <- "https://erikbulow.shinyapps.io/thamortpred/"
-# "https://shpr.registercentrum.se/om-registret-1/forskning/prediktion-av-90-dagarsmortalitet/p/SkyeTsTFB"
+shiny_url <- #"https://erikbulow.shinyapps.io/thamortpred/"
+  "https://shpr.registercentrum.se/om-registret-1/forskning/prediktion-av-90-dagarsmortalitet/p/SkyeTsTFB"
 ```
@@
 
-1.	Department of Orthopaedics, Institute of Surgical Sciences, Uppsala University Hospital, Uppsala, Sweden *(postal: Akademiska sjukhuset, ingång 61, 6 tr, SE-751 85 Uppsala, Sweden)*
+1.	Department of Surgical sciences/Orthopaedics, Institute of Surgical Sciences, Uppsala University Hospital, Uppsala, Sweden *(postal: Akademiska sjukhuset, ingång 61, 6 tr, SE-751 85 Uppsala, Sweden)*
 2.	The Swedish Hip Arthroplasty Register, Gothenburg, Sweden *(postal: Registercentrum Västra Götaland, SE-413 45 Gothenburg, Sweden)*
@@
 **EL:** PhD, statistician, Senior research fellow  
-**AB:** MB, ChB, FRCS, MD, PhD, Head of Bristol Medical School, Professor of Orthopaedic Surgery, orthopaedic surgeon  
+**AB:** MB, ChB, FRCS, MD, PhD, Head of Bristol Medical School, professor of orthopaedic surgery, consultant orthopaedic surgeon  
 **MJW:** MB, ChB, MD, PhD, FRCS, professor of orthopaedics, orthopaedic surgeon  
 **AS:** MSc, statistician, senior research fellow  
-**OR:** MD, PhD, orthopaedic surgeon, professor of orthopaedics, head of department, registry director of the Swedish Hip Arthroplasty Register  
-**NPH:** MD, PhD, orthopaedic surgeon, professor of orthopaedics, research leader
+**OR:** MD, PhD, consultant orthopaedic surgeon, professor of orthopaedics, head of department, registry director of the Swedish Hip Arthroplasty Register  
+**NPH:** MD, PhD, consultant orthopaedic surgeon, professor of orthopaedics, head of department
 
@@
 Shared decision making has evolved into an integral part of patient-physician interactions prior to surgical interventions, and the weighing of risks against benefits is central to this process.
-90-day mortality after THA performed due to osteoarthritis is low, ranging from 0.2 % to 0.6 %.[@Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.
+90-day mortality after THA performed due to osteoarthritis is low, ranging from 0.2&nbsp;% to 0.6&nbsp;%.[@Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.
 
@@
 The model was then validated externally on patients from England and Wales recorded in the National Joint Registry for England, Wales, Northern Ireland, the Isle of Man and the States of Guernsey (NJR).
-We focused on patients with cemented THA due to osteoarthritis.
-~~We thus excluded patients where either the femoral stem and/or the acetabular cup were uncemented (uncemented, hybrids and reverse hybrids).~~ Cementation is the most common fixation technique used in Sweden, thus used as inclusion criteria to decrease heterogeneity.
-Only the last operated hip was accounted for in patients with bilateral THA (Figure \@ref(fig:flowchartshar)).[@Bulow2020]
+We focused on patients with cemented THA due to osteoarthritis ~~since~~ cementation is the most common fixation technique used in Sweden, ~~and we thus excluded patients where either the femoral stem and/or the acetabular cup were cementless (cementless, hybrids, and reverse hybrids).~~ Only the last operated hip was accounted for in patients with bilateral THA (Figure \@ref(fig:flowchartshar)).[@Bulow2020]
 
@@
 
-~~All public and private Swedish hospitals report their THAs to SHAR, yielding a coverage of 100 %, with an individual patient completeness of 98 %.[@Karrholm2018]~~ Our inclusion period started 2008 since the American Society for Anaesthesiologists (ASA) class and Body Mass Index (BMI) were systematically recorded in SHAR from then on.
-The observation period ended 2015, since we had access to comorbidity from the national patient register up to this point.
-Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration.
-This data linkage has been described previously.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital and year of surgery were collected from the SHAR.
-Data on educational level were recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labour market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status, were also collected from LISA and comorbidity during the year preceding index surgery from the National Patient Register (NPR).
+~~All public and private Swedish hospitals report their THAs to SHAR, yielding a coverage of 100&nbsp;%, with an individual patient completeness of 98&nbsp;%.[@Karrholm2018]~~ Our inclusion period started 2008 since the American Society for Anaesthesiologists (ASA) class and Body Mass Index (BMI) were systematically recorded in SHAR from then on.
+The observation period ended 2015, since we ~~only~~ had access to ~~information on~~ comorbidity from the national patient register up to this point.
+Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration, ~~which~~ has been described previously.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital and year of surgery were collected from the SHAR.
+Data on educational level were recorded for more than 98&nbsp;% of the population with 85&nbsp;% accuracy in the longitudinal integration database for health insurance and labour market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status, were also collected from LISA and comorbidity during the year preceding index surgery from the National Patient Register (NPR).
 NPR contains diagnoses coded by ICD-10 from in- and outpatient episodes in all private and public hospitals.
-Completeness for NPR is above 99 % and 85-95 % of all diagnostic codes are valid.[@Ludvigsson2011] Death dates were linked from the national population register.
+Completeness for NPR is above 99&nbsp;% and 85-95&nbsp;% of all diagnostic codes are valid.[@Ludvigsson2011] Death dates were linked from the national population register.
 
 Patients with missing information on BMI or a measurement above 50 were excluded, as were patients with missing information on ASA class or class IV and above, as well as patients with unknown educational level or type of hospital.
-`r format(nrow(df), big.mark = ",")` patients (mean age `r round(mean(df$P_Age))` years, range `r paste(range(df$P_Age), collapse = " - ")` and `r round(mean(df$P_Sex == "Female") * 100)` % females) were included (Figure \@ref(fig:flowchartshar)).
+`r format(nrow(df), big.mark = ",")` patients (mean age `r round(mean(df$P_Age))` years, range `r paste(range(df$P_Age), collapse = " - ")` and `r round(mean(df$P_Sex == "Female") * 100)`&nbsp;% females) were included (Figure \@ref(fig:flowchartshar)).
 
@@
 
-~~Reporting to NJR is mandatory for all hospitals run by the National Health Service (NHS) in England and Wales.[@NationalJointRegistryNJR] The register coverage is thus 100&nbsp;% for such hospitals, although privately founded THAs are not included.
+~~Reporting to NJR is mandatory for all hospitals whether run by the National Health Service (NHS) or privately in England and Wales.
 Patient participation is based on informed consent.
-Consent rates varies slightly between years and regions but were 92.3&nbsp;% in both England and Wales in 2018.[@TheNJREditorialBoard]~~ NJR data was linked to the secondary uses service database from the National Health Service (NHS) with comorbidity from the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Thorn2016] Out-patient deaths were linked from the office of national statistics based on NHS-numbers, as well as patient name, age, sex and address.[@Sayers2016]
+Consent rates vary slightly between years and regions but it was 92.3 % in both England and Wales in 2018.[@TheNJREditorialBoard; @NationalJointRegistryNJR]~~ NJR data was linked to the secondary uses service database from the National Health Service (NHS) with comorbidity from the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Thorn2016] Out-patient deaths were linked from the office of national statistics based on NHS-numbers, as well as patient name, age, sex and address.[@Sayers2016]
 
@@
 Comorbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process.
-100 bootstrap samples were drawn from the observed data set.
-We used 10-fold cross validation for every bootstrap sample with a range of potential penalty values ($\lambda$:s) in a logistic LASSO model.
+~~One hundred~~ bootstrap samples were drawn from the observed data set.
+We used ~~ten~~-fold cross validation for every bootstrap sample with a range of potential penalty values ($\lambda$:s) in a logistic LASSO model.
 We  kept $\lambda$:s minimizing the mean cross-validated deviances and those were used to estimate model coefficients for each potential predictor.
@@
 Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.
-Odds ratios for the final model were estimated with 95 % confidence intervals and $p$-values below 0.05 were considered statistically significant.
+Odds ratios for the final model were estimated with 95&nbsp;% confidence intervals and $p$-values below 0.05 were considered statistically significant.
 
@@
 Receiver operating characteristic (ROC) curves were drawn and the area under those curves (AUC) were used as a measure of discriminative ability.
-Models with a lower bootstrap-based 95 % confidence limit above 0.7, were considered good.
+Models with a lower bootstrap-based 95&nbsp;% confidence limit above 0.7, were considered good.
 We used the bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples to adjust for optimism.[@Miller1991] Calibration bands were made to graphically assess model calibration, comparing predicted probabilities and observed proportions.[@Nattino2016] The reduced model was then evaluated externally.
-An AUC with 95 % CI was calculated for the model as-is.
+An AUC with 95&nbsp;% CI was calculated for the model as-is.
 Re-calibration of the model intercept was then performed to account for different mortality rates in Sweden compared to England and Wales.
@@
 We used `r substr(R.version.string, 1, 15)` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC and shiny.
-All R-scripts and necessary configurations (but no personal data) is available at http://doi.org/10.5281/zenodo.3732852.
+All R-scripts and necessary configurations (but no personal data) ~~are~~ available at http://doi.org/10.5281/zenodo.3732852.
 A linked Binder environment is also available for interactive usage.
@@
 `r sprintf("%d [%s]", sum(df$death90), surv90d)` of the patients from SHAR died within 90 days and no one was censored before that.
-~~The median age of patients who died were `r age_median` years (range `r age_range`) with distribution in Table \@ref(tab:agedist).~~ Patients from NJR had an unadjusted risk of 90 day mortality of `r njr_surv`.
+~~The median age of patients who died was `r age_median` years (range `r age_range`; Table \@ref(tab:agedist).~~ Patients from ~~the~~ NJR had an unadjusted risk of 90-day mortality of `r njr_surv`.
 Further cohort characteristics are presented in Table \@ref(tab:tab1).
@@
 We therefore considered the reduced model as superior due to its simplicity.
-Traditional models performed poorly with 95 % confidence intervals not above 0.7: The Charlson comorbidity index had an AUC of `r half_paran("Charlson")` and the Elixhauser comorbidity an AUC of `r gsub("\\)", ";", half_paran("Elixhauser"))` Figure \@ref(fig:rocvalidation) and Figure \@ref(fig:aucci)).
+Traditional models performed poorly with 95&nbsp;% confidence intervals not above 0.7: The Charlson comorbidity index had an AUC of `r half_paran("Charlson")` and the Elixhauser comorbidity an AUC of `r gsub("\\)", ";", half_paran("Elixhauser"))` Figure \@ref(fig:rocvalidation) and Figure \@ref(fig:aucci)).
 
 The ability of the reduced model to estimate probabilities of death within 90 days is further illustrated in Figure \@ref(fig:separationhist) and \@ref(fig:separationdens).
-Model calibration was good for estimated probabilities up to 3 % and acceptable up to 5 % (Figure \@ref(fig:calibration)).
+Model calibration was good for estimated probabilities up to 3&nbsp;% and acceptable up to 5&nbsp;% (Figure \@ref(fig:calibration)).
 Estimated model coefficients and corresponding odds ratios for the reduced model are presented in Table \@ref(tab:brlreduced).
-Omitting cancer from the reduced model did not affect the AUC or calibration for estimated probabilities below 3 %, but calibration outside this range deteriorated, and we thus retained cancer as an important predictor.
+Omitting cancer from the reduced model did not affect the AUC or calibration for estimated probabilities below 3&nbsp;%, but calibration outside this range deteriorated, and we thus retained cancer as an important predictor.
 
@@
 Calibration of the re-calibrated model was slightly inferior compared to the internal calibration.
-Predicted probabilities between 0.5 % and 1.5 % were lower than observed proportions and the estimated 95 % confidence bands were wider.
-Over-all, calibration was good for predicted probabilities below 3 % and acceptable below 5 % (Figure \@ref(fig:calibration)).
+Predicted probabilities between 0.5&nbsp;% and 1.5&nbsp;% were lower than observed proportions and the estimated 95&nbsp;% confidence bands were wider.
+Over-all, calibration was good for predicted probabilities below 3&nbsp;% and acceptable below 5&nbsp;% (Figure \@ref(fig:calibration)).
 
@@
 The resulting model predicted the probability of death within 90 days as: $\hat p = 1 / [1 +\exp(-\hat \beta x)]$ where $\hat \beta$ is the vector of estimated coefficients (including intercept) presented in table \@ref(tab:brlreduced).
-This formula is considered valid for patients aged ~~`r age_range`~~ years and for predicted probabilities up to 5 %.
+This formula is considered valid for patients aged ~~`r age_range`~~ years and for predicted probabilities up to 5&nbsp;%.
 
@@
 A ~~`r substr(age_range, 4, 5)`~~-year-old man with ASA class III and cancer, would have a risk of ~~`r get_risk("sickest_man")`~~.
-Note, however, that covariate patterns with observed probabilities above 5 % were rare (`r mean(pred_probs > 5) * 100` %, n = `r sum(pred_probs > 5)`).
-Risks above 5 % are therefore subject to extrapolation.
+Note, however, that covariate patterns with observed probabilities above 5&nbsp;% were rare (`r mean(pred_probs > 5) * 100`&nbsp;%, n = `r sum(pred_probs > 5)`).
+Risks above 5&nbsp;% are therefore subject to extrapolation.
 The true proportions are likely lower (as indicated by Figure \@ref(fig:calibration)).
@@
 ~~Diagnosed~~ obesity was not statistically significant ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity.
-We noted that the proportion of patients in the SHAR with BMI above 30, the WHO definition of obesity, was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %, n = `r format(sum(df$P_BMI > 30 & !is.na(df$P_BMI)), big.mark = ",")`) than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the NPR (`r round(mean(df$c_obesity) * 100, 1)` %, n = `r format(sum(df$c_obesity & !is.na(df$c_obesity)), big.mark = ",")`).
-BMI is systematically recorded in the SHAR ~~with only around 5 % missing data,~~ but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.
+We noted that the proportion of patients in the SHAR with BMI above 30, the WHO definition of obesity, was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)`&nbsp;%, n = `r format(sum(df$P_BMI > 30 & !is.na(df$P_BMI)), big.mark = ",")`) than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the NPR (`r round(mean(df$c_obesity) * 100, 1)`&nbsp;%, n = `r format(sum(df$c_obesity & !is.na(df$c_obesity)), big.mark = ",")`).
+BMI is systematically recorded in the SHAR ~~with only around 5&nbsp;% missing data,~~ but the diagnosis of obesity ~~in patients registered in the NJR~~ might not have been made during previous contacts with the healthcare system.
 
-~~It should be noted that short-term mortality after elective THA is rare.
-As such, it might not be the most immediate issue for patients making an informed decision on whether to operate or not.
-Other important outcomes includes adverse events after surgery such as prosthetic joint infections or dislocation.
-Our next step is therefore to extend the web calculator to such events as well.~~
+~~It should be noted that short-term mortality after elective THA is rare and might therefore not be the most important issue for patients making an informed decision on whether to consent to THA surgery.
+Other important outcomes such as adverse events after surgery, including prosthetic joint infections (PJI) or dislocation, are much more common and have a considerable impact on quality of life and the necessity of subsequent surgeries.
+Our next step is therefore to extend the web calculator to adverse events such as PJI and dislocation.~~
 
@@
 The nationwide design with data from both Sweden, England and Wales is a strength.
-The Swedish registers are valid with low proportions of missing data.[@Karrholm2018; @Ludvigsson2011] Some concerns have been raised regarding the HES database from UK, but a systematic review found that the overall median diagnostic accuracy (comparing ICD-codes to individual case notes) was 80 %.[@Burns2012] Both in- and outpatient comorbidity data were available for Sweden, but only in-patient data for England and Wales.
-The linkage between NJR and HES has been previously described,[@Smith2012] where it was noted that privately funded patients were not included in HES.
-17 % of the patients had missing personal data or did not allow linkage, and 6 % were not found in HES although their data were available from the NJR.
+The Swedish registers are valid with low proportions of missing data.[@Karrholm2018; @Ludvigsson2011] Some concerns have been raised regarding the HES database from UK, but a systematic review found that the overall median diagnostic accuracy (comparing ICD-codes to individual case notes) was 80&nbsp;%.[@Burns2012] Both in- and outpatient comorbidity data were available for Sweden, but only in-patient data for England and Wales.
+The linkage between NJR and HES has been previously described,[@Smith2012] where it was noted that ~~patients operated at~~ privately funded ~~units~~ were not included in HES.
+17&nbsp;% of the patients had missing personal data or did not allow linkage, and 6&nbsp;% were not found in HES although their data were available from the NJR.
 
-The indications for different fixation techniques differ between the cohorts.
-Cementation is commonly used in Sweden but not so in England and Wales, where younger patients are more likely to be operated with cementless implants.
-To only include cemented THA therefore implies an over-representation of older and frailer patients from England and Wales.
-Those patients might be more likely to die within 90 days, and this assumption would be in agreement with the calibration plot where estimated probabilities below 1.5 % underestimated the observed proportion of deaths in the external validation cohort.
-~~We tried to also validate the derived model on the Swedish cohort of patients with uncemented, hybrid or reverse hybrid THAs.
-This was not possible for each group separately since we had too few deaths in each cohort.
-It was numerically possible for all patients combined, however, for which we had 28 deaths in total.
-The 95 % CI for the AUC was very wide, however (0.54--0.76), and therefore of less relevance.
-Possible adaptations of the model to other settings or countries should always be preceded by additional external validation with data from that target population.~~
+~~A weakness of our study is that it is restricted to patients operated with cemented THA which impacts generalizability of our findings to countries with a much higher sage of cementless or hybrid techniques.~~ Cementation is ~~the most~~ commonly used ~~mode of THA fixation~~ in Sweden, but not so in England and Wales where ~~especially~~ younger patients are more likely to be operated with cementless implants.
+To only include cemented THA therefore implies an over-representation of older and frailer patients from England and Wales ~~and such~~ patients ~~are~~ more likely to die within 90 days.
+This ~~interpretation agrees~~ with the calibration plot where estimated probabilities below 1.5&nbsp;% underestimated the observed proportion of deaths in the external validation cohort.
 
-We investigated whether the effect of age on mortality was non-linear using restricted cubic splines.
-A similar approach might be relevant for BMI.
+~~We attempted at developing a similar model on a Swedish cohort of patients with uncemented, hybrid or reverse hybrid THAs.
+Unfortunately, it was impossible to analyze each mode of fixation separately since too few deaths occurred in each subgroup.
+It was possible to develop a model for all combined fixation modes other than totally cemented, but the 95&nbsp;% CI obtained for the estimated AUC was very wide (0.54--0.76) and therefore of little relevance.
+Possible adaptations of our model to other settings or countries should therefore always be preceded by additional external validation with data from that target population.~~
+
+We investigated whether the effect of age on mortality was non-linear using restricted cubic splines ~~and~~ a similar approach might be relevant for BMI.
 A U-shaped association between mortality and BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or overweight/obese (above 25 kg/m^2^) has been reported.[@Mouchti2018] Our recordings of BMI might be imprecise however.
@@
 Therefore, we aimed to avoid overfitting BMI based on too elaborate transformations.
-~~To not include BMI in the external validation is, however, beneficial, since the completeness of this variable has traditionaly been low; 60 % in 2010 but since then increasing up to 80 % in 2018.
-Instead, the related ICD-10 code was used to capture diagnosed obesity from NPR and HES respectivelly.~~
+~~To not include BMI treated as a continuous variable during  external validation was, however, beneficial, since the completeness of this variable has traditionally been low in the NJR; 60&nbsp;% in 2010 but since then increasing up to 80&nbsp;% in 2018.
+Instead, the related ICD-10 code was used to capture diagnosed obesity from the Swedish NPR and the UK HES, respectively.~~
 
@@
 The proximity in time however, the maximum of 90 days from THA to death, is an indication that the operation might be the main cause of death for the majority of deceased patients.
-~~This was also indicated by a study performed by Cnudde et al.
-who showed that the relative survival for Swedish patients with THA was above 100 % within a year after surgery.
-Hence, patients with elective THA live longer on average compared to the general population of the same age, sex, and year of birth.
-An elevated risk of death (a relative survival below 100 %) was seen, however, for cemented THA during the first 90 days after surgery, indicating that the event itself is associated with increased short-term mortality.[@Cnudde2018b]~~
+~~This notion is supported by a study performed by Cnudde et al.
+who showed that patients selected for elective THA have a reduced one-year mortality risk when compared to the general population of the same age, sex, and year of birth.
+However, within the first 90 days after cemented THA surgery, the risk of death is increased when compared to the reference population, indicating that the surgery itself is associated with increased short-term mortality.[@Cnudde2018b; @Garland2017]~~
 
@@
 
-We thank the patients and staff of all the hospitals in England, Wales and Northern Ireland who have contributed data to the National Joint Registry.
-We are grateful to the Healthcare Quality Improvement Partnership (HQIP), the NJR Research Committee and staff at the NJR Centre for facilitating this work.
-We also thank all contributors to the Swedish Hip Arthroplasty Register (SHAR), especially Szilard Nemes, previous senior statistician, for his involvement in the study.
+We thank the patients and staff of all hospitals in Sweden, England, Wales and Northern Ireland who have contributed data to the ~~Swedish Hip Arthroplasty Register (SHAR) and the~~ National Joint Registry ~~for England, Wales, Northern Ireland, the Isle of Man and the States of Guernsey (NJR).~~ We are grateful to the Healthcare Quality Improvement Partnership (HQIP), the NJR Research Committee and staff at the NJR Centre for facilitating this work.
+We also thank all contributors at SHAR, especially Szilard Nemes, previous senior statistician, for his involvement in the study.
 The authors have conformed to the NJR's standard protocol for data access and publication.
 The views expressed represent those of the authors and do not necessarily reflect those of the National Joint Registry Steering Committee or the Healthcare Quality Improvement Partnership (HQIP) who do not vouch for how the information is presented.
+~~This work was supported by a grant of the Swedish Research Council (VR 2018-02612) to NPH.~~
 

### 2020-09-05 reports/tabs.Rmd
updates according to NHs suggestions. Fix #60

@@
+# Table headings
+
 ##### PAGE BREAK
@@
 (ref:comorbidity) Categorization of comorbidities based on diagnostic groups according to the Charlson and Elixhauser comorbidity indices.
-~~Each sub-condition was first classified based on data from the Swedish National patient register and from the Hospital Episodes statistics registry in England and Wales, coded by the International classification of diseases version 10 (ICD-10), with categorization defined in table 1 and 2 from Quan et al.[@Quan2005]~~ (CNS = central nervous system.)
+~~Each main  comorbidity category was first defined based on comorbidity data from the Swedish National patient register and from the Hospital Episodes statistics registry in England and Wales, coded along the International Classification of Diseases version 10 (ICD-10) and categorized according to Quan et al.[@Quan2005]~~ (CNS = central nervous system.)
 
@@
 
-(ref:agedist) ~~Age distribution among patients who died (range `r age_range`years).~~
+(ref:agedist) ~~Age distribution among patients who died in the Swedish derivation cohort (range `r age_range`years).~~
 ```{r agedist}
@@
   caption = "Variables selected by the bootstrap ranking procedure, and therefore kept in the main model.
-Variables chosen at least 33 times out of 100 were kept in the reduced model as well.
-ASA class II and III were kept in both models, since those are usually distinguished anyway.
-To simplify the models further would be possible by lumping class I and II together.
-Variables chosen more frequently are likely more important predictors.
-This is not necessarily true however since one of several strongly correlated variables might be chosen spuriously.
+Only variables chosen at least 33 times out of 100 were kept in the reduced model.
+ASA classes II and III were, however, kept in both models.
+Variables chosen more frequently are likely to be more important predictors.
 (CNS = central nervous system.)")

### 2020-09-05 reports/article.Rmd
Revision submitted. Fix #47.

@@
 `r sprintf("%d [%s]", sum(df$death90), surv90d)` of the patients from SHAR died within 90 days and no one was censored before that.
-~~The median age of patients who died was `r age_median` years (range `r age_range`; Table \@ref(tab:agedist).~~ Patients from ~~the~~ NJR had an unadjusted risk of 90-day mortality of `r njr_surv`.
+~~The median age of patients who died was `r age_median` years (range `r age_range`; Table \@ref(tab:agedist)).~~ Patients from ~~the~~ NJR had an unadjusted risk of 90-day mortality of `r njr_surv`.
 Further cohort characteristics are presented in Table \@ref(tab:tab1).

### 2020-09-09 reports/article.Rmd
fixed wrong fig ref.

@@
 We therefore considered the reduced model as superior due to its simplicity.
-Traditional models performed poorly with 95&nbsp;% confidence intervals not above 0.7: The Charlson comorbidity index had an AUC of `r half_paran("Charlson")` and the Elixhauser comorbidity an AUC of `r gsub("\\)", ";", half_paran("Elixhauser"))` Figure \@ref(fig:rocvalidation) and Figure \@ref(fig:aucci)).
+Traditional models performed poorly with 95&nbsp;% confidence intervals not above 0.7: The Charlson comorbidity index had an AUC of `r half_paran("Charlson")` and the Elixhauser comorbidity an AUC of `r gsub("\\)", ";", half_paran("Elixhauser"))` Figure \@ref(fig:rocderivation) and Figure \@ref(fig:aucci)).
 
@@
 We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` was better at predicting death within 90 days after THA than traditional models based on the Charlson or Elixhauser comorbidity indices.
-The resulting model predicted the probability of death within 90 days as: $\hat p = 1 / [1 +\exp(-\hat \beta x)]$ where $\hat \beta$ is the vector of estimated coefficients (including intercept) presented in table \@ref(tab:brlreduced).
+The resulting model predicted the probability of death within 90 days as: $\hat p = 1 / [1 +\exp(-\hat \beta' x)]$ where $\hat \beta$ is the vector of estimated coefficients (including intercept) presented in table \@ref(tab:brlreduced).
 This formula is considered valid for patients aged ~~`r age_range`~~ years and for predicted probabilities up to 5&nbsp;%.
